### 2014-02-07 paper/paper.tex
first import

\documentclass{article} % For LaTeX2e
\usepackage[a4paper,margin=1.5in]{geometry}
\usepackage[mathlines]{lineno}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{latexsym,amsbsy,amssymb,color,xspace, booktabs}

\include{macros}
\include{local_macros}
\def\linenumberfont{\normalfont\small\sffamily}

\title{Multiple Output Regression}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}
\maketitle
\linenumbers

\section{Model Description}
We describe a multiple-output regression model.
Suppose that we have inputs $\X = \{\vx_n\}_{n=1}^N$ and outputs $\vy = \{\vy_n\}_{n=1}^N$, with $\vx_n \in \calR^D$ and $\vy_n \in \calR^P$, i.e. the output is $P$-dimensional.
For ease of exposition, we assume that there are no missing values in any output dimension.
The most dominant approach in GP-based multiple output regression is to compose the different outputs as some \textit{linear} mixing (combination) of some basic processes (random functions).
Our model is also underpinned by this composition approach.
However, to really address the issue of scalability, this model is built upon sparse GPs from the ground up.

\subsection{Prior and Likelihood}
We use a simple composition:
\begin{align}
f_i(\vx) = w_i g(\vx) + h_i(\vx), \quad i = 1 \hdots P,
\end{align}
where each (latent) output is the sum of two functions: one scaled latent function that is shared by all outputs and one latent function unique to the output.
This allows the outputs to be correlated via the shared function $g(\vx)$ while still having their own independent processes.
Furthermore we assume that $g(\vx) \sim \GP(0, k^g(\vx,\vx'; \vectheta^g))$ and $h_i(\vx) \sim \GP(0, k_i(\vx,\vx'; \vectheta_i))$, and that $h_i$ are independent GPs with their own set of hyperparameters. \\

\noindent The auto-variance of the latent output using this composition is
\begin{align}
\nonumber
cov[f_i(\vx), f_i(\vx')] 
&= cov \left[ \left(w_i g(\vx) + h_i(\vx) \right) \left(w_{i} g(\vx') + h_i(\vx') \right) \right] \\
&= w_i^2 k^g(\vx, \vx') + k_i(\vx,\vx').
\end{align}
\\
\noindent The cross-covariance of the latent outputs is
\begin{align}
\nonumber
cov[f_i(\vx), f_{i'}(\vx')] 
&= cov \left[ \left(w_i g(\vx) + h_i(\vx) \right) \left(w_{i'} g(\vx') + h_{i'}(\vx') \right) \right] \\
&= w_i w_{i'} k^g(\vx, \vx').
\end{align}

\noindent It can be seen that the induced covariance function of $f_i(\vx)$ is very similar to that in the MTGP model.
Specifically, in MTGP the task-correlation matrix is a free-form positive definite matrix $B$ whereas here it is the matrix $\vw \vw^T$ where $\vw = [w_1, \hdots, w_P]^T$.
However this model is more flexible compared to MTGP because each output is also influenced by its own independent process $h_i(\vx)$.
Note that we can also use more than one shared function $g(\vx)$, in which case it becomes the semiparametric latent force model (SLFM), again with the addition of independent process in each output dimension. \\

\noindent 
\textbf{Augmented sparse GPs}
Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
This adds much flexibility to the model as the roles of $g(\vx)$ and $h_i(\vx)$ can be quite different, so it is necessary that each has its own inducing inputs.
Furthermore, $g(\vx)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\vx)$ operates only on the inputs of the $i$-th output, which can be smaller than that of $g(\vx)$.
If more than one $g(\vx)$ is used, each $g(\vx)$ may capture a different pattern so they should also have separate inducing inputs. \\

\noindent
\textbf{Prior}
\newcommand{\Zg}{\mat{Z}^g}
\newcommand{\Zi}{\mat{Z}_i}
Let $\vg = g(\X), \vh_i = h_i(\X), \vh = \{\vh_i \}_{i=1}^P$. Let $\Zg$ and $\mat{Z}^h = \{\Zi \}_{i=1}^P$ be the set of inducing inputs for $g(\vx)$ and $h_i(\vx)$; the corresponding inducing points $\vu^g$ and $\vu^h = \{\vu_i \}_{i=1}^P$.
The prior of the augmented model is given by:
\begin{align}
p(\vg | \vu^g) &= \Normal(\vg; \BigMu^g, \tilde{\K}^g )\\
p(\vu^g) &= \Normal(\vu^g; \vec{0}, k^g(\Zg, \Zg)) \\
p(\vh | \vu^h) &= \prod_{i=1}^P \Normal(\vh_i; \BigMu_i, \tilde{\K}_i)\\
p(\vu^h) &= \prod_{i=1}^P \Normal(\vu_i; \vec{0}, k_i(\Zi, \Zi)),
\end{align}
where $\BigMu^g = k^g(\X,\Zg)k(\Zg,\Zg)^{-1}\vu^g$ and $\tilde{\K}^g = k^g(\X,\X) - k(\X,\Zg)k(\Zg,\Zg)^{-1}k(\Zg,\X)$ and $\BigMu_i, \tilde{\K}_i$ are similarly defined.

\noindent
\textbf{Likelihood}
The likelihood as usual follows the standard iid Gaussian likelihood,
\begin{align}
p(\vy | \vg, \vh ) = \prod_{i=1}^P p( \vy_i ; \vg, \vh_i) = \prod_{i=1}^P \Normal( \vy_i ; \vg + \vh_i, \beta_i^{-1} \I).
\end{align}

\subsection{Variational Inference in Sparse GPs Revisited}
In this section we review inference in sparse GPs (as presented in Titsias).
The posterior in an augmented model is $p(\vf, \vu | \vy) = p(\vf | \vu, \vy) p(\vu | \vy)$.
The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\vu$, the latent values $\vf$ are independent with any other set of latent values (e.g. the test set).
In the optimal setting when $\vu$ is the sufficient statistics of $\vf$, it should hold that $p(\vf | \vu, \vy) = p(\vf | \vu)$ as $\vy$ is only the noisy version of $\vf$.
This leads to choosing a variational approximation of the posterior which factorizes as $q(\vf, \vu | \vy) = p(\vf | \vu) q(\vu | \vy)$.
Since the conditional $p(\vf | \vu)$ is known, variational inference becomes learning an optimal posterior $q(\vu | \vy)$ only. \\

\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ is
\begin{align}
\nonumber
p(\vfstar | \vy, \X_*) &= \int p(\vfstar | \vf, \vu, \X_*) q(\vf, \vu | \vy) \der \vf \der \vu \\
\nonumber
&= \int p(\vfstar | \vu) q(\vu| \vy) p(\vf | \vu) \der \vf \der \vu \\
\label{eq:sorprediction}
&= \int p(\vfstar | \vu) q(\vu| \vy) \der \vu
\end{align}
which is the same as the prediction of a SoR model (of course with a different posterior in $q(\vu | \vy)$).

\subsection{Variational Inference}
\newcommand{\ug}{\vu^g}
\newcommand{\uh}{\vu^h}
Our goal of inference is to find the posterior $p(\vg, \vh, \ug, \uh | \vy)$. 
Following the previous discussion, we assume a variational distribution which factorizes as:
\begin{align}
\nonumber
q(\vg, \vh, \ug, \uh | \vy)
 &= p(\vg | \ug) q(\ug | \vy) \prod_{i=1}^P p(\vh_i | \vu_i) q(\vu_i | \vy) \\
 &=  q(\ug, \uh) p(\vg | \ug) \prod_{i=1}^P p(\vh_i | \vu_i)
\end{align}
Since the conditionals $p(\vg | \ug)$ and $p(\vh_i | \vu_i)$ are given, we need only to find the optimum $q(\ug, \uh) = q(\ug | \vy) \prod_{i=1}^P q(\vu_i | \vy)$.
Let $q(\ug | \vy) = \Normal(\ug; \vm^g, \S^g)$ and $q(\vu_i | \vy) = \Normal(\vu_i; \vm_i, \S_i)$. \\

\noindent
To find the optimum $q(\ug, \uh)$ we optimize the evidence lowerbound of the log marginal,
\begin{align}
\nonumber
\log p(\vy) \ge& \int q(\ug, \uh) \log \frac{p(\vy | \ug, \uh) p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
\nonumber
=& \int q(\ug, \uh) \log p(\vy | \ug, \uh)  \der \ug \der \uh 
+ \int q(\ug, \uh) \log \frac{p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
\label{eq:elbo}
=& \int q(\ug, \uh) \log p(\vy | \ug, \uh)  \der \ug \der \uh 
- \left( KL[q(\ug) || p(\ug)] + \sum_{i=1}^P KL[q(\vu_i) || p(\vu_i)] \right),
\end{align}
where the last equality occurs because both of $q(\ug, \uh)$ and $p(\ug, \uh)$ fully factorize.
Since $q(\ug), q(\vu_i), p(\ug), p(\vu_i)$ are all multivariate Gaussian distribution, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g(\vx)$ and $h_i(\vx)$ use separate set of inducing inputs). \\

\noindent To compute the above equation we first focus on the term:
\newcommand{\llangle}{\left\langle}
\newcommand{\rrangle}{\right\rangle}
\begin{align}
\nonumber
\log p(\vy | \ug, \uh)
 &= \log \llangle p(\vy | \vg, \vh) \rrangle_{p(\vg,\vh | \ug, \uh)} \\
 \nonumber
&\ge \llangle \log p(\vy | \vg, \vh) \rrangle_{p(\vg,\vh | \ug, \uh)} \quad &\text{(Jensen's inequality)} \\
\nonumber
&=  \llangle \sum_{i=1}^P \sum_{n=1}^N \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\vg,\vh | \ug, \uh)}  \quad &\text{(factorized likelihood)} \\
&= \sum_{i=1}^P \sum_{n=1}^N \llangle \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\vg,\vh_i | \ug, \vu_i)} \quad &\text{(linear operation)}
\end{align}
Observe that this is very similar to the SVI for standard GP in Hensman et al. 
In this multiple-output model, the outputs factorize given $\vg$ and $\vh$ hence the lowerbound can be seen as sum of the lowerbounds in the single-output setting.

\noindent Each individual term $l_{in} = \llangle \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\vg,\vh_i | \ug, \vu_i)}$ can be computed as follows:
\begin{align}
\nonumber
l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\vg | \ug) p(\vh_i | \vu_i) \der \vg \der \vh_i \\
\nonumber
&= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
\Normal(g_n ; \mu^g_n, \tilde{k}^g_{nn})
\Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
\nonumber
&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta (y_{in} - w_i g_n - h_{in})
\Normal(g_n ; \mu^g_n, \tilde{k}^g_{nn})
\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn} \der g_n \der h_{in} \\
\nonumber
&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu^g_n) \beta (y_{in} - h_{in} - w_i \mu^g_n) + w_i^2 \beta \tilde{k}^g_{nn} \right] 
\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn} \der h_{in} \\
\nonumber
&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta \tilde{k}^g_{nn}
- \frac{1}{2} \beta \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu^g_n - \mu_{in}) \beta (y_{in} - w_i \mu^g_n - \mu_{in}) \\
&= \log \Normal(y_{in}; w_i \mu^g_n + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}^g_{nn} + \tilde{k}_{inn}).
\end{align}
Notice that the above expression is very similar to the single-output case.
Hence our derivation here can be seen as the \textit{generalization} of SVI from the standard GP regression to multiple-output regression.

\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying similar integration we get,
\begin{align}
\log p(\vy)
\ge& \sum_{i=1}^P \sum_{n=1}^N
 \left( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta^{-1})
         - \frac{1}{2} \beta_i (w_i^2 \tilde{k}^g_{nn} + \tilde{k}_{inn})
         - \frac{1}{2} \trace \left( w_i^2 \S^g \mat{\Lambda}^g_n + \S_i \mat{\Lambda}_{in} \right)
\right) \\
&- \left( KL[q(\ug) || p(\ug)] + \sum_{i=1}^P KL[q(\vu_i) || p(\vu_i)] \right),
\end{align}
where 
\begin{align}
\tilde{\mu}_{in}
&= w_i k^g(\vx_n, \Zg)k^g(\Zg,\Zg)^{-1}\vm^g + k_i(\vx_n, \Zi)k_i(\Zi,\Zi)^{-1}\vm_i \\
\mat{\Lambda}^g_n
&= \beta k^g(\Zg,\Zg)^{-1} k^g(\Zg, \vx_n) k^g(\vx_n, \Zg) k^g(\Zg,\Zg)^{-1} \\
\mat{\Lambda}_{in}
&= \beta k_i(\Zi,\Zi)^{-1} k_i(\Zi, \vx_n) k_i(\vx_n, \Zg) k_i(\Zg,\Zg)^{-1}.
\end{align}

\noindent Again, this clearly shows that this model generalizes the standard GP regression. This can be verified by setting $P = 1$, $w_i = 1$ and $h_i(\vx) = 0$ to recover the bound in Hensman et al.
Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.

\subsection{Prediction}
The predictive distribution of the $i$-th output for a test input $\vx_*$ is 
\begin{align}
p(\fstar | \vy, \vx_*) = \int \Normal(\fstar; w_i g_* + h_{i*}, 0) p(g_* | \vy, \vx_*) p(h_{i*} | \vy, \vx_*) \der g_* \der h_{i*},
\end{align}
where $p(g_* | \vy, \vx_*) = \Normal(g_*; \mu^g_*, s^g_*)$ and $p(h_{i*} | \vy, \vx_*) = p(h_{i*}; \mu_{i*}, s_{i*})$ are the predictive distributions of the sparse GPs as given in \ref{eq:sorprediction}.
Therefore we have:
\begin{align}
p(\fstar | \vy, \vx_*) = \Normal(\fstar; w_i \mu^g_* + \mu_{i*}, s^g_* + s_{i*}). 
\end{align}

%\begin{linenomath}
%\begin{align}
%\sum_{i,h} \lambda_i \lambda_h cov [f(\vx, \vx')]
%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\vx_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\vx_h - \vs_j') \\
%\end{align}
%\end{linenomath}

\end{document}

### 2014-02-07 src/identities/checkIdentities.m

@@ -1,0 +1,28 @@
+% Check some identities
+%
+
+%% (ABA')_n,n = a(n,:)*B*a(n,:)'
+A = rand(200,100);
+B = rand(100,100);
+ABA = A*B*A';
+totaldiff = 0;
+for n=1:size(A,1)
+  assert(abs(ABA(n,n) - A(n,:)*B*A(n,:)') < 1e-10, ['diff = ' num2str(abs(ABA(n,n) - A(n,:)*B*A(n,:)'))]);
+  totaldiff = totaldiff + abs(ABA(n,n) - A(n,:)*B*A(n,:)');
+end
+disp('total diff')
+disp(totaldiff)
+
+%% AB'BA = \sum_{n=1}^N AB(n,:)'B(n,:)A
+% A : M x M (like k_MM)
+% B : N x M (like k_NM)
+A = rand(1000,1000);
+B = rand(2000,1000);
+ABBA = A*(B')*B*A;
+result = zeros(size(A));
+for n=1:size(B,1)
+  result = result + A*(B(n,:)')*B(n,:)*A;
+end
+disp('total diff')
+disp(norm(ABBA - result, 2))
+

### 2014-02-07 src/lightgps.m

@@ -1,0 +1,99 @@
+% Comparison of different light gp models. Note though that it is hard to
+% compare them in an absolutely fair way. The reason being the covariance
+% functions (hence effectively the GPs) are of different forms. Perhaps the
+% best comparison is to evaluate their average performance over multpile
+% runs.
+%
+% The spare models include:
+% - gp with fitc approximation (snelson)
+% - convolved gp with discrete approximation (higdon & myself)
+% - convolved gp with latent independence (alvarez)
+%
+% Performance depends on:
+%   - covariance function (different covariance functions are created)
+%   - transformation of outputs (e.g. zero mean) or inputs (scaling etc)
+%   - initial inducing inputs (Xu) and whether they are learned (optimized)
+%   - initial value of hypers (theta) and whether they are learned
+%   (optimized)
+% 
+% Trung V. Nguyen
+% 18/07/2013
+
+clear; close all;
+% test data
+x = load('train_inputs');
+y = load('train_outputs');
+ymean = mean(y);
+y0 = y - ymean; % zero-mean for all methods
+xt = load('test_inputs');
+
+% inducing inputs
+num_induc = 10;
+[induc_indices, xu] = select_inducing('r', x, num_induc);
+
+%% spgp
+[N,dim] = size(x);
+hyp_init(1:dim,1) = -2*log((max(x)-min(x))'/2); % log 1/(lengthscales)^2
+hyp_init(dim+1,1) = log(var(y0,1)); % log size 
+hyp_init(dim+2,1) = log(var(y0,1)/4); % log noise
+
+%% cgp
+nTasks = size(y, 2);
+Xtrain = cell(1,nTasks); Ytrain = cell(1,nTasks);
+Xtest = cell(1,nTasks); Ytest = cell(1,nTasks);
+for task = 1:nTasks
+  Xtrain{task} = x;  Ytrain{task} = y(:,task);
+  Xtest{task} = xt;
+  %Ytest{task} = yt(task,:)';
+end
+
+options = multigpOptions('fitc'); % fully independent training conditional
+options.kernType = 'gg'; % also the default
+options.optimiser = 'scg'; % also the default
+options.nlf = 1; % also the default
+%options.learnScales = false;
+options.numActive = num_induc;
+options.fixInducing = false;
+options.fixIndices = induc_indices;
[…]

### 2014-02-07 src/mycgp/cgp_prediction.m

@@ -1,0 +1,96 @@
+% CGP_PREDICTION Demonstrate multigp convolution model.
+% 
+% Requires library in libs/cgp.
+%
+%	Description:
+%	the FULL covariance matrix.
+% 	demGgJura.m SVN version 312
+% 	last update 2009-04-08T10:39:37.000000Z
+
+rng(100, 'twister');
+
+% XTemp, yTemp = training
+% XTestTemp, yTestTemp = validation
+% think of them as X, y, XTest, yTest
+basedir = 'data';
+basefile = 'gprn31';
+[x, y, xt, yt] = load_data(basedir, basefile);
+
+% Convert the data format to that used by multigp implementation
+% from current data set: XTemp: Ndata x D, yTemp: Q x Ndata, similarly for
+% XTestTemp, yTestTemp
+nTasks = size(y, 1);
+Xtrain = cell(1,nTasks);
+Ytrain = cell(1,nTasks);
+Xtest = cell(1,nTasks);
+Ytest = cell(1,nTasks);
+for task = 1:nTasks
+  Xtrain{task} = x;
+  Ytrain{task} = y(task,:)';
+  Xtest{task} = xt;
+  Ytest{task} = yt(task,:)';
+end
+
+scaleVal = zeros(1,nTasks);
+biasVal = zeros(1,nTasks);
+for k =1:nTasks,
+    biasVal(k) = mean(Ytrain{k});
+    scaleVal(k) = sqrt(var(Ytrain{k}));
+end
+
+options = multigpOptions('ftc'); % full training condition
+options = multigpOptions('fitc'); % fully independent training conditional
+options.kernType = 'gg';
+options.optimiser = 'scg';
+options.nlf = 1;
+options.beta = ones(1, size(Ytrain, 2));
+options.bias =  [zeros(1, options.nlf) biasVal];
+options.scale = [zeros(1, options.nlf) scaleVal];
+
+q = size(Xtrain{1}, 2); % input dimension (spatical coordinates)
+d = nTasks + options.nlf; % 4 = #number of outputs + #latent funcs
+
+% containing data from both the latent functions and training/observed data
+X = cell(nTasks+options.nlf, 1);
+y = cell(nTasks+options.nlf, 1);
+
+% from 1 to options.nlf are indice of latent functions data
+for j=1:options.nlf
+    y{j} = [];
+    X{j} = zeros(1, q); % no inducing value (for ftc)
+end
[…]

### 2014-02-07 src/mycgp/cgp_utility_funcs.m

@@ -1,0 +1,239 @@
+% Sampling utility functions using the multi-output dependent gaussian
+% processes (in Neil's paper)
+function cgp_utility_funcs()
+
+  % Construct the covariance matrix K between all points
+  %rng(0, 'twister'); % to keep the GP parameters consistent between runs
+  D = 1; % input dimensions
+  R = 2; % latent functions
+  Q = 2; % number of users
+  N = length(-2:0.005:2); % number of items
+  axes = [-2, 2, -3, 3];
+  % Set S = Q x R for each of the smoothing kernel k_qr
+  S = 5 + 10 .* abs(randn(Q, R)) % each S_qr ~ N(5, 10)
+  % Same setting for S as in the paper (dependent outputs)
+  % S = [1; 2; 5; 10];
+  % Setting for independent outputs (does it? still share the same latent
+  % function r)
+  %S(1, 1) = 1; S(2, 1) = 2; S(3, 1) = 5; S(4, 1) = 10;
+  
+  % Set L{qr} for each smoothing kernel function k_qr
+  LQR = cell(Q, R);
+  for u=1:Q
+    for r=1:R
+      % Trung: Set similar LQR{q, r} to get similar utility functions
+      lower = sampleLowerDiagonal(D);
+      % Use cholesky decomposition to get a pd matrix
+      LQR{u, r} = lower * lower';
+      % Use diagonal LQR{q, r} for testing with multi-gp
+      LQR{u, r} = LQR{u, r} .* eye(D)
+    end
+  end
+  
+  % Set 1, 2 to be similar and 3, 4 to be similar; otherwise different
+%   for r=1:R
+%     LQR{2, r} = LQR{1, r} + normrnd(0, 3); % add noise 1
+%     LQR{3, r} = LQR{3, r} .* abs(randn(D, D)); % make 1 and 3 different
+%     LQR{4, r} = LQR{3, r} + normrnd(0, 3); 
+%   end
+
+  % Independent setting
+  %LQR{1, 1} = 10, LQR{2, 1} = 100, LQR{3, 1} = 200, LQR{4, 1} = 300;
+  % Same setting as in the paper
+%   LQR{1, 1} = 50;
+%   LQR{2, 1} = 50;
+%   LQR{3, 1} = 300;
[…]

### 2014-02-07 src/mycgp/cgp_utility_funcs1.m

@@ -1,0 +1,93 @@
+% Experiment 1: Utilities functions that can be generated with different
+% number of latent functions using the convolved GPs.
+
+% Trung Nguyen
+% May 2012
+
+function cgp_utility_funcs()
+
+  rng(0, 'twister'); % to keep the GP parameters consistent between runs
+  D = 1; % input dimensions
+  R = 1; % latent functions
+  Q = 1; % number of users
+  N = 400; % number of items
+
+  % S_qr: s mall : constant function, s large: curvy function
+  S = zeros(Q, 4);
+  S(1,1) = 15;
+  S(1,2) = 5;
+  S(1,3) = 10;
+  S(1,4) = 5;
+  % LQR{q,r}: DxD :
+  LQR = cell(Q, 4);
+  LQR{1,1} = 500;
+  LQR{1,2} = 300;
+  LQR{1,3} = 508;
+  LQR{1,4} = 412;
+  % L{r} : DxD - latent functions we want; small => large step-length,
+  % constant func; large => small step-length, wavy latent
+  % small length-step such as 50 is suitable for utility function
+  L = cell(1,4);
+  L{1} = 59;
+  L{2} = 500;
+  L{3} = 230;
+  L{4} = 310;
+  
+  % Generate random item features (input points)
+  items = zeros(N, D);
+  rngSeed = 30;
+  rngType = 'twister';
+  rng(rngSeed, rngType);
+  for itemIdx=1:N
+    items(itemIdx, :) = rand(1, D); % each feature ~ N(0, 1)
+  end
+  
+  % initialize the cumulative covariance matrix
+  cumMatrix = cell(Q);
+  cumMatrix{1,1} = zeros(N, N);
+  
+  % Compute covariance for each latent process and sum them up
+  for r=1:R
+    covr = constructCovMatrix(r);
+    cumMatrix{1,1} = cumMatrix{1,1} + covr{1,1};
+  end
+  
+  % The covariance matrix K(q,x,q',x') of size QN * QN
+  % where first row is the covariance b.w q=1,x=1 and all others:
+  % [1,1,q=1,x=1 ... 1,1,q=1,x=N, 1,1,q=2,x=1, ...,
+  % 1,1,q=2,x=N, ..., 1,1,q=Q,x=1, ... 1,1,q=Q,x=N]
+  % (must consist of Q*N elements)!
+  
+  % Sampling from Gaussian N(0, K) gives an utility function 
+  K = cell2mat(cumMatrix);
[…]

### 2014-02-07 src/mycgp/laplace_approx_cgp.m

@@ -1,0 +1,43 @@
+% Given the parameters of the convolved GP model (LQR, L), compute the
+% parameters Kf, C for multi-task GP that approximates the convolved
+% processes with Laplace approximation.
+
+% Trung Nguyen
+% May 2012
+function [Kf C Cavg] = laplace_approx_cgp(S, LQR, L)
+  Q = size(LQR, 1);
+  D = size(LQR{1, 1}, 1);
+  R = length(L);
+  C = cell(Q, Q);
+  Cavg = zeros(D, D);
+  Kf = zeros(Q);
+  SSL = zeros(R);
+  LLL = cell(R, 1);
+  % Pre-compute inv(L{r}) for re-usability
+  Linv = cell(R, 1);
+  for r = 1:R
+    Linv{r} = pinv(L{r});
+  end
+  for q = 1:Q
+    for s = 1:Q
+      % Compute SSL(r) and LLL{r} for this pair of q and s
+      % SSL(r) = S_{qr}S_{sr}det(inv(L{r}))^0.5
+      % LLL{r} = inv(LQR{q}) + inv(LQR{s}) + inv(L{r});
+      partition = 0;
+      C{q, s} = zeros(D, D);
+      for r = 1:R
+        SSL(r) = S(q,r) * S(s,r) * sqrt(det(Linv{r}));
+        Kf(q,s) = Kf(q,s) + SSL(r);
+        LLL{r} = pinv(LQR{q,r}) + pinv(LQR{s,r}) + Linv{r};
+        weight = SSL(r) / sqrt(det(LLL{r}));
+        C{q, s} = C{q, s} + weight * pinv(LLL{r});
+        partition = partition + weight;
+      end
+      C{q, s} = C{q, s} / partition;
+      Cavg = Cavg + C{q, s};
+    end
+  end
+  % Approximate C from all C for all pair of tasks by taking their mean
+  Cavg = Cavg / (Q * Q);
+  
+end

### 2014-02-07 src/mycgp/mydemGgJura.m

@@ -1,0 +1,143 @@
+rand('twister', 1e6);
+randn('state', 1e6);
+
+%% select data set and convert to cgp format
+dataset = 'jura';
+log_transform = true;
+[x,y,xtest,ytest] = load_data('data/jura', [dataset 'Cd']);
+numOutputs = 3;
+XTemp = cell(1,numOutputs); yTemp = cell(1,numOutputs);
+XTestTemp = cell(1,numOutputs); yTestTemp = cell(1,numOutputs);
+if strcmp(dataset, 'jura')
+  features = 1:2;
+elseif strcmp(dataset, 'concrete')
+  features = 3:5;
+end
+for i=1:numOutputs
+  XTemp{i} = x(:,features);
+  yTemp{i} = y(:,i);
+  if strcmp(dataset, 'jura') && i > 1
+    XTemp{i} = [XTemp{i}; xtest(:,features)];
+    yTemp{i} = [yTemp{i}; ytest(:,i)];
+  end
+  if log_transform
+    yTemp{i} = log(yTemp{i});
+  end
+  XTestTemp{i} = xtest(:,features);
+  yTestTemp{i} = ytest(:,i);
+end
+
+%%
+scaleVal = zeros(1,size(yTemp, 2));
+biasVal = zeros(1,size(yTemp, 2));
+for k =1:size(yTemp, 2),
+    biasVal(k) = mean(yTemp{k});
+    scaleVal(k) = std(yTemp{k});
+end
+
+nlfs = [2];
+for nlf=nlfs
+% ftc = full training condition
+options = multigpOptions('ftc');
+options.kernType = 'gg';
+options.optimiser = 'scg';
+options.nlf = nlf;
+options.beta = ones(1, size(yTemp, 2));
+options.bias =  [zeros(1, options.nlf) biasVal];
+options.scale = [zeros(1, options.nlf) scaleVal];
+
+q = size(XTemp{1}, 2); % input dimension (spatical coordinates)
+d = size(yTemp, 2) + options.nlf; % 4 = #number of outputs + #latent funcs
+
+% containing data from both the latent functions and training/observed data
+X = cell(size(yTemp, 2)+options.nlf,1);
+y = cell(size(yTemp, 2)+options.nlf,1);
+
+% from 1 to options.nlf are indice of latent functions data
+for j=1:options.nlf
+    y{j} = [];
+    X{j} = zeros(1, q); % no inducing value (for ftc)
+end
[…]

### 2014-02-07 src/mymtgp/cgp_to_mtgp.m

@@ -1,0 +1,24 @@
+% Given the parameters of the convolved GP model (LQR, L), compute the
+% parameters Kf, C for multi-task GP that approximates the convolved
+% processes. Here we assume only one latent function.
+%
+% Deprecated.
+% This is only a special case of the approximateConvolvedGP.m function
+
+function [Kf C Cavg] = convolvedGPToMultiTaskGP(S, LQR, L)
+  Linv = pinv(L{1});
+  Q = size(LQR, 1);
+  D = size(LQR{1, 1}, 1);
+  C = cell(Q, Q);  
+  Cavg = zeros(D, D);
+  Kf = zeros(Q);
+  for q = 1:Q
+    for s = 1:Q
+      C{q, s} = pinv(LQR{q,1}) + pinv(LQR{s,1}) + Linv;
+      Kf(q, s) = S(q,1) * S(s,1) * (det(Linv)^0.5);
+      Cavg = Cavg + C{q, s};
+    end
+  end
+  % Approximate C from all C for all pair of tasks by taking their mean
+  Cavg = Cavg / (Q * Q);
+end

### 2014-02-07 src/mymtgp/mtgp.m

@@ -1,0 +1,138 @@
+% Generate utility functions using the factorized kernel (with approximation for the convolved GPs).
+
+% Trung Nguyen
+% April 2012
+function mtgp()
+
+  % Construct the covariance matrix K between all points
+  D = 1; % input dimensions = 1
+  Q = 2; % number of users
+  N = 400; % number of items
+  
+  [C, Kf, items] = randomizeData(N, D);
+  %file = 'convolvedGPParams-latents1-seed30.mat';
+  %[S, LQR, L, items] = loadConvolvedGPParams(file, N, D);
+  %[Kf, C, Cavg] = convolvedGPToMultiTaskGP(S, LQR, L);
+  %[Kf, C, Cavg] = approximateConvolvedGP(S, LQR, L);
+  
+  % The covariance matrix K(q,x,q',x') of size QN * QN
+  % where first row is [q=1,x=1,q=1,x=1 ... q=1,x=1,q=1,x=N, q=1,x=1,q=2,x=1, ...,
+  % q=1,x=1,q=2,x=N, ..., q=1,x=1,q=Q,x=1, ... q=1,x=1,q=Q,x=N] (Q*N
+  % elements)!
+  % In other words, K(q,x,q',x') = kron(Kf, Kx)
+  
+  % Use Q^2 different covariance matrix
+  %K = cell2mat(constructCovMatrix(Kf, C, items, items, true));
+  
+  % use the average Cavg
+  %K = kron(Kf, gaussianCovariance(items, items, Cavg, true));
+  
+  % use randomized data
+  K = kron(Kf, gaussianCovariance(items, items, C, true));
+  
+  % Sampling from Gaussian N(0, K) gives an utility function
+  fqx = gsamp(zeros(size(K, 1), 1), K, 1); % fqx has size 1xQN (1 row)
+  % Reshape fqx to get Q utility functions for Q users
+  fq = reshape(fqx, N, Q)';
+  
+  % plot the utility functions of all Q users
+  figure(2); hold on;
+  color = ['r', 'y', 'b', 'k'];
+  for q=1:Q
+    plot(items, fq(q, :), ['.' color(q)], 'LineWidth', 2, 'MarkerSize', 5);
+  end
+  title('Utility functions (factorized kernel) (exact) for 2 latent funcs');
+  legend(num2str([1; 2]));
+  axis([0,1,-30,30]);
+  
+%   hold off;
+%   for q=1:Q
[…]

### 2014-02-07 src/mymtgp/mtgp_pred_jura.m

@@ -1,0 +1,133 @@
+function mtgp_pred_jura()
+% function jura_pred_mtgp()
+% 1. Load data from Jura swiss data set
+% 2. Sets parameters for learning (including initialization) 
+% 3. Learns hyperparameters with minimize function
+% 4. Makes predictions at all points on all tasks
+% 5. Plots the predictions
+%%
+% rand('state',18);
+% randn('state',20);
+M = 3;    % 3 tasks, one primary and two secondary
+D = 2;    % input space = spatial coordinates of locations (x, y)
+covfunc_x = 'covSEard'; % Covariance function on input space
+
+% Load data from Jura swiss data set
+[XTemp, yTemp, XTestTemp, yTestTemp] = mapLoadData(['juraData' 'Cd']);
+x = XTemp{2};
+[N D]= size(x); % N = number of sample per tasks
+n = N*M;        % n = N * M = total number of output points
+% N x M matrix where each row is the output of M tasks
+Y = [[yTemp{1}; yTestTemp{1}], yTemp{2}(1:N), yTemp{3}(1:N)];
+%Y = [yTemp{1}(1:N), yTemp{2}(1:N), yTemp{3}(1:N)];
+ymean = zeros(1,M);
+ystd = zeros(1,M);
+%NOTE: Y(260:end,1) not in smale scale, but it should not be used
+%[Y(1:259,1) ymean(1) ystd(1)] = standardize(log(Y(1:259,1)),1,[],[]);
+[Y(:,1) ymean(1) ystd(1)] = standardize(log(Y(:,1)),1,[],[]);
+[Y(:,2:3) ymean(2:3) ystd(2:3)] = standardize(log(Y(:,2:3)),1,[],[]);
+Y(260:end,1) = nan;
+y = Y(:);
+v = repmat((1:M),N,1);
+ind_kf = v(:); % indices to tasks
+v = repmat((1:N)',1,M);
+ind_kx = v(:); % indices to input space data-points
+
+%%
+% 2. Set up data points for training
+nx = ones(n-100,1); % Number of observations on each task-input point
+xtrain = x;
+ytrain = y;
+ind_kx_train = ind_kx;
+ind_kf_train = ind_kf;
+
+ytrain = y(~isnan(y));
+Y_isnan = isnan(Y); Y_isnan = Y_isnan(:);
+ind_kf_train = ind_kf(~Y_isnan);
[…]

### 2014-02-07 src/mymtgp/mtgp_prediction.m

@@ -1,0 +1,153 @@
+% Using the factorized kernel to fit the data generated by some other processes or models
+% (such as convolved gps, gprn, etc). Then generate some functions from the
+% approximated kernels.
+% 1. Generates sample from true MTGP model
+% 2. Selects data-points for training
+% 3. Plots training data
+% 4. Sets parameters for learning (including initialization) 
+% 5. Learns hyperparameters with minimize function
+% 6. Makes predictions at all points on all tasks
+% 7. Plots the predictions
+function mtgp_prediction()
+
+% IMPLEMENTATION NOTE:
+% - use help covFunctions to see how parameters are stored in covfunctions,
+% particularly for covSEard. Note that the length-scales matrix is L =
+% [ell_1^2 ... ell_D^2] and the hyperparameters it expect is [log(ell_1)
+% ... log(ell_D)]. Care should be taken when convert back and forth between
+% the values in log-scale and normal scale
+% - mtgp pass in the wrong order of hyperparameters for covSEard so should
+% be careful to re-arrange the hyperparameters for correctness.
+
+Q = 4;
+covfunc_x = 'covSEard';
+axes = [0, 2, 0, 10];
+color = ['b', 'g', 'm', 'r', 'c'];
+markersize = 1.6;
+
+basedir = 'data'; basefile = 'gprnx';
+[xtrain, ytrain, xtest, ytest] = load_data(basedir, basefile);
+x = xtrain;
+y = ytrain'; % N x Q matrix where each row is the output of M tasks
+[N D]= size(x); % N = number of sample per tasks
+n = N*Q;        % n = N * M = total number of output points
+y = y(:);
+v = repmat((1:Q),N,1); % why Q x N? probably for Kf x Kx
+ind_kf_train = v(:); % indices to tasks
+v = repmat((1:N)',1,Q);
+ind_kx_train = v(:); % indices to input space data-points
+
+% 2. Set up data points for training
+% ntrain = n in our case
+nx = ones(n,1); % Number of observations on each task-input point
+ytrain = y; xtrain = x;
+
[…]

### 2014-02-07 src/mymtgp/mtgp_utility_funcs.m

@@ -1,0 +1,29 @@
+% Generate functions by the factorized kernel.
+%
+% Author: Trung V. Nguyen
+% Last update: 30/07/2012
+
+function mtgp_utility_funcs()
+% D: input dimensions, Q : num users, N : num items
+D = 1; Q = 1; N = 400;
+X = linspace(-2, 2, N)';
+hyp = log(rand(D + 1));
+
+Kf = 1;
+Kx = covSEard(hyp, X);
+K = kron(Kf, Kx);
+  
+% Sampling from Gaussian N(0, K) gives f = [f1 ... fQ]
+%rng(30, 'twister');
+Kchol = jit_chol(K)';
+Z = randn(N*Q,1);
+fqx = Kchol*Z;
+fq = reshape(fqx, N, Q)';
+  
+% plot the utility functions of all Q users
+figure(4); hold on;
+plot(X, fq(1,:), '.k', 'LineWidth', 2, 'MarkerSize', 5);
+title('MTGP utility functions');
+
+end
+

### 2014-02-07 src/select_inducing.m

@@ -1,0 +1,92 @@
+function [ind, xu] = select_inducing(method,x,M,hyp,logsn)
+%SELECTINDUCING [ind, xu] = select_inducing(method,x,M,hyp)
+%   
+% Select the inducing inputs from the set of (training) input x.
+%
+% INPUT
+%   - method : the inducing point selection methdo
+%       'r'       : random
+%       'kmeans'  : the k-centres in k-means clustering
+%       'rsde'    : the exact reduce density estimator (threshold for alpha
+%                   must be set)
+%       'meb'     : the standard meb 
+%       'ccmeb'   : the approximate rsde (solved using ccmeb)
+%       'ivm'     : the informative vector machine approach
+%       'hybrid'  : 'rsde' and 'meb' combine
+%   - x : N x d
+%   - M : number of inducing points
+%   - hyp : hyperparameters of kernel (for rsde, meb, ccmeb) and covariance
+%           function (ivm) (empty for 'r' and 'kmeans')
+%   - logsn : log sigma_noise (for ivm)
+%
+% OUTPUT
+%   - ind : indices of inducing inputs ([] when inducing inputs are not
+%   subset of training e.g. kmeans)
+%   - xu : M x d
+D = size(x,2); ind = []; xu = [];
+switch method
+  case 'r'
+    ind = randsample(size(x,1),M,false);
+    xu = x(ind,:);
+  case 'kmeans'
+    [~,xu] = kmeans(x,M);
+  case {'rsde','meb','ccmeb'}
+    model.X = x;     model.kern_hyp = hyp;    model.eps = 1e-6;
+    model.max_coreset = 2000;
+    if strcmp(method,'rsde')
+      model.eta = 10;
+      model = RSDESolver(model);
+      [~,sorted_ind] = sort(model.alpha,'descend');
+      xu = x(sorted_ind(1:M),:);
+      ind = sorted_ind(1:M);
+    else
+      model.solver = upper(method);
+      model = approxMEB(model);
+      if ~isempty(model)
+        xu = x(model.in_coreset,:);
+      end;
+      ind = model.in_coreset;
+      model = []; % free memory
+    end
+  case 'ivm'
+    if ~isempty(logsn),      sigma2_noise = exp(2*logsn);
+    else      sigma2_noise = 1e-6;    end
[…]

### 2014-02-11 src/gpsvi.m
derivatives of inducing inputs

@@ -1,0 +1,91 @@
+clear all; clc;
+rng(1110,'twister');
+
+%% General settings
+covfunc   = 'covSEard';
+PTRAIN    = 0.8; % Proportion of data-points for training
+PIND      = 0.2; % Proportion inducing points
+KMEANS    = 0;   % Use K-means for inducing locations
+SIGMA2N   = 1e-2;
+MAXITER   = 100;
+JITTER    = 1e-7;
+BETA      = 1/0.01; % init for learning, not the true BETA
+
+cf.tol       = 1e-3;
+cf.lrate     = 0.01;
+cf.lrate_hyp = 1e-5;
+cf.momentum = 0.9;
+cf.nbatch    = 5; % batch size 
+cf.covfunc = covfunc;
+cf.beta   = BETA;
+
+%% Generate samples from a GP
+x         = linspace(-10,10)';
+[n, D]    = size(x);
+nhyper    = eval(feval(covfunc));
+loghyp  = log(ones(nhyper,1));
+cf.loghyp = log(ones(D+1,1)); % random initial covariance hyperparamter
+% K = feval(covfunc, loghyp, x);
+% y = jit_chol(K)'*randn(n, 1) + sqrt(SIGMA2N)*randn(n, 1);
+
+linfunc = @(x) sin(x) + cos(x) + JITTER;
+% x         = linspace(-10,10)';
+% D         = size(x,2);
+% nhyper    = eval(feval(covfunc));
+% loghyper  = log(ones(nhyper,1));
+% cf.loghyp  = loghyper;
+% %f         = sample_gp(x, covfunc, loghyper, cf.jitter);
+f	        = linfunc(x);
+y         = f + sqrt(SIGMA2N)*randn(size(f)); % adds isotropic gaussian noise
+
+%% Training, testing
+Nall      = size(x,1);
+N         = ceil(PTRAIN*Nall);
+idx       = randperm(Nall);
+idx_train = idx(1:N);
+idx_test  = idx(N+1:Nall);
+xtrain    = x(idx_train,:); ytrain = y(idx_train);
+xtest     = x(idx_test,:);  ytest  = y(idx_test);
+
+%% Inducing point locations
+M     = ceil(PIND*N);
+idx_z = randperm(N);
+idx_z = idx_z(1:M);
[…]

### 2014-02-11 src/sviELBO.m

@@ -1,5 +1,5 @@
-function [logL,dloghyp,dbeta] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A)
-%SVIUPDATE [fval,dloghyp,dbeta] = sviELBO(x,y,z,m,S,mem,cf)
+function [logL,dloghyp,dbeta,dz] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A)
+%SVIELBO [fval,dloghyp,dbeta,dz] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A)
 %   
 % Lowerbound as a function of the hyperparameters (and inducing inputs) and
 % their derivatives.
@@ -20,17 +20,17 @@ function [logL,dloghyp,dbeta] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A)
 % 10/02/14
 N = size(x,1);
 M = size(z,1);
-betaval = cf.betaval;
+betaval = cf.beta;
 if isempty(Lmm)
   Kmm = feval(cf.covfunc, cf.loghyp, z);
-  Lmm = jit_chol(Kmm,3); clear Kmm;
+  Lmm = jit_chol(Kmm,3);
   Kmminv = invChol(Lmm);
   Knm = feval(cf.covfunc, cf.loghyp, x, z);
   A = Knm*Kmminv;
 end
 
 %TODO
-% 1) optimize some code if possible
+% 1) optimize code if possible
 % 2) remove terms independent of hypers: S, M, N,pi
 diagKnn = feval(cf.covfunc, cf.loghyp, x, 'diag');
 yMinusAm = y - A*m;
@@ -57,7 +57,40 @@ if nargout >= 2
   end
   % noise precision betaval
   dbeta = 0.5*N/betaval - 0.5*sum(yMinusAm.^2) -logTilde/betaval - logTrace/betaval;
+  
+  % inducing inputs
+  dz = zeros(size(z));
+  ell2 = exp(2*cf.loghyp(1:end-1));
+  ASKmminv = A*S*Kmminv;
+  D1 = yMinusAm*m'*Kmminv + A - ASKmminv;
+  D1 = betaval*D1;
+  D2 = -betaval*Kmminv*m*yMinusAm'*A - 0.5*betaval*(A')*A + betaval*ASKmminv'*A;
[…]

### 2014-02-11 src/sviUpdate.m

@@ -1,4 +1,4 @@
-function [m,S] = sviUpdate(x,y,z,m,S,Lmm,Kmminv,cf)
+function [m,S,cf] = sviUpdate(x,y,z,m,S,Lmm,Kmminv,cf)
 %SVIUPDATE [m,S] = sviUpdate(x,y,z,m,S,Lmm,Kmminv,cf)
 %   
 % Update the variational parameters (of the posterior q(u|y) = N(u; m, S))
@@ -21,6 +21,7 @@ function [m,S] = sviUpdate(x,y,z,m,S,Lmm,Kmminv,cf)
 %
 % OUTPUT
 %   - m, S : mean and covariance of q(u)
+%   - cf : loghyp and beta updated
 %
 % TODO: add momentum term for loghyp
 N = size(x,1);
@@ -28,7 +29,7 @@ M = size(z,1);
 Sinv = invChol(jit_chol(S,4));
 theta1_old = Sinv*m;
 theta2_old = -0.5*Sinv;
-betaval = cf.betaval;
+betaval = cf.beta;
 if isempty(Kmminv)
   Kmm = feval(cf.covfunc, cf.loghyp, z);
   Lmm = jit_chol(Kmm,4); clear Kmm;
@@ -46,13 +47,18 @@ theta1 = theta1_old + cf.lrate*(betaval*A'*y - theta1_old);
 %TODO replace with jit_chol and invChol
 [VV DD]     = eig(theta2);
 invTheta2 = VV*diag(1./diag(DD))*VV';
+logOld = sviELBO(x,y,z,m,S,cf,[],[],[],[]);
 S           = - 0.5*invTheta2;
 m           = S*theta1;
 
 % VB-M step for hyperparameters and perform gradient check
 % Note covSEard returns dK / dloghyper 
-[logL,dloghyp] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A);
-%cf.loghyp = cf.loghyp + cf.lrate_covhyp*dloghyp;
-% logL(oldm,oldS) < logL < logL(newhyp)
+[logL,dloghyp,dbeta] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A);
+cf.loghyp = cf.loghyp + cf.lrate_hyp*dloghyp;
+% faster learn rate for the noise 
+cf.beta = cf.beta + cf.lrate_hyp*dbeta;
[…]

### 2014-02-11 src/test/chkGradELBO.m

@@ -6,18 +6,19 @@ function chkGradELBO()
 % See also
 %   sviELBO
 
-%rng(1110, 'twister');
-N = 100; M = 20; D = 1;
+rng(1110, 'twister');
+N = 1000; M = 50; D = 2;
 x = 10*rand(N,D);
 y = 2*rand(N,1);
 z = 10*rand(M,D);
 m = rand(M,1);
 L = rand(M,M);
 S = L'*L;
-cf.betaval = 1e-5;
+cf.beta = 1e-2;
 cf.covfunc = 'covSEard';
 cf.loghyp = [log(rand(D,1)); 1+rand];
-theta = [cf.loghyp; cf.betaval];
+% test for hyperparameters
+theta = [cf.loghyp; cf.beta];
 [mygrad, delta] = gradchek(theta', @f, @grad, x,y,z,m,S,cf);
 delta = abs(delta);
 [valDiff,idx] = max(delta);
@@ -26,7 +27,22 @@ if (valDiff > 1e-5 && percentageDiff > 10)
   fprintf('test sviELBO() failed with valDiff = %.4f\n', valDiff);
   fprintf('percentage (valDiff / gradient) = %.2f\n', percentageDiff);
   Kmm = feval(cf.covfunc, cf.loghyp, z);  
-  fprintf('conditional number = %.4f\n', cond(Kmm));
+  fprintf('rcond = %.10f\n', rcond(Kmm));
+else
+  disp('test sviELBO() passed');
+end
+
+% test for inducing inputs
+theta = z(:);
+[mygrad, delta] = gradchek(theta', @fz, @gradz, x,y,m,S,cf);
+delta = abs(delta);
+[valDiff,idx] = max(delta);
+percentageDiff = (valDiff*100/abs(mygrad(idx)));
+if (valDiff > 1e-4 && percentageDiff > 10)
+  fprintf('test sviELBO() failed with valDiff = %.4f\n', valDiff);
+  fprintf('percentage (valDiff / gradient) = %.2f\n', percentageDiff);
+  Kmm = feval(cf.covfunc, cf.loghyp, z);  
+  fprintf('rcond = %.10f\n', rcond(Kmm));
 else
   disp('test sviELBO() passed');
 end
@@ -35,14 +51,24 @@ end
 
 function fval = f(theta,x,y,z,m,S,cf)
[…]

### 2014-02-11 paper/paper.tex
derivatives of inducing inputs

@@
 \begin{align}
+\nonumber
 \log p(\vy)
@@
 
+\section{Appendix}
+Here we derive the gradients of the lower bound wrt the  hyperparameters.
+Consider the lower bound in Hensman et al (as a function of the hyperparameters):
+\begin{align}
+\nonumber
+\calL
+=& \log \Normal(\vy; \K_{NM} \K_{MM}^{-1} \vm, \beta^{-1}\I)
+ - \frac{1}{2} \beta \trace \tilde{\K}
+ - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
+&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\vm \vm^T + \S)) \right) \\ \nonumber
+=& \underbrace{\log \Normal(\vy; \A \vm, \beta^{-1}\I)}_{\calL_1}
+ - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
+ - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
+&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\vm \vm^T + \S)) \right)}_{\calL_4},
+\end{align}
+where $\A = \K_{NM} \K_{MM}^{-1}$.
+Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.
+
+\subsection{Derivative of the Noise Hyperparameter}
+The derivative of the noise hyperparameter $\beta$ is easily computed as:
+\begin{align}
+\frac{d \calL}{d\beta} = \frac{N}{2\beta} - \frac{1}{2} (\vy - \A\vm)^T (\vy - \A\vm) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
+\end{align}
+\subsection{Derivatives of the Covariance Hyperparameters}  
+To simplify the math, we utilize the matrix $\A$ defined above.
+Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
+\begin{align}
+\frac{d \A}{dt} = \left(\frac{d\K_{NM}}{dt} - \A \frac{d \K_{MM}}{dt}\right)\K_{MM}^{-1}.
+\end{align}
+The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
+\begin{align}
+\frac{d \calL_1}{dt} &= \beta (\vy - \A\vm)^T \frac{d \A}{dt} \vm \\
+\frac{d \calL_2}{dt} &= \frac{1}{2}\beta \trace \left(\frac{d\K_{NN}}{dt} - \A \frac{d\K_{MN}}{dt} - \frac{d\A}{dt} \K_{MN}\right) \\
+\frac{d \calL_3}{dt} &= \beta \trace \left(\A \S \frac{d\A^T}{dt} \right) \\
+\frac{d \calL_4}{dt} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \frac{d \K_{MM}}{dt}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \frac{d \K_{MM}}{dt} \K_{MM}^{-1} (\vm \vm^T + \S) \right) 
+\end{align}
+The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.
+
+\subsection{Derivatives of the Inducing Inputs}
+To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\vz_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
+Hence the derivative wrt a dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\frac{d \calL}{dt}$.
+
+%Since $MD$ parameters are needed for the inducing inputs, it appears that the derivatives of all inducing inputs would require $\calO(MD \times M^3)$ in computation.
+%However, this complexity can actually be reduced to $\calO(DM^3)$ using the following lemma. \\
+%
+%\noindent \textbf{Lemma} Let $A, B$ be two matrices of size $N \times M$ and $M \times N$, respectively.
+Furthermore, $B$ has the property that only one of its rows or columns is non-zero.
+The complexity of $\trace(AB)$ is only $\calO(N)$.
+\\
+%
+%\noindent \textbf{Proof} Let $m <= M$ be the non-zero row of $B$. We have
+%\begin{align}
+%\trace (AB) = \sum_{i=1}^N \sum_{j=1}^M A_{ij} B_{ji} = \sum_{i=1}^N A_{im} B_{mi},
+%\end{align}
+%which clearly takes $\calO(N)$. It is easy to see that the lemma also holds when $B$ is symmetric and only one of its row and the corresponding column is non-zero. \\
+%
+%\noindent To exploit the property in the Lemma, we re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):
+
+\noindent We re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):
+\begin{align}
+\nonumber
+% dL1
+\frac{d \calL_1}{dt}
+ &= \beta \trace (\vy - \A\vm)^T \left(\frac{d\K_{NM}}{dt} -  \A \frac{d \K_{MM}}{dt}\right)\K_{MM}^{-1} \vm \\
+&= \beta \trace \K_{MM}^{-1} \vm (\vy - \A\vm)^T \frac{d\K_{NM}}{dt} 
+-\beta \trace \K_{MM}^{-1} \vm (\vy - \A\vm)^T \A \frac{d \K_{MM}}{dt} \\
+%dL2
+\frac{d \calL_2}{dt}
+&= - \beta \trace \A^T \frac{d \K_{NM}}{dt}
+ + \frac{1}{2} \beta \trace \A^T \A \frac{d \K_{MM}}{dt}  \\
+% dL3
+\frac{d \calL_3}{dt}
+&= \beta \trace \K_{MM}^{-1} \S \A^T \frac{d \K_{NM}}{dt}
+ - \beta \trace \K_{MM}^{-1} \S \A^T \A \frac{d\K_{MM}}{dt}\\
+% dL4
+\frac{d \calL_4}{dt}
+ &= \frac{1}{2}  \trace \K_{MM}^{-1} \frac{d \K_{MM}}{dt}
+  - \frac{1}{2} \trace \K_{MM}^{-1} (\vm \vm^T + \S) \K_{MM}^{-1} \frac{d \K_{MM}}{dt} 
+\end{align}
+From the above equations we see that,
+\begin{align}
+\frac{d\calL}{dt} = \trace \D_1 \frac{d\K_{NM}}{dt} + \trace \D_2 \frac{d \K_{MM}}{dt},
+\end{align}
+where 
+\begin{align}
+\D_1 =& \beta \K_{MM}^{-1} \vm (\vy - \A\vm)^T
+ + \beta \A^T
+ - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
+\D_2 =& -\beta \trace \K_{MM}^{-1} \vm (\vy - \A\vm)^T \A
+ - \frac{1}{2} \beta \A^T \A
+  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
+  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\vm \vm^T + \S) \K_{MM}^{-1}
+\end{align}
+Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
+The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of $\trace \D_2 \frac{d \K_{MM}}{dt}$ is $\calO(M)$ due to the fact that only one row (and its corresponding column) of $\frac{d \K_{MM}}{dt}$ is non-zero.
+This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.
+
 \end{document}

### 2014-02-12 paper/paper.tex
variatinoal parameter derivatives

--- 2014-02-07 paper/paper.tex
@@
 We describe a multiple-output regression model.
-Suppose that we have inputs $\X = \{\vx_n\}_{n=1}^N$ and outputs $\vy = \{\vy_n\}_{n=1}^N$, with $\vx_n \in \calR^D$ and $\vy_n \in \calR^P$, i.e. the output is $P$-dimensional.
+Suppose that we have inputs $\X = \{\x_n\}_{n=1}^N$ and outputs $\y = \{\y_n\}_{n=1}^N$, with $\x_n \in \calR^D$ and $\y_n \in \calR^P$, i.e. the output is $P$-dimensional.
 For ease of exposition, we assume that there are no missing values in any output dimension.
@@
 \begin{align}
-f_i(\vx) = w_i g(\vx) + h_i(\vx), \quad i = 1 \hdots P,
+f_i(\x) = w_i g(\x) + h_i(\x), \quad i = 1 \hdots P,
 \end{align}
 where each (latent) output is the sum of two functions: one scaled latent function that is shared by all outputs and one latent function unique to the output.
-This allows the outputs to be correlated via the shared function $g(\vx)$ while still having their own independent processes.
-Furthermore we assume that $g(\vx) \sim \GP(0, k^g(\vx,\vx'; \vectheta^g))$ and $h_i(\vx) \sim \GP(0, k_i(\vx,\vx'; \vectheta_i))$, and that $h_i$ are independent GPs with their own set of hyperparameters.
+This allows the outputs to be correlated via the shared function $g(\x)$ while still having their own independent processes.
+Furthermore we assume that $g(\x) \sim \GP(0, k_g(\x,\x'; \vectheta_g))$ and $h_i(\x) \sim \GP(0, k_i(\x,\x'; \vectheta_i))$, and that $h_i$ are independent GPs with their own set of hyperparameters.
 \\
@@
 \nonumber
-cov[f_i(\vx), f_i(\vx')] 
-&= cov \left[ \left(w_i g(\vx) + h_i(\vx) \right) \left(w_{i} g(\vx') + h_i(\vx') \right) \right] \\
-&= w_i^2 k^g(\vx, \vx') + k_i(\vx,\vx').
+cov[f_i(\x), f_i(\x')] 
+&= cov \left[ \left(w_i g(\x) + h_i(\x) \right) \left(w_{i} g(\x') + h_i(\x') \right) \right] \\
+&= w_i^2 k_g(\x, \x') + k_i(\x,\x').
 \end{align}
@@
 \nonumber
-cov[f_i(\vx), f_{i'}(\vx')] 
-&= cov \left[ \left(w_i g(\vx) + h_i(\vx) \right) \left(w_{i'} g(\vx') + h_{i'}(\vx') \right) \right] \\
-&= w_i w_{i'} k^g(\vx, \vx').
+cov[f_i(\x), f_{i'}(\x')] 
+&= cov \left[ \left(w_i g(\x) + h_i(\x) \right) \left(w_{i'} g(\x') + h_{i'}(\x') \right) \right] \\
+&= w_i w_{i'} k_g(\x, \x').
 \end{align}
 
-\noindent It can be seen that the induced covariance function of $f_i(\vx)$ is very similar to that in the MTGP model.
-Specifically, in MTGP the task-correlation matrix is a free-form positive definite matrix $B$ whereas here it is the matrix $\vw \vw^T$ where $\vw = [w_1, \hdots, w_P]^T$.
-However this model is more flexible compared to MTGP because each output is also influenced by its own independent process $h_i(\vx)$.
-Note that we can also use more than one shared function $g(\vx)$, in which case it becomes the semiparametric latent force model (SLFM), again with the addition of independent process in each output dimension.
+\noindent It can be seen that the induced covariance function of $f_i(\x)$ is very similar to that in the MTGP model.
+Specifically, in MTGP the task-correlation matrix is a free-form positive definite matrix $B$ whereas here it is the matrix $\w \w^T$ where $\w = [w_1, \hdots, w_P]^T$.
+However this model is more flexible compared to MTGP because each output is also influenced by its own independent process $h_i(\x)$.
+Note that we can also use more than one shared function $g(\x)$, in which case it becomes the semiparametric latent force model (SLFM), again with the addition of independent process in each output dimension.
 \\
@@
 Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
-This adds much flexibility to the model as the roles of $g(\vx)$ and $h_i(\vx)$ can be quite different, so it is necessary that each has its own inducing inputs.
-Furthermore, $g(\vx)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\vx)$ operates only on the inputs of the $i$-th output, which can be smaller than that of $g(\vx)$.
-If more than one $g(\vx)$ is used, each $g(\vx)$ may capture a different pattern so they should also have separate inducing inputs. \\
+This adds much flexibility to the model as the roles of $g(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
+Furthermore, $g(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be smaller than that of $g(\x)$.
+If more than one $g(\x)$ is used, each $g(\x)$ may capture a different pattern so they should also have separate inducing inputs. \\
 
@@
 \textbf{Prior}
-\newcommand{\Zg}{\mat{Z}^g}
-\newcommand{\Zi}{\mat{Z}_i}
-Let $\vg = g(\X), \vh_i = h_i(\X), \vh = \{\vh_i \}_{i=1}^P$.
-Let $\Zg$ and $\mat{Z}^h = \{\Zi \}_{i=1}^P$ be the set of inducing inputs for $g(\vx)$ and $h_i(\vx)$; the corresponding inducing points $\vu^g$ and $\vu^h = \{\vu_i \}_{i=1}^P$.
+\newcommand{\Zg}{\Z_g}
+\newcommand{\Zi}{\Z_i}
+Let $\g = g(\X), \h_i = h_i(\X), \h = \{\h_i \}_{i=1}^P$.
+Let $\Zg$ and $\mat{Z}^h = \{\Zi \}_{i=1}^P$ be the set of inducing inputs for $g(\x)$ and $h_i(\x)$; the corresponding inducing points $\u_g$ and $\u^h = \{\u_i \}_{i=1}^P$.
 The prior of the augmented model is given by:
 \begin{align}
-p(\vg | \vu^g) &= \Normal(\vg; \BigMu^g, \tilde{\K}^g )\\
-p(\vu^g) &= \Normal(\vu^g; \vec{0}, k^g(\Zg, \Zg)) \\
-p(\vh | \vu^h) &= \prod_{i=1}^P \Normal(\vh_i; \BigMu_i, \tilde{\K}_i)\\
-p(\vu^h) &= \prod_{i=1}^P \Normal(\vu_i; \vec{0}, k_i(\Zi, \Zi)),
+p(\g | \u_g) &= \Normal(\g; \BigMu_g, \tilde{\K}_g )\\
+p(\u_g) &= \Normal(\u_g; \vec{0}, k_g(\Zg, \Zg)) \\
+p(\h | \u^h) &= \prod_{i=1}^P \Normal(\h_i; \BigMu_i, \tilde{\K}_i)\\
+p(\u^h) &= \prod_{i=1}^P \Normal(\u_i; \vec{0}, k_i(\Zi, \Zi)),
 \end{align}
-where $\BigMu^g = k^g(\X,\Zg)k(\Zg,\Zg)^{-1}\vu^g$ and $\tilde{\K}^g = k^g(\X,\X) - k(\X,\Zg)k(\Zg,\Zg)^{-1}k(\Zg,\X)$ and $\BigMu_i, \tilde{\K}_i$ are similarly defined.
+where $\BigMu_g = k_g(\X,\Zg)k(\Zg,\Zg)^{-1}\u_g$ and $\tilde{\K}_g = k_g(\X,\X) - k(\X,\Zg)k(\Zg,\Zg)^{-1}k(\Zg,\X)$ and $\BigMu_i, \tilde{\K}_i$ are similarly defined.
 
@@
 \begin{align}
-p(\vy | \vg, \vh ) = \prod_{i=1}^P p( \vy_i ; \vg, \vh_i) = \prod_{i=1}^P \Normal( \vy_i ; \vg + \vh_i, \beta_i^{-1} \I).
+p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; w_i \g + \h_i, \beta_i^{-1} \I).
 \end{align}
@@
 In this section we review inference in sparse GPs (as presented in Titsias).
-The posterior in an augmented model is $p(\vf, \vu | \vy) = p(\vf | \vu, \vy) p(\vu | \vy)$.
-The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\vu$, the latent values $\vf$ are independent with any other set of latent values (e.g.
+The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
+The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g.
 the test set).
-In the optimal setting when $\vu$ is the sufficient statistics of $\vf$, it should hold that $p(\vf | \vu, \vy) = p(\vf | \vu)$ as $\vy$ is only the noisy version of $\vf$.
-This leads to choosing a variational approximation of the posterior which factorizes as $q(\vf, \vu | \vy) = p(\vf | \vu) q(\vu | \vy)$.
-Since the conditional $p(\vf | \vu)$ is known, variational inference becomes learning an optimal posterior $q(\vu | \vy)$ only. \\
+In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
+This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
+Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
 
@@
 \nonumber
-p(\vfstar | \vy, \X_*) &= \int p(\vfstar | \vf, \vu, \X_*) q(\vf, \vu | \vy) \der \vf \der \vu \\
+p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
 \nonumber
-&= \int p(\vfstar | \vu) q(\vu| \vy) p(\vf | \vu) \der \vf \der \vu \\
+&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\
 \label{eq:sorprediction}
-&= \int p(\vfstar | \vu) q(\vu| \vy) \der \vu
+&= \int p(\vfstar | \u) q(\u| \y) \der \u
 \end{align}
-which is the same as the prediction of a SoR model (of course with a different posterior in $q(\vu | \vy)$).
+which is the same as the prediction of a SoR model (of course with a different posterior in $q(\u | \y)$).
 
 \subsection{Variational Inference}
-\newcommand{\ug}{\vu^g}
-\newcommand{\uh}{\vu^h}
-Our goal of inference is to find the posterior $p(\vg, \vh, \ug, \uh | \vy)$. 
+\newcommand{\ug}{\u_g}
+\newcommand{\uh}{\u^h}
+Our goal of inference is to find the posterior $p(\g, \h, \ug, \uh | \y)$. 
 Following the previous discussion, we assume a variational distribution which factorizes as:
@@
 \nonumber
-q(\vg, \vh, \ug, \uh | \vy)
- &= p(\vg | \ug) q(\ug | \vy) \prod_{i=1}^P p(\vh_i | \vu_i) q(\vu_i | \vy) \\
- &=  q(\ug, \uh) p(\vg | \ug) \prod_{i=1}^P p(\vh_i | \vu_i)
+q(\g, \h, \ug, \uh | \y)
+ &= p(\g | \ug) q(\ug | \y) \prod_{i=1}^P p(\h_i | \u_i) q(\u_i | \y) \\
+ &=  q(\ug, \uh) p(\g | \ug) \prod_{i=1}^P p(\h_i | \u_i)
 \end{align}
-Since the conditionals $p(\vg | \ug)$ and $p(\vh_i | \vu_i)$ are given, we need only to find the optimum $q(\ug, \uh) = q(\ug | \vy) \prod_{i=1}^P q(\vu_i | \vy)$.
-Let $q(\ug | \vy) = \Normal(\ug; \vm^g, \S^g)$ and $q(\vu_i | \vy) = \Normal(\vu_i; \vm_i, \S_i)$. \\
+Since the conditionals $p(\g | \ug)$ and $p(\h_i | \u_i)$ are given, we need only to find the optimum $q(\ug, \uh) = q(\ug | \y) \prod_{i=1}^P q(\u_i | \y)$.
+Let $q(\ug | \y) = \Normal(\ug; \m_g, \S_g)$ and $q(\u_i | \y) = \Normal(\u_i; \m_i, \S_i)$. \\
 
@@
 \nonumber
-\log p(\vy) \ge& \int q(\ug, \uh) \log \frac{p(\vy | \ug, \uh) p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
+\log p(\y) \ge& \int q(\ug, \uh) \log \frac{p(\y | \ug, \uh) p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
 \nonumber
-=& \int q(\ug, \uh) \log p(\vy | \ug, \uh)  \der \ug \der \uh 
+=& \int q(\ug, \uh) \log p(\y | \ug, \uh)  \der \ug \der \uh 
 + \int q(\ug, \uh) \log \frac{p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
 \label{eq:elbo}
-=& \int q(\ug, \uh) \log p(\vy | \ug, \uh)  \der \ug \der \uh 
-- \left( KL[q(\ug) || p(\ug)] + \sum_{i=1}^P KL[q(\vu_i) || p(\vu_i)] \right),
+=& \int q(\ug, \uh) \log p(\y | \ug, \uh)  \der \ug \der \uh 
+- \left( \KL[q(\ug) || p(\ug)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right),
 \end{align}
 where the last equality occurs because both of $q(\ug, \uh)$ and $p(\ug, \uh)$ fully factorize.
-Since $q(\ug), q(\vu_i), p(\ug), p(\vu_i)$ are all multivariate Gaussian distribution, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g(\vx)$ and $h_i(\vx)$ use separate set of inducing inputs).
+Since $q(\ug), q(\u_i), p(\ug), p(\u_i)$ are all multivariate Gaussian distribution, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g(\x)$ and $h_i(\x)$ use separate set of inducing inputs).
 \\
@@
 \nonumber
-\log p(\vy | \ug, \uh)
- &= \log \llangle p(\vy | \vg, \vh) \rrangle_{p(\vg,\vh | \ug, \uh)} \\
+\log p(\y | \ug, \uh)
+ &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \ug, \uh)} \\
  \nonumber
-&\ge \llangle \log p(\vy | \vg, \vh) \rrangle_{p(\vg,\vh | \ug, \uh)} \quad &\text{(Jensen's inequality)} \\
+&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \ug, \uh)} \quad &\text{(Jensen's inequality)} \\
 \nonumber
-&=  \llangle \sum_{i=1}^P \sum_{n=1}^N \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\vg,\vh | \ug, \uh)}  \quad &\text{(factorized likelihood)} \\
-&= \sum_{i=1}^P \sum_{n=1}^N \llangle \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\vg,\vh_i | \ug, \vu_i)} \quad &\text{(linear operation)}
+&=  \Eb{\sum_{i=1}^P \sum_{n=1}^N \log p(y_{in} | g_n, h_{in}) }_{p(\g,\h | \ug, \uh)}  \quad &\text{(factorized likelihood)} \\
+&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | g_n, h_{in}) }_{p(\g,\h_i | \ug, \u_i)} \quad &\text{(linear operation)}
 \end{align}
 Observe that this is very similar to the SVI for standard GP in Hensman et al. 
-In this multiple-output model, the outputs factorize given $\vg$ and $\vh$ hence the lowerbound can be seen as sum of the lowerbounds in the single-output setting.
+In this multiple-output model, the outputs factorize given $\g$ and $\h$ hence the lowerbound can be seen as sum of the lowerbounds in the single-output setting.
 
-\noindent Each individual term $l_{in} = \llangle \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\vg,\vh_i | \ug, \vu_i)}$ can be computed as follows:
+\noindent Each individual term $l_{in} = \llangle \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\g,\h_i | \ug, \u_i)}$ can be computed as follows:
 \begin{align}
 \nonumber
-l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\vg | \ug) p(\vh_i | \vu_i) \der \vg \der \vh_i \\
+l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\g | \ug) p(\h_i | \u_i) \der \g \der \h_i \\
 \nonumber
 &= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
-\Normal(g_n ; \mu^g_n, \tilde{k}^g_{nn})
+\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
 \Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
@@
 &= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta (y_{in} - w_i g_n - h_{in})
-\Normal(g_n ; \mu^g_n, \tilde{k}^g_{nn})
-\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn} \der g_n \der h_{in} \\
+\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
+\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der g_n \der h_{in} \\
 \nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu^g_n) \beta (y_{in} - h_{in} - w_i \mu^g_n) + w_i^2 \beta \tilde{k}^g_{nn} \right] 
-\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn} \der h_{in} \\
+&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta \tilde{k}_{gnn} \right] 
+\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
 \nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta \tilde{k}^g_{nn}
-- \frac{1}{2} \beta \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu^g_n - \mu_{in}) \beta (y_{in} - w_i \mu^g_n - \mu_{in}) \\
-&= \log \Normal(y_{in}; w_i \mu^g_n + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}^g_{nn} + \tilde{k}_{inn}).
+&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta \tilde{k}_{gnn}
+- \frac{1}{2} \beta \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
+&= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
 \end{align}
+where $\tilde{k}_{gnn} = (\tilde{\K}_g)_{n,n}, \tilde{k}_{inn} = (\tilde{\K}_i)_{n,n}, \mu_{gn} = (\Mu_g)_n, \text{ and } \mu_{in} = (\Mu_i)_n$.
 Notice that the above expression is very similar to the single-output case.
@@
 \begin{align}
-\log p(\vy)
+\nonumber
+\log p(\y)
 \ge& \sum_{i=1}^P \sum_{n=1}^N
  \left( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta^{-1})
-         - \frac{1}{2} \beta_i (w_i^2 \tilde{k}^g_{nn} + \tilde{k}_{inn})
-         - \frac{1}{2} \trace \left( w_i^2 \S^g \mat{\Lambda}^g_n + \S_i \mat{\Lambda}_{in} \right)
+         - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn})
+         - \frac{1}{2} \trace \left( w_i^2 \S_g \mat{\Lambda}_{gn} + \S_i \mat{\Lambda}_{in} \right)
 \right) \\
-&- \left( KL[q(\ug) || p(\ug)] + \sum_{i=1}^P KL[q(\vu_i) || p(\vu_i)] \right),
+&- \left( \KL[q(\ug) || p(\ug)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right) \define \calL,
 \end{align}
@@
 \tilde{\mu}_{in}
-&= w_i k^g(\vx_n, \Zg)k^g(\Zg,\Zg)^{-1}\vm^g + k_i(\vx_n, \Zi)k_i(\Zi,\Zi)^{-1}\vm_i \\
-\mat{\Lambda}^g_n
-&= \beta k^g(\Zg,\Zg)^{-1} k^g(\Zg, \vx_n) k^g(\vx_n, \Zg) k^g(\Zg,\Zg)^{-1} \\
+&= w_i k_g(\x_n, \Zg)k_g(\Zg,\Zg)^{-1}\m_g + k_i(\x_n, \Zi)k_i(\Zi,\Zi)^{-1}\m_i \\
+\mat{\Lambda}_{gn}
+&= \beta k_g(\Zg,\Zg)^{-1} k_g(\Zg, \x_n) k_g(\x_n, \Zg) k_g(\Zg,\Zg)^{-1} \\
 \mat{\Lambda}_{in}
-&= \beta k_i(\Zi,\Zi)^{-1} k_i(\Zi, \vx_n) k_i(\vx_n, \Zg) k_i(\Zg,\Zg)^{-1}.
+&= \beta k_i(\Zi,\Zi)^{-1} k_i(\Zi, \x_n) k_i(\x_n, \Zg) k_i(\Zg,\Zg)^{-1}.
 \end{align}
 
-\noindent Again, this clearly shows that this model generalizes the standard GP regression. This can be verified by setting $P = 1$, $w_i = 1$ and $h_i(\vx) = 0$ to recover the bound in Hensman et al.
+\noindent Again, this clearly shows that this model generalizes the standard GP regression. This can be verified by setting $P = 1$, $w_i = 1$ and $h_i(\x) = 0$ to recover the bound in Hensman et al.
 Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.
 
+\subsubsection{Variational Parameters Derivatives}
+\newcommand{\oi}{\vec{o}_i}
+% some notation
+Before diving into the details, we first define the indexing operator of a matrix: $\B(\vec{r},\vec{c})$ extracts the submatrix in rows $\vec{r}$ and columns $\vec{c}$ of $\B$.
+To index all columns we use $\B(\vec{r},:)$ and similarly for all rows $\B(:,\vec{c})$.
+Readers familiar with this operator will recognize that this is the MATLAB indexing operator.
+
+The optimal posteriors are found by setting the gradients of the lowerbound $\calL$ wrt to the parameters of $q(\ug)$ and $q(\u_i)$.
+Recall that in the model, different outputs can be observed at different inputs (i.e. the case of missing values).
+Let $\oi$ be the indice of the observed inputs of the output dimension $i$.
+We denote its set of observed inputs and targets as: $\X_i = \X(\oi,:)$ and $\y_i = y_{i}(\oi)$.
+
+\noindent As a function of the parameters of $q(\ug)$, the lowerbound $\calL$ is:
+\begin{align}
+\nonumber
+\calL_g \define&
+ \sum_{i=1}^P \left( \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)
+ - \frac{1}{2} \beta_i \trace w_i^2 \tilde{\K}_g(\oi,\oi)
+ - \frac{1}{2} \beta_i \trace w_i^2 \S_g \A_g^T \A_g 
+  \right) \\
+  &- \frac{1}{2} \log |k_g(\Zg,\Zg) \S_g^{-1}| -\frac{1}{2} \trace k_g(\Zg,\Zg)^{-1} (\m_g \m_g^T + \S_g) ,
+\end{align}
+where $\A_g = k_g(\X,\Z_g)k_g(\Z_g,\Z_g)^{-1}$, which gives $\A_g(\oi,:) = k_g(\X_i,\Zg) k_g(\Z_g,\Z_g)^{-1}$, and  
+$\A_i = k_i(\X_i,\Z_i)k_i(\Z_i,\Z_i)^{-1}$. \\
+
+%------------------------------------------
+% derivatives of q(u_g)
+\noindent The derivatives of $\calL_q$ wrt $\m_g$ and $\S_g$ are given by:
+\begin{align}
+\deriv{\calL_g}{\m_g}
+& = \sum_{i=1}^P \beta_i w_i \A_g(\oi,:)^T (\y_i - \A_i \m_i) - \left[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i \A_g(\oi,:)^T \A_g(\oi,:) \right] \m_g \\
+\deriv{\calL_g}{\S_g} 
+&= \frac{1}{2} \S_g^{-1} - \frac{1}{2} \left[ k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i \A_g(\oi,:)^T \A_g(\oi,:) \right].
+\end{align}
+
+%-------------------------------------------
+%  derivatives of q(u_i)
+\noindent As a function of the parameters of $q(\u_i)$, the lowerbound $\calL$ is:
+\begin{align}
+\nonumber
+\calL_i \define&
+ \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)
+ - \frac{1}{2} \beta_i \trace \tilde{\K}_i(\oi,\oi)
+ - \frac{1}{2} \beta_i \trace \S_i \A_i^T \A_i
+ \\
+  &- \frac{1}{2} \log |k_i(\Zi,\Zi) \S_i^{-1}| -\frac{1}{2} \trace k_i(\Zi,\Zi)^{-1} (\m_i \m_i^T + \S_i) ,
+\end{align}
+
+\noindent The derivatives of $\calL_i$ wrt $\m_i$ and $\S_i$ are given by:
+\begin{align}
+\deriv{\calL_g}{\m_i}
+& = \beta_i \A_i^T (\y_i - w_i \A_g(\oi,:)^T \m_g) - \left[k_i(\Zi,\Zi)^{-1} +  \beta_i \A_i^T \A_i \right] \m_i \\
+\deriv{\calL_g}{\S_g} 
+&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \left[ k_i(\Zi,\Zi)^{-1} + \beta_i \A_i^T \A_i \right].
+\end{align}
+
+% comment on computation
+\noindent It can be seen that the derivatives of the parameters of $q(\u_i)$ only involve the observations of the output dimension $i$.
+The derivatives of the parameters of $q(\u_g)$ involve the observations across all output dimensions but decompose as a sum of contributions from individual outputs.
+Therefore, computation of the derivatives (and hence the update equations) can be distributed or parallelized easily.
+This attractive property allows the model to scale to a very large number of outputs.
+
+%\subsubsection{Update Equations}
+% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
 \subsection{Prediction}
-The predictive distribution of the $i$-th output for a test input $\vx_*$ is 
+The predictive distribution of the $i$-th output for a test input $\x_*$ is 
 \begin{align}
-p(\fstar | \vy, \vx_*) = \int \Normal(\fstar; w_i g_* + h_{i*}, 0) p(g_* | \vy, \vx_*) p(h_{i*} | \vy, \vx_*) \der g_* \der h_{i*},
+p(\fstar | \y, \x_*) = \int \Normal(\fstar; w_i g_* + h_{i*}, 0) p(g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der g_* \der h_{i*},
 \end{align}
-where $p(g_* | \vy, \vx_*) = \Normal(g_*; \mu^g_*, s^g_*)$ and $p(h_{i*} | \vy, \vx_*) = p(h_{i*}; \mu_{i*}, s_{i*})$ are the predictive distributions of the sparse GPs as given in \ref{eq:sorprediction}.
+where $p(g_* | \y, \x_*) = \Normal(g_*; \mu_{g*}, s_{g*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu_{i*}, s_{i*})$ are the predictive distributions of the sparse GPs as given in \ref{eq:sorprediction}.
 Therefore we have:
 \begin{align}
-p(\fstar | \vy, \vx_*) = \Normal(\fstar; w_i \mu^g_* + \mu_{i*}, s^g_* + s_{i*}). 
+p(\fstar | \y, \x_*) = \Normal(\fstar; w_i \mu_{g*} + \mu_{i*}, s_{g*} + s_{i*}). 
 \end{align}
@@
 %\begin{align}
-%\sum_{i,h} \lambda_i \lambda_h cov [f(\vx, \vx')]
-%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\vx_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\vx_h - \vs_j') \\
+%\sum_{i,h} \lambda_i \lambda_h cov [f(\x, \x')]
+%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\x_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\x_h - \vs_j') \\
 %\end{align}
@@
 
+\section{Appendix}
+Here we derive the gradients of the lower bound wrt the  hyperparameters.
+As a template, consider the lower bound in Hensman et al (as a function of the hyperparameters):
+\begin{align}
+\nonumber
+\calL
+=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I)
+ - \frac{1}{2} \beta \trace \tilde{\K}
+ - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
+&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right) \\ \nonumber
+=& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
+ - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
+ - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
+&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4},
+\end{align}
+where $\A = \K_{NM} \K_{MM}^{-1}$.
+Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.
+
+\subsection{Derivative of the Noise Hyperparameter}
+The derivative of the noise hyperparameter $\beta$ is easily computed as:
+\begin{align}
+\deriv{\calL}{\beta} = \frac{N}{2\beta} - \frac{1}{2} (\y - \A\m)^T (\y - \A\m) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
+\end{align}
+\subsection{Derivatives of the Covariance Hyperparameters}  
+To simplify the math, we utilize the matrix $\A$ defined above.
+Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
+\begin{align}
+\deriv{\A}{t} = \left(\deriv{\K_{NM}}{t} - \A \deriv{\K_{MM}}{t}\right)\K_{MM}^{-1}.
+\end{align}
+The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
+\begin{align}
+\deriv{\calL_1}{t} &= \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
+\deriv{\calL_2}{t} &= \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
+\deriv{\calL_3}{t} &= \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
+\deriv{\calL_4}{t} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
+\end{align}
+The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.
+
+\subsection{Derivatives of the Inducing Inputs}
+To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\z_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
+Hence the derivative wrt a single dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\deriv{ \calL}{t}$.
+
+%Since $MD$ parameters are needed for the inducing inputs, it appears that the derivatives of all inducing inputs would require $\calO(MD \times M^3)$ in computation.
+%However, this complexity can actually be reduced to $\calO(DM^3)$ using the following lemma. \\
+%
+%\noindent \textbf{Lemma} Let $A, B$ be two matrices of size $N \times M$ and $M \times N$, respectively.
+Furthermore, $B$ has the property that only one of its rows or columns is non-zero.
+The complexity of $\trace(AB)$ is only $\calO(N)$.
+\\
+%
+%\noindent \textbf{Proof} Let $m <= M$ be the non-zero row of $B$. We have
+%\begin{align}
+%\trace (AB) = \sum_{i=1}^N \sum_{j=1}^M A_{ij} B_{ji} = \sum_{i=1}^N A_{im} B_{mi},
+%\end{align}
+%which clearly takes $\calO(N)$. It is easy to see that the lemma also holds when $B$ is symmetric and only one of its row and the corresponding column is non-zero. \\
+%
+%\noindent To exploit the property in the Lemma, we re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):
+
+\noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
+\begin{align}
+\nonumber
+% dL1
+\deriv{\calL_1}{t}
+ &= \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
+&= \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} 
+-\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t} \\
+%dL2
+\deriv{\calL_2}{t}
+&= - \beta \trace \A^T \deriv{\K_{NM}}{t}
+ + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}  \\
+% dL3
+\deriv{\calL_3}{t}
+&= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
+ - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}\\
+% dL4
+\deriv{\calL_4}{t}
+ &= \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t}
+  - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
+\end{align}
+From the above equations we get,
+\begin{align}
+\deriv{\calL}{t} = \trace \D_1 \deriv{\K_{NM}}{t} + \trace \D_2 \deriv{ \K_{MM}}{t},
+\end{align}
+where 
+\begin{align}
+\D_1 =& \beta \K_{MM}^{-1} \m (\y - \A\m)^T
+ + \beta \A^T
+ - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
+\D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
+ - \frac{1}{2} \beta \A^T \A
+  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
+  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
+\end{align}
+Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
+The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two $\trace$ operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
+This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.
+
 \end{document}

### 2014-02-12 src/sviELBO.m
restructure: put all parameters in structure

@@ -1,38 +1,40 @@
-function [logL,dloghyp,dbeta,dz] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A)
-%SVIELBO [fval,dloghyp,dbeta,dz] = sviELBO(x,y,z,m,S,cf,Lmm,Kmminv,Knm,A)
+function [logL,dloghyp,dbeta,dz] = sviELBO(x,y,params,cf,Lmm,Kmminv,Knm,A)
+%SVIELBO [fval,dloghyp,dbeta,dz] = sviELBO(x,y,params,cf,Lmm,Kmminv,Knm,A)
 %   
 % Lowerbound as a function of the hyperparameters (and inducing inputs) and
-% their derivatives.
+% their derivatives. Note that the inducing derivatives are for the
+% SEard covariance function only.
 %
 % INPUT
 %   - x : inputs (in SVI, this is a mini-batch)
 %   - y : outputs (of inputs x)
-%   - z : inducing inputs
-%   - m,S : current variational parameters
-%   - cf : options for the optimization procedure and hyperparameters
+%   - params : parameters structure
+%   - cf : cf.covfunc 
 %   - Lmm, Kmminv, Knm, A : [optional] previously computed values of those
 %   (see also SVIUPDATE).
 %
 % OUTPUT
-%   - logL, dloghyp : objective value and its derivatives
+%   - logL, dloghyp, dbeta, dz : objective value and its derivatives
 %
 % Trung Nguyen
-% 10/02/14
+% 11/02/14
+z = params.z;
+m = params.m;
+S = params.S;
+betaval = params.beta;
+loghyp = params.loghyp;
 N = size(x,1);
 M = size(z,1);
-betaval = cf.beta;
+Kmm = feval(cf.covfunc, loghyp, z);
 if isempty(Lmm)
-  Kmm = feval(cf.covfunc, cf.loghyp, z);
   Lmm = jit_chol(Kmm,3);
   Kmminv = invChol(Lmm);
-  Knm = feval(cf.covfunc, cf.loghyp, x, z);
+  Knm = feval(cf.covfunc, loghyp, x, z);
   A = Knm*Kmminv;
 end
 
-%TODO
-% 1) optimize code if possible
-% 2) remove terms independent of hypers: S, M, N,pi
[…]

### 2014-02-12 src/sviUpdate.m

@@ -1,5 +1,5 @@
-function [m,S,cf] = sviUpdate(x,y,z,m,S,Lmm,Kmminv,cf)
-%SVIUPDATE [m,S] = sviUpdate(x,y,z,m,S,Lmm,Kmminv,cf)
+function params = sviUpdate(x,y,params,cf,Lmm,Kmminv)
+%SVIUPDATE params = sviUpdate(x,y,params,cf,Lmm,Kmminv)
 %   
 % Update the variational parameters (of the posterior q(u|y) = N(u; m, S))
 % and the covariance hyperparameters of a standard GP regression using SVI.
@@ -13,30 +13,33 @@ function [m,S,cf] = sviUpdate(x,y,z,m,S,Lmm,Kmminv,cf)
 % INPUT
 %   - x : inputs (in SVI, this is a mini-batch)
 %   - y : outputs (of inputs x)
-%   - z : inducing inputs
-%   - m,S : current variational parameters
+%   - params : structure containing variational parameters (m,S), inducing
+%   inputs, and hyperparameters
+%       params.m, params.S : variatinal parameters of the posterior
+%       params.z : inducing inputs
+%       params.loghyp : cov hyperparameters
+%       params.beta : noise hyperparameter
+%       params.delta_z,delta_beta,delta_loghyp: for SGD learning
 %   - Lmm : [optional] chol(Kmm)
 %   - Kmminv : [optional] Kmm^{-1} (must update if hyper is not fixed)
-%   - cf : options for the optimization procedure and hyperparameters
+%   - cf : options for the optimization procedure
 %
 % OUTPUT
-%   - m, S : mean and covariance of q(u)
-%   - cf : loghyp and beta updated
+%   - params : updated parameters
 %
-% TODO: add momentum term for loghyp
-N = size(x,1);
-M = size(z,1);
-Sinv = invChol(jit_chol(S,4));
-theta1_old = Sinv*m;
+% Trung V. Nguyen
+% 11/02/14
+Sinv = invChol(jit_chol(params.S,4));
+theta1_old = Sinv*params.m;
 theta2_old = -0.5*Sinv;
-betaval = cf.beta;
+betaval = params.beta;
 if isempty(Kmminv)
-  Kmm = feval(cf.covfunc, cf.loghyp, z);
[…]

### 2014-02-12 src/test/chkGradELBO.m

@@ -6,20 +6,20 @@ function chkGradELBO()
 % See also
 %   sviELBO
 
-rng(1110, 'twister');
+%rng(1110, 'twister');
 N = 1000; M = 50; D = 2;
 x = 10*rand(N,D);
 y = 2*rand(N,1);
-z = 10*rand(M,D);
-m = rand(M,1);
-L = rand(M,M);
-S = L'*L;
-cf.beta = 1e-2;
 cf.covfunc = 'covSEard';
-cf.loghyp = [log(rand(D,1)); 1+rand];
+params.z = 10*rand(M,D);
+params.m = rand(M,1);
+L = rand(M,M);
+params.S = L'*L;
+params.beta = 1e-2;
+params.loghyp = [log(rand(D,1)); 1+rand];
 % test for hyperparameters
-theta = [cf.loghyp; cf.beta];
-[mygrad, delta] = gradchek(theta', @f, @grad, x,y,z,m,S,cf);
+theta = [params.loghyp; params.beta];
+[mygrad, delta] = gradchek(theta', @f, @grad, x,y,params,cf);
 delta = abs(delta);
 [valDiff,idx] = max(delta);
 percentageDiff = (valDiff*100/abs(mygrad(idx)));
@@ -33,8 +33,8 @@ else
 end
 
 % test for inducing inputs
-theta = z(:);
-[mygrad, delta] = gradchek(theta', @fz, @gradz, x,y,m,S,cf);
+theta = params.z(:);
+[mygrad, delta] = gradchek(theta', @fz, @gradz, x,y,params,cf);
 delta = abs(delta);
 [valDiff,idx] = max(delta);
 percentageDiff = (valDiff*100/abs(mygrad(idx)));
@@ -49,26 +49,28 @@ end
 
 end
 
-function fval = f(theta,x,y,z,m,S,cf)
-  cf.loghyp = theta(1:end-1)';
-  cf.beta = theta(end);
-  fval = sviELBO(x,y,z,m,S,cf,[],[],[],[]);
+function fval = f(theta,x,y,params,cf)
+  params.loghyp = theta(1:end-1)';
+  params.beta = theta(end);
+  fval = sviELBO(x,y,params,cf,[],[],[],[]);
 end
 
-function g = grad(theta,x,y,z,m,S,cf)
-  cf.loghyp = theta(1:end-1)';
-  cf.beta = theta(end);
[…]

### 2014-02-13 src/learn_fitc.m
wrapper for gpsvi and fitc

@@ -1,0 +1,49 @@
+function [z,z0,hyp,mu,s2] = learn_fitc(x,y,xtest,M,z0)
+%LEARN_FITC [z,z0,hyp,mu,s2] = learn_fitc(x,y,xtest,M,z0)
+%   Wrapper for FITC.
+% 
+% INPUT
+%   - x, y, xtest : train inputs, train outputs, test inputs (optional)
+%   - M : #inducing points
+%   - z0 : initial inducing inputs (optional)
+%
+% OUTPUT
+%   - z, z0: learned and initial inducing inputs
+%   - hyp : learned hyp
+%   - mu : predictive mean 
+%   - s2 : predictive variance
+
+me_y = mean(y); y0 = y - me_y; % zero mean the data
+[N,dim] = size(x);
+
+% initialize pseudo-inputs to a random subset of training inputs
+if isempty(z0)
+  [dum,I] = sort(rand(N,1)); clear dum;
+  I = I(1:M);
+  z0 = x(I,:);
+end
+
+% initialize hyperparameters sensibly (see spgp_lik for how
+% the hyperparameters are encoded)
+% hyp_init(1:dim,1) = -2*log((max(x)-min(x))'/2); % log 1/(lengthscales)^2
+% hyp_init(dim+1,1) = log(var(y0,1)); % log size 
+% hyp_init(dim+2,1) = log(var(y0,1)/2); % log noise
+hyp_init(1:dim+2,1) = log(ones(dim+2,1));
+
+% optimize hyperparameters and pseudo-inputs
+w_init = [reshape(z0,M*dim,1);hyp_init];
+[w,f] = minimize(w_init,'spgp_lik',-500,y0,x,M);
+% [w,f] = lbfgs(w_init,'spgp_lik',200,10,y0,x,M); % an alternative
+z = reshape(w(1:M*dim,1),M,dim);
+hyp = w(M*dim+1:end,1);
+
+% PREDICTION
+if ~isempty(xtest)
+  [mu0,s2] = spgp_pred(y0,x,z,xtest,hyp);
+  mu = mu0 + me_y; % add the mean back on
+  % if you want predictive variances to include noise variance add noise:
+  %s2 = s2 + exp(hyp(end));
+end  
+
+return;
+

### 2014-02-13 src/learn_gpsvi.m

@@ -1,0 +1,59 @@
+function [mu,s2,elbo,params] = learn_gpsvi(x,y,xtest,M,cf,z0)
+%LEARN_GPSVI learn_gpsvi(x,y,xtest,M,cf,z0)
+%   Quick wrapper to run GPSVI.
+% 
+% INPUT
+%   - x,y,xtest : train inputs, train outputs, test outputs
+%   - M : #inducing inputs
+%   - cf : configuration
+%
+% OUTPUT
+%   - mu, s2 : mean and variance prediction
+%   - elbo : the bound
+%   - params : learned parameters
+[N D] = size(x);
+idx_z = randperm(N,M);
+if isempty(z0)
+  z0 = x(idx_z,:);
+  if cf.init_kmeans
+    z0 = kmeans(z0, x, foptions());
+  end
+end
+idx = randperm(N,M);
+m =  y(idx);  % shouldn't this be the y of idx_z?
+%m = y(idx_z);
+Sinv = 0.01*(1/var(y))*eye(M);
+S = inv(Sinv);
+params.m = m;
+params.S = S;
+params.z = z0;
+params.z0 = z0;
+nhyper = eval(feval(cf.covfunc));
+loghyp = log(ones(nhyper,1));
+%sigma2n = var(y-mean(y),1)/10;
+beta0 = 1/0.01;
+params.beta   = beta0; % should be init using the data
+params.loghyp = loghyp;
+params.delta_hyp = zeros(size(params.loghyp));
+params.delta_beta = 0;
+params.delta_z = zeros(size(z0));
+
+elbo = zeros(numel(cf.maxiter),1);
+for i = 1 : cf.maxiter
+  idx = randperm(N, cf.nbatch);
+  params = sviUpdate(x(idx,:),y(idx),params,cf,[],[]);
+  elbo(i) = sviELBO(x,y,params,cf,[],[],[],[]);
+end
+
+Kmm = feval(cf.covfunc, params.loghyp, params.z);
+Lmm = jit_chol(Kmm,3);
+Kmminv = invChol(Lmm);
+if ~isempty(xtest)
+  [mu s2] = predict_gpsvi(Kmminv, cf.covfunc, params.loghyp, params.m, params.S, params.z, xtest);
+end
+
+disp('initial hyp .vs learned hyp (including sigma2n)')
+fprintf('%.5f\t%.5f\n', [[loghyp;1/beta0],[params.loghyp;1/params.beta]]');
+
+end
+

### 2014-02-13 src/gpsvi.m
remove files

@@ -1,91 +1,0 @@
-clear all; clc;
-rng(1110,'twister');
-
-%% General settings
-covfunc   = 'covSEard';
-PTRAIN    = 0.8; % Proportion of data-points for training
-PIND      = 0.2; % Proportion inducing points
-KMEANS    = 0;   % Use K-means for inducing locations
-SIGMA2N   = 1e-2;
-MAXITER   = 100;
-JITTER    = 1e-7;
-BETA      = 1/0.01; % init for learning, not the true BETA
-
-cf.tol       = 1e-3;
-cf.lrate     = 0.01;
-cf.lrate_hyp = 1e-5;
-cf.momentum = 0.9;
-cf.nbatch    = 5; % batch size 
-cf.covfunc = covfunc;
-cf.beta   = BETA;
-
-%% Generate samples from a GP
-x         = linspace(-10,10)';
-[n, D]    = size(x);
-nhyper    = eval(feval(covfunc));
-loghyp  = log(ones(nhyper,1));
-cf.loghyp = log(ones(D+1,1)); % random initial covariance hyperparamter
-% K = feval(covfunc, loghyp, x);
-% y = jit_chol(K)'*randn(n, 1) + sqrt(SIGMA2N)*randn(n, 1);
-
-linfunc = @(x) sin(x) + cos(x) + JITTER;
-% x         = linspace(-10,10)';
-% D         = size(x,2);
-% nhyper    = eval(feval(covfunc));
-% loghyper  = log(ones(nhyper,1));
-% cf.loghyp  = loghyper;
-% %f         = sample_gp(x, covfunc, loghyper, cf.jitter);
-f	        = linfunc(x);
-y         = f + sqrt(SIGMA2N)*randn(size(f)); % adds isotropic gaussian noise
-
-%% Training, testing
-Nall      = size(x,1);
-N         = ceil(PTRAIN*Nall);
-idx       = randperm(Nall);
-idx_train = idx(1:N);
-idx_test  = idx(N+1:Nall);
-xtrain    = x(idx_train,:); ytrain = y(idx_train);
-xtest     = x(idx_test,:);  ytest  = y(idx_test);
-
-%% Inducing point locations
-M     = ceil(PIND*N);
-idx_z = randperm(N);
-idx_z = idx_z(1:M);
[…]

### 2014-02-13 src/identities/checkIdentities.m

@@ -1,28 +1,0 @@
-% Check some identities
-%
-
-%% (ABA')_n,n = a(n,:)*B*a(n,:)'
-A = rand(200,100);
-B = rand(100,100);
-ABA = A*B*A';
-totaldiff = 0;
-for n=1:size(A,1)
-  assert(abs(ABA(n,n) - A(n,:)*B*A(n,:)') < 1e-10, ['diff = ' num2str(abs(ABA(n,n) - A(n,:)*B*A(n,:)'))]);
-  totaldiff = totaldiff + abs(ABA(n,n) - A(n,:)*B*A(n,:)');
-end
-disp('total diff')
-disp(totaldiff)
-
-%% AB'BA = \sum_{n=1}^N AB(n,:)'B(n,:)A
-% A : M x M (like k_MM)
-% B : N x M (like k_NM)
-A = rand(1000,1000);
-B = rand(2000,1000);
-ABBA = A*(B')*B*A;
-result = zeros(size(A));
-for n=1:size(B,1)
-  result = result + A*(B(n,:)')*B(n,:)*A;
-end
-disp('total diff')
-disp(norm(ABBA - result, 2))
-

### 2014-02-13 src/demo_fitc.m
some analysis of learning inducing inputs

@@ -1,0 +1,12 @@
+rng(1110, 'twister');
+% load demo data set (1D inputs for easy visualization -
+
+%[x,y,xtest,ytest] = load_data('.','toy');
+x = load('train_inputs');
+y = load('train_outputs');
+xtest = load('test_inputs');
+M = 10;
+[z,z0,~,mu,s2] = learn_fitc(x,y,xtest,M);
+
+plot_all(x,y,xtest,mu,s2,z0,z,'FITC');
+

### 2014-02-13 src/plot_all.m

@@ -1,0 +1,19 @@
+function plot_all(x,y,xtest,mu,var,z0,z,titlestr)
+%PLOT_ALL plot_all(x,y,xtest,mu,var,z0,z)
+%   Plot for analysis.
+%
+% INPUT
+%   - x, y, xtest
+%   - mu, var : predictive mean and variance
+%   - z0, z : initial and learned inducing
+%   - stitle : title string
+figure; hold on
+plot(xtest,mu,'b') % mean predictions in blue
+plotMeanAndStd(xtest,mu,2*sqrt(var),[7 7 7]/7.5);
+plot(z0, -0.1 + min(y)*ones(size(z,1), 1), 'k+', 'MarkerSize', 12);
+plot(z, -0.2 + min(y)*ones(size(z,1), 1), 'm+', 'MarkerSize', 12);
+plot(x,y,'.m') % data points in magenta
+ylabel('Predictive distribution');
+title(titlestr);
+
+return;

### 2014-02-13 src/toyme.m

@@ -1,0 +1,66 @@
+clear all; clc; close all;
+rng(1111,'twister');
+
+CHOICE = 2;
+switch CHOICE
+  case 1
+    % The function in SPGP 
+    x = load('train_inputs');
+    y = load('train_outputs');
+    xtest = load('test_inputs');
+    func = 'func1';
+    theaxis = [-4 10 -2.5 1.5];
+  case 2
+    % sin(x) + cos(x)
+    x = linspace(-10,10)';
+    xtest = linspace(-11,11)';
+    linfunc = @(x) sin(x) + cos(x) + 1e-7;
+    f	= linfunc(x);
+    y = f + sqrt(1e-2)*randn(size(f)); % adds isotropic gaussian noise
+    func = 'func2';
+    theaxis = [-11 11 -2.5 2.5];
+  case 3
+    % sample from a gp
+    x = linspace(-10,10)';
+    xtest = linspace(-11,11)';
+    f = sample_gp(x, 'covSEard', log([2;1]), 1);
+    y = f + sqrt(1e-2)*randn(size(f));
+    func = 'func3';
+    theaxis = [-11 11 -2.5 2.5];
+end
+
+M = 10;
+z0 = x(randperm(size(x,1),M),:);
+
+% fitc
+[z,z0,~,mu,s2] = learn_fitc(x,y,xtest,M,z0);
+plot_all(x,y,xtest,mu,s2,z0,z,'FITC');
+axis(theaxis);
+saveas(gcf, ['results/figures/' func '-fitc.eps'],'epsc');
+
+% gpsvi
+covfunc   = 'covSEard';
+cf.covfunc = covfunc;
+cf.lrate     = 1e-2;
+cf.lrate_hyp = 1e-4;
+cf.lrate_beta = 1e-4;
+cf.lrate_z   = 1e-3;
+cf.momentum  = 0.9;
+cf.momentum_z = 0.0;
+cf.learn_z   = true;
+cf.init_kmeans = false;
+cf.maxiter = 500;
+cf.nbatch = 5;
+
+[mu,s2,elbo,params] = learn_gpsvi(x,y,xtest,M,cf,z0);
+plot_all(x,y,xtest,mu,s2,params.z0,params.z,['GPSVI lrate = ' num2str(cf.lrate_z)]);
+axis(theaxis);
+saveas(gcf, ['results/figures/' func '-svi-lrate' num2str(cf.lrate_z) '.eps'],'epsc');
+
[…]

### 2014-02-13 paper/paper.tex
discussion on learning of inducing inputs

@@
 
+\subsection{Analysis on Learning of the Inducing Inputs}
+In this section we present some analysis on learning of the inducing inputs under the stochastic variational inference procedure on some toy problems.
+We use three real-valued truth functions of scalar inputs: the first one was used in Snelson et al, the second one is a function $f(x) = \sin(x) + \cos(x)$, and the third one is a sample function generated from a GP with squared exponential with ARD covariance function with a lengthscale of 2 and signal variance of 1.
+
+\noindent \paragraph{Experimental Settings} We use gpsvi to optimize for \textit{all} of the parameters in the model, i.e.
+the variational parameters of the posterior, the covariance hyperparameters, the noise hyperparameter, and the inducing inputs.
+For the hyperparameters, we use a momentum of $0.9$ and a fixed learn rate of $1e-5$. 
+For the variational parameters, we use a learn rate of $0.01$ (no momentum was used).
+We vary the learn rate of the inducing inputs from $1e{-2}$ to $1e{-6}$.
+Using momentum for the inducing inputs does not seem to have much effects.
+The maximum number of iterations is 500, the batch size 5, and the number of inducing values 10.
+
+% effect of learn rate on inducing inputs
+\noindent \paragraph{Effects of the Learn Rates }
+In Figure \ref{fig1}, \ref{fig2}, and \ref{fig3}, we show the lower bounds, predictive distributions, and the learned inducing inputs using stochastic variational inference with varying learn rates for the inducing inputs.
+We also show the results with FITC for comparison.
+It can be seen from the figures that, as typical in stochastic optimization, the behavior of GPSVI is very sensitive to the learn rate.
+In particular, when the rate is slow e.g. $1e{-5}$, almost no progress was made in learning the inducing inputs.
+However, when the rate increases to e.g. $1e{-3}$, the inducing inputs are spread out more evenly in the input space compared to the initial locations.
+Although not shown the in figures, when the learn rate is too large, e.g. $1e{-02}$, the inducing inputs may spread to regions far outside of the training intervals.
+An example of this phenomenon can be observed in the top left figure in Figure \ref{fig1}.
+The learn rates also affect the evidence lower bound which seems to wiggle more compared to not learning the inducing inputs.
+The predictive distributions are sensible particularly because the toy functions exhibit sparsity and are easy to learn.  \\
+
+\noindent The results of this qualitative analysis do not seem to suggest that learning of the inducing inputs does not work in GPSVI.
+Its effectiveness perhaps rests upon empirical performance on real datasets.
+
+\noindent \paragraph{The Noise Parameter} It seems harder to set the initial value and the learning rate of the noise parameter compared to standard GP. 
+While a large range of values of noise can be used in standard gp or fitc (e.g. 0.5), a very small value of noise $1e{-02}$, which is the true noise, is required for gpsvi to work well.
+It appears that an adaptive learning rate may be required for the noise parameter.
+
+\begin{figure*}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03.eps} &
+\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04.eps} &
+\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05.eps} \\
+\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03-bound.eps} &
+\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04-bound.eps} &
+\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05-bound.eps} \\ 
+(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
+\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func1-fitc.eps}}
+\end{tabular}
+\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the first function.
+Top row: GPSVI with different learning rates (the rates decreases from left to right).
+Bottom row: FITC.
+Magenta dots : training points.
+Solid blue line: predictive mean.
+Grey-shaded area and solid black lines: two standard deviations of the predictive distributions.
+Black (+) crosses: initial locations of inducing inputs.
+Magenta (+) crosses: learned locations of inducing inputs.
+Middle row: the evidence lower bound of gpsvi vs.
+iteration.}
+\label{fig1}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03.eps} &
+\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04.eps} &
+\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05.eps} \\
+\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03-bound.eps} &
+\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04-bound.eps} &
+\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05-bound.eps} \\ 
+(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
+\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func2-fitc.eps}}
+\end{tabular}
+\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the second function. The legends are same as in Figure 1.}
+\label{fig2}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03.eps} &
+\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04.eps} &
+\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05.eps} \\
+\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03-bound.eps} &
+\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04-bound.eps} &
+\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05-bound.eps} \\ 
+(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
+\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func3-fitc.eps}}
+\end{tabular}
+\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the third function. The legends are same as in Figure 1.}
+\label{fig3}
+\end{figure*}
+
 \end{document}

### 2014-02-18 src/computeKnmKmminv.m
1. refactor code 2. implementation of structured gpsvi (not including learning of weight w_i yet)

@@ -1,0 +1,11 @@
+function A = computeKnmKmminv(covfunc,loghyp,x,z)
+%COMPUTEKNMKMMINV A = computeKnmKmminv(covfunc,loghyp,x,z)
+%   Compute the term A = k(x,z)k(z,z)^{-1} which is commonly encountered.
+%
+Kmm = feval(covfunc, loghyp, z);
+Lmm = jit_chol(Kmm,4);
+Kmminv = invChol(Lmm);
+Knm = feval(covfunc, loghyp, x, z);
+A = Knm*Kmminv;
+end
+

### 2014-02-18 src/demo_ssvi.m

@@ -1,0 +1,69 @@
+function demo_ssvi()
+clear all; clc; close all;
+rng(1111,'twister');
+
+% sin(x) + cos(x)
+x = linspace(-10,10)';
+N = size(x,1);
+xtest = linspace(-11,11)';
+fn1 = @(x) sin(x) + 1e-7;
+fn2 = @(x) sin(x) + 1e-7;
+y1 = fn1(x) + sqrt(1e-4)*randn(N,1);
+y2 = fn2(x) + sqrt(1e-4)*randn(N,1);
+
+ind1 = -8 < x & x < -4;
+ind2 = 4 < x & x < 8;
+y1miss = y1;
+y1miss(ind1) = nan;
+y2miss = y2;
+y2miss(ind2) = nan;
+y = [y1miss, y2miss];
+%y = y1;
+M = 15;
+
+% ssvi
+cf.covfunc_g = 'covSEard';
+cf.covfunc_h = 'covSEard';
+cf.lrate     = 1e-2;
+cf.lrate_hyp = 1e-5;
+cf.lrate_beta = 1e-10;
+cf.lrate_z   = 1e-3;
+cf.momentum  = 0.9;
+cf.momentum_z = 0.0;
+cf.learn_z   = true;
+cf.init_kmeans = false;
+cf.maxiter = 500;
+cf.nbatch = 5;
+
+[elbo,params] = ssvi_learn(x,y,M,cf,[]);
+[mu,var,mu_g,var_g] = ssvi_predict(cf.covfunc_g,cf.covfunc_h,params,xtest);
+
+plot_all(x,y1,xtest,mu(:,1),var(:,1),params.g.z,params.task{1}.z,'y1');
+axis([-11 11 -4 4]);
+plot_all(x,y2,xtest,mu(:,2),var(:,2),params.g.z,params.task{2}.z,'y2');
+axis([-11 11 -4 4]);
+plot_all(x,y1,xtest,mu_g(:,1),var_g(:,1),params.g.z,params.task{1}.z,'y1 by g');
+axis([-11 11 -4 4]);
+
+% figure;
+% semilogy(1:numel(elbo),elbo);
+% ylabel('elbo')
+% xlabel('iteration')
+% title(['elbo vs. iteration, lrate = ' num2str(cf.lrate_z)])
+% disp('elbo = ')
+% disp(elbo(end))
+
+%% independent gpsvi
+cf.covfunc = 'covSEard';
[…]

### 2014-02-18 src/gpsvi_predict.m

@@ -1,0 +1,38 @@
+function [mu var] = gpsvi_predict(covfunc, loghyper, m, S, z, xstar)
+% Makes predictions with gpsvi
+% z are the inducing point lcoations
+% xstar are the test-pints
+
+Kmm = feval(covfunc, loghyper, z);
+Lmm = jit_chol(Kmm,3);
+Kmminv = invChol(Lmm);
+
+if size(xstar,1) > 2000
+  nbatch = 10;
+else
+  nbatch = 1;
+end
+bsize = size(xstar,1)/nbatch;
+mu = [];
+var= [];
+for i=1:nbatch
+  pos = (i-1)*bsize + 1;
+  if i < nbatch
+    bxstar = xstar(pos:i*bsize,:);
+  else
+    bxstar = xstar(pos:end,:);
+  end
+  Kss = feval(covfunc, loghyper, bxstar, 'diag');
+  Kms = feval(covfunc, loghyper, z, bxstar);
+  Ksm = Kms';
+  mu = [mu; Ksm*(Kmminv*m)];
+
+  % we can also compute full covariance at a higher cost
+  % diag(Ksm * kmminv * S * Kmmonv *Kms) 
+  var_1 =  sum(Kms.*(Kmminv*S*Kmminv*Kms),1)';
+  var_2 =  sum(Kms.*(Kmminv*Kms),1)';   
+  var = [var; var_1 + Kss - var_2];
+end
+
+return;
+

### 2014-02-18 src/ssvi_elbo.m

@@ -1,0 +1,84 @@
+function [elbo,dloghyp,dz] = ssvi_elbo(x,y,params,cf)
+%SSVI_ELBO [elbo,dloghyp,dz] = ssvi_elbo(x,y,params,cf)
+%   
+% Lowerbound as a function of the hyperparameters (and inducing inputs)
+% for STRUCTURED (multiple-output) gps. Gradients are wrt to the shared
+% function g(x).
+%
+% INPUT
+%   - x : inputs (in SVI, this is a mini-batch)
+%   - y : outputs (of inputs x)
+%   - params : parameters structure
+%   - cf : cf.covfunc 
+%
+% OUTPUT
+%   - elbo, dloghyp, dz : objective value and its derivatives
+%
+% Trung Nguyen
+% 17/02/14
+
+% TODO:
+% 1) add weight w_i
+% 2) also output derivatives of independent processes h_i in here 
+z_g = params.g.z;
+m_g = params.g.m;
+S_g = params.g.S;
+P = size(y,2); % # outputs
+loghyp_g = params.g.loghyp;
+
+% contribution from g
+Kmm = feval(cf.covfunc_g, loghyp_g, z_g);
+Lmm = jit_chol(Kmm,3);
+Kmminv = invChol(Lmm);
+Knm = feval(cf.covfunc_g, loghyp_g, x, z_g);
+A = Knm*Kmminv;
+diagKnn = feval(cf.covfunc_g, loghyp_g, x, 'diag');
+lkl = 0.5*(logdetChol(Lmm) - logdet(S_g) + traceABsym(Kmminv,S_g) + m_g'*Kmminv*m_g); % kl part
+elbo = - lkl;
+for i=1:P
+  indice = params.idx(:,i);
+  xi = x(indice,:);
+  y_minus_g = y(indice,i) - A(indice,:)*m_g;
+  elbo = elbo + svi_elbo(xi,y_minus_g,params.task{i},cf.covfunc_h,[],[],[],[]);
+  ltilde = 0.5*params.task{i}.beta*sum(diagKnn(indice) - diagProd(A(indice,:),Knm(indice,:)')); % Ktilde part
+  ltrace = 0.5*params.task{i}.beta*traceABsym(S_g,A(indice,:)'*A(indice,:)); % trace(SA'A) part
+  elbo = elbo - ltilde - ltrace;
+end
+
+if nargout >= 2
[…]

### 2014-02-18 src/ssvi_elbo_g.m

@@ -1,0 +1,72 @@
+function [elbo,dloghyp,dz] = ssvi_elbo_g(x,y,params,covfunc,n_outputs)
+%SVI_ELBO [elbo,dloghyp,dz] = ssvi_elbo_g(x,y,params,covfunc,n_outputs)
+%   
+% For g, use only 1/P KL, all other thing same.
+%
+% INPUT
+%   - x : inputs (in SVI, this is a mini-batch)
+%   - y : outputs (of inputs x)
+%   - params : parameters structure
+%   - covfunc : covariance function
+%   - n_outputs : #outputs (1/P term)
+%
+% OUTPUT
+%   - elbo, dloghyp, dz : objective value and its derivatives
+%
+% Trung Nguyen
+% 17/02/14
+z = params.z;
+m = params.m;
+S = params.S;
+betaval = params.beta;
+loghyp = params.loghyp;
+N = size(x,1);
+M = size(z,1);
+Kmm = feval(covfunc, loghyp, z);
+Lmm = jit_chol(Kmm,3);
+Kmminv = invChol(Lmm);
+Knm = feval(covfunc, loghyp, x, z);
+A = Knm*Kmminv;
+
+diagKnn = feval(covfunc, loghyp, x, 'diag');
+yMinusAm = y - A*m;
+logN = -0.5*N*log(2*pi/betaval) - 0.5*betaval*sum(yMinusAm.^2); % \sum logN() part
+ltilde = 0.5*betaval*sum(diagKnn - diagProd(A,Knm')); % Ktilde part
+ltrace = 0.5*betaval*traceABsym(S,A'*A); % trace(SA'A) part
+lkl = 0.5*(logdetChol(Lmm) - logdet(S) + traceABsym(Kmminv,S) + m'*Kmminv*m - M); % kl-divergence part
+elbo = logN - ltilde - ltrace - lkl/n_outputs;
+
+if nargout >= 2     %covariance derivatives
+  dloghyp = zeros(size(params.loghyp));
+  for i=1:numel(params.loghyp)
+    % covSEard returns dK / dloghyper 
+    dKnm = feval(covfunc, loghyp, x, z, i);
+    dKmm = feval(covfunc, loghyp, z, [], i);
[…]

### 2014-02-18 src/ssvi_learn.m

@@ -1,0 +1,49 @@
+function [elbo,params] = ssvi_learn(x,y,M,cf,z0)
+%SSVI_LEARN ssvi_learn(x,y,M,cf,z0)
+%   Quick wrapper to run structured GPSVI.
+% 
+% INPUT
+%   - x,y : train inputs, train outputs
+%   - M : #inducing inputs
+%   - cf : configuration
+%
+% OUTPUT
+%   - mu, s2 : mean and variance prediction
+%   - elbo : the bound
+%   - params : learned parameters
+[N D] = size(x);
+% get observation indice for each output
+P = size(y,2);
+ytmp = y;
+observed = ~isnan(y);
+for i=1:P
+  yi = y(observed(:,i),i);
+  ytmp(~observed(:,i),i) = mean(yi);
+end
+% for each input, the mean of all outputs are used to initialize m_g
+ymean = mean(ytmp,2);
+nhyper_g = eval(feval(cf.covfunc_g));
+nhyper_h = eval(feval(cf.covfunc_h));
+params.g = init_params(x,ymean,M,nhyper_g,0,[]);
+params.g.m = zeros(size(params.g.m));
+params.g.beta = []; params.g.delta_beta = [];
+for i=1:P
+  %params.task{i} = init_params(x(observed(:,i),:),y(observed(:,i),i),M,nhyper,0,[]);
+  params.task{i} = init_params(x(observed(:,i),:),y(observed(:,i),i),M,nhyper_h,0,[]);
+  params.task{i}.m = zeros(size(params.task{i}.m));
+  params.task{i}.S = diag(1*ones(numel(params.task{i}.m),1));
+end
+elbo = zeros(numel(cf.maxiter),1);
+for i = 1 : cf.maxiter
+  idx = randperm(N, cf.nbatch);
+  params.idx = ~isnan(y(idx,:));
+  params = ssvi_update(x(idx,:),y(idx,:),params,cf,i);
+  params.idx = observed;
+  elbo(i) = ssvi_elbo(x,y,params,cf);
+end
+ 
+% disp('initial hyp .vs learned hyp (including sigma2n)')
+% fprintf('%.5f\t%.5f\n', [[loghyp;1/beta0],[params.loghyp;1/params.beta]]');
+
+end
+

### 2014-02-18 src/ssvi_predict.m

@@ -1,0 +1,18 @@
+function [mu var, mu_g, var_g] = ssvi_predict(covfunc_g, covfunc_h, params, xstar)
+%SSVI_PREDCIT [mu var, mu_g, var_g] = ssvi_predict(covfunc_g, covfunc_h, params, xstar)
+%   Makes predictions with ssvi.
+%
+g = params.g;
+[mu_g, var_g] = gpsvi_predict(covfunc_g, g.loghyp, g.m, g.S, g.z, xstar);
+P = numel(params.task);
+mu = zeros(size(xstar,1),P);
+var = mu;
+for i=1:P
+  h = params.task{i};
+  [mu(:,i), var(:,i)] = gpsvi_predict(covfunc_h, h.loghyp, h.m, h.S, h.z, xstar);
+end
+mu_g = repmat(mu_g,1,P);
+var_g = repmat(var_g,1,P);
+mu = mu_g + mu;
+var = var_g + var;
+

### 2014-02-18 src/ssvi_update.m

@@ -1,0 +1,84 @@
+function params = ssvi_update(x,y,params,cf,iter)
+%SSVI_UPDATE params = ssvi_update(x,y,params,cf,iter)
+%   
+% Update all parameters of the structured gp model.
+% This update is similar to variational EM:
+%   - first update the variational parameters of q(u_g) and q(u_i) (E-step)
+%   - then update the hyperparameters of u_g and u_i (M-step)
+%
+% We can analyze the effect of the order of optimization: do we update
+% q(u_g) and then its hyperparameters, or do we update q(u_g) and then
+% q(u_i) and then all hyperparameters?
+%
+% If x is the entire training, this becomes the (natural) variational
+% inference. 
+%
+% INPUT
+%   - x : inputs (in SVI, this is a mini-batch)
+%   - y : outputs (of inputs x)
+%   - params
+%   - cf : options for the optimization procedure
+%
+% OUTPUT
+%   - params : updated parameters
+%
+% Trung V. Nguyen
+% 17/02/14
+P = size(y,2);
+
+% some common entities
+Kmm = feval(cf.covfunc_g, params.g.loghyp, params.g.z);
+Knm = feval(cf.covfunc_g, params.g.loghyp, x, params.g.z);
+diagKnn = feval(cf.covfunc_g, params.g.loghyp, x, 'diag');
+Lmm = jit_chol(Kmm,4); clear Kmm;
+Kmminv = invChol(Lmm);
+A = Knm*Kmminv;
+Lambda = Kmminv;
+tmp = zeros(size(params.g.m));
+for i=1:P
+  betaval = params.task{i}.beta;
+  indice = params.idx(:,i);
+  Lambda = Lambda + betaval*(A(indice,:)')*A(indice,:);
+  Ai = computeKnmKmminv(cf.covfunc_h, params.task{i}.loghyp, x(indice,:), params.task{i}.z);
+  y_minus_hi = y(indice,i) - Ai*params.task{i}.m;
+  tmp = tmp + betaval*A(indice,:)'*y_minus_hi;
+end
+
+% Update for g
+m_g = params.g.m;
+S_g = params.g.S;
+Sinv = invChol(jit_chol(S_g,4));
+theta1_old = Sinv*m_g;
+theta2_old = -0.5*Sinv;
[…]

### 2014-02-18 src/svi_elbo.m

@@ -1,5 +1,5 @@
-function [logL,dloghyp,dbeta,dz] = sviELBO(x,y,params,cf,Lmm,Kmminv,Knm,A)
-%SVIELBO [fval,dloghyp,dbeta,dz] = sviELBO(x,y,params,cf,Lmm,Kmminv,Knm,A)
+function [elbo,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,Lmm,Kmminv,Knm,A)
+%SVI_ELBO [elbo,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,Lmm,Kmminv,Knm,A)
 %   
 % Lowerbound as a function of the hyperparameters (and inducing inputs) and
 % their derivatives. Note that the inducing derivatives are for the
@@ -9,15 +9,15 @@ function [logL,dloghyp,dbeta,dz] = sviELBO(x,y,params,cf,Lmm,Kmminv,Knm,A)
 %   - x : inputs (in SVI, this is a mini-batch)
 %   - y : outputs (of inputs x)
 %   - params : parameters structure
-%   - cf : cf.covfunc 
+%   - covfunc : covariance function
 %   - Lmm, Kmminv, Knm, A : [optional] previously computed values of those
 %   (see also SVIUPDATE).
 %
 % OUTPUT
-%   - logL, dloghyp, dbeta, dz : objective value and its derivatives
+%   - elbo, dloghyp, dbeta, dz : objective value and its derivatives
 %
 % Trung Nguyen
-% 11/02/14
+% 18/02/14
 z = params.z;
 m = params.m;
 S = params.S;
@@ -25,31 +25,31 @@ betaval = params.beta;
 loghyp = params.loghyp;
 N = size(x,1);
 M = size(z,1);
-Kmm = feval(cf.covfunc, loghyp, z);
+Kmm = feval(covfunc, loghyp, z);
 if isempty(Lmm)
   Lmm = jit_chol(Kmm,3);
   Kmminv = invChol(Lmm);
-  Knm = feval(cf.covfunc, loghyp, x, z);
+  Knm = feval(covfunc, loghyp, x, z);
   A = Knm*Kmminv;
 end
 
 %TODO optimize code if possible
-diagKnn = feval(cf.covfunc, loghyp, x, 'diag');
+diagKnn = feval(covfunc, loghyp, x, 'diag');
 yMinusAm = y - A*m;
[…]

### 2014-02-18 src/svi_update.m

@@ -1,5 +1,5 @@
-function params = sviUpdate(x,y,params,cf,Lmm,Kmminv)
-%SVIUPDATE params = sviUpdate(x,y,params,cf,Lmm,Kmminv)
+function params = svi_update(x,y,params,cf,covfunc,Lmm,Kmminv)
+%SVI_UPDATE params = svi_update(x,y,params,cf,covfunc,Lmm,Kmminv)
 %   
 % Update the variational parameters (of the posterior q(u|y) = N(u; m, S))
 % and the covariance hyperparameters of a standard GP regression using SVI.
@@ -20,57 +20,53 @@ function params = sviUpdate(x,y,params,cf,Lmm,Kmminv)
 %       params.loghyp : cov hyperparameters
 %       params.beta : noise hyperparameter
 %       params.delta_z,delta_beta,delta_loghyp: for SGD learning
+%   - cf : options for the optimization procedure
+%   - covfunc: 
 %   - Lmm : [optional] chol(Kmm)
 %   - Kmminv : [optional] Kmm^{-1} (must update if hyper is not fixed)
-%   - cf : options for the optimization procedure
 %
 % OUTPUT
 %   - params : updated parameters
 %
 % Trung V. Nguyen
-% 11/02/14
+% 17/02/14
 Sinv = invChol(jit_chol(params.S,4));
 theta1_old = Sinv*params.m;
 theta2_old = -0.5*Sinv;
 betaval = params.beta;
 if isempty(Kmminv)
-  Kmm = feval(cf.covfunc, params.loghyp, params.z);
+  Kmm = feval(covfunc, params.loghyp, params.z);
   Lmm = jit_chol(Kmm,4); clear Kmm;
   Kmminv = invChol(Lmm);
 end
 
-Knm = feval(cf.covfunc, params.loghyp, x, params.z);
+Knm = feval(covfunc, params.loghyp, x, params.z);
 A = Knm*Kmminv;
 Lambda = betaval*(A'*A) + Kmminv;
 
 % Stochastic natural ascent: VB-E update for m,S
 theta2 = theta2_old + cf.lrate*(-0.5*Lambda - theta2_old); clear Lambda;
 theta1 = theta1_old + cf.lrate*(betaval*A'*y - theta1_old);
-
-%TODO replace with jit_chol and invChol
-[VV DD]     = eig(theta2);
+[VV DD] = eig(theta2);
[…]

### 2014-02-18 src/test/chkgrad_ssvi_elbo.m

@@ -1,0 +1,151 @@
+function chkgrad_ssvi_elbo()
+%CHKGRAD_SSVI_ELBO chkgrad_ssvi_elbo()
+%  Check gradient of the evidence lowerbound wrt covariance hyperparameters
+%  and inducing inputs.
+%
+% See also
+%   ssvi_elbo
+
+%rng(1110, 'twister');
+
+% note on gradient:
+% 1. delbo/dg is from ssvi_elbo
+% 2. delbo/dhi is from svi_elbo except that y is discounted contribution
+% from g and beta also has contribution from g.
+% TODO: output derivatives wrt h_i directly from ssvi_elbo.
+%
+N = 100; M = 5; D = 2; P = 2;
+cf.covfunc_g = 'covSEard';
+cf.covfunc_h = 'covSEard';
+cf.use_h = false;
+nhyper_g = eval(feval(cf.covfunc_g));
+nhyper_h = eval(feval(cf.covfunc_h));
+cf.n_outputs = P;
+x = 10*rand(N,D);
+y1 = 2*rand(N,1);
+y2 = 2*rand(N,1);
+Nmiss = 10;
+randind = randperm(N,2*Nmiss);
+y1(randind(1:Nmiss)) = nan;
+y2(randind(Nmiss+1:end)) = nan;
+y = [y1,y2];
+z_g = 10*rand(M,D);
+params.idx = ~isnan(y);
+% init params for g and h_i
+params.g = init_params(x,rand(N,1),M,nhyper_g,0,z_g);
+params.task = cell(P,1);
+for i=1:P
+  z = 10*rand(M,D);
+  params.task{i} = init_params(x,rand(N,1),M,nhyper_h,0,z);
+end
+
+% test for hyperparameters of g
+theta = params.g.loghyp;
+[mygrad, delta] = gradchek(theta', @elbo_g_hyp, @delbo_g_hyp, x,y,params,cf);
+Kmm = feval(cf.covfunc_g, params.g.loghyp, z);
+myassert(mygrad,delta,rcond(Kmm),'test ssvi_elbo() for hyperparamters of g passed');
+
+% test for inducing inputs of g
+theta = params.g.z(:);
+[mygrad, delta] = gradchek(theta', @elbo_g_z, @delbo_g_z, x,y,params,cf);
+myassert(mygrad,delta,rcond(Kmm),'test ssvi_elbo() for inducing inputs of g passed');
+
+% test for m_g
[…]

### 2014-02-18 src/test/chkgrad_svi_elbo.m

@@ -1,10 +1,10 @@
-function chkGradELBO()
-%CHKGRADELBO chkGradELBO()
+function chkgrad_svi_elbo()
+%CHKGRADELBO chkgrad_svi_elbo()
 %  Check gradient of the evidence lowerbound wrt covariance hyperparameters
 %  and inducing inputs.
 %
 % See also
-%   sviELBO
+%   svi_elbo
 
 %rng(1110, 'twister');
 N = 1000; M = 50; D = 2;
@@ -24,12 +24,12 @@ delta = abs(delta);
 [valDiff,idx] = max(delta);
 percentageDiff = (valDiff*100/abs(mygrad(idx)));
 if (valDiff > 1e-5 && percentageDiff > 10)
-  fprintf('test sviELBO() failed with valDiff = %.4f\n', valDiff);
+  fprintf('test svi_elbo() failed with valDiff = %.4f\n', valDiff);
   fprintf('percentage (valDiff / gradient) = %.2f\n', percentageDiff);
-  Kmm = feval(cf.covfunc, cf.loghyp, z);  
+  Kmm = feval(cf.covfunc, params.loghyp, z);  
   fprintf('rcond = %.10f\n', rcond(Kmm));
 else
-  disp('test sviELBO() passed');
+  disp('test svi_elbo() passed');
 end
 
 % test for inducing inputs
@@ -39,12 +39,12 @@ delta = abs(delta);
 [valDiff,idx] = max(delta);
 percentageDiff = (valDiff*100/abs(mygrad(idx)));
 if (valDiff > 1e-4 && percentageDiff > 10)
-  fprintf('test sviELBO() failed with valDiff = %.4f\n', valDiff);
+  fprintf('test svi_elbo() failed with valDiff = %.4f\n', valDiff);
   fprintf('percentage (valDiff / gradient) = %.2f\n', percentageDiff);
   Kmm = feval(cf.covfunc, cf.loghyp, z);  
   fprintf('rcond = %.10f\n', rcond(Kmm));
 else
-  disp('test sviELBO() passed');
+  disp('test svi_elbo() passed');
 end
 
 end
@@ -52,25 +52,25 @@ end
 function fval = f(theta,x,y,params,cf)
   params.loghyp = theta(1:end-1)';
   params.beta = theta(end);
-  fval = sviELBO(x,y,params,cf,[],[],[],[]);
[…]

### 2014-02-18 src/gpsvi/predict_gpsvi.m
misc utility code

@@ -29,7 +29,7 @@ for i=1:nbatch
 
   %% variance of predictive distribution
   % we can also compute full covariance at a higher cost
-  % diag(Ksm * kmminv * S * Kmmonv *Kms) 
+  % diag(Ksm * kmminv * S * Kmminv *Kms) 
   var_1 =  sum(Kms.*(Kmminv*S*Kmminv*Kms),1)';
   var_2 =  sum(Kms.*(Kmminv*Kms),1)';   
   %var = var_1 + Kss - var_2;

### 2014-02-18 src/init_params.m

@@ -1,0 +1,31 @@
+function params = init_params(x,y,M,nhyper,init_kmeans,z0)
+%INIT_PARAMS params = init_params(x,y,M)
+%   Init parameters of a single augmented sparse gp.
+%
+[N D] = size(x);
+idx_z = randperm(N,M);
+if isempty(z0)
+  z0 = x(idx_z,:);
+  if init_kmeans
+    z0 = kmeans(z0, x, foptions());
+  end
+end  
+idx = randperm(N,M);
+m =  y(idx);  % shouldn't this be the y of idx_z?
+Sinv = 0.1*(1/var(y))*eye(M);
+S = inv(Sinv);
+params.m = m;
+params.S = S;
+params.z = z0;
+params.z0 = z0;
+loghyp = log(ones(nhyper,1));
+%sigma2n = var(y-mean(y),1)/10;
+beta0 = 1/0.01;
+params.beta   = beta0; % should be init using the data
+params.loghyp = loghyp;
+params.delta_hyp = zeros(size(params.loghyp));
+params.delta_beta = 0;
+params.delta_z = zeros(size(z0));
+
+end
+

### 2014-02-18 src/learn_gpsvi.m

@@ -12,48 +12,18 @@ function [mu,s2,elbo,params] = learn_gpsvi(x,y,xtest,M,cf,z0)
 %   - elbo : the bound
 %   - params : learned parameters
 [N D] = size(x);
-idx_z = randperm(N,M);
-if isempty(z0)
-  z0 = x(idx_z,:);
-  if cf.init_kmeans
-    z0 = kmeans(z0, x, foptions());
-  end
-end
-idx = randperm(N,M);
-m =  y(idx);  % shouldn't this be the y of idx_z?
-%m = y(idx_z);
-Sinv = 0.01*(1/var(y))*eye(M);
-S = inv(Sinv);
-params.m = m;
-params.S = S;
-params.z = z0;
-params.z0 = z0;
 nhyper = eval(feval(cf.covfunc));
-loghyp = log(ones(nhyper,1));
-%sigma2n = var(y-mean(y),1)/10;
-beta0 = 1/0.01;
-params.beta   = beta0; % should be init using the data
-params.loghyp = loghyp;
-params.delta_hyp = zeros(size(params.loghyp));
-params.delta_beta = 0;
-params.delta_z = zeros(size(z0));
-
+params = init_params(x,y,M,nhyper,0,z0);
 elbo = zeros(numel(cf.maxiter),1);
 for i = 1 : cf.maxiter
   idx = randperm(N, cf.nbatch);
-  params = sviUpdate(x(idx,:),y(idx),params,cf,[],[]);
-  elbo(i) = sviELBO(x,y,params,cf,[],[],[],[]);
+  params = svi_update(x(idx,:),y(idx),params,cf,cf.covfunc,[],[]);
+  elbo(i) = svi_elbo(x,y,params,cf.covfunc,[],[],[],[]);
 end
 
-Kmm = feval(cf.covfunc, params.loghyp, params.z);
-Lmm = jit_chol(Kmm,3);
-Kmminv = invChol(Lmm);
 if ~isempty(xtest)
-  [mu s2] = predict_gpsvi(Kmminv, cf.covfunc, params.loghyp, params.m, params.S, params.z, xtest);
+  [mu s2] = gpsvi_predict(cf.covfunc, params.loghyp, params.m, params.S, params.z, xtest);
 end
-
-disp('initial hyp .vs learned hyp (including sigma2n)')
-fprintf('%.5f\t%.5f\n', [[loghyp;1/beta0],[params.loghyp;1/params.beta]]');
 
 end
 

### 2014-02-18 src/meannan.m

@@ -1,0 +1,14 @@
+function m = meannan(x)
+%MEANNAN  m = meannan(x)
+%   Replacement for the mean() function where nan is regarded as the mean
+%   of all valid elements (i.e. not nan) in the same column.
+%
+
+% column wise
+for i=1:size(x,2)
+  valid = ~isnan(x(:,i));
+  x(isnan(x(:,i)),i) = mean(x(valid,i));
+end
+m = mean(x,2);
+end
+

### 2014-02-18 src/plot_all.m

@@ -10,8 +10,10 @@ function plot_all(x,y,xtest,mu,var,z0,z,titlestr)
 figure; hold on
 plot(xtest,mu,'b') % mean predictions in blue
 plotMeanAndStd(xtest,mu,2*sqrt(var),[7 7 7]/7.5);
-plot(z0, -0.1 + min(y)*ones(size(z,1), 1), 'k+', 'MarkerSize', 12);
-plot(z, -0.2 + min(y)*ones(size(z,1), 1), 'm+', 'MarkerSize', 12);
+if ~isempty(z0)
+  plot(z0, -0.1 + min(y)*ones(size(z0,1), 1), 'k+', 'MarkerSize', 12);
+  plot(z, -0.2 + min(y)*ones(size(z,1), 1), 'm+', 'MarkerSize', 12);
+end  
 plot(x,y,'.m') % data points in magenta
 ylabel('Predictive distribution');
 title(titlestr);

### 2014-02-18 src/sumbeta.m

@@ -1,0 +1,11 @@
+function s = sumbeta(params)
+%SUMBETA  s = sumbeta(params)
+%   s = \sum params.task{i}.beta
+P = size(params.task,1);
+s = 0;
+for i=1:P
+  s = s + params.task{i}.beta;  
+end
+
+end
+

### 2014-02-18 src/toyme.m

@@ -1,7 +1,7 @@
 clear all; clc; close all;
 rng(1111,'twister');
 
-CHOICE = 2;
+CHOICE = 1;
 switch CHOICE
   case 1
     % The function in SPGP 
@@ -36,7 +36,7 @@ z0 = x(randperm(size(x,1),M),:);
 [z,z0,~,mu,s2] = learn_fitc(x,y,xtest,M,z0);
 plot_all(x,y,xtest,mu,s2,z0,z,'FITC');
 axis(theaxis);
-saveas(gcf, ['results/figures/' func '-fitc.eps'],'epsc');
+%saveas(gcf, ['results/figures/' func '-fitc.eps'],'epsc');
 
 % gpsvi
 covfunc   = 'covSEard';
@@ -44,7 +44,7 @@ cf.covfunc = covfunc;
 cf.lrate     = 1e-2;
 cf.lrate_hyp = 1e-4;
 cf.lrate_beta = 1e-4;
-cf.lrate_z   = 1e-3;
+cf.lrate_z   = 1e-4;
 cf.momentum  = 0.9;
 cf.momentum_z = 0.0;
 cf.learn_z   = true;
@@ -55,12 +55,13 @@ cf.nbatch = 5;
 [mu,s2,elbo,params] = learn_gpsvi(x,y,xtest,M,cf,z0);
 plot_all(x,y,xtest,mu,s2,params.z0,params.z,['GPSVI lrate = ' num2str(cf.lrate_z)]);
 axis(theaxis);
-saveas(gcf, ['results/figures/' func '-svi-lrate' num2str(cf.lrate_z) '.eps'],'epsc');
+%saveas(gcf, ['results/figures/' func '-svi-lrate' num2str(cf.lrate_z) '.eps'],'epsc');
 
 figure;
 semilogy(1:numel(elbo),elbo);
 ylabel('elbo')
 xlabel('iteration')
 title(['elbo vs. iteration, lrate = ' num2str(cf.lrate_z)])
-saveas(gcf, ['results/figures/' func '-svi-lrate' num2str(cf.lrate_z) '-bound.eps'],'epsc');
-
+%saveas(gcf, ['results/figures/' func '-svi-lrate' num2str(cf.lrate_z) '-bound.eps'],'epsc');
+disp('elbo = ')
+disp(elbo(end))

### 2014-02-18 paper/paper.tex
fix equations

@@
 
-\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ is
+\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
 \begin{align}
@@
 \nonumber
-&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\
+&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
+&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
 \label{eq:sorprediction}
-&= \int p(\vfstar | \u) q(\u| \y) \der \u
+&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
 \end{align}
-which is the same as the prediction of a SoR model (of course with a different posterior in $q(\u | \y)$).
+where,
+\begin{align}
+\nonumber
+\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
+\nonumber
+\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
+\end{align}
+ Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.
 
@@
 \nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta (y_{in} - w_i g_n - h_{in})
+&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta_i (y_{in} - w_i g_n - h_{in})
 \Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
@@
 \nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta \tilde{k}_{gnn} \right] 
+&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta_i (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta_i \tilde{k}_{gnn} \right] 
 \Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
 \nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta \tilde{k}_{gnn}
-- \frac{1}{2} \beta \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
+&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta_i \tilde{k}_{gnn}
+- \frac{1}{2} \beta_i \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta_i (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
 &= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
@@
 where 
+%TODO \Lambda_gn depends on the output i so need a better notation
 \begin{align}
@@
 \mat{\Lambda}_{gn}
-&= \beta k_g(\Zg,\Zg)^{-1} k_g(\Zg, \x_n) k_g(\x_n, \Zg) k_g(\Zg,\Zg)^{-1} \\
+&= \beta_i k_g(\Zg,\Zg)^{-1} k_g(\Zg, \x_n) k_g(\x_n, \Zg) k_g(\Zg,\Zg)^{-1} \\
 \mat{\Lambda}_{in}
-&= \beta k_i(\Zi,\Zi)^{-1} k_i(\Zi, \x_n) k_i(\x_n, \Zg) k_i(\Zg,\Zg)^{-1}.
+&= \beta_i k_i(\Zi,\Zi)^{-1} k_i(\Zi, \x_n) k_i(\x_n, \Zg) k_i(\Zg,\Zg)^{-1}.
 \end{align}
@@
 \calL_g \define&
- \sum_{i=1}^P \left( \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)
- - \frac{1}{2} \beta_i \trace w_i^2 \tilde{\K}_g(\oi,\oi)
- - \frac{1}{2} \beta_i \trace w_i^2 \S_g \A_g^T \A_g 
-  \right) \\
+ \sum_{i=1}^P \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)  \\
+ &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_i^2 \tilde{\K}_g(\oi,\oi) 
+ + \beta_i \trace w_i^2 \S_g \A_g(\oi,:)^T \A_g(\oi,:) \bigg)
+ \\
  &- \frac{1}{2} \log |k_g(\Zg,\Zg) \S_g^{-1}| -\frac{1}{2} \trace k_g(\Zg,\Zg)^{-1} (\m_g \m_g^T + \S_g) ,
@@
 \begin{align}
-\deriv{\calL_g}{\m_i}
+\deriv{\calL_i}{\m_i}
 & = \beta_i \A_i^T (\y_i - w_i \A_g(\oi,:)^T \m_g) - \left[k_i(\Zi,\Zi)^{-1} +  \beta_i \A_i^T \A_i \right] \m_i \\
-\deriv{\calL_g}{\S_g} 
+\deriv{\calL_i}{\S_i} 
 &= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \left[ k_i(\Zi,\Zi)^{-1} + \beta_i \A_i^T \A_i \right].
@@
 \begin{align}
-p(\fstar | \y, \x_*) = \Normal(\fstar; w_i \mu_{g*} + \mu_{i*}, s_{g*} + s_{i*}). 
+p(\fstar | \y, \x_*) = \Normal(\fstar; w_i \mu_{g*} + \mu_{i*}, w_i^2 s_{g*} + s_{i*}). 
 \end{align}

### 2014-02-20 src/computeKnmKmminv.m
add the weight w_i

@@ -1,5 +1,5 @@
-function [A,Knm,Kmminv] = computeKnmKmminv(covfunc,loghyp,x,z)
-%COMPUTEKNMKMMINV [A,Knm,Kmminv] = computeKnmKmminv(covfunc,loghyp,x,z)
+function [A,Knm,Kmminv,Lmm,Kmm] = computeKnmKmminv(covfunc,loghyp,x,z)
+%COMPUTEKNMKMMINV [A,Knm,Kmminv,Lmm,Kmm] = computeKnmKmminv(covfunc,loghyp,x,z)
 %   Compute the term A = k(x,z)k(z,z)^{-1} which is commonly encountered.
 %
 Kmm = feval(covfunc, loghyp, z);

### 2014-02-20 src/ssvi_elbo.m

@@ -1,5 +1,5 @@
-function [elbo,dloghyp,dz] = ssvi_elbo(x,y,params,cf)
-%SSVI_ELBO [elbo,dloghyp,dz] = ssvi_elbo(x,y,params,cf)
+function [elbo,dloghyp,dw,dz] = ssvi_elbo(x,y,params,cf)
+%SSVI_ELBO [elbo,dloghyp,dw,dz] = ssvi_elbo(x,y,params,cf)
 %   
 % Lowerbound as a function of the hyperparameters (and inducing inputs)
 % for STRUCTURED (multiple-output) gps. Gradients are wrt to the shared
@@ -15,11 +15,8 @@ function [elbo,dloghyp,dz] = ssvi_elbo(x,y,params,cf)
 %   - elbo, dloghyp, dz : objective value and its derivatives
 %
 % Trung Nguyen
-% 17/02/14
+% 21/02/14
 
-% TODO:
-% 1) add weight w_i
-% 2) also output derivatives of independent processes h_i in here 
 z_g = params.g.z;
 m_g = params.g.m;
 S_g = params.g.S;
@@ -27,41 +24,67 @@ P = size(y,2); % # outputs
 loghyp_g = params.g.loghyp;
 
 % contribution from g
-Kmm = feval(cf.covfunc_g, loghyp_g, z_g);
-Lmm = jit_chol(Kmm,3);
-Kmminv = invChol(Lmm);
-Knm = feval(cf.covfunc_g, loghyp_g, x, z_g);
-A = Knm*Kmminv;
+[A,Knm,Kmminv,Lmm] = computeKnmKmminv(cf.covfunc_g, loghyp_g, x, z_g);
 diagKnn = feval(cf.covfunc_g, loghyp_g, x, 'diag');
 lkl = 0.5*(logdetChol(Lmm) - logdet(S_g) + traceABsym(Kmminv,S_g) + m_g'*Kmminv*m_g); % kl part
+ltilde = zeros(P,1);
+ltrace = ltilde;
 elbo = - lkl;
 for i=1:P
   indice = params.idx(:,i);
+  betaval = params.task{i}.beta;
   xi = x(indice,:);
-  y_minus_g = y(indice,i) - A(indice,:)*m_g;
+  w = params.w(i); w2 = w*w;
+  y_minus_g = y(indice,i) - w*A(indice,:)*m_g;
   elbo = elbo + svi_elbo(xi,y_minus_g,params.task{i},cf.covfunc_h,[],[],[],[]);
[…]

### 2014-02-20 src/ssvi_elbo_g.m

@@ -18,21 +18,17 @@ function [elbo,dloghyp,dz] = ssvi_elbo_g(x,y,params,covfunc,n_outputs)
 z = params.z;
 m = params.m;
 S = params.S;
+w = params.w; w2 = w*w;
 betaval = params.beta;
 loghyp = params.loghyp;
 N = size(x,1);
 M = size(z,1);
-Kmm = feval(covfunc, loghyp, z);
-Lmm = jit_chol(Kmm,3);
-Kmminv = invChol(Lmm);
-Knm = feval(covfunc, loghyp, x, z);
-A = Knm*Kmminv;
-
+[A,Knm,Kmminv,Lmm,Kmm] = computeKnmKmminv(covfunc,loghyp,x,z);
 diagKnn = feval(covfunc, loghyp, x, 'diag');
-yMinusAm = y - A*m;
+yMinusAm = y - w*A*m;
 logN = -0.5*N*log(2*pi/betaval) - 0.5*betaval*sum(yMinusAm.^2); % \sum logN() part
-ltilde = 0.5*betaval*sum(diagKnn - diagProd(A,Knm')); % Ktilde part
-ltrace = 0.5*betaval*traceABsym(S,A'*A); % trace(SA'A) part
+ltilde = 0.5*betaval*w2*sum(diagKnn - diagProd(A,Knm')); % Ktilde part
+ltrace = 0.5*betaval*w2*traceABsym(S,A'*A); % trace(SA'A) part
 lkl = 0.5*(logdetChol(Lmm) - logdet(S) + traceABsym(Kmminv,S) + m'*Kmminv*m - M); % kl-divergence part
 elbo = logN - ltilde - ltrace - lkl/n_outputs;
 
@@ -44,9 +40,9 @@ if nargout >= 2     %covariance derivatives
     dKmm = feval(covfunc, loghyp, z, [], i);
     dKnn = feval(covfunc, loghyp, x, 'diag', i);
     dA = (dKnm - A*dKmm)*Kmminv;
-    dlogN = betaval*(yMinusAm'*dA)*m;
-    dTilde = 0.5*betaval*(sum(dKnn) - sum(diagProd(A,dKnm')) - sum(diagProd(dA,Knm')));
-    dTrace = betaval*traceAB(A*S,dA');
+    dlogN = betaval*w*(yMinusAm'*dA)*m;
[…]

### 2014-02-20 src/ssvi_learn.m

@@ -33,6 +33,7 @@ for i=1:P
   params.task{i}.m = zeros(size(params.task{i}.m));
 %  params.task{i}.S = diag(1*ones(numel(params.task{i}.m),1));
 end
+params.w = ones(P,1);
 elbo = zeros(numel(cf.maxiter),1);
 for i = 1 : cf.maxiter
   idx = randperm(N, cf.nbatch);

### 2014-02-20 src/ssvi_update.m

@@ -30,12 +30,13 @@ P = size(y,2);
 Lambda = Kmminv;
 tmp = zeros(size(params.g.m));
 for i=1:P
+  w = params.w(i); w2 = w*w;
   betaval = params.task{i}.beta;
   indice = params.idx(:,i);
-  Lambda = Lambda + betaval*(A(indice,:)')*A(indice,:);
+  Lambda = Lambda + betaval*w2*(A(indice,:)')*A(indice,:);
   Ai = computeKnmKmminv(cf.covfunc_h, params.task{i}.loghyp, x(indice,:), params.task{i}.z);
   y_minus_hi = y(indice,i) - Ai*params.task{i}.m;
-  tmp = tmp + betaval*A(indice,:)'*y_minus_hi;
+  tmp = tmp + betaval*w*A(indice,:)'*y_minus_hi;
 end
 
 Sinv = invChol(jit_chol(params.g.S,4));
@@ -58,14 +59,15 @@ end
 % Update h_i
 diagKnn = feval(cf.covfunc_g, params.g.loghyp, x, 'diag');
 if iter >= 0
-  A = computeKnmKmminv(cf.covfunc_g, params.g.loghyp, x, params.g.z);
+  [A,Knm] = computeKnmKmminv(cf.covfunc_g, params.g.loghyp, x, params.g.z);
   for i=1:P
     indice = params.idx(:,i);
-    y_discounted = y(indice,i) - A(indice,:)*params.g.m;
-    % DON'T FORGET w_i
-    params.task{i} = svi_update(x(indice,:),y_discounted,params.task{i},cf,cf.covfunc_h,[],[]);
+    w = params.w(i); w2 = w*w;
+    y_minus_g = y(indice,i) - w*A(indice,:)*params.g.m;
+    params.task{i} = svi_update(x(indice,:),y_minus_g,params.task{i},cf,cf.covfunc_h,[],[]);
     % contribution from g
-    dbeta_g = -0.5*sum(diagKnn(indice)) - 0.5*traceABsym(params.g.S,A(indice,:)'*A(indice,:));
+    dbeta_g = -0.5*w2*sum(diagKnn(indice)-diagProd(A(indice,:),Knm(indice,:)'));
+    dbeta_g = dbeta_g - 0.5*w2*traceABsym(params.g.S,A(indice,:)'*A(indice,:));
     params.task{i}.delta_beta = params.task{i}.delta_beta + cf.lrate_beta*dbeta_g;
[…]

### 2014-02-20 src/test/chkgrad_ssvi_elbo.m

@@ -14,7 +14,7 @@ function chkgrad_ssvi_elbo()
 % from g and beta also has contribution from g.
 % TODO: output derivatives wrt h_i directly from ssvi_elbo.
 %
-N = 100; M = 5; D = 2; P = 2;
+N = 200; M = 10; D = 2; P = 2;
 cf.covfunc_g = 'covSEard';
 cf.covfunc_h = 'covSEard';
 cf.use_h = false;
@@ -30,7 +30,6 @@ y1(randind(1:Nmiss)) = nan;
 y2(randind(Nmiss+1:end)) = nan;
 y = [y1,y2];
 z_g = 10*rand(M,D);
-params.idx = ~isnan(y);
 % init params for g and h_i
 params.g = init_params(x,rand(N,1),M,nhyper_g,0,z_g);
 params.task = cell(P,1);
@@ -38,71 +37,63 @@ for i=1:P
   z = 10*rand(M,D);
   params.task{i} = init_params(x,rand(N,1),M,nhyper_h,0,z);
 end
-
-% test for hyperparameters of g
-theta = params.g.loghyp;
-[mygrad, delta] = gradchek(theta', @elbo_g_hyp, @delbo_g_hyp, x,y,params,cf);
+params.idx = ~isnan(y);
+params.w = 0.1+rand(P,1);
 Kmm = feval(cf.covfunc_g, params.g.loghyp, z);
-myassert(mygrad,delta,rcond(Kmm),'test ssvi_elbo() for hyperparamters of g passed');
 
 % test for inducing inputs of g
-theta = params.g.z(:);
-[mygrad, delta] = gradchek(theta', @elbo_g_z, @delbo_g_z, x,y,params,cf);
-myassert(mygrad,delta,rcond(Kmm),'test ssvi_elbo() for inducing inputs of g passed');
+theta = [params.g.loghyp; params.w; params.g.z(:)];
+[mygrad, delta] = gradchek(theta', @elbo_g, @delbo_g, x,y,params,cf);
+assert_helper(mygrad,delta,rcond(Kmm),'test ssvi_elbo() for parameters of g passed');
 
-% test for m_g
-% theta = params.g.m;
-% [mygrad,delta] = gradchek(theta', @elbo_g_m, @delbo_g_m, x,y,params,cf);
-% myassert(mygrad,delta,rcond(Kmm),'test ssvi_elbo() for m_g passed');
+% test for m_g,S_g
[…]

### 2014-02-21 src/svi_elbo.m
refactor for svigp: separate update for variational parameters and hyperparameters

@@ -1,5 +1,5 @@
-function [elbo,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,Lmm,Kmminv,Knm,A)
-%SVI_ELBO [elbo,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,Lmm,Kmminv,Knm,A)
+function [elbo,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,A,Knm,Kmminv,Lmm)
+%SVI_ELBO [elbo,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,A,Knm,Kmminv,Lmm)
 %   
 % Lowerbound as a function of the hyperparameters (and inducing inputs) and
 % their derivatives. Note that the inducing derivatives are for the
@@ -10,8 +10,8 @@ function [elbo,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,Lmm,Kmminv,Knm,A)
 %   - y : outputs (of inputs x)
 %   - params : parameters structure
 %   - covfunc : covariance function
-%   - Lmm, Kmminv, Knm, A : [optional] previously computed values of those
-%   (see also SVIUPDATE).
+%   - Lmm, Kmminv, Knm, A : [optional] previously saved computation
+%   (see also SVI_UPDATE).
 %
 % OUTPUT
 %   - elbo, dloghyp, dbeta, dz : objective value and its derivatives
@@ -25,15 +25,11 @@ betaval = params.beta;
 loghyp = params.loghyp;
 N = size(x,1);
 M = size(z,1);
-Kmm = feval(covfunc, loghyp, z);
-if isempty(Lmm)
-  Lmm = jit_chol(Kmm,3);
-  Kmminv = invChol(Lmm);
-  Knm = feval(covfunc, loghyp, x, z);
-  A = Knm*Kmminv;
+if isempty(A)
+  [A,Knm,Kmminv,Lmm,Kmm] = computeKnmKmminv(covfunc, loghyp, x, z);
+else
+  Kmm = feval(covfunc, loghyp, z);
 end
-
-%TODO optimize code if possible
 diagKnn = feval(covfunc, loghyp, x, 'diag');
 yMinusAm = y - A*m;
 logN = -0.5*N*log(2*pi/betaval) - 0.5*betaval*sum(yMinusAm.^2); % \sum logN() part

### 2014-02-21 src/svi_learn.m

@@ -1,4 +1,4 @@
-function [mu,s2,elbo,params] = learn_gpsvi(x,y,xtest,M,cf,z0)
+function [mu,s2,elbo,params] = svi_learn(x,y,xtest,M,cf,z0)
 %LEARN_GPSVI learn_gpsvi(x,y,xtest,M,cf,z0)
 %   Quick wrapper to run GPSVI.
 % 
@@ -17,7 +17,16 @@ params = init_params(x,y,M,nhyper,0,z0);
 elbo = zeros(numel(cf.maxiter),1);
 for i = 1 : cf.maxiter
   idx = randperm(N, cf.nbatch);
-  params = svi_update(x(idx,:),y(idx),params,cf,cf.covfunc,[],[]);
+  xi = x(idx,:); yi = y(idx);
+  [params,A,Knm,Kmminv,Lmm] = svi_update(xi,yi,params,cf,cf.covfunc);
+  if cf.learn_z
+    [~,dloghyp,dbeta,dz] = svi_elbo(xi,yi,params,cf.covfunc,A,Knm,Kmminv,Lmm);
+    params = stochastic_update(params,cf,dloghyp,dbeta,dz);
+  else
+    [~,dloghyp,dbeta] = svi_elbo(xi,yi,params,cf.covfunc,A,Knm,Kmminv,Lmm);
+    params = stochastic_update(params,cf,dloghyp,dbeta,[]);
+  end
+  % compute elbo using updated parameters
   elbo(i) = svi_elbo(x,y,params,cf.covfunc,[],[],[],[]);
 end
 

### 2014-02-21 src/svi_update.m

@@ -1,11 +1,9 @@
-function params = svi_update(x,y,params,cf,covfunc,Lmm,Kmminv)
-%SVI_UPDATE params = svi_update(x,y,params,cf,covfunc,Lmm,Kmminv)
+function [params,A,Knm,Kmminv,Lmm] = svi_update(x,y,params,cf,covfunc)
+%SVI_UPDATE [params,A,Knm,Kmminv,Lmm] = svi_update(x,y,params,cf,covfunc)
 %   
-% Update the variational parameters (of the posterior q(u|y) = N(u; m, S))
-% and the covariance hyperparameters of a standard GP regression using SVI.
-% This update is composed of a VB-update (of the variational parameter using
-% natural gradient) and a standard VB-update of the covaraince
-% hyperparamters.
+% Update the variational parameters (of the posterior q(u|y) = N(u; m, S)).
+% 
+% This update corresponds to the E-step in a variational EM procedure.
 %
 % If x is the entire training, this becomes the (natural) variational
 % inference. 
@@ -21,46 +19,26 @@ function params = svi_update(x,y,params,cf,covfunc,Lmm,Kmminv)
 %       params.beta : noise hyperparameter
 %       params.delta_z,delta_beta,delta_loghyp: for SGD learning
 %   - cf : options for the optimization procedure
-%   - covfunc: 
-%   - Lmm : [optional] chol(Kmm)
-%   - Kmminv : [optional] Kmm^{-1} (must update if hyper is not fixed)
+%   - covfunc: covariance function
 %
 % OUTPUT
 %   - params : updated parameters
+%   - A,Knm,Kmminv,Lmm: saved computation
 %
 % Trung V. Nguyen
-% 17/02/14
+% 20/02/14
+betaval = params.beta;
+[A,Knm,Kmminv,Lmm] = computeKnmKmminv(covfunc, params.loghyp, x, params.z);
+Lambda = betaval*(A'*A) + Kmminv;
 Sinv = invChol(jit_chol(params.S,4));
 theta1_old = Sinv*params.m;
 theta2_old = -0.5*Sinv;
-betaval = params.beta;
-if isempty(Kmminv)
-  Kmm = feval(covfunc, params.loghyp, params.z);
-  Lmm = jit_chol(Kmm,4); clear Kmm;
-  Kmminv = invChol(Lmm);
-end
[…]

### 2014-02-21 src/svi_update_hyper.m

@@ -1,0 +1,34 @@
+function params = svi_update_hyper(x,y,params,cf,covfunc,A,Knm,Kmminv,Lmm)
+%SVI_UPDATE_HYPER params = svi_update_hyper(x,y,params,cf,covfunc,A,Knm,Kmminv,Lmm)
+%   
+% Update the hyperparameters (including the inducing inputs) of a standard
+% GP regression using stochastic variational inference.
+% This update is the M-step in the variational EM procedure.
+%
+% INPUT
+%   - x : inputs (in SVI, this is a mini-batch)
+%   - y : outputs (of inputs x)
+%   - params : structure containing variational parameters (m,S), inducing
+%   inputs, and hyperparameters
+%       params.m, params.S : variatinal parameters of the posterior
+%       params.z : inducing inputs
+%       params.loghyp : cov hyperparameters
+%       params.beta : noise hyperparameter
+%       params.delta_z,delta_beta,delta_loghyp: for SGD learning
+%   - cf : options for the optimization procedure
+%   - covfunc: 
+%   - Lmm,Kmminv,Knm,A : saved computation
+%
+% OUTPUT
+%   - params : updated parameters
+%
+% Trung V. Nguyen
+% 20/02/14
+if cf.learn_z
+  [logL,dloghyp,dbeta,dz] = svi_elbo(x,y,params,covfunc,A,Knm,Kmminv,Lmm);
+  params = stochastic_update(params,cf,dloghyp,dbeta,dz);
+else
+  [logL,dloghyp,dbeta] = svi_elbo(x,y,params,covfunc,A,Knm,Kmminv,Lmm);
+  params = stochastic_update(params,cf,dloghyp,dbeta,[]);
+end
+

### 2014-02-21 src/test/chkgrad_svi_elbo.m

@@ -7,7 +7,7 @@ function chkgrad_svi_elbo()
 %   svi_elbo
 
 %rng(1110, 'twister');
-N = 1000; M = 50; D = 2;
+N = 500; M = 20; D = 2;
 x = 10*rand(N,D);
 y = 2*rand(N,1);
 cf.covfunc = 'covSEard';
@@ -17,60 +17,41 @@ L = rand(M,M);
 params.S = L'*L;
 params.beta = 1e-2;
 params.loghyp = [log(rand(D,1)); 1+rand];
-% test for hyperparameters
-theta = [params.loghyp; params.beta];
+Kmm = feval(cf.covfunc, params.loghyp, params.z);  
+rconditional = rcond(Kmm);
+
+theta = [params.loghyp; params.beta; params.z(:)];
 [mygrad, delta] = gradchek(theta', @f, @grad, x,y,params,cf);
-delta = abs(delta);
-[valDiff,idx] = max(delta);
-percentageDiff = (valDiff*100/abs(mygrad(idx)));
-if (valDiff > 1e-5 && percentageDiff > 10)
-  fprintf('test svi_elbo() failed with valDiff = %.4f\n', valDiff);
-  fprintf('percentage (valDiff / gradient) = %.2f\n', percentageDiff);
-  Kmm = feval(cf.covfunc, params.loghyp, z);  
-  fprintf('rcond = %.10f\n', rcond(Kmm));
-else
-  disp('test svi_elbo() passed');
-end
-
-% test for inducing inputs
-theta = params.z(:);
-[mygrad, delta] = gradchek(theta', @fz, @gradz, x,y,params,cf);
-delta = abs(delta);
-[valDiff,idx] = max(delta);
-percentageDiff = (valDiff*100/abs(mygrad(idx)));
-if (valDiff > 1e-4 && percentageDiff > 10)
-  fprintf('test svi_elbo() failed with valDiff = %.4f\n', valDiff);
-  fprintf('percentage (valDiff / gradient) = %.2f\n', percentageDiff);
-  Kmm = feval(cf.covfunc, cf.loghyp, z);  
-  fprintf('rcond = %.10f\n', rcond(Kmm));
-else
-  disp('test svi_elbo() passed');
-end
-
+assert_helper(mygrad,delta,rconditional, 'gradient check for svi_elbo() passed');
 end
 
 function fval = f(theta,x,y,params,cf)
[…]

### 2014-02-21 src/ssvi_elbo.m
refactor: seperate learning of variational and hyperparameters in structured gpsvi

@@ -1,5 +1,5 @@
-function [elbo,dloghyp,dw,dz] = ssvi_elbo(x,y,params,cf)
-%SSVI_ELBO [elbo,dloghyp,dw,dz] = ssvi_elbo(x,y,params,cf)
+function [elbo,dloghyp,dw,dz] = ssvi_elbo(x,y,params,cf,A,Knm,Kmminv,Lmm)
+%SSVI_ELBO [elbo,dloghyp,dw,dz] = ssvi_elbo(x,y,params,cf,A,Knm,Kmminv,Lmm)
 %   
 % Lowerbound as a function of the hyperparameters (and inducing inputs)
 % for STRUCTURED (multiple-output) gps. Gradients are wrt to the shared
@@ -10,9 +10,10 @@ function [elbo,dloghyp,dw,dz] = ssvi_elbo(x,y,params,cf)
 %   - y : outputs (of inputs x)
 %   - params : parameters structure
 %   - cf : cf.covfunc 
+%   - A,Knm,Kmminv,Lmm: saved computation (covariance function of g)
 %
 % OUTPUT
-%   - elbo, dloghyp, dz : objective value and its derivatives
+%   - elbo, dloghyp, dw, dz : objective value and its derivatives
 %
 % Trung Nguyen
 % 21/02/14
@@ -22,9 +23,10 @@ m_g = params.g.m;
 S_g = params.g.S;
 P = size(y,2); % # outputs
 loghyp_g = params.g.loghyp;
-
 % contribution from g
-[A,Knm,Kmminv,Lmm] = computeKnmKmminv(cf.covfunc_g, loghyp_g, x, z_g);
+if nargin == 4 || isempty(A)
+  [A,Knm,Kmminv,Lmm] = computeKnmKmminv(cf.covfunc_g, loghyp_g, x, z_g);
+end  
 diagKnn = feval(cf.covfunc_g, loghyp_g, x, 'diag');
 lkl = 0.5*(logdetChol(Lmm) - logdet(S_g) + traceABsym(Kmminv,S_g) + m_g'*Kmminv*m_g); % kl part
 ltilde = zeros(P,1);

### 2014-02-21 src/ssvi_learn.m

@@ -28,19 +28,83 @@ params.g = init_params(x,ymean,M,nhyper_g,0,[]);
 params.g.m = zeros(size(params.g.m));
 params.g.beta = []; params.g.delta_beta = [];
 for i=1:P
-  %params.task{i} = init_params(x(observed(:,i),:),y(observed(:,i),i),M,nhyper,0,[]);
-  params.task{i} = init_params(x(observed(:,i),:),y(observed(:,i),i),M,nhyper_h,0,[]);
+  %params.task{i} = init_params(x(observed(:,i),:),y(observed(:,i),i),M,nhyper_h,0,[]);
+  % this allows inducing inputs to appear in even unobserved region
+  yi = y(:,i); yi(isnan(yi)) = 0;
+  params.task{i} = init_params(x,yi,M,nhyper_h,0,[]);
   params.task{i}.m = zeros(size(params.task{i}.m));
 %  params.task{i}.S = diag(1*ones(numel(params.task{i}.m),1));
 end
 params.w = ones(P,1);
+params.delta_w = zeros(P,1);
 elbo = zeros(numel(cf.maxiter),1);
-for i = 1 : cf.maxiter
+for iter = 1:cf.maxiter
   idx = randperm(N, cf.nbatch);
   params.idx = ~isnan(y(idx,:));
-  params = ssvi_update(x(idx,:),y(idx,:),params,cf,i);
+  xbatch = x(idx,:); ybatch = y(idx,:);
+  % variational parameters of g
+  [params,A,Knm,Kmminv,Lmm] = ssvi_update_g(xbatch,ybatch,params,cf);
+  % variational of h_i 
+  A_g = computeKnmKmminv(cf.covfunc_g, params.g.loghyp, xbatch, params.g.z);
+  for i=1:P
+    indice = params.idx(:,i);
+    w = params.w(i);
+    y_minus_g = ybatch(indice,i) - w*A_g(indice,:)*params.g.m;
+    params.task{i}= svi_update(xbatch(indice,:),y_minus_g,params.task{i},cf,cf.covfunc_h);
+  end
+  
+  % hyperparameters of g
+  if cf.learn_z
+    [~,dloghyp,dw,dz] = ssvi_elbo(xbatch,ybatch,params,cf,A,Knm,Kmminv,Lmm);
+    params.g = stochastic_update(params.g,cf,dloghyp,[],dz);
+  else
[…]

### 2014-02-21 src/ssvi_predict.m

@@ -11,8 +11,11 @@ for i=1:P
   h = params.task{i};
   [mu(:,i), var(:,i)] = gpsvi_predict(covfunc_h, h.loghyp, h.m, h.S, h.z, xstar);
 end
-mu_g = repmat(mu_g,1,P);
-var_g = repmat(var_g,1,P);
-mu = mu_g + mu;
-var = var_g + var;
+%mu_g = repmat(mu_g,1,P);
+%var_g = repmat(var_g,1,P);
+for i=1:P
+  w = params.w(i);
+  mu(:,i) = w*mu_g + mu(:,i);
+  var(:,i) = w*w*var_g + var(:,i);
+end
 

### 2014-02-21 src/ssvi_update.m

@@ -1,77 +1,0 @@
-function params = ssvi_update(x,y,params,cf,iter)
-%SSVI_UPDATE params = ssvi_update(x,y,params,cf,iter)
-%   
-% Update all parameters of the structured gp model.
-% This update is similar to variational EM:
-%   - first update the variational parameters of q(u_g) and q(u_i) (E-step)
-%   - then update the hyperparameters of u_g and u_i (M-step)
-%
-% We can analyze the effect of the order of optimization: do we update
-% q(u_g) and then its hyperparameters, or do we update q(u_g) and then
-% q(u_i) and then all hyperparameters?
-%
-% If x is the entire training, this becomes the (natural) variational
-% inference. 
-%
-% INPUT
-%   - x : inputs (in SVI, this is a mini-batch)
-%   - y : outputs (of inputs x)
-%   - params
-%   - cf : options for the optimization procedure
-%
-% OUTPUT
-%   - params : updated parameters
-%
-% Trung V. Nguyen
-% 20/02/14
-P = size(y,2);
-% Update for g
-[A,~,Kmminv] = computeKnmKmminv(cf.covfunc_g, params.g.loghyp, x, params.g.z);
-Lambda = Kmminv;
-tmp = zeros(size(params.g.m));
-for i=1:P
-  w = params.w(i); w2 = w*w;
-  betaval = params.task{i}.beta;
-  indice = params.idx(:,i);
-  Lambda = Lambda + betaval*w2*(A(indice,:)')*A(indice,:);
-  Ai = computeKnmKmminv(cf.covfunc_h, params.task{i}.loghyp, x(indice,:), params.task{i}.z);
-  y_minus_hi = y(indice,i) - Ai*params.task{i}.m;
-  tmp = tmp + betaval*w*A(indice,:)'*y_minus_hi;
-end
-
-Sinv = invChol(jit_chol(params.g.S,4));
-theta1_old = Sinv*params.g.m;
-theta2_old = -0.5*Sinv;
-theta2 = theta2_old + cf.lrate*(-0.5*Lambda - theta2_old);
-theta1 = theta1_old + cf.lrate*(tmp - theta1_old);
-[VV DD] = eig(theta2);
-invTheta2 = VV*diag(1./diag(DD))*VV';
-params.g.S = - 0.5*invTheta2;
[…]

### 2014-02-21 src/ssvi_update_g.m

@@ -1,0 +1,44 @@
+function [params,A,Knm,Kmminv,Lmm] = ssvi_update_g(x,y,params,cf)
+%SSVI_UPDATE_G [params,A,Knm,Kmminv,Lmm] = ssvi_update_g(x,y,params,cf)
+%   
+% Update the variational parameters of the shared process g in a structured
+% gp model.
+%
+% INPUT
+%   - x : inputs (in SVI, this is a mini-batch)
+%   - y : outputs (of inputs x)
+%   - params
+%   - cf : options for the optimization procedure
+%
+% OUTPUT
+%   - params : updated parameters
+%   - A,Knm,Kmminv,Lmm : saved computation (covariance of g)
+%
+% Trung V. Nguyen
+% 20/02/14
+P = size(y,2);
+[A,Knm,Kmminv,Lmm] = computeKnmKmminv(cf.covfunc_g, params.g.loghyp, x, params.g.z);
+Lambda = Kmminv;
+tmp = zeros(size(params.g.m));
+for i=1:P
+  w = params.w(i); w2 = w*w;
+  betaval = params.task{i}.beta;
+  indice = params.idx(:,i);
+  Lambda = Lambda + betaval*w2*(A(indice,:)')*A(indice,:);
+  Ai = computeKnmKmminv(cf.covfunc_h, params.task{i}.loghyp, x(indice,:), params.task{i}.z);
+  y_minus_hi = y(indice,i) - Ai*params.task{i}.m;
+  tmp = tmp + betaval*w*A(indice,:)'*y_minus_hi;
+end
+
+Sinv = invChol(jit_chol(params.g.S,4));
+theta1_old = Sinv*params.g.m;
+theta2_old = -0.5*Sinv;
+theta2 = theta2_old + cf.lrate*(-0.5*Lambda - theta2_old);
+theta1 = theta1_old + cf.lrate*(tmp - theta1_old);
+[VV DD] = eig(theta2);
+invTheta2 = VV*diag(1./diag(DD))*VV';
+params.g.S = - 0.5*invTheta2;
+params.g.m = params.g.S*theta1;
+
+end
+

### 2014-02-21 src/svi_elbo.m

@@ -25,7 +25,7 @@ betaval = params.beta;
 loghyp = params.loghyp;
 N = size(x,1);
 M = size(z,1);
-if isempty(A)
+if nargin == 4 || isempty(A)
   [A,Knm,Kmminv,Lmm,Kmm] = computeKnmKmminv(covfunc, loghyp, x, z);
 else
   Kmm = feval(covfunc, loghyp, z);

### 2014-02-21 src/test/chkgrad_ssvi_elbo.m

@@ -118,7 +118,7 @@ diagKnn = feval(cf.covfunc_g, params.g.loghyp, x, 'diag');
 S_g = params.g.S;
 w = params.w(i); w2 = w*w;
 y_minus_g = y(indice,i) - w*A*params.g.m;
-[~,dloghyp,dbeta,dz] = svi_elbo(x(indice,:),y_minus_g,params.task{i},cf.covfunc_h,[],[]);
+[~,dloghyp,dbeta,dz] = svi_elbo(x(indice,:),y_minus_g,params.task{i},cf.covfunc_h);
 dbeta_g = -0.5*w2*sum(diagKnn(indice)-diagProd(A,Knm'));
 dbeta_g = dbeta_g - 0.5*w2*traceABsym(S_g,(A')*A);
 dbeta = dbeta + dbeta_g;

### 2014-02-21 src/test/chkgrad_svi_elbo.m

@@ -30,7 +30,7 @@ nhyp = numel(params.loghyp);
 params.loghyp = theta(1:nhyp)';
 params.beta = theta(nhyp+1);
 params.z = reshape(theta(nhyp+2:end),numel(params.m),[]);
-fval = svi_elbo(x,y,params,cf.covfunc,[],[],[],[]);
+fval = svi_elbo(x,y,params,cf.covfunc);
 end
 
 function g = grad(theta,x,y,params,cf)
@@ -38,7 +38,7 @@ nhyp = numel(params.loghyp);
 params.loghyp = theta(1:nhyp)';
 params.beta = theta(nhyp+1);
 params.z = reshape(theta(nhyp+2:end),numel(params.m),[]);
-[~,dloghyp,dbeta,dz] = svi_elbo(x,y,params,cf.covfunc,[],[],[],[]);
+[~,dloghyp,dbeta,dz] = svi_elbo(x,y,params,cf.covfunc);
 g = [dloghyp; dbeta; dz(:)]';
 end
 

### 2014-02-21 paper/paper.tex
small fix in equations

@@
 \deriv{\calL_g}{\m_g}
-& = \sum_{i=1}^P \beta_i w_i \A_g(\oi,:)^T (\y_i - \A_i \m_i) - \left[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i \A_g(\oi,:)^T \A_g(\oi,:) \right] \m_g \\
+& = \sum_{i=1}^P \beta_i w_i \A_g(\oi,:)^T (\y_i - \A_i \m_i) - \left[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i w_i^2 \A_g(\oi,:)^T \A_g(\oi,:) \right] \m_g \\
 \deriv{\calL_g}{\S_g} 
-&= \frac{1}{2} \S_g^{-1} - \frac{1}{2} \left[ k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i \A_g(\oi,:)^T \A_g(\oi,:) \right].
+&= \frac{1}{2} \S_g^{-1} - \frac{1}{2} \left[ k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i w_i^2 \A_g(\oi,:)^T \A_g(\oi,:) \right].
 \end{align}
@@
 %\end{linenomath}
+
+\section{Toy Experiments}
+The first toy experiment uses two identical outputs which are noisy version of the same version plus some noise: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
+In this case the shared function $g(x)$ is $sin(x)$ and the independent processes $h_1(x)$ and $h_2(x)$ are just white noises.
+Each output has missing values in one region of the input space.
+The predictive distributions of the multiple-gp model compared to that of independent gps are shown in \ref{fig4}.
+
+\noindent The second toy experiment uses similar setting as the first one, except that now $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$. 
+In this case the shared function is still $g(x) = sin(x)$, but the weights should be opposite i.e. $w_1 = 1$ and $w_2 = 1$.
+The learning procedure was indeed able to recovered this relation and learned that $w_1 = 1.2$ and $w_2 = -1.3$.
+The predictive distributions are shown in \ref{fig5}.
+
+\begin{figure*}
+\centering
+\begin{tabular}{cc}
+\includegraphics[scale=0.5]{figures/ssvi-y1.eps} &
+\includegraphics[scale=0.5]{figures/ssvi-svi1.eps} \\
+\includegraphics[scale=0.5]{figures/ssvi-y2.eps} &
+\includegraphics[scale=0.5]{figures/ssvi-svi2.eps} \\
+\multicolumn{2}{c}{\includegraphics[scale=0.5]{figures/ssvi-y1byg.eps} }
+\end{tabular}
+\label{fig4}
+\caption{Predictive distributions of the multipe-output gps (left column) and independent gps (right column) for the first toy example.
+The predictive distribution by $g(x)$ for $y_1(x)$ is shown in the bottom figure.}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\begin{tabular}{cc}
+\includegraphics[scale=0.5]{figures/ssvi2-y1.eps} &
+\includegraphics[scale=0.5]{figures/ssvi2-svi1.eps} \\
+\includegraphics[scale=0.5]{figures/ssvi2-y2.eps} &
+\includegraphics[scale=0.5]{figures/ssvi2-svi2.eps} \\
+\multicolumn{2}{c}{\includegraphics[scale=0.5]{figures/ssvi-y1byg.eps} }
+\end{tabular}
+\label{fig5}
+\caption{Predictive distributions of the multipe-output gps (left column) and independent gps (right column) for the second toy experiment.
+The predictive distribution by $g(x)$ for $y_1(x)$ is shown in the bottom figure.}
+\end{figure*}
 

### 2014-03-04 src/gpsvi_predict.m
new scripts

@@ -12,7 +12,7 @@ if size(xstar,1) > 2000
 else
   nbatch = 1;
 end
-bsize = size(xstar,1)/nbatch;
+bsize = ceil(size(xstar,1)/nbatch);
 mu = [];
 var= [];
 for i=1:nbatch

### 2014-03-04 src/init_params.m

@@ -2,7 +2,7 @@ function params = init_params(x,y,M,nhyper,init_kmeans,z0)
 %INIT_PARAMS params = init_params(x,y,M)
 %   Init parameters of a single augmented sparse gp.
 %
-y(isnan(y)) = 0; %mean(y(~isnan(y))) or 0? for missing data
+y(isnan(y)) = mean(y(~isnan(y))); %mean(y(~isnan(y))) or 0? for missing data
 [N D] = size(x);
 idx_z = randperm(N,M);
 if isempty(z0)

### 2014-03-04 src/scripts/jura.m

@@ -2,7 +2,7 @@
 clear all; clc; %close all;
 rng(1110,'twister');
 
-[x,y,xtest,ytest] = read_juraCd();
+[x,y,xtest,ytest] = load_data('data/jura','juraCd');
 x = x(1:259,:); y = y(1:259,1);
 ytest = ytest(:,1);
 y = log(y);
@@ -28,9 +28,9 @@ plot(Ms,mae_mean,'x-','MarkerSize',15);
 errorbar(Ms,mae_mean,mae_std);
 xlabel('number of inducing points');
 ylabel('mae');
-title('MAE by SPGP (averaged over 10 runs)');
-saveas(gcf, 'jura-fitc.eps','epsc');
-save('jura-fitc.mat','maes');
+title('fitc, mae vs. inducing inputs');
+saveas(gcf, 'juraCu-fitc.eps','epsc');
+save('juraCu-fitc.mat','maes');
 
 %% standard gp
 model = standard_gp([],x,y,xtest,[],false);
@@ -45,16 +45,16 @@ cf.covfunc = covfunc;
 cf.lrate     = 1e-2;
 cf.lrate_hyp = 1e-5;
 cf.lrate_beta = 1e-4;
-cf.lrate_z   = 1e-4;
+cf.lrate_z   = 1e-6;
 cf.momentum  = 0.9;
 cf.momentum_z = 0.9;
 cf.learn_z   = true;
 cf.init_kmeans = true;
 cf.maxiter = 1000;
-cf.nbatch = 100;
+cf.nbatch = 50;
 
-M = 100;
-%z0 = x(randperm(size(x,1),M),:);
+M = 150;
+z0 = x(randperm(size(x,1),M),:);
 [mu,s2,elbo,params] = svi_learn(x,y,xtest,M,cf,[]);
 mu = exp(mu*ystd + ymean);
 disp('mae = ')
@@ -74,23 +74,23 @@ disp(mean(abs(ytest - mu)));
 %     disp(elbo(end))
 %   end
 % end
-
+% 
 % mae_mean = mean(maes,2);
 % mae_std = 2*std(maes,0,2);
 % plot(Ms,mae_mean,'x-','MarkerSize',15);
 % errorbar(Ms,mae_mean,mae_std);
 % xlabel('number of inducing points');
 % ylabel('mae');
-% title('svigp fixed z, mae vs. inducing inputs');
-% saveas(gcf, 'jura-svigp-batch100.eps','epsc');
[…]

### 2014-03-04 src/scripts/ms.m

@@ -12,7 +12,7 @@ LM = LinearModel.fit(x,y(:,1));
 ypred = LM.predict(xtest);
 
 % gp model
-GM = standard_gp([],x,y(:,1),xtest,[],true);
+GM = standard_gp([],x,y(:,1),xtest,[],false);
 disp('lengthscales')
 disp(exp(GM.hyp.cov));
 
@@ -28,7 +28,6 @@ LM = LinearModel.fit(x,y(:,2));
 disp('prediction for Pulse2')
 ypred = LM.predict(xtest);
 
-gp
 GM = standard_gp([],x,y(:,2),xtest,[],false);
 disp('lengthscales')
 disp(exp(GM.hyp.cov));

### 2014-03-04 src/scripts/sarcos.m

@@ -1,0 +1,97 @@
+clear all; clc;
+load sarcos_inv
+load sarcos_inv_test
+fmae = @(ytrue,ypred) mean(abs(ypred-ytrue(:,1)));
+frmse = @(ytrue,ypred) sqrt(mean((ypred-ytrue(:,1)).^2));
+
+x = sarcos_inv(:,1:21);
+y = sarcos_inv(:,22:end);
+xtest = sarcos_inv_test(:,1:21);
+ytest = sarcos_inv_test(:,22:end);
+outputs = [2,3,4,7];
+y = y(:,outputs);
+ytest = ytest(:,outputs);
+[x,xmean,xstd] = standardize(x,[],[]);
+xtest = standardize(xtest,xmean,xstd);
+[y,ymean,ystd] = standardize(y,[],[]);
+clear sarcos_inv
+clear sarcos_inv_test
+
+%% Exploratory analysis of sarcos
+correlation = zeros(7,7);
+for i=1:7
+  for j=1:7
+    correlation(i,j) = corr(y(:,i),y(:,j));
+  end
+end
+
+disp('correlation: ')
+correlation
+
+outputs = [2,3,4,7];
+for i=1:numel(outputs)-1
+  for j=i+1:numel(outputs)
+    figure; scatter(y(:,i),y(:,j)); title(['output ' num2str(i) ' and ' num2str(j)']);
+  end
+end
+
+%% independent models - fitc
+rng(1110,'twister');
+M = 200;
+runs = 5;
+maes = zeros(runs,size(y,2));
+rmses = maes;
+smses = maes;
+for n=1:runs
+  for i=1:size(y,2)
+    [z,z0,~,mu,s2] = learn_fitc(x,y(:,i),xtest,M,[]);
+    mu = mu*ystd(i)+ymean(i);
+    disp('mae    rmse')
+    fprintf('%.4f   %.4f\n', fmae(ytest(:,i),mu), frmse(ytest(:,i),mu));
+    maes(n,i) = fmae(ytest(:,i),mu);
+    rmses(n,i) = frmse(ytest(:,i),mu);
+    smses(n,i) = mysmse(ytest(:,i),mu,ymean(i));
+  end  
+end  
+save(['sarcos-fitc-M' num2str(M)],'maes','rmses','smses');
+disp('average mae, rmse, smse')
+disp(mean(maes))
+disp(mean(rmses))
[…]

### 2014-03-04 src/scripts/slfm_fx.m

@@ -1,0 +1,65 @@
+clear all; %clc; %close all;
+rng(1110,'twister');
+
+% pre-process y
+[x,y,xtest,ytest,y0] = read_fx();
+[y,ymean,ystd] = standardize(y,[],[]);
+
+%% ssvi
+cf.covfunc_g  = 'covSEard';
+cf.lrate      = 1e-2;
+cf.momentum   = 0.9;
+cf.lrate_hyp  = 1e-5;
+cf.lrate_beta = 1e-4;
+cf.momentum_w = 0.9;
+cf.lrate_w    = 1e-5;
+cf.learn_z    = false;
+cf.momentum_z = 0.0;
+cf.lrate_z    = 1e-4;
+cf.init_kmeans = false;
+cf.maxiter = 500;
+cf.nbatch = 200;
+cf.beta = 1/0.1;
+cf.initz = 'random';
+
+M = 100;
+Q = 1;
+par.g = cell(Q,1);
+[elbo,par] = slfm2_learn(x,y,M,par,cf);
+[mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest,size(y,2));
+mu =mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1);
+
+disp('elbo = ')
+disp(elbo(end))
+disp('learned w = ')
+disp(par.w)
+
+figure;
+%plot(1:numel(elbo),elbo);
+semilogy(1:numel(elbo),-elbo);
+ylabel('elbo')
+xlabel('iteration')
+title(['elbo vs. iteration, lrate = ' num2str(cf.lrate_z)])
+
+%%
+outputs = [4,6,9];
+titles = {'CAD','JPY','AUD'};
+xtest = cell(3,1);
+xtest{1} = (50:100)';
+xtest{2} = (100:150)';
+xtest{3} = (150:200)';
+disp('smse of CAD, JPY, AUD')
+theaxis{1} = [0 250 0.75 1.15];
+theaxis{2} = [0 250 7.8*1e-3 9.4*1e-3];
+theaxis{3} = [0 250 0.7 0.95];
+smses = zeros(3,1);
+for i=1:3
+  t = outputs(i);
+  plot_all(x,y0(:,t),x,mu(:,t),(ystd(t).^2)*var(:,t),[],[],titles{i});
+  axis(theaxis{i});
[…]

### 2014-03-04 src/scripts/slfm_jura.m

@@ -1,14 +1,13 @@
 clear all; clc; %close all;
 rng(1110,'twister');
 
-[x,y,xtest,ytest] = read_juraCd();
 % pre-process y
 y0 = y;
 y = log(y);
 [y,ymean,ystd] = standardize(y,[],[]);
 
 %M = size(x,1);
-M.g = 20; M.h = 20;
+M.g = 100; M.h = 20;
 % ssvi
 cf.covfunc_g  = 'covSEard';
 cf.covfunc_h  = 'covNoise';
@@ -23,7 +22,7 @@ cf.momentum_z = 0.0;
 cf.lrate_z    = 1e-4;
 cf.init_kmeans = true;
 cf.maxiter = 500;
-cf.nbatch = 5;
+cf.nbatch = 100;
 
 Q = 2;
 par.task = cell(size(y,2),1);
@@ -56,13 +55,13 @@ disp(elbo(end))
 disp('learned w = ')
 disp(par.w)
 
-figure; hold on;
-scatter(x(:,1),x(:,2))
-scatter(par.g{1}.z(:,1),par.g{1}.z(:,2),'xr')
-scatter(par.g{1}.z0(:,1),par.g{1}.z0(:,2),'sr')
-scatter(par.g{2}.z(:,1),par.g{2}.z(:,2),'xm')
-scatter(par.g{2}.z0(:,1),par.g{2}.z0(:,2),'sm')
-title('inducing of g')
+% figure; hold on;
+% scatter(x(:,1),x(:,2))
+% scatter(par.g{1}.z(:,1),par.g{1}.z(:,2),'xr')
+% scatter(par.g{1}.z0(:,1),par.g{1}.z0(:,2),'sr')
+% scatter(par.g{2}.z(:,1),par.g{2}.z(:,2),'xm')
+% scatter(par.g{2}.z0(:,1),par.g{2}.z0(:,2),'sm')
+% title('inducing of g')
 
 % for i=1:size(ytest,2)
 % figure; hold on;

### 2014-03-04 src/svi_learn.m

@@ -17,22 +17,25 @@ par = init_params(x,y,M,nhyper,0,z0);
 par.beta = 1/0.05;
 par.delta_beta = 0;
 elbo = zeros(numel(cf.maxiter),1);
-for i = 1 : cf.maxiter
+for iter = 1 : cf.maxiter
   idx = randperm(N, cf.nbatch);
   xi = x(idx,:); yi = y(idx);
   [par,A,Knm,Kmminv,Lmm] = svi_update(xi,yi,par,cf,cf.covfunc);
-  [~,dloghyp,dbeta] = svi_elbo(xi,yi,par,cf.covfunc,A,Knm,Kmminv,Lmm);
-  [par.loghyp,par.delta_hyp] = stochastic_update(par.loghyp,par.delta_hyp,...
-    dloghyp, cf.momentum, cf.lrate_hyp);
-  [par.beta,par.delta_beta] = stochastic_update(par.beta,par.delta_beta,...
-    dbeta, cf.momentum, cf.lrate_beta);
-  if cf.learn_z
-    [~,~,~,dz] = svi_elbo(xi,yi,par,cf.covfunc);
-    [par.z,par.delta_z] = stochastic_update(par.z, par.delta_z, dz, ...
-      cf.momentum_z, cf.lrate_z);
+  if iter > N/cf.nbatch   % fix covariance hyperparameter in the first eposch
+    [~,dloghyp,dbeta] = svi_elbo(xi,yi,par,cf.covfunc,A,Knm,Kmminv,Lmm);
+    [par.loghyp,par.delta_hyp] = stochastic_update(par.loghyp,par.delta_hyp,...
+      dloghyp, cf.momentum, cf.lrate_hyp);
+    [par.beta,par.delta_beta] = stochastic_update(par.beta,par.delta_beta,...
+      dbeta, cf.momentum, cf.lrate_beta);
+    if cf.learn_z
+      [~,~,~,dz] = svi_elbo(xi,yi,par,cf.covfunc);
+      [par.z,par.delta_z] = stochastic_update(par.z, par.delta_z, dz, ...
+        cf.momentum_z, cf.lrate_z);
+    end
   end
   % compute elbo using updated parameters
-  elbo(i) = svi_elbo(x,y,par,cf.covfunc,[],[],[],[]);
+  elbo(iter) = svi_elbo(x,y,par,cf.covfunc,[],[],[],[]);
[…]

### 2014-03-04 src/tmp/slfm2_learn.m

@@ -1,0 +1,71 @@
+function [elbo,par] = slfm2_learn(x,y,M,par,cf)
+%SLFM_LEARN slfm_learn(x,y,M,cf)
+%   Quick wrapper to run structured GPSVI.
+% 
+% INPUT
+%   - x,y : train inputs, train outputs
+%   - M : #inducing inputs
+%   - cf : configuration
+%
+% OUTPUT
+%   - elbo : the bound
+%   - params : learned parameters
+[N D] = size(x);
+% get observation indice for each output
+P = size(y,2); Q = numel(par.g);
+par.w = ones(P,Q);
+par.delta_w = zeros(P,Q);
+par.beta = cf.beta*ones(P,1);
+par.delta_beta = zeros(P,1);
+observed = ~isnan(y);
+nhyper_g = eval(feval(cf.covfunc_g));
+ytmp = y;
+for i=1:P
+  yi = y(observed(:,i),i);
+  ytmp(~observed(:,i),i) = mean(yi);
+end
+ymean = mean(ytmp,2);
+for j=1:Q
+  par.g{j} = init_params(x,ymean,M,nhyper_g,0,initz(x,M,cf.initz));
+  par.g{j}.m = zeros(size(par.g{j}.m));
+  if strcmp(cf.initz,'espaced')
+    par.g{j}.m = par.g{j}.m + 1e-4*randn(M,1); % small pertubance to break symmetry
+  end
+end
+elbo = zeros(numel(cf.maxiter),1);
+for iter = 1:cf.maxiter
+  idx = randperm(N, cf.nbatch);
+  par.idx = ~isnan(y(idx,:));
+  xbatch = x(idx,:); ybatch = y(idx,:);
+  
+  % variational parameters of g_j
+  par = slfm2_update_g(xbatch,ybatch,par,cf);
+  
+  % hyperparameters of g_j
+  [~,dbeta,dw,dloghyp] = slfm2_elbo(xbatch,ybatch,par,cf);
+  for j=1:Q
+    [par.g{j}.loghyp,par.g{j}.delta_hyp] = stochastic_update(...
+      par.g{j}.loghyp,par.g{j}.delta_hyp,dloghyp{j},cf.momentum,cf.lrate_hyp);
+  end
+  
+  % update w, beta
+  dw = reshape(cell2mat(dw),P,Q);
+  [par.w,par.delta_w] = stochastic_update(par.w,par.delta_w,dw,cf.momentum_w,cf.lrate_w);
[…]

### 2014-03-04 src/tmp/slfm2_predict.m

@@ -1,0 +1,21 @@
+function [mu, var, mu_g, var_g] = slfm2_predict(covfunc_g, par, xstar, n_outputs)
+%SLFM2_PREDCIT [mu var, mu_g, var_g] = slfm2_predict(covfunc_g, params, xstar, P)
+%   Makes predictions with ssvi.
+%
+Q = numel(par.g);
+mu_g = zeros(size(xstar,1),Q);
+var_g = mu_g;
+for j=1:Q
+  g = par.g{j};
+  [mu_g(:,j), var_g(:,j)] = gpsvi_predict(covfunc_g, g.loghyp, g.m, g.S, g.z, xstar);
+end
+mu = zeros(size(xstar,1),n_outputs);
+var = mu;
+for i=1:n_outputs
+  for j=1:Q
+    w = par.w(i,j);
+    mu(:,i) = w*mu_g(:,j) + mu(:,i);
+    var(:,i) = w*w*var_g(:,j) + var(:,i);
+  end
+end
+

### 2014-03-04 src/ssvi_learn2.m
more clean

@@ -1,95 +1,0 @@
-function [elbo,params] = ssvi_learn(x,y,M,cf,z0)
-%SSVI_LEARN ssvi_learn(x,y,M,cf,z0)
-%   Quick wrapper to run structured GPSVI.
-% 
-% INPUT
-%   - x,y : train inputs, train outputs
-%   - M : #inducing inputs
-%   - cf : configuration
-%
-% OUTPUT
-%   - mu, s2 : mean and variance prediction
-%   - elbo : the bound
-%   - params : learned parameters
-[N D] = size(x);
-% get observation indice for each output
-P = size(y,2);
-ytmp = y;
-observed = ~isnan(y);
-for i=1:P
-  yi = y(observed(:,i),i);
-  ytmp(~observed(:,i),i) = mean(yi);
-end
-% for each input, the mean of all outputs are used to initialize m_g
-ymean = mean(ytmp,2);
-nhyper_g = eval(feval(cf.covfunc_g));
-nhyper_h = eval(feval(cf.covfunc_h));
-params.g = init_params(x,ymean,M.g,nhyper_g,0,[]);
-%params.g.m = zeros(size(params.g.m));
-params.g.beta = []; params.g.delta_beta = [];
-for i=1:P
-  %params.task{i} = init_params(x(observed(:,i),:),y(observed(:,i),i),M,nhyper_h,0,[]);
-  % this allows inducing inputs to appear in even unobserved region
-  yi = y(:,i); yi(isnan(yi)) = 0;
-  params.task{i} = init_params(x,yi,M.h,nhyper_h,0,[]);
-  %params.task{i}.m = zeros(size(params.task{i}.m));
-  if strcmp(cf.covfunc_h,'covNoise')
-    params.task{i}.loghyp = log(0.01); % learning of noise is hard so set it small here
-  end
-%  params.task{i}.S = diag(1*ones(numel(params.task{i}.m),1));
-end
-params.w = ones(P,1);
-params.delta_w = zeros(P,1);
-elbo = zeros(numel(cf.maxiter),1);
-for iter = 1:cf.maxiter
-  idx = randperm(N, cf.nbatch);
-  params.idx = ~isnan(y(idx,:));
-  xbatch = x(idx,:); ybatch = y(idx,:);
-  % variational parameters of g
[…]

### 2014-03-06 paper/experiments.tex
convert paper to uai proceeding

--- 2014-02-21 paper/paper.tex
@@
-\documentclass{article} % For LaTeX2e
-\usepackage[a4paper,margin=1.5in]{geometry}
-\usepackage[mathlines]{lineno}
-\usepackage{hyperref}
-\usepackage{url}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{bm}
-\usepackage{hyperref}
-\usepackage{natbib}
-\usepackage{latexsym,amsbsy,amssymb,color,xspace, booktabs}
-
-\include{macros}
-\include{local_macros}
-\def\linenumberfont{\normalfont\small\sffamily}
-
-\title{Multiple Output Regression}
-
-\newcommand{\fix}{\marginpar{FIX}}
-\newcommand{\new}{\marginpar{NEW}}
-
-%\nipsfinalcopy % Uncomment for camera-ready version
-
-\begin{document}
-\maketitle
-\linenumbers
-
-\section{Model Description}
-We describe a multiple-output regression model.
-Suppose that we have inputs $\X = \{\x_n\}_{n=1}^N$ and outputs $\y = \{\y_n\}_{n=1}^N$, with $\x_n \in \calR^D$ and $\y_n \in \calR^P$, i.e. the output is $P$-dimensional.
-For ease of exposition, we assume that there are no missing values in any output dimension.
-The most dominant approach in GP-based multiple output regression is to compose the different outputs as some \textit{linear} mixing (combination) of some basic processes (random functions).
-Our model is also underpinned by this composition approach.
-However, to really address the issue of scalability, this model is built upon sparse GPs from the ground up.
-
-\subsection{Prior and Likelihood}
-We use a simple composition:
-\begin{align}
-f_i(\x) = w_i g(\x) + h_i(\x), \quad i = 1 \hdots P,
-\end{align}
-where each (latent) output is the sum of two functions: one scaled latent function that is shared by all outputs and one latent function unique to the output.
-This allows the outputs to be correlated via the shared function $g(\x)$ while still having their own independent processes.
-Furthermore we assume that $g(\x) \sim \GP(0, k_g(\x,\x'; \vectheta_g))$ and $h_i(\x) \sim \GP(0, k_i(\x,\x'; \vectheta_i))$, and that $h_i$ are independent GPs with their own set of hyperparameters.
-\\
-
-\noindent The auto-variance of the latent output using this composition is
-\begin{align}
-\nonumber
-cov[f_i(\x), f_i(\x')] 
-&= cov \left[ \left(w_i g(\x) + h_i(\x) \right) \left(w_{i} g(\x') + h_i(\x') \right) \right] \\
-&= w_i^2 k_g(\x, \x') + k_i(\x,\x').
-\end{align}
-\\
-\noindent The cross-covariance of the latent outputs is
-\begin{align}
-\nonumber
-cov[f_i(\x), f_{i'}(\x')] 
-&= cov \left[ \left(w_i g(\x) + h_i(\x) \right) \left(w_{i'} g(\x') + h_{i'}(\x') \right) \right] \\
-&= w_i w_{i'} k_g(\x, \x').
-\end{align}
-
-\noindent It can be seen that the induced covariance function of $f_i(\x)$ is very similar to that in the MTGP model.
-Specifically, in MTGP the task-correlation matrix is a free-form positive definite matrix $B$ whereas here it is the matrix $\w \w^T$ where $\w = [w_1, \hdots, w_P]^T$.
-However this model is more flexible compared to MTGP because each output is also influenced by its own independent process $h_i(\x)$.
-Note that we can also use more than one shared function $g(\x)$, in which case it becomes the semiparametric latent force model (SLFM), again with the addition of independent process in each output dimension.
-\\
-
-\noindent 
-\textbf{Augmented sparse GPs}
-Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
-This adds much flexibility to the model as the roles of $g(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
-Furthermore, $g(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be smaller than that of $g(\x)$.
-If more than one $g(\x)$ is used, each $g(\x)$ may capture a different pattern so they should also have separate inducing inputs. \\
-
-\noindent
-\textbf{Prior}
-\newcommand{\Zg}{\Z_g}
-\newcommand{\Zi}{\Z_i}
-Let $\g = g(\X), \h_i = h_i(\X), \h = \{\h_i \}_{i=1}^P$.
-Let $\Zg$ and $\mat{Z}^h = \{\Zi \}_{i=1}^P$ be the set of inducing inputs for $g(\x)$ and $h_i(\x)$; the corresponding inducing points $\u_g$ and $\u^h = \{\u_i \}_{i=1}^P$.
-The prior of the augmented model is given by:
-\begin{align}
-p(\g | \u_g) &= \Normal(\g; \BigMu_g, \tilde{\K}_g )\\
-p(\u_g) &= \Normal(\u_g; \vec{0}, k_g(\Zg, \Zg)) \\
-p(\h | \u^h) &= \prod_{i=1}^P \Normal(\h_i; \BigMu_i, \tilde{\K}_i)\\
-p(\u^h) &= \prod_{i=1}^P \Normal(\u_i; \vec{0}, k_i(\Zi, \Zi)),
-\end{align}
-where $\BigMu_g = k_g(\X,\Zg)k(\Zg,\Zg)^{-1}\u_g$ and $\tilde{\K}_g = k_g(\X,\X) - k(\X,\Zg)k(\Zg,\Zg)^{-1}k(\Zg,\X)$ and $\BigMu_i, \tilde{\K}_i$ are similarly defined.
-
-\noindent
-\textbf{Likelihood}
-The likelihood as usual follows the standard iid Gaussian likelihood,
-\begin{align}
-p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; w_i \g + \h_i, \beta_i^{-1} \I).
-\end{align}
-
-\subsection{Variational Inference in Sparse GPs Revisited}
-In this section we review inference in sparse GPs (as presented in Titsias).
-The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
-The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g.
-the test set).
-In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
-This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
-Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
-
-\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
-\begin{align}
-\nonumber
-p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
-\nonumber
-&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
-&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
-\label{eq:sorprediction}
-&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
-\end{align}
-where,
-\begin{align}
-\nonumber
-\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
-\nonumber
-\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
-\end{align}
- Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.
-
-\subsection{Variational Inference}
-\newcommand{\ug}{\u_g}
-\newcommand{\uh}{\u^h}
-Our goal of inference is to find the posterior $p(\g, \h, \ug, \uh | \y)$. 
-Following the previous discussion, we assume a variational distribution which factorizes as:
-\begin{align}
-\nonumber
-q(\g, \h, \ug, \uh | \y)
- &= p(\g | \ug) q(\ug | \y) \prod_{i=1}^P p(\h_i | \u_i) q(\u_i | \y) \\
- &=  q(\ug, \uh) p(\g | \ug) \prod_{i=1}^P p(\h_i | \u_i)
-\end{align}
-Since the conditionals $p(\g | \ug)$ and $p(\h_i | \u_i)$ are given, we need only to find the optimum $q(\ug, \uh) = q(\ug | \y) \prod_{i=1}^P q(\u_i | \y)$.
-Let $q(\ug | \y) = \Normal(\ug; \m_g, \S_g)$ and $q(\u_i | \y) = \Normal(\u_i; \m_i, \S_i)$. \\
-
-\noindent
-To find the optimum $q(\ug, \uh)$ we optimize the evidence lowerbound of the log marginal,
-\begin{align}
-\nonumber
-\log p(\y) \ge& \int q(\ug, \uh) \log \frac{p(\y | \ug, \uh) p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
-\nonumber
-=& \int q(\ug, \uh) \log p(\y | \ug, \uh)  \der \ug \der \uh 
-+ \int q(\ug, \uh) \log \frac{p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
-\label{eq:elbo}
-=& \int q(\ug, \uh) \log p(\y | \ug, \uh)  \der \ug \der \uh 
-- \left( \KL[q(\ug) || p(\ug)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right),
-\end{align}
-where the last equality occurs because both of $q(\ug, \uh)$ and $p(\ug, \uh)$ fully factorize.
-Since $q(\ug), q(\u_i), p(\ug), p(\u_i)$ are all multivariate Gaussian distribution, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g(\x)$ and $h_i(\x)$ use separate set of inducing inputs).
-\\
-
-\noindent To compute the above equation we first focus on the term:
-\newcommand{\llangle}{\left\langle}
-\newcommand{\rrangle}{\right\rangle}
-\begin{align}
-\nonumber
-\log p(\y | \ug, \uh)
- &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \ug, \uh)} \\
- \nonumber
-&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \ug, \uh)} \quad &\text{(Jensen's inequality)} \\
-\nonumber
-&=  \Eb{\sum_{i=1}^P \sum_{n=1}^N \log p(y_{in} | g_n, h_{in}) }_{p(\g,\h | \ug, \uh)}  \quad &\text{(factorized likelihood)} \\
-&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | g_n, h_{in}) }_{p(\g,\h_i | \ug, \u_i)} \quad &\text{(linear operation)}
-\end{align}
-Observe that this is very similar to the SVI for standard GP in Hensman et al. 
-In this multiple-output model, the outputs factorize given $\g$ and $\h$ hence the lowerbound can be seen as sum of the lowerbounds in the single-output setting.
-
-\noindent Each individual term $l_{in} = \llangle \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\g,\h_i | \ug, \u_i)}$ can be computed as follows:
-\begin{align}
-\nonumber
-l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\g | \ug) p(\h_i | \u_i) \der \g \der \h_i \\
-\nonumber
-&= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
-\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
-\Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
-\nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta_i (y_{in} - w_i g_n - h_{in})
-\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
-\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der g_n \der h_{in} \\
-\nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta_i (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta_i \tilde{k}_{gnn} \right] 
-\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
-\nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta_i \tilde{k}_{gnn}
-- \frac{1}{2} \beta_i \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta_i (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
-&= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
-\end{align}
-where $\tilde{k}_{gnn} = (\tilde{\K}_g)_{n,n}, \tilde{k}_{inn} = (\tilde{\K}_i)_{n,n}, \mu_{gn} = (\Mu_g)_n, \text{ and } \mu_{in} = (\Mu_i)_n$.
-Notice that the above expression is very similar to the single-output case.
-Hence our derivation here can be seen as the \textit{generalization} of SVI from the standard GP regression to multiple-output regression.
-
-\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying similar integration we get,
-\begin{align}
-\nonumber
-\log p(\y)
-\ge& \sum_{i=1}^P \sum_{n=1}^N
- \left( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta^{-1})
-         - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn})
-         - \frac{1}{2} \trace \left( w_i^2 \S_g \mat{\Lambda}_{gn} + \S_i \mat{\Lambda}_{in} \right)
-\right) \\
-&- \left( \KL[q(\ug) || p(\ug)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right) \define \calL,
-\end{align}
-where 
-%TODO \Lambda_gn depends on the output i so need a better notation
-\begin{align}
-\tilde{\mu}_{in}
-&= w_i k_g(\x_n, \Zg)k_g(\Zg,\Zg)^{-1}\m_g + k_i(\x_n, \Zi)k_i(\Zi,\Zi)^{-1}\m_i \\
-\mat{\Lambda}_{gn}
-&= \beta_i k_g(\Zg,\Zg)^{-1} k_g(\Zg, \x_n) k_g(\x_n, \Zg) k_g(\Zg,\Zg)^{-1} \\
-\mat{\Lambda}_{in}
-&= \beta_i k_i(\Zi,\Zi)^{-1} k_i(\Zi, \x_n) k_i(\x_n, \Zg) k_i(\Zg,\Zg)^{-1}.
-\end{align}
-
-\noindent Again, this clearly shows that this model generalizes the standard GP regression. This can be verified by setting $P = 1$, $w_i = 1$ and $h_i(\x) = 0$ to recover the bound in Hensman et al.
-Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.
-
-\subsubsection{Variational Parameters Derivatives}
-\newcommand{\oi}{\vec{o}_i}
-% some notation
-Before diving into the details, we first define the indexing operator of a matrix: $\B(\vec{r},\vec{c})$ extracts the submatrix in rows $\vec{r}$ and columns $\vec{c}$ of $\B$.
-To index all columns we use $\B(\vec{r},:)$ and similarly for all rows $\B(:,\vec{c})$.
-Readers familiar with this operator will recognize that this is the MATLAB indexing operator.
-
-The optimal posteriors are found by setting the gradients of the lowerbound $\calL$ wrt to the parameters of $q(\ug)$ and $q(\u_i)$.
-Recall that in the model, different outputs can be observed at different inputs (i.e. the case of missing values).
-Let $\oi$ be the indice of the observed inputs of the output dimension $i$.
-We denote its set of observed inputs and targets as: $\X_i = \X(\oi,:)$ and $\y_i = y_{i}(\oi)$.
-
-\noindent As a function of the parameters of $q(\ug)$, the lowerbound $\calL$ is:
-\begin{align}
-\nonumber
-\calL_g \define&
- \sum_{i=1}^P \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)  \\
- &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_i^2 \tilde{\K}_g(\oi,\oi) 
- + \beta_i \trace w_i^2 \S_g \A_g(\oi,:)^T \A_g(\oi,:) \bigg)
- \\
- &- \frac{1}{2} \log |k_g(\Zg,\Zg) \S_g^{-1}| -\frac{1}{2} \trace k_g(\Zg,\Zg)^{-1} (\m_g \m_g^T + \S_g) ,
-\end{align}
-where $\A_g = k_g(\X,\Z_g)k_g(\Z_g,\Z_g)^{-1}$, which gives $\A_g(\oi,:) = k_g(\X_i,\Zg) k_g(\Z_g,\Z_g)^{-1}$, and  
-$\A_i = k_i(\X_i,\Z_i)k_i(\Z_i,\Z_i)^{-1}$. \\
-
-%------------------------------------------
-% derivatives of q(u_g)
-\noindent The derivatives of $\calL_q$ wrt $\m_g$ and $\S_g$ are given by:
-\begin{align}
-\deriv{\calL_g}{\m_g}
-& = \sum_{i=1}^P \beta_i w_i \A_g(\oi,:)^T (\y_i - \A_i \m_i) - \left[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i w_i^2 \A_g(\oi,:)^T \A_g(\oi,:) \right] \m_g \\
-\deriv{\calL_g}{\S_g} 
-&= \frac{1}{2} \S_g^{-1} - \frac{1}{2} \left[ k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i w_i^2 \A_g(\oi,:)^T \A_g(\oi,:) \right].
-\end{align}
-
-%-------------------------------------------
-%  derivatives of q(u_i)
-\noindent As a function of the parameters of $q(\u_i)$, the lowerbound $\calL$ is:
-\begin{align}
-\nonumber
-\calL_i \define&
- \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)
- - \frac{1}{2} \beta_i \trace \tilde{\K}_i(\oi,\oi)
- - \frac{1}{2} \beta_i \trace \S_i \A_i^T \A_i
- \\
-  &- \frac{1}{2} \log |k_i(\Zi,\Zi) \S_i^{-1}| -\frac{1}{2} \trace k_i(\Zi,\Zi)^{-1} (\m_i \m_i^T + \S_i) ,
-\end{align}
-
-\noindent The derivatives of $\calL_i$ wrt $\m_i$ and $\S_i$ are given by:
-\begin{align}
-\deriv{\calL_i}{\m_i}
-& = \beta_i \A_i^T (\y_i - w_i \A_g(\oi,:)^T \m_g) - \left[k_i(\Zi,\Zi)^{-1} +  \beta_i \A_i^T \A_i \right] \m_i \\
-\deriv{\calL_i}{\S_i} 
-&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \left[ k_i(\Zi,\Zi)^{-1} + \beta_i \A_i^T \A_i \right].
-\end{align}
-
-% comment on computation
-\noindent It can be seen that the derivatives of the parameters of $q(\u_i)$ only involve the observations of the output dimension $i$.
-The derivatives of the parameters of $q(\u_g)$ involve the observations across all output dimensions but decompose as a sum of contributions from individual outputs.
-Therefore, computation of the derivatives (and hence the update equations) can be distributed or parallelized easily.
-This attractive property allows the model to scale to a very large number of outputs.
-
-%\subsubsection{Update Equations}
-% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
-\subsection{Prediction}
-The predictive distribution of the $i$-th output for a test input $\x_*$ is 
-\begin{align}
-p(\fstar | \y, \x_*) = \int \Normal(\fstar; w_i g_* + h_{i*}, 0) p(g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der g_* \der h_{i*},
-\end{align}
-where $p(g_* | \y, \x_*) = \Normal(g_*; \mu_{g*}, s_{g*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu_{i*}, s_{i*})$ are the predictive distributions of the sparse GPs as given in \ref{eq:sorprediction}.
-Therefore we have:
-\begin{align}
-p(\fstar | \y, \x_*) = \Normal(\fstar; w_i \mu_{g*} + \mu_{i*}, w_i^2 s_{g*} + s_{i*}). 
-\end{align}
-
-%\begin{linenomath}
-%\begin{align}
-%\sum_{i,h} \lambda_i \lambda_h cov [f(\x, \x')]
-%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\x_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\x_h - \vs_j') \\
-%\end{align}
-%\end{linenomath}
-
 \section{Toy Experiments}
@@
 \end{figure*}
-
-\section{Appendix}
-Here we derive the gradients of the lower bound wrt the  hyperparameters.
-As a template, consider the lower bound in Hensman et al (as a function of the hyperparameters):
-\begin{align}
-\nonumber
-\calL
-=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I)
- - \frac{1}{2} \beta \trace \tilde{\K}
- - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
-&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right) \\ \nonumber
-=& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
- - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
- - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
-&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4},
-\end{align}
-where $\A = \K_{NM} \K_{MM}^{-1}$.
-Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.
-
-\subsection{Derivative of the Noise Hyperparameter}
-The derivative of the noise hyperparameter $\beta$ is easily computed as:
-\begin{align}
-\deriv{\calL}{\beta} = \frac{N}{2\beta} - \frac{1}{2} (\y - \A\m)^T (\y - \A\m) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
-\end{align}
-\subsection{Derivatives of the Covariance Hyperparameters}  
-To simplify the math, we utilize the matrix $\A$ defined above.
-Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
-\begin{align}
-\deriv{\A}{t} = \left(\deriv{\K_{NM}}{t} - \A \deriv{\K_{MM}}{t}\right)\K_{MM}^{-1}.
-\end{align}
-The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
-\begin{align}
-\deriv{\calL_1}{t} &= \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
-\deriv{\calL_2}{t} &= \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
-\deriv{\calL_3}{t} &= \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
-\deriv{\calL_4}{t} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
-\end{align}
-The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.
-
-\subsection{Derivatives of the Inducing Inputs}
-To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\z_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
-Hence the derivative wrt a single dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\deriv{ \calL}{t}$.
-
-%Since $MD$ parameters are needed for the inducing inputs, it appears that the derivatives of all inducing inputs would require $\calO(MD \times M^3)$ in computation.
-%However, this complexity can actually be reduced to $\calO(DM^3)$ using the following lemma. \\
-%
-%\noindent \textbf{Lemma} Let $A, B$ be two matrices of size $N \times M$ and $M \times N$, respectively.
-Furthermore, $B$ has the property that only one of its rows or columns is non-zero.
-The complexity of $\trace(AB)$ is only $\calO(N)$.
-\\
-%
-%\noindent \textbf{Proof} Let $m <= M$ be the non-zero row of $B$. We have
-%\begin{align}
-%\trace (AB) = \sum_{i=1}^N \sum_{j=1}^M A_{ij} B_{ji} = \sum_{i=1}^N A_{im} B_{mi},
-%\end{align}
-%which clearly takes $\calO(N)$. It is easy to see that the lemma also holds when $B$ is symmetric and only one of its row and the corresponding column is non-zero. \\
-%
-%\noindent To exploit the property in the Lemma, we re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):
-
-\noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
-\begin{align}
-\nonumber
-% dL1
-\deriv{\calL_1}{t}
- &= \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
-&= \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} 
--\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t} \\
-%dL2
-\deriv{\calL_2}{t}
-&= - \beta \trace \A^T \deriv{\K_{NM}}{t}
- + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}  \\
-% dL3
-\deriv{\calL_3}{t}
-&= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
- - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}\\
-% dL4
-\deriv{\calL_4}{t}
- &= \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t}
-  - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
-\end{align}
-From the above equations we get,
-\begin{align}
-\deriv{\calL}{t} = \trace \D_1 \deriv{\K_{NM}}{t} + \trace \D_2 \deriv{ \K_{MM}}{t},
-\end{align}
-where 
-\begin{align}
-\D_1 =& \beta \K_{MM}^{-1} \m (\y - \A\m)^T
- + \beta \A^T
- - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
-\D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
- - \frac{1}{2} \beta \A^T \A
-  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
-  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
-\end{align}
-Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
-The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two $\trace$ operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
-This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.
-
-\subsection{Analysis on Learning of the Inducing Inputs}
-In this section we present some analysis on learning of the inducing inputs under the stochastic variational inference procedure on some toy problems.
-We use three real-valued truth functions of scalar inputs: the first one was used in Snelson et al, the second one is a function $f(x) = \sin(x) + \cos(x)$, and the third one is a sample function generated from a GP with squared exponential with ARD covariance function with a lengthscale of 2 and signal variance of 1.
-
-\noindent \paragraph{Experimental Settings} We use gpsvi to optimize for \textit{all} of the parameters in the model, i.e.
-the variational parameters of the posterior, the covariance hyperparameters, the noise hyperparameter, and the inducing inputs.
-For the hyperparameters, we use a momentum of $0.9$ and a fixed learn rate of $1e-5$. 
-For the variational parameters, we use a learn rate of $0.01$ (no momentum was used).
-We vary the learn rate of the inducing inputs from $1e{-2}$ to $1e{-6}$.
-Using momentum for the inducing inputs does not seem to have much effects.
-The maximum number of iterations is 500, the batch size 5, and the number of inducing values 10.
-
-% effect of learn rate on inducing inputs
-\noindent \paragraph{Effects of the Learn Rates }
-In Figure \ref{fig1}, \ref{fig2}, and \ref{fig3}, we show the lower bounds, predictive distributions, and the learned inducing inputs using stochastic variational inference with varying learn rates for the inducing inputs.
-We also show the results with FITC for comparison.
-It can be seen from the figures that, as typical in stochastic optimization, the behavior of GPSVI is very sensitive to the learn rate.
-In particular, when the rate is slow e.g. $1e{-5}$, almost no progress was made in learning the inducing inputs.
-However, when the rate increases to e.g. $1e{-3}$, the inducing inputs are spread out more evenly in the input space compared to the initial locations.
-Although not shown the in figures, when the learn rate is too large, e.g. $1e{-02}$, the inducing inputs may spread to regions far outside of the training intervals.
-An example of this phenomenon can be observed in the top left figure in Figure \ref{fig1}.
-The learn rates also affect the evidence lower bound which seems to wiggle more compared to not learning the inducing inputs.
-The predictive distributions are sensible particularly because the toy functions exhibit sparsity and are easy to learn.  \\
-
-\noindent The results of this qualitative analysis do not seem to suggest that learning of the inducing inputs does not work in GPSVI.
-Its effectiveness perhaps rests upon empirical performance on real datasets.
-
-\noindent \paragraph{The Noise Parameter} It seems harder to set the initial value and the learning rate of the noise parameter compared to standard GP. 
-While a large range of values of noise can be used in standard gp or fitc (e.g. 0.5), a very small value of noise $1e{-02}$, which is the true noise, is required for gpsvi to work well.
-It appears that an adaptive learning rate may be required for the noise parameter.
-
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05.eps} \\
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03-bound.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04-bound.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05-bound.eps} \\ 
-(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
-\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func1-fitc.eps}}
-\end{tabular}
-\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the first function.
-Top row: GPSVI with different learning rates (the rates decreases from left to right).
-Bottom row: FITC.
-Magenta dots : training points.
-Solid blue line: predictive mean.
-Grey-shaded area and solid black lines: two standard deviations of the predictive distributions.
-Black (+) crosses: initial locations of inducing inputs.
-Magenta (+) crosses: learned locations of inducing inputs.
-Middle row: the evidence lower bound of gpsvi vs.
-iteration.}
-\label{fig1}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05.eps} \\
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03-bound.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04-bound.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05-bound.eps} \\ 
-(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
-\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func2-fitc.eps}}
-\end{tabular}
-\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the second function. The legends are same as in Figure 1.}
-\label{fig2}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05.eps} \\
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03-bound.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04-bound.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05-bound.eps} \\ 
-(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
-\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func3-fitc.eps}}
-\end{tabular}
-\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the third function. The legends are same as in Figure 1.}
-\label{fig3}
-\end{figure*}
-
-\end{document}

### 2014-03-06 paper/model.tex
convert paper to uai proceeding

We describe a multiple-output regression model.
Suppose that we have inputs $\X = \{\x_n\}_{n=1}^N$ and outputs $\y = \{\y_n\}_{n=1}^N$, with $\x_n \in \calR^D$ and $\y_n \in \calR^P$, i.e. the output is $P$-dimensional.
For ease of exposition, we assume that there are no missing values in any output dimension.
The most dominant approach in GP-based multiple output regression is to compose the different outputs as some \textit{linear} mixing (combination) of some basic processes (random functions).
Our model is also underpinned by this composition approach.

\noindent Consider the semiparametric latent factor models (SLFM) \cite{teh-et-al-aistats-05}.
It assumes $Q$ basic independent processes, $g_j(\x) \sim \GP(0, k(\x,\x';\vectheta_j))$, where $\vectheta_j$ is the hyperparameters of the covariance function of $g_j(\x)$.
Each (latent) output is a weighted combination of these basic processes, i.e.
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x), \quad i = 1 \hdots P,
\end{align}
In \cite{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x) + h_i(\x), 
\end{align}
We emphasize that, although $k$ denotes covariance functions in general, each independent process can have its own covariance function.

\noindent 
\textbf{Augmented sparse GPs}
Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
This adds much flexibility to the model as each of the shared process $g_j(\x)$ can model a different pattern in the data with its own covariance function and inducing inputs. The roles of $g_j(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
Furthermore, $g_j(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be a subspace of the input.

%However, to really address the issue of scalability, this model is built upon sparse GPs from the ground up.

\subsection{Prior and Likelihood}
In this section we specify the prior and likelihood of this model. 
\newcommand{\Zj}{\Z_j}
\newcommand{\Zhi}{\Z^h_i}
Let the set of inducing inputs for $g_j(\x)$ and $h_i(\x)$ be $\Z_j$ and $\Z^h_i$, and the corresponding inducing points $\u_j$ and $\v_i$, respectively.
We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
The subscripts are $i = 1 \hdots P$, $j = 1 \hdots Q$, and $n = 1 \hdots N$.

\noindent
\textbf{Prior}
The prior of the augmented model is given by:
\begin{align}
p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)),
\end{align}
where $\BigMu_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}\u$ and $\tilde{\K}_j = k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X)$ and $\BigMu^h_i, \tilde{\K}^h_i$ are similarly defined.
To avoid notational clutter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.

\noindent
\textbf{Likelihood}
The likelihood as usual follows the standard iid Gaussian likelihood,
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I).
\end{align}

### 2014-03-06 paper/paper.tex
convert paper to uai proceeding

--- 2014-02-13 paper/paper.tex
@@
-\documentclass{article} % For LaTeX2e
-\usepackage[a4paper,margin=1.5in]{geometry}
-\usepackage[mathlines]{lineno}
+% Sample LaTeX file for creating a paper in the Morgan Kaufmannn two
+% column, 8 1/2 by 11 inch proceedings format.
+
+\documentclass[]{article}
+\usepackage{proceed2e}
 \usepackage{hyperref}
@@
 \include{local_macros}
-\def\linenumberfont{\normalfont\small\sffamily}
 
-\title{Multiple Output Regression}
+\title{Stochastic Multiple Output Gaussian Processes}
 
-\newcommand{\fix}{\marginpar{FIX}}
-\newcommand{\new}{\marginpar{NEW}}
+\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
+          % UAI  reviewing is double-blind.
 
-%\nipsfinalcopy % Uncomment for camera-ready version
+% The author names and affiliations should appear only in the accepted paper.
+%
+%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
+%alternate address.}} \\
+%Computer Science Dept. \\
+%Cranberry University\\
+%Pittsburgh, PA 15213 \\
+%\And
+%{\bf Coauthor}  \\
+%Affiliation          \\
+%Address \\
+%\And
+%{\bf Coauthor}   \\
+%Affiliation \\
+%Address    \\
+%(if needed)\\
+%}
 
 \begin{document}
+
 \maketitle
-\linenumbers
 
-\section{Model Description}
-We describe a multiple-output regression model.
-Suppose that we have inputs $\X = \{\x_n\}_{n=1}^N$ and outputs $\y = \{\y_n\}_{n=1}^N$, with $\x_n \in \calR^D$ and $\y_n \in \calR^P$, i.e. the output is $P$-dimensional.
-For ease of exposition, we assume that there are no missing values in any output dimension.
-The most dominant approach in GP-based multiple output regression is to compose the different outputs as some \textit{linear} mixing (combination) of some basic processes (random functions).
-Our model is also underpinned by this composition approach.
-However, to really address the issue of scalability, this model is built upon sparse GPs from the ground up.
+\begin{abstract}
+The Abstract paragraph 
+\end{abstract}
 
-\subsection{Prior and Likelihood}
-We use a simple composition:
-\begin{align}
-f_i(\x) = w_i g(\x) + h_i(\x), \quad i = 1 \hdots P,
-\end{align}
-where each (latent) output is the sum of two functions: one scaled latent function that is shared by all outputs and one latent function unique to the output.
-This allows the outputs to be correlated via the shared function $g(\x)$ while still having their own independent processes.
-Furthermore we assume that $g(\x) \sim \GP(0, k_g(\x,\x'; \vectheta_g))$ and $h_i(\x) \sim \GP(0, k_i(\x,\x'; \vectheta_i))$, and that $h_i$ are independent GPs with their own set of hyperparameters.
-\\
+\section{INTRODUCTION}
+\input{intro}
 
-\noindent The auto-variance of the latent output using this composition is
-\begin{align}
-\nonumber
-cov[f_i(\x), f_i(\x')] 
-&= cov \left[ \left(w_i g(\x) + h_i(\x) \right) \left(w_{i} g(\x') + h_i(\x') \right) \right] \\
-&= w_i^2 k_g(\x, \x') + k_i(\x,\x').
-\end{align}
-\\
-\noindent The cross-covariance of the latent outputs is
-\begin{align}
-\nonumber
-cov[f_i(\x), f_{i'}(\x')] 
-&= cov \left[ \left(w_i g(\x) + h_i(\x) \right) \left(w_{i'} g(\x') + h_{i'}(\x') \right) \right] \\
-&= w_i w_{i'} k_g(\x, \x').
-\end{align}
+\subsubsection{Citations in Text}
 
-\noindent It can be seen that the induced covariance function of $f_i(\x)$ is very similar to that in the MTGP model.
-Specifically, in MTGP the task-correlation matrix is a free-form positive definite matrix $B$ whereas here it is the matrix $\w \w^T$ where $\w = [w_1, \hdots, w_P]^T$.
-However this model is more flexible compared to MTGP because each output is also influenced by its own independent process $h_i(\x)$.
-Note that we can also use more than one shared function $g(\x)$, in which case it becomes the semiparametric latent force model (SLFM), again with the addition of independent process in each output dimension.
-\\
+Citations within the text should include the author's last name and
+year, e.g., (Cheesman, 1985). Reference style should follow the style
+that you are used to using, as long as the citation style is
+consistent.
 
-\noindent 
-\textbf{Augmented sparse GPs}
-Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
-This adds much flexibility to the model as the roles of $g(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
-Furthermore, $g(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be smaller than that of $g(\x)$.
-If more than one $g(\x)$ is used, each $g(\x)$ may capture a different pattern so they should also have separate inducing inputs. \\
+\section{MODEL DESCRIPTION}
+\input{model}
 
-\noindent
-\textbf{Prior}
-\newcommand{\Zg}{\Z_g}
-\newcommand{\Zi}{\Z_i}
-Let $\g = g(\X), \h_i = h_i(\X), \h = \{\h_i \}_{i=1}^P$.
-Let $\Zg$ and $\mat{Z}^h = \{\Zi \}_{i=1}^P$ be the set of inducing inputs for $g(\x)$ and $h_i(\x)$; the corresponding inducing points $\u_g$ and $\u^h = \{\u_i \}_{i=1}^P$.
-The prior of the augmented model is given by:
-\begin{align}
-p(\g | \u_g) &= \Normal(\g; \BigMu_g, \tilde{\K}_g )\\
-p(\u_g) &= \Normal(\u_g; \vec{0}, k_g(\Zg, \Zg)) \\
-p(\h | \u^h) &= \prod_{i=1}^P \Normal(\h_i; \BigMu_i, \tilde{\K}_i)\\
-p(\u^h) &= \prod_{i=1}^P \Normal(\u_i; \vec{0}, k_i(\Zi, \Zi)),
-\end{align}
-where $\BigMu_g = k_g(\X,\Zg)k(\Zg,\Zg)^{-1}\u_g$ and $\tilde{\K}_g = k_g(\X,\X) - k(\X,\Zg)k(\Zg,\Zg)^{-1}k(\Zg,\X)$ and $\BigMu_i, \tilde{\K}_i$ are similarly defined.
+\subsection{THIS SHOULD BE CAP}
 
-\noindent
-\textbf{Likelihood}
-The likelihood as usual follows the standard iid Gaussian likelihood,
-\begin{align}
-p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; w_i \g + \h_i, \beta_i^{-1} \I).
-\end{align}
+\section{INFERENCE}
+\input{inference}
 
-\subsection{Variational Inference in Sparse GPs Revisited}
-In this section we review inference in sparse GPs (as presented in Titsias).
-The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
-The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g.
-the test set).
-In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
-This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
-Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
+\section{EXPERIMENTS}
+\input{experiments}
 
-\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ is
-\begin{align}
-\nonumber
-p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
-\nonumber
-&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\
-\label{eq:sorprediction}
-&= \int p(\vfstar | \u) q(\u| \y) \der \u
-\end{align}
-which is the same as the prediction of a SoR model (of course with a different posterior in $q(\u | \y)$).
+\section{DISCUSSION}
+\input{discussion}
 
-\subsection{Variational Inference}
-\newcommand{\ug}{\u_g}
-\newcommand{\uh}{\u^h}
-Our goal of inference is to find the posterior $p(\g, \h, \ug, \uh | \y)$. 
-Following the previous discussion, we assume a variational distribution which factorizes as:
-\begin{align}
-\nonumber
-q(\g, \h, \ug, \uh | \y)
- &= p(\g | \ug) q(\ug | \y) \prod_{i=1}^P p(\h_i | \u_i) q(\u_i | \y) \\
- &=  q(\ug, \uh) p(\g | \ug) \prod_{i=1}^P p(\h_i | \u_i)
-\end{align}
-Since the conditionals $p(\g | \ug)$ and $p(\h_i | \u_i)$ are given, we need only to find the optimum $q(\ug, \uh) = q(\ug | \y) \prod_{i=1}^P q(\u_i | \y)$.
-Let $q(\ug | \y) = \Normal(\ug; \m_g, \S_g)$ and $q(\u_i | \y) = \Normal(\u_i; \m_i, \S_i)$. \\
+\subsubsection{Figures}
 
-\noindent
-To find the optimum $q(\ug, \uh)$ we optimize the evidence lowerbound of the log marginal,
-\begin{align}
-\nonumber
-\log p(\y) \ge& \int q(\ug, \uh) \log \frac{p(\y | \ug, \uh) p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
-\nonumber
-=& \int q(\ug, \uh) \log p(\y | \ug, \uh)  \der \ug \der \uh 
-+ \int q(\ug, \uh) \log \frac{p(\ug, \uh)}{q(\ug, \uh)} \der \ug \der \uh \\
-\label{eq:elbo}
-=& \int q(\ug, \uh) \log p(\y | \ug, \uh)  \der \ug \der \uh 
-- \left( \KL[q(\ug) || p(\ug)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right),
-\end{align}
-where the last equality occurs because both of $q(\ug, \uh)$ and $p(\ug, \uh)$ fully factorize.
-Since $q(\ug), q(\u_i), p(\ug), p(\u_i)$ are all multivariate Gaussian distribution, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g(\x)$ and $h_i(\x)$ use separate set of inducing inputs).
-\\
+All artwork must be centered, neat, clean, and legible. Figure number
+and caption always appear below the figure.  Leave 2 line spaces
+between the figure and the caption. The figure caption is initial caps
+and each figure numbered consecutively.
 
-\noindent To compute the above equation we first focus on the term:
-\newcommand{\llangle}{\left\langle}
-\newcommand{\rrangle}{\right\rangle}
-\begin{align}
-\nonumber
-\log p(\y | \ug, \uh)
- &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \ug, \uh)} \\
- \nonumber
-&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \ug, \uh)} \quad &\text{(Jensen's inequality)} \\
-\nonumber
-&=  \Eb{\sum_{i=1}^P \sum_{n=1}^N \log p(y_{in} | g_n, h_{in}) }_{p(\g,\h | \ug, \uh)}  \quad &\text{(factorized likelihood)} \\
-&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | g_n, h_{in}) }_{p(\g,\h_i | \ug, \u_i)} \quad &\text{(linear operation)}
-\end{align}
-Observe that this is very similar to the SVI for standard GP in Hensman et al. 
-In this multiple-output model, the outputs factorize given $\g$ and $\h$ hence the lowerbound can be seen as sum of the lowerbounds in the single-output setting.
+Make sure that the figure caption does not get separated from the
+figure. Leave extra white space at the bottom of the page rather than
+splitting the figure and figure caption.
+\begin{figure}[h]
+\vspace{1in}
+\caption{Sample Figure Caption}
+\end{figure}
 
-\noindent Each individual term $l_{in} = \llangle \log p(y_{in} | g_n, h_{in}) \rrangle_{p(\g,\h_i | \ug, \u_i)}$ can be computed as follows:
-\begin{align}
-\nonumber
-l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\g | \ug) p(\h_i | \u_i) \der \g \der \h_i \\
-\nonumber
-&= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
-\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
-\Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
-\nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta (y_{in} - w_i g_n - h_{in})
-\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
-\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der g_n \der h_{in} \\
-\nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta \tilde{k}_{gnn} \right] 
-\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
-\nonumber
-&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta \tilde{k}_{gnn}
-- \frac{1}{2} \beta \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
-&= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
-\end{align}
-where $\tilde{k}_{gnn} = (\tilde{\K}_g)_{n,n}, \tilde{k}_{inn} = (\tilde{\K}_i)_{n,n}, \mu_{gn} = (\Mu_g)_n, \text{ and } \mu_{in} = (\Mu_i)_n$.
-Notice that the above expression is very similar to the single-output case.
-Hence our derivation here can be seen as the \textit{generalization} of SVI from the standard GP regression to multiple-output regression.
+\subsubsection{Tables}
 
-\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying similar integration we get,
-\begin{align}
-\nonumber
-\log p(\y)
-\ge& \sum_{i=1}^P \sum_{n=1}^N
- \left( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta^{-1})
-         - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn})
-         - \frac{1}{2} \trace \left( w_i^2 \S_g \mat{\Lambda}_{gn} + \S_i \mat{\Lambda}_{in} \right)
-\right) \\
-&- \left( \KL[q(\ug) || p(\ug)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right) \define \calL,
-\end{align}
-where 
-\begin{align}
-\tilde{\mu}_{in}
-&= w_i k_g(\x_n, \Zg)k_g(\Zg,\Zg)^{-1}\m_g + k_i(\x_n, \Zi)k_i(\Zi,\Zi)^{-1}\m_i \\
-\mat{\Lambda}_{gn}
-&= \beta k_g(\Zg,\Zg)^{-1} k_g(\Zg, \x_n) k_g(\x_n, \Zg) k_g(\Zg,\Zg)^{-1} \\
-\mat{\Lambda}_{in}
-&= \beta k_i(\Zi,\Zi)^{-1} k_i(\Zi, \x_n) k_i(\x_n, \Zg) k_i(\Zg,\Zg)^{-1}.
-\end{align}
+\begin{table}[h]
+\caption{Sample Table Title}
+\label{sample-table}
+\begin{center}
+\begin{tabular}{ll}
+\multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION} \\
+\hline \\
+Dendrite         &Input terminal \\
+Axon             &Output terminal \\
+Soma             &Cell body (contains cell nucleus) \\
+\end{tabular}
+\end{center}
+\end{table}
 
-\noindent Again, this clearly shows that this model generalizes the standard GP regression. This can be verified by setting $P = 1$, $w_i = 1$ and $h_i(\x) = 0$ to recover the bound in Hensman et al.
-Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.
+\newpage
+\section{APPENDIX}
+Perhaps this should go to a supplementary material.
+\input{appendix}
 
-\subsubsection{Variational Parameters Derivatives}
-\newcommand{\oi}{\vec{o}_i}
-% some notation
-Before diving into the details, we first define the indexing operator of a matrix: $\B(\vec{r},\vec{c})$ extracts the submatrix in rows $\vec{r}$ and columns $\vec{c}$ of $\B$.
-To index all columns we use $\B(\vec{r},:)$ and similarly for all rows $\B(:,\vec{c})$.
-Readers familiar with this operator will recognize that this is the MATLAB indexing operator.
+%\subsubsection*{Acknowledgements}
 
-The optimal posteriors are found by setting the gradients of the lowerbound $\calL$ wrt to the parameters of $q(\ug)$ and $q(\u_i)$.
-Recall that in the model, different outputs can be observed at different inputs (i.e. the case of missing values).
-Let $\oi$ be the indice of the observed inputs of the output dimension $i$.
-We denote its set of observed inputs and targets as: $\X_i = \X(\oi,:)$ and $\y_i = y_{i}(\oi)$.
-
-\noindent As a function of the parameters of $q(\ug)$, the lowerbound $\calL$ is:
-\begin{align}
-\nonumber
-\calL_g \define&
- \sum_{i=1}^P \left( \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)
- - \frac{1}{2} \beta_i \trace w_i^2 \tilde{\K}_g(\oi,\oi)
- - \frac{1}{2} \beta_i \trace w_i^2 \S_g \A_g^T \A_g 
-  \right) \\
-  &- \frac{1}{2} \log |k_g(\Zg,\Zg) \S_g^{-1}| -\frac{1}{2} \trace k_g(\Zg,\Zg)^{-1} (\m_g \m_g^T + \S_g) ,
-\end{align}
-where $\A_g = k_g(\X,\Z_g)k_g(\Z_g,\Z_g)^{-1}$, which gives $\A_g(\oi,:) = k_g(\X_i,\Zg) k_g(\Z_g,\Z_g)^{-1}$, and  
-$\A_i = k_i(\X_i,\Z_i)k_i(\Z_i,\Z_i)^{-1}$. \\
-
-%------------------------------------------
-% derivatives of q(u_g)
-\noindent The derivatives of $\calL_q$ wrt $\m_g$ and $\S_g$ are given by:
-\begin{align}
-\deriv{\calL_g}{\m_g}
-& = \sum_{i=1}^P \beta_i w_i \A_g(\oi,:)^T (\y_i - \A_i \m_i) - \left[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i \A_g(\oi,:)^T \A_g(\oi,:) \right] \m_g \\
-\deriv{\calL_g}{\S_g} 
-&= \frac{1}{2} \S_g^{-1} - \frac{1}{2} \left[ k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i \A_g(\oi,:)^T \A_g(\oi,:) \right].
-\end{align}
-
-%-------------------------------------------
-%  derivatives of q(u_i)
-\noindent As a function of the parameters of $q(\u_i)$, the lowerbound $\calL$ is:
-\begin{align}
-\nonumber
-\calL_i \define&
- \log \Normal(\y_i; w_i \A_g(\oi,:) \m_g + \A_i \m_i, \beta_i^{-1} \I)
- - \frac{1}{2} \beta_i \trace \tilde{\K}_i(\oi,\oi)
- - \frac{1}{2} \beta_i \trace \S_i \A_i^T \A_i
- \\
-  &- \frac{1}{2} \log |k_i(\Zi,\Zi) \S_i^{-1}| -\frac{1}{2} \trace k_i(\Zi,\Zi)^{-1} (\m_i \m_i^T + \S_i) ,
-\end{align}
-
-\noindent The derivatives of $\calL_i$ wrt $\m_i$ and $\S_i$ are given by:
-\begin{align}
-\deriv{\calL_g}{\m_i}
-& = \beta_i \A_i^T (\y_i - w_i \A_g(\oi,:)^T \m_g) - \left[k_i(\Zi,\Zi)^{-1} +  \beta_i \A_i^T \A_i \right] \m_i \\
-\deriv{\calL_g}{\S_g} 
-&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \left[ k_i(\Zi,\Zi)^{-1} + \beta_i \A_i^T \A_i \right].
-\end{align}
-
-% comment on computation
-\noindent It can be seen that the derivatives of the parameters of $q(\u_i)$ only involve the observations of the output dimension $i$.
-The derivatives of the parameters of $q(\u_g)$ involve the observations across all output dimensions but decompose as a sum of contributions from individual outputs.
-Therefore, computation of the derivatives (and hence the update equations) can be distributed or parallelized easily.
-This attractive property allows the model to scale to a very large number of outputs.
-
-%\subsubsection{Update Equations}
-% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
-\subsection{Prediction}
-The predictive distribution of the $i$-th output for a test input $\x_*$ is 
-\begin{align}
-p(\fstar | \y, \x_*) = \int \Normal(\fstar; w_i g_* + h_{i*}, 0) p(g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der g_* \der h_{i*},
-\end{align}
-where $p(g_* | \y, \x_*) = \Normal(g_*; \mu_{g*}, s_{g*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu_{i*}, s_{i*})$ are the predictive distributions of the sparse GPs as given in \ref{eq:sorprediction}.
-Therefore we have:
-\begin{align}
-p(\fstar | \y, \x_*) = \Normal(\fstar; w_i \mu_{g*} + \mu_{i*}, s_{g*} + s_{i*}). 
-\end{align}
-
-%\begin{linenomath}
-%\begin{align}
-%\sum_{i,h} \lambda_i \lambda_h cov [f(\x, \x')]
-%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\x_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\x_h - \vs_j') \\
-%\end{align}
-%\end{linenomath}
-
-\section{Appendix}
-Here we derive the gradients of the lower bound wrt the  hyperparameters.
-As a template, consider the lower bound in Hensman et al (as a function of the hyperparameters):
-\begin{align}
-\nonumber
-\calL
-=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I)
- - \frac{1}{2} \beta \trace \tilde{\K}
- - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
-&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right) \\ \nonumber
-=& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
- - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
- - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
-&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4},
-\end{align}
-where $\A = \K_{NM} \K_{MM}^{-1}$.
-Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.
-
-\subsection{Derivative of the Noise Hyperparameter}
-The derivative of the noise hyperparameter $\beta$ is easily computed as:
-\begin{align}
-\deriv{\calL}{\beta} = \frac{N}{2\beta} - \frac{1}{2} (\y - \A\m)^T (\y - \A\m) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
-\end{align}
-\subsection{Derivatives of the Covariance Hyperparameters}  
-To simplify the math, we utilize the matrix $\A$ defined above.
-Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
-\begin{align}
-\deriv{\A}{t} = \left(\deriv{\K_{NM}}{t} - \A \deriv{\K_{MM}}{t}\right)\K_{MM}^{-1}.
-\end{align}
-The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
-\begin{align}
-\deriv{\calL_1}{t} &= \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
-\deriv{\calL_2}{t} &= \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
-\deriv{\calL_3}{t} &= \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
-\deriv{\calL_4}{t} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
-\end{align}
-The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.
-
-\subsection{Derivatives of the Inducing Inputs}
-To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\z_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
-Hence the derivative wrt a single dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\deriv{ \calL}{t}$.
-
-%Since $MD$ parameters are needed for the inducing inputs, it appears that the derivatives of all inducing inputs would require $\calO(MD \times M^3)$ in computation.
-%However, this complexity can actually be reduced to $\calO(DM^3)$ using the following lemma. \\
-%
-%\noindent \textbf{Lemma} Let $A, B$ be two matrices of size $N \times M$ and $M \times N$, respectively.
-Furthermore, $B$ has the property that only one of its rows or columns is non-zero.
-The complexity of $\trace(AB)$ is only $\calO(N)$.
-\\
-%
-%\noindent \textbf{Proof} Let $m <= M$ be the non-zero row of $B$. We have
-%\begin{align}
-%\trace (AB) = \sum_{i=1}^N \sum_{j=1}^M A_{ij} B_{ji} = \sum_{i=1}^N A_{im} B_{mi},
-%\end{align}
-%which clearly takes $\calO(N)$. It is easy to see that the lemma also holds when $B$ is symmetric and only one of its row and the corresponding column is non-zero. \\
-%
-%\noindent To exploit the property in the Lemma, we re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):
-
-\noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
-\begin{align}
-\nonumber
-% dL1
-\deriv{\calL_1}{t}
- &= \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
-&= \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} 
--\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t} \\
-%dL2
-\deriv{\calL_2}{t}
-&= - \beta \trace \A^T \deriv{\K_{NM}}{t}
- + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}  \\
-% dL3
-\deriv{\calL_3}{t}
-&= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
- - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}\\
-% dL4
-\deriv{\calL_4}{t}
- &= \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t}
-  - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
-\end{align}
-From the above equations we get,
-\begin{align}
-\deriv{\calL}{t} = \trace \D_1 \deriv{\K_{NM}}{t} + \trace \D_2 \deriv{ \K_{MM}}{t},
-\end{align}
-where 
-\begin{align}
-\D_1 =& \beta \K_{MM}^{-1} \m (\y - \A\m)^T
- + \beta \A^T
- - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
-\D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
- - \frac{1}{2} \beta \A^T \A
-  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
-  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
-\end{align}
-Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
-The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two $\trace$ operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
-This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.
-
-\subsection{Analysis on Learning of the Inducing Inputs}
-In this section we present some analysis on learning of the inducing inputs under the stochastic variational inference procedure on some toy problems.
-We use three real-valued truth functions of scalar inputs: the first one was used in Snelson et al, the second one is a function $f(x) = \sin(x) + \cos(x)$, and the third one is a sample function generated from a GP with squared exponential with ARD covariance function with a lengthscale of 2 and signal variance of 1.
-
-\noindent \paragraph{Experimental Settings} We use gpsvi to optimize for \textit{all} of the parameters in the model, i.e.
-the variational parameters of the posterior, the covariance hyperparameters, the noise hyperparameter, and the inducing inputs.
-For the hyperparameters, we use a momentum of $0.9$ and a fixed learn rate of $1e-5$. 
-For the variational parameters, we use a learn rate of $0.01$ (no momentum was used).
-We vary the learn rate of the inducing inputs from $1e{-2}$ to $1e{-6}$.
-Using momentum for the inducing inputs does not seem to have much effects.
-The maximum number of iterations is 500, the batch size 5, and the number of inducing values 10.
-
-% effect of learn rate on inducing inputs
-\noindent \paragraph{Effects of the Learn Rates }
-In Figure \ref{fig1}, \ref{fig2}, and \ref{fig3}, we show the lower bounds, predictive distributions, and the learned inducing inputs using stochastic variational inference with varying learn rates for the inducing inputs.
-We also show the results with FITC for comparison.
-It can be seen from the figures that, as typical in stochastic optimization, the behavior of GPSVI is very sensitive to the learn rate.
-In particular, when the rate is slow e.g. $1e{-5}$, almost no progress was made in learning the inducing inputs.
-However, when the rate increases to e.g. $1e{-3}$, the inducing inputs are spread out more evenly in the input space compared to the initial locations.
-Although not shown the in figures, when the learn rate is too large, e.g. $1e{-02}$, the inducing inputs may spread to regions far outside of the training intervals.
-An example of this phenomenon can be observed in the top left figure in Figure \ref{fig1}.
-The learn rates also affect the evidence lower bound which seems to wiggle more compared to not learning the inducing inputs.
-The predictive distributions are sensible particularly because the toy functions exhibit sparsity and are easy to learn.  \\
-
-\noindent The results of this qualitative analysis do not seem to suggest that learning of the inducing inputs does not work in GPSVI.
-Its effectiveness perhaps rests upon empirical performance on real datasets.
-
-\noindent \paragraph{The Noise Parameter} It seems harder to set the initial value and the learning rate of the noise parameter compared to standard GP. 
-While a large range of values of noise can be used in standard gp or fitc (e.g. 0.5), a very small value of noise $1e{-02}$, which is the true noise, is required for gpsvi to work well.
-It appears that an adaptive learning rate may be required for the noise parameter.
-
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05.eps} \\
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03-bound.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04-bound.eps} &
-\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05-bound.eps} \\ 
-(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
-\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func1-fitc.eps}}
-\end{tabular}
-\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the first function.
-Top row: GPSVI with different learning rates (the rates decreases from left to right).
-Bottom row: FITC.
-Magenta dots : training points.
-Solid blue line: predictive mean.
-Grey-shaded area and solid black lines: two standard deviations of the predictive distributions.
-Black (+) crosses: initial locations of inducing inputs.
-Magenta (+) crosses: learned locations of inducing inputs.
-Middle row: the evidence lower bound of gpsvi vs.
-iteration.}
-\label{fig1}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05.eps} \\
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03-bound.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04-bound.eps} &
-\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05-bound.eps} \\ 
-(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
-\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func2-fitc.eps}}
-\end{tabular}
-\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the second function. The legends are same as in Figure 1.}
-\label{fig2}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05.eps} \\
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03-bound.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04-bound.eps} &
-\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05-bound.eps} \\ 
-(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
-\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func3-fitc.eps}}
-\end{tabular}
-\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the third function. The legends are same as in Figure 1.}
-\label{fig3}
-\end{figure*}
+\bibliographystyle{abbrv}
+\bibliography{references}
 

### 2014-03-06 paper/experiments.tex
paper

@@
-\section{Toy Experiments}
-The first toy experiment uses two identical outputs which are noisy version of the same version plus some noise: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
-In this case the shared function $g(x)$ is $sin(x)$ and the independent processes $h_1(x)$ and $h_2(x)$ are just white noises.
+We perform three sets of experiments to evaluate the model and the stochastic variational inference proposed. 
+First, we use a toy problem to demonstrate the effectiveness of the proposed method in learning the model parameters.
+Second, we assess the predictive accuracy of the model on two medium-sized datasets and compare with existing multiple output models.
+In the final experiments, we evaluate the model on large-scale learning with multiple tasks.
+
+\subsection{TOY PROBLEM}
+In this toy problem we use two related outputs generated from the same latent function $sin(x)$ but are corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
 Each output has missing values in one region of the input space.
-The predictive distributions of the multiple-gp model compared to that of independent gps are shown in \ref{fig4}.
-
-\noindent The second toy experiment uses similar setting as the first one, except that now $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$. 
-In this case the shared function is still $g(x) = sin(x)$, but the weights should be opposite i.e. $w_1 = 1$ and $w_2 = 1$.
-The learning procedure was indeed able to recovered this relation and learned that $w_1 = 1.2$ and $w_2 = -1.3$.
-The predictive distributions are shown in \ref{fig5}.
-
-\begin{figure*}
-\centering
-\begin{tabular}{cc}
-\includegraphics[scale=0.5]{figures/ssvi-y1.eps} &
-\includegraphics[scale=0.5]{figures/ssvi-svi1.eps} \\
-\includegraphics[scale=0.5]{figures/ssvi-y2.eps} &
-\includegraphics[scale=0.5]{figures/ssvi-svi2.eps} \\
-\multicolumn{2}{c}{\includegraphics[scale=0.5]{figures/ssvi-y1byg.eps} }
-\end{tabular}
-\label{fig4}
-\caption{Predictive distributions of the multipe-output gps (left column) and independent gps (right column) for the first toy example.
-The predictive distribution by $g(x)$ for $y_1(x)$ is shown in the bottom figure.}
-\end{figure*}
+% Say that this is from a single gpsvi
+The predictive distributions of the model compared to that of independent gps are shown in \ref{fig5}.
 
@@
 \includegraphics[scale=0.5]{figures/ssvi2-svi2.eps} \\
-\multicolumn{2}{c}{\includegraphics[scale=0.5]{figures/ssvi-y1byg.eps} }
 \end{tabular}
 \label{fig5}
-\caption{Predictive distributions of the multipe-output gps (left column) and independent gps (right column) for the second toy experiment.
+\caption{Predictive distributions of the multioutput gps (left column) and independent gps (right column) for the second toy experiment.
 The predictive distribution by $g(x)$ for $y_1(x)$ is shown in the bottom figure.}
 \end{figure*}
+
+\subsection{EVALUATION OF THE PREDICTIVE ACCURACY}
+\subsubsection{Foregin Exchange Rate Prediction}
+\subsubsection{Weather Data Prediction}
+
+\subsection{LARGE-SCALE EXPERIMENTS}
+\subsubsection{Robort Arms}
+\subsubsection{Foreign Exchange Rate Prediction}

### 2014-03-06 paper/paper.tex
paper

@@
 \input{discussion}
-
-\subsubsection{Figures}
-
-All artwork must be centered, neat, clean, and legible. Figure number
-and caption always appear below the figure.  Leave 2 line spaces
-between the figure and the caption. The figure caption is initial caps
-and each figure numbered consecutively.
-
-Make sure that the figure caption does not get separated from the
-figure. Leave extra white space at the bottom of the page rather than
-splitting the figure and figure caption.
-\begin{figure}[h]
-\vspace{1in}
-\caption{Sample Figure Caption}
-\end{figure}
-
-\subsubsection{Tables}
-
-\begin{table}[h]
-\caption{Sample Table Title}
-\label{sample-table}
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION} \\
-\hline \\
-Dendrite         &Input terminal \\
-Axon             &Output terminal \\
-Soma             &Cell body (contains cell nucleus) \\
-\end{tabular}
-\end{center}
-\end{table}
-
 \newpage

### 2014-03-06 src/datautils/clean_weather.m
1. add weather dataset 2. add figures

@@ -1,0 +1,31 @@
+clear all; clc
+
+csvs = {'newbra.csv','newcam.csv','newchi.csv','newsot.csv'};
+mats = {'bra','cam','chi','sot'};
+day = 1:15;
+hour = 0:23;
+minute = 0:5:55;
+for i=1:numel(csvs)
+  xy = load(csvs{i});
+  data = [];
+  for d = day
+    for h = hour
+      for m = minute
+        thetime = [d,h,m];
+        rows = findRows(xy(:,1:3),thetime);
+        assert(sum(rows) <= 1)
+        if sum(rows) == 1
+          data = [data; xy(rows,4:end)];
+        else
+          data = [data; -1*ones(1,4)];
+        end
+      end
+    end
+  end
+  disp('missing data: ')
+  disp(sum(data(:,1)==-1));
+  x = (1:size(data,1))'/(24*12);
+  y = data;
+  save(['data/weather/' mats{i} '.mat'],'x','y')
+end
+

### 2014-03-06 src/datautils/read_weather.m

@@ -1,0 +1,29 @@
+function [x,y,xtest,ytest,y0] = read_weather()
+%READ_WEATHER  [x,y,xtest,ytest] = read_weather()
+%   Reads the weather data.
+clear all;
+output = 4; % air temperature
+load bra
+ybra = y(:,output);
+load cam
+ycam = y(:,output);
+load chi
+ychi = y(:,output);
+load sot
+ysot = y(:,output);
+
+range = x >= 10 & x <= 15;
+x = x(range);
+y = [ybra,ycam,ychi,ysot];
+y = y(range,:);
+
+y(y == -1) = nan; % missing data
+y0 = y;
+
+xtest = x;
+ytest = y;
+
+% impute missing data
+y(x >= 10.2 & x <= 10.8,2) = nan;
+y(x >= 13.5 & x <= 14.2,3) = nan;
+

### 2014-03-06 paper/appendix.tex

\subsection{Appendix 1: Derivatives}
Here we derive the gradients of the lower bound wrt the  hyperparameters.
As a template, consider the lower bound in Hensman et al (as a function of the hyperparameters):
\begin{align}
\nonumber
\calL
=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I)
 - \frac{1}{2} \beta \trace \tilde{\K}
 - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right) \\ \nonumber
=& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
 - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
 - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4},
\end{align}
where $\A = \K_{NM} \K_{MM}^{-1}$.
Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.

\subsection{Derivative of the Noise Hyperparameter}
The derivative of the noise hyperparameter $\beta$ is easily computed as:
\begin{align}
\deriv{\calL}{\beta} = \frac{N}{2\beta} - \frac{1}{2} (\y - \A\m)^T (\y - \A\m) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
\end{align}
\subsection{Derivatives of the Covariance Hyperparameters}  
To simplify the math, we utilize the matrix $\A$ defined above.
Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
\begin{align}
\deriv{\A}{t} = \left(\deriv{\K_{NM}}{t} - \A \deriv{\K_{MM}}{t}\right)\K_{MM}^{-1}.
\end{align}
The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
\begin{align}
\deriv{\calL_1}{t} &= \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
\deriv{\calL_2}{t} &= \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
\deriv{\calL_3}{t} &= \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
\deriv{\calL_4}{t} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
\end{align}
The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.

\subsection{Derivatives of the Inducing Inputs}
To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\z_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
Hence the derivative wrt a single dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\deriv{ \calL}{t}$.

%Since $MD$ parameters are needed for the inducing inputs, it appears that the derivatives of all inducing inputs would require $\calO(MD \times M^3)$ in computation.
%However, this complexity can actually be reduced to $\calO(DM^3)$ using the following lemma. \\
%
%\noindent \textbf{Lemma} Let $A, B$ be two matrices of size $N \times M$ and $M \times N$, respectively. Furthermore, $B$ has the property that only one of its rows or columns is non-zero. The complexity of $\trace(AB)$ is only $\calO(N)$. \\
%
%\noindent \textbf{Proof} Let $m <= M$ be the non-zero row of $B$. We have
%\begin{align}
%\trace (AB) = \sum_{i=1}^N \sum_{j=1}^M A_{ij} B_{ji} = \sum_{i=1}^N A_{im} B_{mi},
%\end{align}
%which clearly takes $\calO(N)$. It is easy to see that the lemma also holds when $B$ is symmetric and only one of its row and the corresponding column is non-zero. \\
%
%\noindent To exploit the property in the Lemma, we re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):

\noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
\begin{align}
\nonumber
% dL1
\deriv{\calL_1}{t}
 &= \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
&= \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} 
-\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t} \\
%dL2
\deriv{\calL_2}{t}
&= - \beta \trace \A^T \deriv{\K_{NM}}{t}
 + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}  \\
% dL3
\deriv{\calL_3}{t}
&= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
 - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}\\
% dL4
\deriv{\calL_4}{t}
 &= \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t}
  - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
\end{align}
From the above equations we get,
\begin{align}
\deriv{\calL}{t} = \trace \D_1 \deriv{\K_{NM}}{t} + \trace \D_2 \deriv{ \K_{MM}}{t},
\end{align}
where 
\begin{align}
\D_1 =& \beta \K_{MM}^{-1} \m (\y - \A\m)^T
 + \beta \A^T
 - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
\D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
 - \frac{1}{2} \beta \A^T \A
  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
\end{align}
Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two $\trace$ operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.

\subsection{Appendix 2: Gaussian Identity}
Let $\y$, $\g = \{\g_j\}_{j=1}^q$, and $\h$ be random variables with multivariate Gaussian distributions: 
$p(\y | \g, \h) = \Normal(\y; \sum_{j=1}^Q \W_j \g_j + \W \h, \beta^{-1} \I)$, $p(\g_j) = \Normal(\g_j; \m_j, \S_j)$, and $p(\h) = \Normal(\h; \m, \S)$.
The following identity is important in deriving the evidence lower bound:
\begin{align}
\nonumber
\int \log p(\y | \g, \h) \prod_{j=1}^q p(\g_j) p(\h) \der \g \der \h
=& \log \Normal(\y; \sum_{j=1}^q \W_j \m_j + \W \m, \beta^{-1} \I) - \frac{1}{2} \beta \trace \W^T \W \S \\
\label{eq:identity}
&- \frac{1}{2} \beta \trace \sum_{j=1}^q \W_j^T \W_j \S_j.
\end{align}
The identity can be proved by using this fact: 
\begin{align}
\nonumber
\int (\W\x - \Mu)^T \BigSigma^{-1} &(\W \x - \Mu) \Normal(\x; \m, \S) \der \x \\
\nonumber
&= (\Mu - \W\m)^T \BigSigma^{-1} (\Mu - \W\m) + \trace \W^T \BigSigma^{-1} \W \S.
\end{align}

### 2014-03-06 paper/inference.tex

\subsection{Variational Inference in Sparse GPs Revisited}
In this section we review inference in sparse GPs (as presented in Titsias).
The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\

\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
\begin{align}
\nonumber
p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
\nonumber
&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
\label{eq:sorprediction}
&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
\end{align}
where,
\begin{align}
\nonumber
\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
\nonumber
\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
\end{align}
 Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.

\subsection{Variational Inference}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
Our goal of inference is to find the posterior $p(\g, \h, \u, \v | \y)$. 
Following the previous discussion, we assume a variational distribution which factorizes as:
\begin{align}
\nonumber
q(\g, \h, \u, \v | \y)
\nonumber
 &= p(\g|\u) p(\h|\v) q(\u,\v)  \\
 &= p(\g|\u) p(\h|\v) \prod_{j=1}^Q q(\u_j) \prod_{i=1}^P  q(\v_i)
\end{align}
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given, we need only to find the optimum $q(\u_j)$ and $q(\v_i)$.
Let $q(\u_j) = \Normal(\u_j; \mgj, \Sgj)$ and $q(\v_i) = \Normal(\v_i; \mhi, \Shi)$. \\

\noindent
To find the optimum $q(\u, \v)$ we optimize the evidence lower bound (ELBO) of the log marginal,
\begin{align}
\nonumber
\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\label{eq:elbo}
=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
- \left(\sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right),
\end{align}
where the last equality occurs because both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g_j(\x)$ and $h_i(\x)$ use separate set of inducing inputs). \\

\noindent To compute the ELBO we first focus on the term:
\newcommand{\llangle}{\left\langle}
\newcommand{\rrangle}{\right\rangle}
\begin{align}
\nonumber
\log p(\y | \u, \v)
 &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
 \nonumber
&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \quad &\text{(Jensen's inequality)} \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_{in} | \v_i)}, \quad 
 &\text{(factorized likelihood)}
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
Each individual term $l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ can be computed using the identity in eq. \ref{eq:identity} given in the appendix to give:
\begin{align}
\nonumber
l_{in} &= \int \log p(y_{in} | \g_n, h_{in}) \prod_{j=1}^Q p(g_{jn} | \u_j) p(h_{in} | \u_i) \der \g_n \der h_{in} \\
&= \log \Normal(y_{in}; \sum_{j=1}^Q w_{ij} \mu_{jn}+ \mu^h_{in}, \beta_i^{-1})
- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} 
- \frac{1}{2} \beta_i \tilde{k}^h_{inn},
\end{align}
where $\tilde{k}_{nn} = (\tilde{\K})_{nn}, \tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}, \mu_{jn} = (\Mu_j)_n, \text{ and } \mu^h_{in} = (\Mu^h_i)_n$.

% previous detailed derivation
%\begin{align}
%\nonumber
%l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\g | \ug) p(\h_i | \u_i) \der \g \der \h_i \\
%\nonumber
%&= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
%\Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
%\nonumber
%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta_i (y_{in} - w_i g_n - h_{in})
%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der g_n \der h_{in} \\
%\nonumber
%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta_i (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta_i \tilde{k}_{gnn} \right] 
%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
%\nonumber
%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta_i \tilde{k}_{gnn}
%- \frac{1}{2} \beta_i \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta_i (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
%&= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
%\end{align}

\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying out the integral using the identity in eq. \ref{eq:identity} we get:
\begin{align}
\nonumber
\log p(\y)
\ge& \sum_{i=1}^P \sum_{n=1}^N
\bigg( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2        \tilde{k}_{nn} - \frac{1}{2} \beta_i \tilde{k}^h_{inn}
 \\ \nonumber
         &\quad \quad \quad \quad - \frac{1}{2} \beta_i \trace  \sum_{j=1}^Q w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
\bigg) \\
&- \left(\sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] + \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)] \right) \define \calL,
\end{align}
where 
%TODO \Lambda_gn depends on the output i so need a better notation
\newcommand{\Zg}{\Z_g}
\newcommand{\Zi}{\Z_i}
\begin{align}
\tilde{\mu}_{in}
&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
\mat{\Lambda}_{jn}
&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
\mat{\Lambda}_{in}
&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}.
\end{align}

\noindent Notice that this ELBO clearly generalizes the standard GP regression. In particular, setting $P = Q = 1$, $w_i = 1$ and $h_i(\x) = 0$ we recover the bound in Hensman et al \cite{hensmangaussian}.
Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.

\subsubsection{Variational Parameters Derivatives}
\newcommand{\oi}{\vec{o}_i}
% some notation
Before diving into the details, we first define the indexing operator of a matrix: $\B(\vec{r},\vec{c})$ extracts the submatrix in rows $\vec{r}$ and columns $\vec{c}$ of $\B$.
To index all columns we use $\B(\vec{r},:)$ and similarly for all rows $\B(:,\vec{c})$.
Readers familiar with this operator will recognize that this is the MATLAB indexing operator.

The optimal posteriors are found by setting the gradients of the lowerbound $\calL$ wrt to the parameters of $q(\u_j)$ and $q(\v_i)$.
Recall that in the model, different outputs can be observed at different inputs (i.e. the case of missing values).
Let $\oi$ be the indice of the observed inputs of the output dimension $i$.
We denote its set of observed inputs and targets as: $\X_i = \X(\oi,:)$ and $\y_i = y_{i}(\oi)$.

\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
\newcommand{\Ahi}{\A^h_i}
\begin{align}
\nonumber
\calL^g_j \define&
 \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
 \nonumber
 &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
 + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
 \\
 &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
\end{align}
where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\

%------------------------------------------
% derivatives of q(u_j)
\newcommand{\Lgj}{\calL^g_j}
\newcommand{\ynoj}{\y_i^{\backslash j}}
\noindent The derivatives of $\Lgj$ wrt $\m_j$ and $\S_j$ are given by:
\begin{align}
\deriv{\Lgj}{\m_j}
& = \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi,:)^T \ynoj - \bigg[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg] \m_j \\
\deriv{\Lgj}{\S_j} 
&= \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ k(\Zj,\Zj)^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg],
\end{align}
where $\y_i^{\backslash j} = \y_i - \Ahi \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi,:) \m_{j'}$.

%-------------------------------------------
%  derivatives of q(v_i)
\newcommand{\Lhi}{\calL^h_i}
\noindent As a function of the parameters of $q(\v_i)$, the lower bound $\calL$ is:
\begin{align}
\nonumber
\Lhi \define&
 \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)
 - \frac{1}{2} \beta_i \trace \tilde{\K}^h_i(\oi,\oi)
 - \frac{1}{2} \beta_i \trace \Shi (\Ahi)^T \Ahi
 \\
  &- \frac{1}{2} \log |k(\Zhi,\Zhi) (\Shi)^{-1}| -\frac{1}{2} \trace k(\Zhi,\Zhi)^{-1} (\mhi (\mhi)^T + \Shi) ,
\end{align}

\noindent The derivatives of $\Lhi$ wrt $\mhi$ and $\Shi$ are given by:
\newcommand{\ynoh}{\y_i^{\backslash h}}
\begin{align}
\deriv{\Lhi}{\mhi}
& = \beta_i \A_i^T \ynoh - \left[k(\Zhi,\Zhi)^{-1} +  \beta_i \A_i^T \A_i \right] \m_i \\
\deriv{\Lhi}{\Shi} 
&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \left[ k(\Zhi,\Zhi)^{-1} + \beta_i \A_i^T \A_i \right] ,
\end{align}
where $\ynoh = \y_i - \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j$.

% comment on computation
\noindent It can be seen that the derivatives of the parameters of $q(\v_i)$ only involve the observations of the output dimension $i$.
The derivatives of the parameters of $q(\u_j)$ involve the observations across all output dimensions but decompose as a sum of contributions from individual outputs.
Therefore, computation of the derivatives (and hence the update equations) can be distributed or parallelized easily.
This attractive property allows the model to scale to a very large number of inputs and outputs.

% Comment on other hyperparameters

%\subsubsection{Update Equations}
% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
\subsection{Prediction}
The predictive distribution of the $i$-th output for a test input $\x_*$ is 
\begin{align}
p(\fstar | \y, \x_*) = \int \Normal(\fstar; \sum_{j=1}^Q w_j g_{j*} + h_{i*}, 0) p(\g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der \g_* \der h_{i*},
\end{align}
where $p(\g_* | \y, \x_*) = \prod_{j=1}^Q \Normal(g_{j*}; \mu_{j*}, s_{j*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu^h_{i*}, s^h_{i*})$ are the predictive distributions of the sparse GPs as given in eq. \ref{eq:sorprediction}.
Therefore we have:
\begin{align}
p(\fstar | \y, \x_*) = \Normal(\fstar; \sum_{j=1}^Q w_{ij} \mu_{j*} + \mu_{i*}, w_{ij}^2 s_{j*} + s_{i*}). 
\end{align}

%\begin{linenomath}
%\begin{align}
%\sum_{i,h} \lambda_i \lambda_h cov [f(\x, \x')]
%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\x_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\x_h - \vs_j') \\
%\end{align}
%\end{linenomath}

### 2014-03-06 paper/note.tex

\documentclass{article} % For LaTeX2e
\usepackage[a4paper,margin=1.5in]{geometry}
\usepackage[mathlines]{lineno}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{latexsym,amsbsy,amssymb,color,xspace, booktabs}

\include{macros}
\include{local_macros}
\def\linenumberfont{\normalfont\small\sffamily}

\title{Multiple Output Gaussian Processes Regression}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}
\maketitle
\linenumbers

\section{Model Description}
We describe a multiple-output regression model.
Suppose that we have inputs $\X = \{\x_n\}_{n=1}^N$ and outputs $\y = \{\y_n\}_{n=1}^N$, with $\x_n \in \calR^D$ and $\y_n \in \calR^P$, i.e. the output is $P$-dimensional.
For ease of exposition, we assume that there are no missing values in any output dimension.
The most dominant approach in GP-based multiple output regression is to compose the different outputs as some \textit{linear} mixing (combination) of some basic processes (random functions).
Our model is also underpinned by this composition approach.

\noindent Consider the semiparametric latent factor models (SLFM) \cite{teh-et-al-aistats-05}.
It assumes $Q$ basic independent processes, $g_j(\x) \sim \GP(0, k(\x,\x';\vectheta_j))$, where $\vectheta_j$ is the hyperparameters of the covariance function of $g_j(\x)$.
Each (latent) output is a weighted combination of these basic processes, i.e.
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x), \quad i = 1 \hdots P,
\end{align}
In \cite{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
\begin{align}
f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x) + h_i(\x), 
\end{align}
We emphasize that, although $k$ denotes covariance functions in general, each independent process can have its own covariance function.

\noindent 
\textbf{Augmented sparse GPs}
Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
This adds much flexibility to the model as each of the shared process $g_j(\x)$ can model a different pattern in the data with its own covariance function and inducing inputs. The roles of $g_j(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
Furthermore, $g_j(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be a subspace of the input.

%However, to really address the issue of scalability, this model is built upon sparse GPs from the ground up.

\subsection{Prior and Likelihood}
In this section we specify the prior and likelihood of this model. 
\newcommand{\Zj}{\Z_j}
\newcommand{\Zhi}{\Z^h_i}
Let the set of inducing inputs for $g_j(\x)$ and $h_i(\x)$ be $\Z_j$ and $\Z^h_i$, and the corresponding inducing points $\u_j$ and $\v_i$, respectively.
We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
The subscripts are $i = 1 \hdots P$, $j = 1 \hdots Q$, and $n = 1 \hdots N$.

\noindent
\textbf{Prior}
The prior of the augmented model is given by:
\begin{align}
p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)),
\end{align}
where $\BigMu_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}\u$ and $\tilde{\K}_j = k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X)$ and $\BigMu^h_i, \tilde{\K}^h_i$ are similarly defined.
To avoid notational clutter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.

\noindent
\textbf{Likelihood}
The likelihood as usual follows the standard iid Gaussian likelihood,
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I).
\end{align}

\subsection{Variational Inference in Sparse GPs Revisited}
In this section we review inference in sparse GPs (as presented in Titsias).
The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g. the test set).
In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\

\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
\begin{align}
\nonumber
p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
\nonumber
&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
\label{eq:sorprediction}
&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
\end{align}
where,
\begin{align}
\nonumber
\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
\nonumber
\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
\end{align}
 Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.

\subsection{Variational Inference}
\newcommand{\ug}{\u_g}
\newcommand{\uh}{\u^h}
\newcommand{\mgj}{\m_j}
\newcommand{\mhi}{\m^h_i}
\newcommand{\Sgj}{\S_j}
\newcommand{\Shi}{\S^h_i}
Our goal of inference is to find the posterior $p(\g, \h, \u, \v | \y)$. 
Following the previous discussion, we assume a variational distribution which factorizes as:
\begin{align}
\nonumber
q(\g, \h, \u, \v | \y)
\nonumber
 &= p(\g|\u) p(\h|\v) q(\u,\v)  \\
 &= p(\g|\u) p(\h|\v) \prod_{j=1}^Q q(\u_j) \prod_{i=1}^P  q(\v_i)
\end{align}
Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given, we need only to find the optimum $q(\u_j)$ and $q(\v_i)$.
Let $q(\u_j) = \Normal(\u_j; \mgj, \Sgj)$ and $q(\v_i) = \Normal(\v_i; \mhi, \Shi)$. \\

\noindent
To find the optimum $q(\u, \v)$ we optimize the evidence lower bound (ELBO) of the log marginal,
\begin{align}
\nonumber
\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\nonumber
=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
\label{eq:elbo}
=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
- \left(\sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right),
\end{align}
where the last equality occurs because both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g_j(\x)$ and $h_i(\x)$ use separate set of inducing inputs). \\

\noindent To compute the ELBO we first focus on the term:
\newcommand{\llangle}{\left\langle}
\newcommand{\rrangle}{\right\rangle}
\begin{align}
\nonumber
\log p(\y | \u, \v)
 &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
 \nonumber
&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \quad &\text{(Jensen's inequality)} \\
&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_{in} | \v_i)}, \quad 
 &\text{(factorized likelihood)}
\end{align}
where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
Each individual term $l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ can be computed using the identity in eq. \ref{eq:identity} given in the appendix to give:
\begin{align}
\nonumber
l_{in} &= \int \log p(y_{in} | \g_n, h_{in}) \prod_{j=1}^Q p(g_{jn} | \u_j) p(h_{in} | \u_i) \der \g_n \der h_{in} \\
&= \log \Normal(y_{in}; \sum_{j=1}^Q w_{ij} \mu_{jn}+ \mu^h_{in}, \beta_i^{-1})
- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} 
- \frac{1}{2} \beta_i \tilde{k}^h_{inn},
\end{align}
where $\tilde{k}_{nn} = (\tilde{\K})_{nn}, \tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}, \mu_{jn} = (\Mu_j)_n, \text{ and } \mu^h_{in} = (\Mu^h_i)_n$.

% previous detailed derivation
%\begin{align}
%\nonumber
%l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\g | \ug) p(\h_i | \u_i) \der \g \der \h_i \\
%\nonumber
%&= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
%\Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
%\nonumber
%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta_i (y_{in} - w_i g_n - h_{in})
%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der g_n \der h_{in} \\
%\nonumber
%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta_i (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta_i \tilde{k}_{gnn} \right] 
%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
%\nonumber
%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta_i \tilde{k}_{gnn}
%- \frac{1}{2} \beta_i \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta_i (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
%&= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
%\end{align}

\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying out the integral using the identity in eq. \ref{eq:identity} we get:
\begin{align}
\nonumber
\log p(\y)
\ge& \sum_{i=1}^P \sum_{n=1}^N
\bigg( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2        \tilde{k}_{nn} - \frac{1}{2} \beta_i \tilde{k}^h_{inn}
 \\ \nonumber
         &\quad \quad \quad \quad - \frac{1}{2} \beta_i \trace  \sum_{j=1}^Q w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
\bigg) \\
&- \left(\sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] + \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)] \right) \define \calL,
\end{align}
where 
%TODO \Lambda_gn depends on the output i so need a better notation
\newcommand{\Zg}{\Z_g}
\newcommand{\Zi}{\Z_i}
\begin{align}
\tilde{\mu}_{in}
&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
\mat{\Lambda}_{jn}
&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
\mat{\Lambda}_{in}
&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}.
\end{align}

\noindent Notice that this ELBO clearly generalizes the standard GP regression. In particular, setting $P = Q = 1$, $w_i = 1$ and $h_i(\x) = 0$ we recover the bound in Hensman et al \cite{hensmangaussian}.
Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.

\subsubsection{Variational Parameters Derivatives}
\newcommand{\oi}{\vec{o}_i}
% some notation
Before diving into the details, we first define the indexing operator of a matrix: $\B(\vec{r},\vec{c})$ extracts the submatrix in rows $\vec{r}$ and columns $\vec{c}$ of $\B$.
To index all columns we use $\B(\vec{r},:)$ and similarly for all rows $\B(:,\vec{c})$.
Readers familiar with this operator will recognize that this is the MATLAB indexing operator.

The optimal posteriors are found by setting the gradients of the lowerbound $\calL$ wrt to the parameters of $q(\u_j)$ and $q(\v_i)$.
Recall that in the model, different outputs can be observed at different inputs (i.e. the case of missing values).
Let $\oi$ be the indice of the observed inputs of the output dimension $i$.
We denote its set of observed inputs and targets as: $\X_i = \X(\oi,:)$ and $\y_i = y_{i}(\oi)$.

\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
\newcommand{\Ahi}{\A^h_i}
\begin{align}
\nonumber
\calL^g_j \define&
 \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
 \nonumber
 &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
 + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
 \\
 &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
\end{align}
where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\

%------------------------------------------
% derivatives of q(u_j)
\newcommand{\Lgj}{\calL^g_j}
\newcommand{\ynoj}{\y_i^{\backslash j}}
\noindent The derivatives of $\Lgj$ wrt $\m_j$ and $\S_j$ are given by:
\begin{align}
\deriv{\Lgj}{\m_j}
& = \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi,:)^T \ynoj - \bigg[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg] \m_j \\
\deriv{\Lgj}{\S_j} 
&= \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ k(\Zj,\Zj)^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg],
\end{align}
where $\y_i^{\backslash j} = \y_i - \Ahi \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi,:) \m_{j'}$.

%-------------------------------------------
%  derivatives of q(v_i)
\newcommand{\Lhi}{\calL^h_i}
\noindent As a function of the parameters of $q(\v_i)$, the lower bound $\calL$ is:
\begin{align}
\nonumber
\Lhi \define&
 \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)
 - \frac{1}{2} \beta_i \trace \tilde{\K}^h_i(\oi,\oi)
 - \frac{1}{2} \beta_i \trace \Shi (\Ahi)^T \Ahi
 \\
  &- \frac{1}{2} \log |k(\Zhi,\Zhi) (\Shi)^{-1}| -\frac{1}{2} \trace k(\Zhi,\Zhi)^{-1} (\mhi (\mhi)^T + \Shi) ,
\end{align}

\noindent The derivatives of $\Lhi$ wrt $\mhi$ and $\Shi$ are given by:
\newcommand{\ynoh}{\y_i^{\backslash h}}
\begin{align}
\deriv{\Lhi}{\mhi}
& = \beta_i \A_i^T \ynoh - \left[k(\Zhi,\Zhi)^{-1} +  \beta_i \A_i^T \A_i \right] \m_i \\
\deriv{\Lhi}{\Shi} 
&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \left[ k(\Zhi,\Zhi)^{-1} + \beta_i \A_i^T \A_i \right] ,
\end{align}
where $\ynoh = \y_i - \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j$.

% comment on computation
\noindent It can be seen that the derivatives of the parameters of $q(\v_i)$ only involve the observations of the output dimension $i$.
The derivatives of the parameters of $q(\u_j)$ involve the observations across all output dimensions but decompose as a sum of contributions from individual outputs.
Therefore, computation of the derivatives (and hence the update equations) can be distributed or parallelized easily.
This attractive property allows the model to scale to a very large number of inputs and outputs.

% Comment on other hyperparameters

%\subsubsection{Update Equations}
% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
\subsection{Prediction}
The predictive distribution of the $i$-th output for a test input $\x_*$ is 
\begin{align}
p(\fstar | \y, \x_*) = \int \Normal(\fstar; \sum_{j=1}^Q w_j g_{j*} + h_{i*}, 0) p(\g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der \g_* \der h_{i*},
\end{align}
where $p(\g_* | \y, \x_*) = \prod_{j=1}^Q \Normal(g_{j*}; \mu_{j*}, s_{j*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu^h_{i*}, s^h_{i*})$ are the predictive distributions of the sparse GPs as given in eq. \ref{eq:sorprediction}.
Therefore we have:
\begin{align}
p(\fstar | \y, \x_*) = \Normal(\fstar; \sum_{j=1}^Q w_{ij} \mu_{j*} + \mu_{i*}, w_{ij}^2 s_{j*} + s_{i*}). 
\end{align}

%\begin{linenomath}
%\begin{align}
%\sum_{i,h} \lambda_i \lambda_h cov [f(\x, \x')]
%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\x_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\x_h - \vs_j') \\
%\end{align}
%\end{linenomath}

\section{Toy Experiments}
The first toy experiment uses two identical outputs which are noisy version of the same version plus some noise: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
In this case the shared function $g(x)$ is $sin(x)$ and the independent processes $h_1(x)$ and $h_2(x)$ are just white noises.
Each output has missing values in one region of the input space.
The predictive distributions of the multiple-gp model compared to that of independent gps are shown in \ref{fig4}.

\noindent The second toy experiment uses similar setting as the first one, except that now $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$. 
In this case the shared function is still $g(x) = sin(x)$, but the weights should be opposite i.e. $w_1 = 1$ and $w_2 = 1$.
The learning procedure was indeed able to recovered this relation and learned that $w_1 = 1.2$ and $w_2 = -1.3$.
The predictive distributions are shown in \ref{fig5}.

\begin{figure*}
\centering
\begin{tabular}{cc}
\includegraphics[scale=0.5]{figures/ssvi-y1.eps} &
\includegraphics[scale=0.5]{figures/ssvi-svi1.eps} \\
\includegraphics[scale=0.5]{figures/ssvi-y2.eps} &
\includegraphics[scale=0.5]{figures/ssvi-svi2.eps} \\
\multicolumn{2}{c}{\includegraphics[scale=0.5]{figures/ssvi-y1byg.eps} }
\end{tabular}
\label{fig4}
\caption{Predictive distributions of the multipe-output gps (left column) and independent gps (right column) for the first toy example. The predictive distribution by $g(x)$ for $y_1(x)$ is shown in the bottom figure.}
\end{figure*}

\begin{figure*}
\centering
\begin{tabular}{cc}
\includegraphics[scale=0.5]{figures/ssvi2-y1.eps} &
\includegraphics[scale=0.5]{figures/ssvi2-svi1.eps} \\
\includegraphics[scale=0.5]{figures/ssvi2-y2.eps} &
\includegraphics[scale=0.5]{figures/ssvi2-svi2.eps} \\
\multicolumn{2}{c}{\includegraphics[scale=0.5]{figures/ssvi-y1byg.eps} }
\end{tabular}
\label{fig5}
\caption{Predictive distributions of the multipe-output gps (left column) and independent gps (right column) for the second toy experiment. The predictive distribution by $g(x)$ for $y_1(x)$ is shown in the bottom figure.}
\end{figure*}

\section{Appendix 1: Derivatives}
Here we derive the gradients of the lower bound wrt the  hyperparameters.
As a template, consider the lower bound in Hensman et al (as a function of the hyperparameters):
\begin{align}
\nonumber
\calL
=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I)
 - \frac{1}{2} \beta \trace \tilde{\K}
 - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
&- \frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right) \\ \nonumber
=& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
 - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
 - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4},
\end{align}
where $\A = \K_{NM} \K_{MM}^{-1}$.
Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.

\subsection{Derivative of the Noise Hyperparameter}
The derivative of the noise hyperparameter $\beta$ is easily computed as:
\begin{align}
\deriv{\calL}{\beta} = \frac{N}{2\beta} - \frac{1}{2} (\y - \A\m)^T (\y - \A\m) - \frac{\calL_2}{\beta} - \frac{\calL_3}{\beta}.
\end{align}
\subsection{Derivatives of the Covariance Hyperparameters}  
To simplify the math, we utilize the matrix $\A$ defined above.
Firstly, the derivative of $\A$ wrt a covariance hyperparameter $t$ is given by:
\begin{align}
\deriv{\A}{t} = \left(\deriv{\K_{NM}}{t} - \A \deriv{\K_{MM}}{t}\right)\K_{MM}^{-1}.
\end{align}
The derivatives of $\calL_1, \calL_2, \calL_3 \text{ and } \calL_4$ are thus given by:
\begin{align}
\deriv{\calL_1}{t} &= \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
\deriv{\calL_2}{t} &= \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
\deriv{\calL_3}{t} &= \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
\deriv{\calL_4}{t} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
\end{align}
The derivatives are then computed by taking the derivatives of the covariance matrices $\K_{NN} (\text{the diagonal only}), \K_{NM} \text{ and }  \K_{MM}$, hence the covariance function, wrt the hyperparameters.

\subsection{Derivatives of the Inducing Inputs}
To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\z_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
Hence the derivative wrt a single dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\deriv{ \calL}{t}$.

%Since $MD$ parameters are needed for the inducing inputs, it appears that the derivatives of all inducing inputs would require $\calO(MD \times M^3)$ in computation.
%However, this complexity can actually be reduced to $\calO(DM^3)$ using the following lemma. \\
%
%\noindent \textbf{Lemma} Let $A, B$ be two matrices of size $N \times M$ and $M \times N$, respectively. Furthermore, $B$ has the property that only one of its rows or columns is non-zero. The complexity of $\trace(AB)$ is only $\calO(N)$. \\
%
%\noindent \textbf{Proof} Let $m <= M$ be the non-zero row of $B$. We have
%\begin{align}
%\trace (AB) = \sum_{i=1}^N \sum_{j=1}^M A_{ij} B_{ji} = \sum_{i=1}^N A_{im} B_{mi},
%\end{align}
%which clearly takes $\calO(N)$. It is easy to see that the lemma also holds when $B$ is symmetric and only one of its row and the corresponding column is non-zero. \\
%
%\noindent To exploit the property in the Lemma, we re-write $\frac{d \calL_1}{dt}, \frac{d \calL_2}{dt}, \frac{d \calL_3}{dt}, \frac{d \calL_4}{dt}$ by expanding $\frac{d\A}{dt}$ (here $t = z_{mj}$):

\noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
\begin{align}
\nonumber
% dL1
\deriv{\calL_1}{t}
 &= \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
&= \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} 
-\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t} \\
%dL2
\deriv{\calL_2}{t}
&= - \beta \trace \A^T \deriv{\K_{NM}}{t}
 + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}  \\
% dL3
\deriv{\calL_3}{t}
&= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
 - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}\\
% dL4
\deriv{\calL_4}{t}
 &= \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t}
  - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
\end{align}
From the above equations we get,
\begin{align}
\deriv{\calL}{t} = \trace \D_1 \deriv{\K_{NM}}{t} + \trace \D_2 \deriv{ \K_{MM}}{t},
\end{align}
where 
\begin{align}
\D_1 =& \beta \K_{MM}^{-1} \m (\y - \A\m)^T
 + \beta \A^T
 - \beta \K_{MM}^{-1} \S \A^T \\ \nonumber
\D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
 - \frac{1}{2} \beta \A^T \A
  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
\end{align}
Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two $\trace$ operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.

\section{Appendix 2: Gaussian Identity}
Let $\y$, $\g = \{\g_j\}_{j=1}^q$, and $\h$ be random variables with multivariate Gaussian distributions: 
$p(\y | \g, \h) = \Normal(\y; \sum_{j=1}^Q \W_j \g_j + \W \h, \beta^{-1} \I)$, $p(\g_j) = \Normal(\g_j; \m_j, \S_j)$, and $p(\h) = \Normal(\h; \m, \S)$.
The following identity is important in deriving the evidence lower bound:
\begin{align}
\nonumber
\int \log p(\y | \g, \h) \prod_{j=1}^q p(\g_j) p(\h) \der \g \der \h
=& \log \Normal(\y; \sum_{j=1}^q \W_j \m_j + \W \m, \beta^{-1} \I) - \frac{1}{2} \beta \trace \W^T \W \S \\
\label{eq:identity}
&- \frac{1}{2} \beta \trace \sum_{j=1}^q \W_j^T \W_j \S_j.
\end{align}
The identity can be proved by using this fact: 
\begin{align}
\nonumber
\int (\W\x - \Mu)^T \BigSigma^{-1} &(\W \x - \Mu) \Normal(\x; \m, \S) \der \x \\
\nonumber
&= (\Mu - \W\m)^T \BigSigma^{-1} (\Mu - \W\m) + \trace \W^T \BigSigma^{-1} \W \S.
\end{align}

\section{Appendix: Analysis on Learning of the Inducing Inputs}
In this section we present some analysis on learning of the inducing inputs under the stochastic variational inference procedure on some toy problems.
We use three real-valued truth functions of scalar inputs: the first one was used in Snelson et al, the second one is a function $f(x) = \sin(x) + \cos(x)$, and the third one is a sample function generated from a GP with squared exponential with ARD covariance function with a lengthscale of 2 and signal variance of 1.

\noindent \paragraph{Experimental Settings} We use gpsvi to optimize for \textit{all} of the parameters in the model, i.e. the variational parameters of the posterior, the covariance hyperparameters, the noise hyperparameter, and the inducing inputs.
For the hyperparameters, we use a momentum of $0.9$ and a fixed learn rate of $1e-5$. 
For the variational parameters, we use a learn rate of $0.01$ (no momentum was used).
We vary the learn rate of the inducing inputs from $1e{-2}$ to $1e{-6}$.
Using momentum for the inducing inputs does not seem to have much effects.
The maximum number of iterations is 500, the batch size 5, and the number of inducing values 10.

% effect of learn rate on inducing inputs
\noindent \paragraph{Effects of the Learn Rates }
In Figure \ref{fig1}, \ref{fig2}, and \ref{fig3}, we show the lower bounds, predictive distributions, and the learned inducing inputs using stochastic variational inference with varying learn rates for the inducing inputs.
We also show the results with FITC for comparison.
It can be seen from the figures that, as typical in stochastic optimization, the behavior of GPSVI is very sensitive to the learn rate.
In particular, when the rate is slow e.g. $1e{-5}$, almost no progress was made in learning the inducing inputs.
However, when the rate increases to e.g. $1e{-3}$, the inducing inputs are spread out more evenly in the input space compared to the initial locations.
Although not shown the in figures, when the learn rate is too large, e.g. $1e{-02}$, the inducing inputs may spread to regions far outside of the training intervals.
An example of this phenomenon can be observed in the top left figure in Figure \ref{fig1}.
The learn rates also affect the evidence lower bound which seems to wiggle more compared to not learning the inducing inputs.
The predictive distributions are sensible particularly because the toy functions exhibit sparsity and are easy to learn.  \\

\noindent The results of this qualitative analysis do not seem to suggest that learning of the inducing inputs does not work in GPSVI.
Its effectiveness perhaps rests upon empirical performance on real datasets.

\noindent \paragraph{The Noise Parameter} It seems harder to set the initial value and the learning rate of the noise parameter compared to standard GP. 
While a large range of values of noise can be used in standard gp or fitc (e.g. 0.5), a very small value of noise $1e{-02}$, which is the true noise, is required for gpsvi to work well.
It appears that an adaptive learning rate may be required for the noise parameter.

\begin{figure*}
\centering
\begin{tabular}{ccc}
\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03.eps} &
\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04.eps} &
\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05.eps} \\
\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-03-bound.eps} &
\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-04-bound.eps} &
\includegraphics[scale=0.3]{figures/func1-svi-lrate1e-05-bound.eps} \\ 
(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func1-fitc.eps}}
\end{tabular}
\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the first function. Top row: GPSVI with different learning rates (the rates decreases from left to right). Bottom row: FITC. Magenta dots : training points. Solid blue line: predictive mean. Grey-shaded area and solid black lines: two standard deviations of the predictive distributions. Black (+) crosses: initial locations of inducing inputs. Magenta (+) crosses: learned locations of inducing inputs.  Middle row: the evidence lower bound of gpsvi vs. iteration.}
\label{fig1}
\end{figure*}

\begin{figure*}
\centering
\begin{tabular}{ccc}
\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03.eps} &
\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04.eps} &
\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05.eps} \\
\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-03-bound.eps} &
\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-04-bound.eps} &
\includegraphics[scale=0.3]{figures/func2-svi-lrate1e-05-bound.eps} \\ 
(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func2-fitc.eps}}
\end{tabular}
\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the second function. The legends are same as in Figure 1.}
\label{fig2}
\end{figure*}

\begin{figure*}
\centering
\begin{tabular}{ccc}
\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03.eps} &
\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04.eps} &
\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05.eps} \\
\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-03-bound.eps} &
\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-04-bound.eps} &
\includegraphics[scale=0.3]{figures/func3-svi-lrate1e-05-bound.eps} \\ 
(a) learn rate = $1e{-3}$ & (b) learn rate = $1e{-4}$ & (c) learn rate = $1e{-5}$ \\
\multicolumn{3}{c}{\includegraphics[scale=0.4]{figures/func3-fitc.eps}}
\end{tabular}
\caption{Predictive distributions and learned inducing inputs by GPSVI and FITC for the third function. The legends are same as in Figure 1.}
\label{fig3}
\end{figure*}

\bibliographystyle{abbrv}
\bibliography{references}

\end{document}

### 2014-03-06 paper/sample2e.tex

% Sample LaTeX file for creating a paper in the Morgan Kaufmannn two
% column, 8 1/2 by 11 inch proceedings format.

\documentclass[]{article}
\usepackage{proceed2e}

\title{Instructions for Authors}

\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
          % UAI  reviewing is double-blind.

% The author names and affiliations should appear only in the accepted paper.
%
%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
%alternate address.}} \\
%Computer Science Dept. \\
%Cranberry University\\
%Pittsburgh, PA 15213 \\
%\And
%{\bf Coauthor}  \\
%Affiliation          \\
%Address \\
%\And
%{\bf Coauthor}   \\
%Affiliation \\
%Address    \\
%(if needed)\\
%}

\begin{document}

\maketitle

\begin{abstract}
The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
both left and right-hand margins.  Use 10~point type, with a vertical
spacing of 11~points.  {\bf Abstract} must be centered, bold, and in
point size 12. Two line spaces precede the Abstract. The Abstract must
be limited to one paragraph.
\end{abstract}

\section{GENERAL FORMATTING INSTRUCTIONS}

Papers are in 2 columns with the overall line width of 6.75~inches
(41~picas).  Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin
is 1~inch (6~picas).  Use 10~point type with a vertical spacing of
11~points.  Times Roman is the preferred typeface throughout.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal
rules.  Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Reviewing is double-blind, so do not include author names, affiliations, or any
other identifying information in the original submission.  If you include urls
to supplementary material, make sure the urls also do not disclose your identity.

After a paper is accepted, for the camera-ready submission, Authors' names are
centered, initial caps.  The lead author's name is to be listed first
(left-most), and the Co-authors' names (if different address) are set to
follow.  If only one co-author, center both the author and co-author,
side-by-side.

One-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold and in point size
12. One line space before the first level heading and 1/2~line space
after the first level heading.

\subsection{SECOND LEVEL HEADING}

Second level headings must be flush left, all caps, bold and in point
size 10. One line space before the second level heading and 1/2~line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings must be flush left, initial caps, bold, and in
point size 10.  One line space before the third level heading and
1/2~line space after the third level heading.

\vskip .5pc
Fourth Level Heading

Fourth level headings must be flush left and initial caps.
One line space before the fourth level heading and 1/2~line space
after the fourth level heading.

\subsection{CITATIONS, FIGURES, REFERENCES}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). Reference style should follow the style
that you are used to using, as long as the citation style is
consistent.

For the original submission, take care not to reveal the authors' identity through
the manner in which one's own previous work is cited.  For example, writing
``In (Bovik, 1970), we studied the problem of AI'' would be inappropriate, as
it reveals the author's identity.  Instead, write ``(Bovik, 1970) studied the
problem of AI.''

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
footnote} in the text. Use 8 point type for footnotes.  Place the
footnotes at the bottom of the page on which they appear.  Precede the
footnote with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible. Figure number
and caption always appear below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{1in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Table number
and title always appear above the table.  See
Table~\ref{sample-table}.

One line space before the table title, one line space after the table
title, and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title}
\label{sample-table}
\begin{center}
\begin{tabular}{ll}
\multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\newpage

\subsubsection*{Acknowledgements}

Use unnumbered third level headings for the acknowledgements title.
All acknowledgements go at the end of the paper.

\subsubsection*{References}

References follow the acknowledgements.  Use unnumbered third level
heading for the references title.  Any choice of citation style is
acceptable as long as you are consistent.

J.~Alspector, B.~Gupta, and R.~B.~Allen  (1989). Performance of a
stochastic learning microchip.  In D. S. Touretzky (ed.), {\it Advances
in Neural Information Processing Systems 1}, 748-760.  San Mateo, Calif.:
Morgan Kaufmann.

F.~Rosenblatt (1962). {\it Principles of Neurodynamics.} Washington,
D.C.: Spartan Books.

G.~Tesauro (1989). Neurogammon wins computer Olympiad.  {\it Neural
Computation} {\bf 1}(3):321-323.

\end{document}

### 2014-03-06 paper/appendix.tex
add figures and tables

@@
+\subsection{Expectation of an individual likelihood term}
+The expectation of an individual likelihood term with respect to the posterior distribution is given by:
+\begin{align}
+\nonumber
+l_{in} =& \int \log p(y_{in} | \g_n, h_{in}) \prod_{j=1}^Q p(g_{jn} | \u_j) p(h_{in} | \u_i) \der \g_n \der h_{in} \\
+\nonumber
+=& \log \Normal(y_{in}; \sum_{j=1}^Q w_{ij} \mu_{jn}+ \mu^h_{in}, \beta_i^{-1}) \\
+&- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} 
+- \frac{1}{2} \beta_i \tilde{k}^h_{inn},
+\label{eq:lin}
+\end{align}
+where $\tilde{k}_{nn} = (\tilde{\K})_{nn}, \tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}, \mu_{jn} = (\Mu_j)_n, \text{ and } \mu^h_{in} = (\Mu^h_i)_n$.
+
+% previous detailed derivation
+%\begin{align}
+%\nonumber
+%l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\g | \ug) p(\h_i | \u_i) \der \g \der \h_i \\
+%\nonumber
+%&= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
+%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
+%\Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
+%\nonumber
+%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta_i (y_{in} - w_i g_n - h_{in})
+%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
+%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der g_n \der h_{in} \\
+%\nonumber
+%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta_i (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta_i \tilde{k}_{gnn} \right] 
+%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
+%\nonumber
+%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta_i \tilde{k}_{gnn}
+%- \frac{1}{2} \beta_i \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta_i (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
+%&= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
+%\end{align}
+
 \subsection{Appendix 1: Derivatives}

### 2014-03-06 paper/experiments.tex
add figures and tables

@@
 % Say that this is from a single gpsvi
-The predictive distributions of the model compared to that of independent gps are shown in \ref{fig5}.
+The predictive distributions of the model compared to that of independent GPs with stochastic variational inference are shown in figure \ref{fig:toy}.
+Furthermore, the inference algorithm correctly learns the correlation between the two functions giving weights $w_{11} = 1.07$ and $w_{21} = -1.06$.
+
+\begin{figure*}
+\centering
+\begin{tabular}{cccc}
+\includegraphics[scale=0.25]{figures/toy-slfm-y1.eps} &
+\includegraphics[scale=0.25]{figures/toy-svigp-y1.eps} &
+\includegraphics[scale=0.25]{figures/toy-slfm-y2.eps} &
+\includegraphics[scale=0.25]{figures/toy-svigp-y2.eps}
+\end{tabular}
+\caption{Predictive distributions of the multioutput GPs (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the  toy problem.
+Solid black line: predictive mean; grey bar: two standard deviations; magenta dots: real observations; blue dots: missing data.
+The black crosses show the locations of the inducing inputs.}
+\label{fig:toy}
+\end{figure*}
+
+\subsection{EVALUATION OF THE PREDICTIVE ACCURACY}
+\subsubsection{Foregin Exchange Rate Prediction}
+In this experiment we predict the foreign exchange rate.
+The smse are shown in Table \ref{tab:fx}.
+
+\begin{table}[h]
+\caption{Performance comparison by SMSE on the fx dataset.}
+\label{tab:fx}
+\begin{center}
+\begin{tabular}{c|c}
+method & smse \\ \hline
+SLFM (Q=2) & 0.2232 \\
+SLFM (Q=1) & 0.48 \\
+CGP & 0.2795 \\
+ICM & 0.3927
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{figure*}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.3]{figures/fxAUD.eps} &
+\includegraphics[scale=0.3]{figures/fxCAD.eps} &
+\includegraphics[scale=0.3]{figures/fxJPY.eps}
+\end{tabular}
+\caption{Real observations and predictive distributions for AUD (left), CAD (middle), and JPY (right). The color coding scheme is the same as in figure \ref{fig:toy}.}
+\label{fig:fx}
+\end{figure*}
+
+\subsubsection{Weather Data Prediction}
 
@@
 \begin{tabular}{cc}
-\includegraphics[scale=0.5]{figures/ssvi2-y1.eps} &
-\includegraphics[scale=0.5]{figures/ssvi2-svi1.eps} \\
-\includegraphics[scale=0.5]{figures/ssvi2-y2.eps} &
-\includegraphics[scale=0.5]{figures/ssvi2-svi2.eps} \\
+\includegraphics[scale=0.3]{figures/weatherCambermet.eps} &
+\includegraphics[scale=0.3]{figures/weatherCambermet.eps} \\
+\includegraphics[scale=0.3]{figures/weatherChimet.eps} &
+\includegraphics[scale=0.3]{figures/weatherChimet.eps}
 \end{tabular}
-\label{fig5}
-\caption{Predictive distributions of the multioutput gps (left column) and independent gps (right column) for the second toy experiment.
-The predictive distribution by $g(x)$ for $y_1(x)$ is shown in the bottom figure.}
+\caption{Predictive distributions of the multioutput GPs (left column) and full independent GPs (right column) for the air temperature problem.}
+\label{fig:toy}
 \end{figure*}
-
-\subsection{EVALUATION OF THE PREDICTIVE ACCURACY}
-\subsubsection{Foregin Exchange Rate Prediction}
-\subsubsection{Weather Data Prediction}
 

### 2014-03-06 paper/inference.tex
add figures and tables

@@
 \end{align}
-Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given, we need only to find the optimum $q(\u_j)$ and $q(\v_i)$.
-Let $q(\u_j) = \Normal(\u_j; \mgj, \Sgj)$ and $q(\v_i) = \Normal(\v_i; \mhi, \Shi)$. \\
+Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given, we need only  find the optimum $q(\u_j) = \Normal(\u_j; \mgj, \Sgj)$ and $q(\v_i) = q(\v_i) = \Normal(\v_i; \mhi, \Shi)$.
 
 \noindent
-To find the optimum $q(\u, \v)$ we optimize the evidence lower bound (ELBO) of the log marginal,
+In variational inference, this is done by optimizing the evidence lower bound (ELBO) of the log marginal:
+\begin{align}
+%\nonumber
+%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
+%\nonumber
+%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
+%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
+\nonumber
+&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
+&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)],
+\label{eq:elbo}
+\end{align}
+which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
+Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ computation.
+To compute the expected likelihood term in the ELBO we first see that
 \begin{align}
 \nonumber
-\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
-\nonumber
-=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
-+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
-\label{eq:elbo}
-=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
-- \left(\sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] + \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)] \right),
+\log \text{ } &p(\y | \u, \v)
+% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
+% \nonumber
+\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
+&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_{in} | \v_i)} 
 \end{align}
-where the last equality occurs because both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
-Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ computation, where $M$ is the largest number of inducing inputs (recall that $g_j(\x)$ and $h_i(\x)$ use separate set of inducing inputs).
-\\
-
-\noindent To compute the ELBO we first focus on the term:
-\newcommand{\llangle}{\left\langle}
-\newcommand{\rrangle}{\right\rangle}
+where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
+The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
+Each individual term $l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ is computed using the identity in eq.
+\ref{eq:identity} and  given in eq \ref{eq:lin} (both are in the Appendix).
+\newcommand{\Ahi}{\A^h_i}
+\newcommand{\Zi}{\Z_i}
+ Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying out the integral using the identity in eq. \ref{eq:identity} again we get:
 \begin{align}
 \nonumber
-\log p(\y | \u, \v)
- &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
- \nonumber
-&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \quad &\text{(Jensen's inequality)} \\
-&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_{in} | \v_i)}, \quad 
- &\text{(factorized likelihood)}
+&\log p(\y)
+\ge \sum_{i,n}
+\bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
+          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} \\ \nonumber
+         &- \frac{1}{2} \beta_i \tilde{k}^h_{inn}
+         - \frac{1}{2} \beta_i \trace  \sum_{j=1}^Q w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
+\bigg) \\
+&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)]  \define \calL,
 \end{align}
-where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
-Each individual term $l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ can be computed using the identity in eq. \ref{eq:identity} given in the appendix to give:
-\begin{align}
-\nonumber
-l_{in} &= \int \log p(y_{in} | \g_n, h_{in}) \prod_{j=1}^Q p(g_{jn} | \u_j) p(h_{in} | \u_i) \der \g_n \der h_{in} \\
-&= \log \Normal(y_{in}; \sum_{j=1}^Q w_{ij} \mu_{jn}+ \mu^h_{in}, \beta_i^{-1})
-- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} 
-- \frac{1}{2} \beta_i \tilde{k}^h_{inn},
-\end{align}
-where $\tilde{k}_{nn} = (\tilde{\K})_{nn}, \tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}, \mu_{jn} = (\Mu_j)_n, \text{ and } \mu^h_{in} = (\Mu^h_i)_n$.
-
-% previous detailed derivation
-%\begin{align}
-%\nonumber
-%l_{in} &= \int \log p(y_{in} | g_n, h_{in}) p(\g | \ug) p(\h_i | \u_i) \der \g \der \h_i \\
-%\nonumber
-%&= \int \log \Normal(y_{in} ; w_i g_n + h_{in}, \beta_i^{-1}) 
-%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
-%\Normal(h_{in} ; \mu_{in}, (\tilde{k}_{inn}) \der g_n \der h_{in} \\
-%\nonumber
-%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int (y_{in} - w_i g_n - h_{in}) \beta_i (y_{in} - w_i g_n - h_{in})
-%\Normal(g_n ; \mu_{gn}, \tilde{k}_{gnn})
-%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der g_n \der h_{in} \\
-%\nonumber
-%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} \int \left[(y_{in} - h_{in} - w_i \mu_{gn}) \beta_i (y_{in} - h_{in} - w_i \mu_{gn}) + w_i^2 \beta_i \tilde{k}_{gnn} \right] 
-%\Normal(h_{in} ; \mu_{in}, \tilde{k}_{inn}) \der h_{in} \\
-%\nonumber
-%&= -\frac{1}{2} \log 2 \pi \beta_i^{-1} - \frac{1}{2} w_i^2 \beta_i \tilde{k}_{gnn}
-%- \frac{1}{2} \beta_i \tilde{k}_{inn} - \frac{1}{2} (y_{in} - w_i \mu_{gn} - \mu_{in}) \beta_i (y_{in} - w_i \mu_{gn} - \mu_{in}) \\
-%&= \log \Normal(y_{in}; w_i \mu_{gn} + \mu_{in}, \beta_i^{-1})  - \frac{1}{2} \beta_i (w_i^2 \tilde{k}_{gnn} + \tilde{k}_{inn}),
-%\end{align}
-
-\noindent Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying out the integral using the identity in eq. \ref{eq:identity} we get:
-\begin{align}
-\nonumber
-\log p(\y)
-\ge& \sum_{i=1}^P \sum_{n=1}^N
-\bigg( \log \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
-          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2        \tilde{k}_{nn} - \frac{1}{2} \beta_i \tilde{k}^h_{inn}
- \\ \nonumber
-         &\quad \quad \quad \quad - \frac{1}{2} \beta_i \trace  \sum_{j=1}^Q w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
-\bigg) \\
-&- \left(\sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] + \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)] \right) \define \calL,
-\end{align}
-where 
-%TODO \Lambda_gn depends on the output i so need a better notation
-\newcommand{\Zg}{\Z_g}
-\newcommand{\Zi}{\Z_i}
+where, with the help of the auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  
+$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$, 
 \begin{align}
 \tilde{\mu}_{in}
-&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
+%&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
+&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \\
 \mat{\Lambda}_{jn}
-&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
+%&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
+&= \A_j(n,:)^T \A_j(n,:)  \\
 \mat{\Lambda}_{in}
-&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}.
+%&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
+&= \Ahi(n,:)^T \Ahi(n,:) 
 \end{align}
+where $\tilde{k}_{nn} = (\tilde{\K})_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$,  $\mu^h_{in} = (\Mu^h_i)_n$, and $\A_j(n,:)$ selects the $n$-th row vector of $\A_j$.
 
+% More on this elbo term
 \noindent Notice that this ELBO clearly generalizes the standard GP regression.
@@
 
-\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
-\newcommand{\Ahi}{\A^h_i}
-\begin{align}
-\nonumber
-\calL^g_j \define&
- \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
- \nonumber
- &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
- + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
- \\
- &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
-\end{align}
-where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
-$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\
+%\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
+%\begin{align}
+%\nonumber
+%\calL^g_j \define&
+% \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
+% \nonumber
+% &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
+% + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
+% \\
+% &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
+%\end{align}
+%where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
+%$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\
 
@@
 \newcommand{\ynoj}{\y_i^{\backslash j}}
-\noindent The derivatives of $\Lgj$ wrt $\m_j$ and $\S_j$ are given by:
+\newcommand{\Kjzz}{\mat{K}_{jzz}}
+\noindent The derivatives of $\calL$ wrt $\m_j$ and $\S_j$ are given by:
 \begin{align}
-\deriv{\Lgj}{\m_j}
-& = \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi,:)^T \ynoj - \bigg[k_g(\Zg,\Zg)^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg] \m_j \\
-\deriv{\Lgj}{\S_j} 
-&= \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ k(\Zj,\Zj)^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg],
+\nonumber
+\deriv{\calL}{\m_j} 
+=& \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi,:)^T \ynoj \\
+&- \bigg[\Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg] \m_j \\
+\deriv{\calL}{\S_j} 
+=& \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg],
 \end{align}
-where $\y_i^{\backslash j} = \y_i - \Ahi \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi,:) \m_{j'}$.
+where $\Kjzz = k(\Zj,\Zj)$ and $\y_i^{\backslash j} = \y_i - \Ahi(\oi,:) \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi,:) \m_{j'}$.
 
@@
 %  derivatives of q(v_i)
+%\noindent As a function of the parameters of $q(\v_i)$, the lower bound $\calL$ is:
+%\begin{align}
+%\nonumber
+%\Lhi \define&
+% \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)
+% - \frac{1}{2} \beta_i \trace \tilde{\K}^h_i(\oi,\oi)
+% - \frac{1}{2} \beta_i \trace \Shi (\Ahi)^T \Ahi
+% \\
+%  &- \frac{1}{2} \log |k(\Zhi,\Zhi) (\Shi)^{-1}| -\frac{1}{2} \trace k(\Zhi,\Zhi)^{-1} (\mhi (\mhi)^T + \Shi) ,
+%\end{align}
 \newcommand{\Lhi}{\calL^h_i}
-\noindent As a function of the parameters of $q(\v_i)$, the lower bound $\calL$ is:
+\newcommand{\Kizz}{\mat{K}_{izz}}
+\noindent The derivatives of $\calL$ wrt $\mhi$ and $\Shi$ are given by:
+\newcommand{\ynoh}{\y_i^{\backslash h}}
 \begin{align}
 \nonumber
-\Lhi \define&
- \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)
- - \frac{1}{2} \beta_i \trace \tilde{\K}^h_i(\oi,\oi)
- - \frac{1}{2} \beta_i \trace \Shi (\Ahi)^T \Ahi
- \\
-  &- \frac{1}{2} \log |k(\Zhi,\Zhi) (\Shi)^{-1}| -\frac{1}{2} \trace k(\Zhi,\Zhi)^{-1} (\mhi (\mhi)^T + \Shi) ,
+\deriv{\calL}{\mhi}
+= & \beta_i (\Ahi(\oi,:))^T \ynoh \\ 
+& - \bigg[\Kizz^{-1} +  \beta_i (\Ahi(\oi,:))^T \Ahi(\oi,:) \bigg] \m_i \\
+\deriv{\calL}{\Shi} 
+&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \bigg[ \Kizz^{-1} + \beta_i \Ahi(\oi,:))^T \Ahi(\oi,:) \bigg] ,
 \end{align}
-
-\noindent The derivatives of $\Lhi$ wrt $\mhi$ and $\Shi$ are given by:
-\newcommand{\ynoh}{\y_i^{\backslash h}}
-\begin{align}
-\deriv{\Lhi}{\mhi}
-& = \beta_i \A_i^T \ynoh - \left[k(\Zhi,\Zhi)^{-1} +  \beta_i \A_i^T \A_i \right] \m_i \\
-\deriv{\Lhi}{\Shi} 
-&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \left[ k(\Zhi,\Zhi)^{-1} + \beta_i \A_i^T \A_i \right] ,
-\end{align}
-where $\ynoh = \y_i - \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j$.
+where $\ynoh = \y_i - \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j$ and $\Kizz = k(\Zhi, \Zhi)$.
 

### 2014-03-07 paper/experiments.tex
updated paper
clean figures

@@
 \subsection{TOY PROBLEM}
-In this toy problem we use two related outputs generated from the same latent function $sin(x)$ but are corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
+In this toy problem we simulate two related outputs from the same latent function $sin(x)$ corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
 Each output has missing values in one region of the input space.
@@
 \begin{tabular}{cccc}
-\includegraphics[scale=0.25]{figures/toy-slfm-y1.eps} &
-\includegraphics[scale=0.25]{figures/toy-svigp-y1.eps} &
-\includegraphics[scale=0.25]{figures/toy-slfm-y2.eps} &
-\includegraphics[scale=0.25]{figures/toy-svigp-y2.eps}
+\includegraphics[scale=0.2]{figures/toy-slfm-y1.eps} &
+\includegraphics[scale=0.2]{figures/toy-svigp-y1.eps} &
+\includegraphics[scale=0.2]{figures/toy-slfm-y2.eps} &
+\includegraphics[scale=0.2]{figures/toy-svigp-y2.eps}
+\label{fig:toy}
 \end{tabular}
@@
 
-\subsection{EVALUATION OF THE PREDICTIVE ACCURACY}
-\subsubsection{Foregin Exchange Rate Prediction}
+\subsection{FOREIGN EXCHANGE RATE PREDICTION}
 In this experiment we predict the foreign exchange rate.
@@
 SLFM (Q=2) & 0.2232 \\
-SLFM (Q=1) & 0.48 \\
+SLFM (Q=1) & 0.48xx \\
 CGP & 0.2795 \\
@@
 
-\subsubsection{Weather Data Prediction}
+\subsection{AIR TEMPERATURES PREDICTION}
+
+\begin{table}[h]
+\caption{Performance comparison by SMSE on the air temperature dataset.}
+\label{tab:fx}
+\begin{center}
+\begin{tabular}{c|c}
+method & smse \\ \hline
+SLFM & 0.0680 \\
+CGP & 0.0994 \\
+IGP & 0.8944
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 \begin{tabular}{cc}
-\includegraphics[scale=0.3]{figures/weatherCambermet.eps} &
+\includegraphics[scale=0.3]{figures/slfm-weatherCambermet.eps} &
 \includegraphics[scale=0.3]{figures/weatherCambermet.eps} \\
-\includegraphics[scale=0.3]{figures/weatherChimet.eps} &
+\includegraphics[scale=0.3]{figures/slfm-weatherChimet.eps} &
 \includegraphics[scale=0.3]{figures/weatherChimet.eps}
@@
 \caption{Predictive distributions of the multioutput GPs (left column) and full independent GPs (right column) for the air temperature problem.}
-\label{fig:toy}
+\label{fig:weather}
 \end{figure*}
 
-\subsection{LARGE-SCALE EXPERIMENTS}
-\subsubsection{Robort Arms}
-\subsubsection{Foreign Exchange Rate Prediction}
+\subsection{ROBOT ARMS}

### 2014-03-07 paper/inference.tex
clean figures

@@
 The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
-The key property of such augmented GPs  is the notation of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g.
+The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g.
 the test set).
@@
 &\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
-&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\u_i) || p(\u_i)],
+&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
 \label{eq:elbo}
@@
 \end{align}
-where, with the help of the auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  
+where, with the help of auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  
 $\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$, 
@@
 \noindent Notice that this ELBO clearly generalizes the standard GP regression.
-In particular, setting $P = Q = 1$, $w_i = 1$ and $h_i(\x) = 0$ we recover the bound in Hensman et al \cite{hensmangaussian}.
+In particular, setting $P = Q = 1$, $w_i = 1$ and $h_i(\x) = 0$ we recover the bound in Hensman et al \citet{hensmangaussian}.
 Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.

### 2014-03-07 paper/model.tex
clean figures

@@
 
-\noindent Consider the semiparametric latent factor models (SLFM) \cite{teh-et-al-aistats-05}.
+\noindent Consider the semiparametric latent factor models (SLFM) \citet{teh-et-al-aistats-05}.
 It assumes $Q$ basic independent processes, $g_j(\x) \sim \GP(0, k(\x,\x';\vectheta_j))$, where $\vectheta_j$ is the hyperparameters of the covariance function of $g_j(\x)$.
@@
 \end{align}
-In \cite{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
+In \citet{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
 \begin{align}
@@
 \begin{align}
-p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I).
+p(\y | \g, \h ) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I),
 \end{align}
+where each term in the product is the likelihood of an output, $p( \y_i ; \g, \h_i)$.

### 2014-03-07 src/tmp/slfm2_jura.m
misc remove

@@ -1,85 +1,0 @@
-clear all; clc; %close all;
-rng(1110,'twister');
-
-[x,y,xtest,ytest] = read_juraCd();
-% pre-process y
-y0 = y;
-y = log(y);
-[y,ymean,ystd] = standardize(y,[],[]);
-
-% ssvi
-cf.covfunc_g  = 'covSEard';
-cf.lrate      = 1e-2;
-cf.momentum   = 0.9;
-cf.lrate_hyp  = 1e-5;
-cf.lrate_beta = 1e-4;
-cf.momentum_w = 0.9;
-cf.lrate_w    = 1e-4;
-cf.learn_z    = false;
-cf.momentum_z = 0.0;
-cf.lrate_z    = 1e-4;
-cf.init_kmeans = false;
-cf.maxiter = 2000;
-cf.nbatch = 100;
-
-Q = 2;
-runs = 10;
-Ms = [5,10,20,50,100,150,200,259];
-maes = zeros(numel(Ms),runs);
-for i=1:numel(Ms)
-  M = Ms(i);
-  for j=1:runs
-    par.g = cell(Q,1);
-    [elbo,par] = slfm2_learn(x,y,M,par,cf);
-    [mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest,size(y,2));
-    mu = exp(mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1));
-    maes(i,j) = mean(abs(ytest(:,1) - mu(:,1)));
-    clear par;
-  end
-end
-
-mae_mean = mean(maes(1:4,:),2);
-mae_std = 2*std(maes(1:4,:),0,2);
-plot(Ms(1:4),mae_mean,'x-','MarkerSize',15);
-errorbar(Ms(1:4),mae_mean,mae_std);
-xlabel('number of inducing points');
-ylabel('mae');
-title('slfm fixed z, mae vs. inducing inputs');
-saveas(gcf, 'juraCd-slfm.eps','epsc');
-save('juraCd-slfm.mat','maes');
-
-% par.g = cell(Q,1);
-% [elbo,par] = slfm2_learn(x,y,M,par,cf);
-% [mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest,size(y,2));
-% mu = exp(mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1));
[…]

### 2014-03-07 src/init_params.m
misc update and refactor

@@ -1,4 +1,4 @@
-function params = init_params(x,y,M,nhyper,init_kmeans,z0)
+function params = init_params(x,y,M,nhyper,z0)
 %INIT_PARAMS params = init_params(x,y,M)
 %   Init parameters of a single augmented sparse gp.
 %
@@ -7,9 +7,6 @@ y(isnan(y)) = mean(y(~isnan(y))); %mean(y(~isnan(y))) or 0? for missing data
 idx_z = randperm(N,M);
 if isempty(z0)
   z0 = x(idx_z,:);
-  if init_kmeans
-    z0 = kmeans(z0, x, foptions());
-  end
 end  
 idx = randperm(N,M);
 m =  y(idx);

### 2014-03-07 src/plot_all.m

@@ -8,16 +8,16 @@ function plot_all(x,y,xtest,mu,var,z0,z,titlestr)
 %   - z0, z : initial and learned inducing
 %   - stitle : title string
 figure; hold on
-plot(xtest,mu,'b') % mean predictions in blue
-plotMeanAndStd(xtest,mu,2*sqrt(var),[7 7 7]/7.5);
+plotMeanAndStd(xtest,mu,2*sqrt(var),[7 7 7]/9);
 if ~isempty(z0)
-  plot(z0, -0.1 + min(y)*ones(size(z0,1), 1), 'k+', 'MarkerSize', 12);
+  plot(z0, -0.5 + min(y)*ones(size(z0,1), 1), 'k+', 'MarkerSize', 12);
 end
 if ~isempty(z)
-  plot(z, -0.2 + min(y)*ones(size(z,1), 1), 'm+', 'MarkerSize', 12);
+  plot(z, -0.5 + min(y)*ones(size(z,1), 1), 'm+', 'MarkerSize', 12);
 end  
-plot(x,y,'.m') % data points in magenta
-ylabel('Predictive distribution');
-title(titlestr);
+plot(x,y,'.m','MarkerSize',14) % data points in magenta
+%ylabel('Predictive distribution');
+title(titlestr,'FontSize',20);
+set(gca, 'FontSize', 20);
 
 return;

### 2014-03-07 src/scripts/demo_slfm.m

@@ -1,12 +1,10 @@
 clear all; clc; %close all;
-%rng(1111,'twister');
+rng(1111,'twister');
 
 % sin(x) + cos(x)
 x = linspace(-10,10,200)';
 N = size(x,1);
 xtest = linspace(-11,11)';
-% fn1 = @(x) 0.2*x + sin(x) + 1e-7;
-% fn2 = @(x) 0.2*x + sin(x) + 1e-7;
 fn1 = @(x) sin(x) + 1e-7;
 fn2 = @(x) -sin(x) + 1e-7;
 y1 = fn1(x) + sqrt(1e-4)*randn(N,1);
@@ -20,7 +18,7 @@ y2miss = y2;
 y2miss(ind2) = nan;
 y = [y1miss, y2miss];
 %y = y1;
-M.g = 100; M.h = 20;
+Mstr.g = 15; Mstr.h = 10;
 
 % ssvi
 cf.covfunc_g  = 'covSEard';
@@ -32,24 +30,34 @@ cf.lrate_beta = 1e-5;
 cf.momentum_w = 0.9;
 cf.lrate_w    = 1e-5;
 cf.learn_z    = false;
-cf.momentum_z = 0.0;
-cf.lrate_z    = 5e-4;
-cf.init_kmeans = true;
+cf.momentum_z = 0.9;
+cf.lrate_z    = 1e-4;
 cf.maxiter = 500;
-cf.nbatch = 50;
+cf.nbatch = 20;
+cf.beta = 1/0.01;
+cf.initz = 'random';
+cf.monitor_elbo = 1;
+cf.w = ones(size(y,2),1);
+cf.fix_first = true;
+M = 5;
 
-Q = 2;
+Q = 1;
 par.task = cell(size(y,2),1);
 par.g = cell(Q,1);
-[elbo,par] = slfm_learn(x,y,M,par,cf);
+[elbo,par] = slfm_learn(x,y,Mstr,par,cf);
 [mu,var,mu_g,var_g] = slfm_predict(cf.covfunc_g,cf.covfunc_h,par,xtest);
+% [elbo,par] = slfm2_learn(x,y,M,par,cf);
+% [mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest,size(y,2));
 
-% plot_all(x,y1,xtest,mu(:,1),var(:,1),[],par.task{1}.z,'y1');
-% axis([-11 11 -4 4]);
[…]

### 2014-03-07 src/scripts/demo_ssvi.m

@@ -1,74 +1,0 @@
-clear all; clc; %close all;
-%rng(1111,'twister');
-
-% sin(x) + cos(x)
-x = linspace(-10,10)';
-N = size(x,1);
-xtest = linspace(-11,11)';
-% fn1 = @(x) 0.2*x + sin(x) + 1e-7;
-% fn2 = @(x) 0.2*x + sin(x) + 1e-7;
-fn1 = @(x) sin(x) + 1e-7;
-fn2 = @(x) -sin(x) + 1e-7;
-y1 = fn1(x) + sqrt(1e-4)*randn(N,1);
-y2 = fn2(x) + sqrt(1e-4)*randn(N,1);
-
-ind1 = -7 < x & x < -3;
-ind2 = 4 < x & x < 8;
-y1miss = y1;
-y1miss(ind1) = nan;
-y2miss = y2;
-y2miss(ind2) = nan;
-y = [y1miss, y2miss];
-%y = y1;
-M.g = 20; M.h = 20;
-
-% ssvi
-cf.covfunc_g  = 'covSEard';
-cf.covfunc_h  = 'covNoise';
-cf.lrate      = 1e-2;
-cf.momentum   = 0.9;
-cf.lrate_hyp  = 1e-5;
-cf.lrate_beta = 1e-4;
-cf.momentum_w = 0.9;
-cf.lrate_w    = 1e-5;
-cf.learn_z    = true;
-cf.momentum_z = 0.0;
-cf.lrate_z    = 1e-4;
-cf.init_kmeans = false;
-cf.maxiter = 500;
-cf.nbatch = 10;
-
-[elbo,params] = ssvi_learn(x,y,M,cf,[]);
-[mu,var,mu_g,var_g] = ssvi_predict(cf.covfunc_g,cf.covfunc_h,params,xtest);
-
-plot_all(x,y1,xtest,mu(:,1),var(:,1),[],params.task{1}.z,'y1');
-axis([-11 11 -4 4]);
-plot_all(x,y2,xtest,mu(:,2),var(:,2),[],params.task{2}.z,'y2');
-axis([-11 11 -4 4]);
-plot_all(x,y1,xtest,mu_g(:,1),var_g(:,1),params.g.z0,params.g.z,'y1 by g');
-axis([-11 11 -4 4]);
-
-figure;
-plot(1:numel(elbo),elbo);
-ylabel('elbo')
-xlabel('iteration')
[…]

### 2014-03-07 src/slfm_learn.m

@@ -15,7 +15,7 @@ function [elbo,par] = slfm_learn(x,y,M,par,cf)
 P = numel(par.task); Q = numel(par.g);
 par.w = ones(P,Q);
 par.delta_w = zeros(P,Q);
-par.beta = (1/0.05)*ones(P,1);
+par.beta = cf.beta*ones(P,1);
 par.delta_beta = zeros(P,1);
 observed = ~isnan(y);
 nhyper_g = eval(feval(cf.covfunc_g));
@@ -26,12 +26,12 @@ for i=1:P
 end
 ymean = mean(ytmp,2);
 for j=1:Q
-  par.g{j} = init_params(x,ymean,M.g,nhyper_g,0,[]);
+  par.g{j} = init_params(x,ymean,M.g,nhyper_g,initz(x,M.g,cf.initz));
   par.g{j}.m = zeros(size(par.g{j}.m));
 end
 nhyper_h = eval(feval(cf.covfunc_h));
 for i=1:P
-  par.task{i} = init_params(x,y(:,i),M.h,nhyper_h,0,[]);
+  par.task{i} = init_params(x,y(:,i),M.h,nhyper_h,initz(x,M.h,cf.initz));
   %par.task{i}.m = zeros(size(par.task{i}.m));
   if strcmp(cf.covfunc_h,'covNoise')
     par.task{i}.loghyp = log(0.05); % learning of noise is hard so set it small here

### 2014-03-07 src/ssvi_learn.m

@@ -1,94 +1,0 @@
-function [elbo,params] = ssvi_learn(x,y,M,cf,z0)
-%SSVI_LEARN ssvi_learn(x,y,M,cf,z0)
-%   Quick wrapper to run structured GPSVI.
-% 
-% INPUT
-%   - x,y : train inputs, train outputs
-%   - M : #inducing inputs
-%   - cf : configuration
-%
-% OUTPUT
-%   - mu, s2 : mean and variance prediction
-%   - elbo : the bound
-%   - params : learned parameters
-[N D] = size(x);
-% get observation indice for each output
-P = size(y,2);
-ytmp = y;
-observed = ~isnan(y);
-for i=1:P
-  yi = y(observed(:,i),i);
-  ytmp(~observed(:,i),i) = mean(yi);
-end
-% for each input, the mean of all outputs are used to initialize m_g
-ymean = mean(ytmp,2);
-nhyper_g = eval(feval(cf.covfunc_g));
-nhyper_h = eval(feval(cf.covfunc_h));
-params.g = init_params(x,ymean,M.g,nhyper_g,0,[]);
-params.g.m = zeros(size(params.g.m));
-params.g.beta = []; params.g.delta_beta = [];
-for i=1:P
-  % this allows inducing inputs to appear in even unobserved region
-  yi = y(:,i); yi(isnan(yi)) = 0;
-  params.task{i} = init_params(x,yi,M.h,nhyper_h,0,[]);
-  %params.task{i}.m = zeros(size(params.task{i}.m));
-  if strcmp(cf.covfunc_h,'covNoise')
-    params.task{i}.loghyp = log(0.01); % learning of noise is hard so set it small here
-  end
-  params.task{i}.beta = 1/0.01;
-  params.task{i}.delta_beta = 0;
-end
-params.w = ones(P,1);
-params.delta_w = zeros(P,1);
-elbo = zeros(numel(cf.maxiter),1);
-for iter = 1:cf.maxiter
-  idx = randperm(N, cf.nbatch);
-  params.idx = ~isnan(y(idx,:));
-  xbatch = x(idx,:); ybatch = y(idx,:);
-  % variational parameters of g
-  [params,A,Knm,Kmminv,Lmm] = ssvi_update_g(xbatch,ybatch,params,cf);
-  % hyperparameters of g
-  if cf.learn_z
[…]

### 2014-03-07 src/ssvi_predict.m

@@ -1,21 +1,0 @@
-function [mu var, mu_g, var_g] = ssvi_predict(covfunc_g, covfunc_h, params, xstar)
-%SSVI_PREDCIT [mu var, mu_g, var_g] = ssvi_predict(covfunc_g, covfunc_h, params, xstar)
-%   Makes predictions with ssvi.
-%
-g = params.g;
-[mu_g, var_g] = gpsvi_predict(covfunc_g, g.loghyp, g.m, g.S, g.z, xstar);
-P = numel(params.task);
-mu = zeros(size(xstar,1),P);
-var = mu;
-for i=1:P
-  h = params.task{i};
-  [mu(:,i), var(:,i)] = gpsvi_predict(covfunc_h, h.loghyp, h.m, h.S, h.z, xstar);
-end
-%mu_g = repmat(mu_g,1,P);
-%var_g = repmat(var_g,1,P);
-for i=1:P
-  w = params.w(i);
-  mu(:,i) = w*mu_g + mu(:,i);
-  var(:,i) = w*w*var_g + var(:,i);
-end
-

### 2014-03-07 src/ssvi_update_g.m

@@ -1,44 +1,0 @@
-function [params,A,Knm,Kmminv,Lmm] = ssvi_update_g(x,y,params,cf)
-%SSVI_UPDATE_G [params,A,Knm,Kmminv,Lmm] = ssvi_update_g(x,y,params,cf)
-%   
-% Update the variational parameters of the shared process g in a structured
-% gp model.
-%
-% INPUT
-%   - x : inputs (in SVI, this is a mini-batch)
-%   - y : outputs (of inputs x)
-%   - params
-%   - cf : options for the optimization procedure
-%
-% OUTPUT
-%   - params : updated parameters
-%   - A,Knm,Kmminv,Lmm : saved computation (covariance of g)
-%
-% Trung V. Nguyen
-% 20/02/14
-P = size(y,2);
-[A,Knm,Kmminv,Lmm] = computeKnmKmminv(cf.covfunc_g, params.g.loghyp, x, params.g.z);
-Lambda = Kmminv;
-tmp = zeros(size(params.g.m));
-for i=1:P
-  w = params.w(i); w2 = w*w;
-  betaval = params.task{i}.beta;
-  indice = params.idx(:,i);
-  Lambda = Lambda + betaval*w2*(A(indice,:)')*A(indice,:);
-  Ai = computeKnmKmminv(cf.covfunc_h, params.task{i}.loghyp, x(indice,:), params.task{i}.z);
-  y_minus_hi = y(indice,i) - Ai*params.task{i}.m;
-  tmp = tmp + betaval*w*A(indice,:)'*y_minus_hi;
-end
-
-Sinv = invChol(jit_chol(params.g.S));
-theta1_old = Sinv*params.g.m;
-theta2_old = -0.5*Sinv;
-theta2 = theta2_old + cf.lrate*(-0.5*Lambda - theta2_old);
-theta1 = theta1_old + cf.lrate*(tmp - theta1_old);
-[VV DD] = eig(theta2);
-invTheta2 = VV*diag(1./diag(DD))*VV';
-params.g.S = - 0.5*invTheta2;
-params.g.m = params.g.S*theta1;
-
-end
-

### 2014-03-07 src/svi_learn.m

@@ -13,10 +13,14 @@ function [mu,s2,elbo,par] = svi_learn(x,y,xtest,M,cf,z0)
 %   - params : learned parameters
 [N D] = size(x);
 nhyper = eval(feval(cf.covfunc));
-par = init_params(x,y,M,nhyper,0,z0);
-par.beta = 1/0.05;
+if isempty(z0)
+  par = init_params(x,y,M,nhyper,initz(x,M,cf.initz));
+else
+  par = init_params(x,y,M,nhyper,z0);
+end
+par.beta = cf.beta;
 par.delta_beta = 0;
-elbo = zeros(numel(cf.maxiter),1);
+elbo = zeros(cf.maxiter,1);
 for iter = 1 : cf.maxiter
   idx = randperm(N, cf.nbatch);
   xi = x(idx,:); yi = y(idx);
@@ -33,9 +37,19 @@ for iter = 1 : cf.maxiter
         cf.momentum_z, cf.lrate_z);
     end
   end
-  % compute elbo using updated parameters
-  elbo(iter) = svi_elbo(x,y,par,cf.covfunc,[],[],[],[]);
-  fprintf('Iteration\t%d:\t%.4f\n',iter,elbo(iter));
+  % compute global elbo using updated parameters
+  if ~isempty(cf.monitor_elbo) && mod(iter,cf.monitor_elbo) == 0
+    nn = ceil(N/2000);
+    for i=1:nn-1
+      indice = ((i-1)*2000+1):(i*2000);
+      elbo(iter) = elbo(iter) + svi_elbo(x(indice,:),y(indice),par,cf.covfunc,[],[],[],[]);
+    end
+    indice = ((nn-1)*2000+1):N;
+    elbo(iter) = elbo(iter) + svi_elbo(x(indice,:),y(indice),par,cf.covfunc,[],[],[],[]);
+    fprintf('Iteration\t%d:\t%.4f\n',iter,elbo(iter));
+  else
+    disp(['finished iter ' num2str(iter)]);
+  end
 end
 
 if ~isempty(xtest)

### 2014-03-07 src/tmp/slfm2_learn.m

@@ -13,7 +13,7 @@ function [elbo,par] = slfm2_learn(x,y,M,par,cf)
 [N D] = size(x);
 % get observation indice for each output
 P = size(y,2); Q = numel(par.g);
-par.w = ones(P,Q);
+par.w = cf.w;
 par.delta_w = zeros(P,Q);
 par.beta = cf.beta*ones(P,1);
 par.delta_beta = zeros(P,1);
@@ -26,13 +26,15 @@ for i=1:P
 end
 ymean = mean(ytmp,2);
 for j=1:Q
-  par.g{j} = init_params(x,ymean,M,nhyper_g,0,initz(x,M,cf.initz));
+  par.g{j} = init_params(x,ymean,M,nhyper_g,initz(x,M,cf.initz));
   par.g{j}.m = zeros(size(par.g{j}.m));
   if strcmp(cf.initz,'espaced')
     par.g{j}.m = par.g{j}.m + 1e-4*randn(M,1); % small pertubance to break symmetry
   end
 end
-elbo = zeros(numel(cf.maxiter),1);
+% par.g{1}.loghyp = log([0.5; 1]);
+% par.g{2}.loghyp = log(ones(size(par.g{2}.loghyp)));
+elbo = zeros(cf.maxiter,1);
 for iter = 1:cf.maxiter
   idx = randperm(N, cf.nbatch);
   par.idx = ~isnan(y(idx,:));
@@ -40,32 +42,46 @@ for iter = 1:cf.maxiter
   
   % variational parameters of g_j
   par = slfm2_update_g(xbatch,ybatch,par,cf);
-  
-  % hyperparameters of g_j
-  [~,dbeta,dw,dloghyp] = slfm2_elbo(xbatch,ybatch,par,cf);
-  for j=1:Q
-    [par.g{j}.loghyp,par.g{j}.delta_hyp] = stochastic_update(...
-      par.g{j}.loghyp,par.g{j}.delta_hyp,dloghyp{j},cf.momentum,cf.lrate_hyp);
-  end
-  
-  % update w, beta
-  dw = reshape(cell2mat(dw),P,Q);
-  [par.w,par.delta_w] = stochastic_update(par.w,par.delta_w,dw,cf.momentum_w,cf.lrate_w);
-  [par.beta,par.delta_beta] = stochastic_update(par.beta,par.delta_beta,dbeta,cf.momentum,cf.lrate_beta);
-
-  % update inducing of g (using new hyperparameters)
-  if cf.learn_z
[…]

### 2014-03-07 src/scripts/slfm_fx.m
script for running fx and weather

@@ -1,3 +1,4 @@
+%% reading data and model setting
 clear all; %clc; %close all;
 rng(1110,'twister');
 
@@ -5,7 +6,6 @@ rng(1110,'twister');
 [x,y,xtest,ytest,y0] = read_fx();
 [y,ymean,ystd] = standardize(y,[],[]);
 
-%% ssvi
 cf.covfunc_g  = 'covSEard';
 cf.lrate      = 1e-2;
 cf.momentum   = 0.9;
@@ -16,17 +16,25 @@ cf.lrate_w    = 1e-5;
 cf.learn_z    = false;
 cf.momentum_z = 0.0;
 cf.lrate_z    = 1e-4;
-cf.init_kmeans = false;
 cf.maxiter = 500;
 cf.nbatch = 200;
 cf.beta = 1/0.1;
 cf.initz = 'random';
-
+cf.w = ones(size(y,2),2);
+cf.monitor_elbo = 50;
+cf.fix_first = false;
 M = 100;
-Q = 1;
+Q = 2;
+xtest = cell(3,1);
+xtest{1} = (50:100)';
+xtest{2} = (100:150)';
+xtest{3} = (150:200)';
+outputs = [4,6,9];
+
+%% single run 
 par.g = cell(Q,1);
 [elbo,par] = slfm2_learn(x,y,M,par,cf);
-[mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest,size(y,2));
+[mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,x,size(y,2));
 mu =mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1);
 
 disp('elbo = ')
@@ -36,18 +44,13 @@ disp(par.w)
 
 figure;
 %plot(1:numel(elbo),elbo);
+elbo = elbo(elbo ~= 0);
 semilogy(1:numel(elbo),-elbo);
 ylabel('elbo')
 xlabel('iteration')
 title(['elbo vs. iteration, lrate = ' num2str(cf.lrate_z)])
 
-%%
-outputs = [4,6,9];
 titles = {'CAD','JPY','AUD'};
-xtest = cell(3,1);
-xtest{1} = (50:100)';
-xtest{2} = (100:150)';
-xtest{3} = (150:200)';
 disp('smse of CAD, JPY, AUD')
 theaxis{1} = [0 250 0.75 1.15];
[…]

### 2014-03-07 src/scripts/slfm_weather.m

@@ -1,0 +1,115 @@
+%% reading data and model setting
+clear all; %clc; %close all;
+rng(1110,'twister');
+
+%explorary analysis
+[x,y,xtest,ytest,y0] = read_weather();
+
+% pre-process y
+[y,ymean,ystd] = standardize(y,[],[]);
+
+cf.covfunc_g  = 'covSEard';
+cf.lrate      = 1e-2;
+cf.momentum   = 0.9;
+cf.lrate_hyp  = 1e-5;
+cf.lrate_beta = 1e-4;
+cf.momentum_w = 0.9;
+cf.lrate_w    = 1e-5;
+cf.learn_z    = false;
+cf.momentum_z = 0.0;
+cf.lrate_z    = 1e-4;
+cf.maxiter = 1500;
+cf.nbatch = 500;
+cf.beta = 1/0.1;
+cf.initz = 'random';
+cf.w = ones(size(y,2),2);
+cf.monitor_elbo = 100;
+cf.fix_first = true;
+M = 150;
+Q = 2;
+test_ind{1} = x >= 10.2 & x <= 10.8;
+test_ind{2} = x >= 13.5 & x <= 14.2;
+outputs = [2,3];
+
+% a good initialization for hyperparameters
+% par.g{1}.loghyp = log([0.5; 1]);
+% par.g{2}.loghyp = log(ones(size(par.g{2}.loghyp)));
+
+%% single run 
+par.g = cell(Q,1);
+[elbo,par] = slfm2_learn(x,y,M,par,cf);
+[mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,x,size(y,2));
+mu =mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1);
+
+disp('elbo = ')
+disp(elbo(end))
+disp('learned w = ')
+disp(par.w)
+
+figure;
+%plot(1:numel(elbo),elbo);
+elbo = elbo(elbo ~= 0);
+semilogy(1:numel(elbo),-elbo);
+ylabel('elbo')
+xlabel('iteration')
+title(['elbo vs. iteration, lrate = ' num2str(cf.lrate_z)])
+
+titles = {'Cambermet','Chimet'};
+smses = zeros(2,1);
+for i=1:2
+  t = outputs(i);
+  plot_all(x,y0(:,t),x,mu(:,t),(ystd(t).^2)*var(:,t),[],[],titles{i});
[…]

### 2014-03-07 src/scripts/sarcos.m
script to run sarcos

@@ -18,46 +18,46 @@ clear sarcos_inv
 clear sarcos_inv_test
 
 %% Exploratory analysis of sarcos
-correlation = zeros(7,7);
-for i=1:7
-  for j=1:7
-    correlation(i,j) = corr(y(:,i),y(:,j));
-  end
-end
-
-disp('correlation: ')
-correlation
-
-outputs = [2,3,4,7];
-for i=1:numel(outputs)-1
-  for j=i+1:numel(outputs)
-    figure; scatter(y(:,i),y(:,j)); title(['output ' num2str(i) ' and ' num2str(j)']);
-  end
-end
+% correlation = zeros(7,7);
+% for i=1:7
+%   for j=1:7
+%     correlation(i,j) = corr(y(:,i),y(:,j));
+%   end
+% end
+% 
+% disp('correlation: ')
+% correlation
+% 
+% outputs = [2,3,4,7];
+% for i=1:numel(outputs)-1
+%   for j=(i+1):numel(outputs)
+%     figure; scatter(y(:,i),y(:,j)); title(['output ' num2str(i) ' and ' num2str(j)']);
+%   end
+% end
 
 %% independent models - fitc
-rng(1110,'twister');
-M = 200;
-runs = 5;
-maes = zeros(runs,size(y,2));
-rmses = maes;
-smses = maes;
-for n=1:runs
-  for i=1:size(y,2)
-    [z,z0,~,mu,s2] = learn_fitc(x,y(:,i),xtest,M,[]);
-    mu = mu*ystd(i)+ymean(i);
-    disp('mae    rmse')
-    fprintf('%.4f   %.4f\n', fmae(ytest(:,i),mu), frmse(ytest(:,i),mu));
-    maes(n,i) = fmae(ytest(:,i),mu);
-    rmses(n,i) = frmse(ytest(:,i),mu);
-    smses(n,i) = mysmse(ytest(:,i),mu,ymean(i));
-  end  
-end  
-save(['sarcos-fitc-M' num2str(M)],'maes','rmses','smses');
-disp('average mae, rmse, smse')
-disp(mean(maes))
-disp(mean(rmses))
-disp(mean(smses))
+% rng(1110,'twister');
+% M = 200;
+% runs = 5;
+% maes = zeros(runs,size(y,2));
+% rmses = maes;
+% smses = maes;
+% for n=1:runs
+%   for i=1:size(y,2)
[…]

### 2014-03-07 src/scripts/sarcos.m
sarcos script

@@ -76,24 +76,32 @@ cf.beta = 1/0.01;
 cf.initz = 'random';
 cf.monitor_elbo = 50;
 
-M = 200;
-runs = 1;
-maes = zeros(runs,size(y,2));
-smses = maes;
-for n=1:runs
-  for i=1:size(y,2)
-    [mu,s2,elbo,params] = svi_learn(x,y(:,i),xtest,M,cf,[]);
-    mu = mu*ystd(i)+ymean(i);
-    disp('mae    smse')
-    fprintf('%.4f   %.4f\n', fmae(ytest(:,i),mu), frmse(ytest(:,i),mu));
-    maes(n,i) = fmae(ytest(:,i),mu);
-    smses(n,i) = mysmse(ytest(:,i),mu,ymean(i));
+Ms = [1000, 800, 600, 400, 200];
+runs = 5;
+for idx=1:numel(Ms)
+  maes = zeros(runs,size(y,2));
+  smses = maes;
+  nlpds = maes;
+  M = Ms(idx);
+  for n=1:runs
+    for i=1:size(y,2)
+      [mu,s2,elbo,params] = svi_learn(x,y(:,i),xtest,M,cf,[]);
+      mu = mu*ystd(i)+ymean(i);
+      fvar = s2*(ystd(i)^2);
+      disp('mae    smse')
+      fprintf('%.4f   %.4f\n', fmae(ytest(:,i),mu), frmse(ytest(:,i),mu));
+      maes(n,i) = fmae(ytest(:,i),mu);
+      smses(n,i) = mysmse(ytest(:,i),mu,ymean(i));
+      nlpds(n,i) = mynlpd(ytest(:,i),mu,fvar);
+    end  
   end  
-end  
+  disp('average mae, smses, nlpd')
+  disp(mean(maes))
+  disp(mean(smses))
+  disp(mean(nlpds))
+  save(['sarcos-svi-M' num2str(M)],'maes','smses','nlpds')
+end
 
-disp('average mae and smses')
-disp(mean(maes))
-disp(mean(smses))
 
 %% independent model - soD
 % rng(1110,'twister');

### 2014-03-11 paper/experiments.tex
updated figures and description of experiment on the fx dataset

@@
-We perform three sets of experiments to evaluate the model and the stochastic variational inference proposed. 
-First, we use a toy problem to demonstrate the effectiveness of the proposed method in learning the model parameters.
-Second, we assess the predictive accuracy of the model on two medium-sized datasets and compare with existing multiple output models.
-In the final experiments, we evaluate the model on large-scale learning with multiple tasks.
+We evaluate the proposed approach using a set of four experiments.
+A toy problem is first used to study the transfer of learning between two related processes through the shared inducing points.
+We then compare the model with existing multioutput models on the tasks of predicting foreign exchange rates and air temperature.
+In the final experiment, we test the model on the prediction of robot arms dynamic on a large-scale dataset. 
+All experiments are executed on an Intel(R) Core(TM) i7-2600 3.40GHz CPU with 8GB of RAM using Matlab R2012a.
+%except for large scale?
 
 \subsection{TOY PROBLEM}
-In this toy problem we simulate two related outputs from the same latent function $sin(x)$ corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
-Each output has missing values in one region of the input space.
-% Say that this is from a single gpsvi
-The predictive distributions of the model compared to that of independent GPs with stochastic variational inference are shown in figure \ref{fig:toy}.
-Furthermore, the inference algorithm correctly learns the correlation between the two functions giving weights $w_{11} = 1.07$ and $w_{21} = -1.06$.
+In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
+The first output has missing observations in the $(-7,-3)$ interval and the second output in the $(4,8)$ interval.
+The predictive distributions by our model ($Q = 1$) and independent stochastic GPs (one for each output) are shown in \ref{fig:toy}.
+It can be seen that our multioutput model gives perfect prediction in the unobserved regions where the independent models fail to interpolate despite having the same set of inducing inputs.
+This confirms that transfer of information from one output to another via collaborative learning of the posterior over the shared inducing inputs.
+Furthermore, the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which precisely reflects the correlation between the two outputs.
 
@@
 \subsection{FOREIGN EXCHANGE RATE PREDICTION}
-In this experiment we predict the foreign exchange rate.
-The smse are shown in Table \ref{tab:fx}.
+The application considered here is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca/data.html}.
+The setting of our experiment described here is identical to that in \citet{alvarez2010efficient}.
+The dataset contains all the data available for the 251 working days in the year of 2007.
+There are 9, 8, and 42 days of missing values  for gold, silver, and platinum, respectively.
+We remove from the data the exchange rate of CAD on days 50-100, JPY on day 100-150, and AUD on day 150-200.
+Note that these 3 currencies are from very different geographical locations. 
+The 153 points extracted is used for testing and the remaining 3051 data points is used for training.
+Since the missing data is long contiguous sections, the objective is to evaluate the capacity of the model to impute the missing currency values based on other currencies.
+%todo: batch size, learn rate (maybe at the beginning of experiments)
+
+For preprocessing we normalized the outputs to have zero mean and unit variance.
+Since the exchange rates are driven by a small number of latent market forces \citet{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
+We used the squared-exponential covariance function for the shared processes and a noise covariance function for the independent process of each output.
+The inducing inputs $(M = 100)$ are randomly selected from the training set and fixed throughout training.
+
+The predictive distributions by our model (Figure \ref{fig:fx}) exhibit similar behavior to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
+In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
+Our investigation of the dataset found that 4 other currencies (GBP, NZD, KRW, MXN) also experienced the same trend during the days 150 - 200.
+This was effectively used by the model to extrapolate the values of the AUD.
+
+%todo: result for independent gps
+We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to that of CGP variants and independent GPs. 
+Our model outperforms both variants of the convolved GPs in terms of standardized mean-squared error (SMSE).
+It has lower test likelihood, as measured by the negative log predictive density (NLPD), which is mainly due to the less conservative predictive variance of the full CGP for the CAD currency.
+Note that the SMSE of CGPVAR is taken from \citet{alvarez2010efficient} while the NLPD was not provided.
+Training took only 10 minutes for our model compared to 1.4 hours of the full CGP model.
 
 \begin{table}[h]
-\caption{Performance comparison by SMSE on the fx dataset.}
+\caption{Performance comparison on the foreign exchange rate dataset.}
 \label{tab:fx}
 \begin{center}
-\begin{tabular}{c|c}
-method & smse \\ \hline
-SLFM (Q=2) & 0.2232 \\
-SLFM (Q=1) & 0.48xx \\
-CGP & 0.2795 \\
-ICM & 0.3927
+\begin{tabular}{c|c|c}
+METHOD & SMSE & NLPD \\ \hline
+COGP  & \textbf{0.2125} & -0.8394 \\
+CGPVAR & 0.2795 & NA \\
+CGP & 0.2427 & \textbf{-2.9474} \\
+%ICM & 0.3927 &
 \end{tabular}
@@
 \begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/fxAUD.eps} &
 \includegraphics[scale=0.3]{figures/fxCAD.eps} &
-\includegraphics[scale=0.3]{figures/fxJPY.eps}
+\includegraphics[scale=0.3]{figures/fxJPY.eps} &
+\includegraphics[scale=0.3]{figures/fxAUD.eps}
 \end{tabular}
-\caption{Real observations and predictive distributions for AUD (left), CAD (middle), and JPY (right). The color coding scheme is the same as in figure \ref{fig:toy}.}
+\caption{Real observations and predictive distributions for CAD (left), JPY (middle), and AUD (right). The color coding scheme is the same as in figure \ref{fig:toy}.}
 \label{fig:fx}
@@
 
-\subsection{AIR TEMPERATURES PREDICTION}
+\subsection{AIR TEMPERATURE PREDICTION}
+Two latent functions, inverse lengthscale 136 capturing global trend of increasing temperature from day 10 - 15 and inverse lengthscale of 0.5 to model the fluctuation within a day.
+Training for full cgp model with 2 latent functions took 3 hours compared to ... in our model.
 
 \begin{table}[h]
-\caption{Performance comparison by SMSE on the air temperature dataset.}
+\caption{Performance comparison by SMSE and NLPD on the air temperature dataset.}
 \label{tab:fx}
 \begin{center}
-\begin{tabular}{c|c}
-method & smse \\ \hline
-SLFM & 0.0680 \\
-CGP & 0.0994 \\
-IGP & 0.8944
+\begin{tabular}{c|c|c}
+method & smse & nlpd \\ \hline
+COGP & \textbf{0.1077} & \textbf{2.1712} \\
+CGP & 0.1125 & 2.2219 \\
+IGP & 0.8944 & 12.5319
 \end{tabular}
@@
 \centering
-\begin{tabular}{cc}
+\begin{tabular}{ccc}
 \includegraphics[scale=0.3]{figures/slfm-weatherCambermet.eps} &
-\includegraphics[scale=0.3]{figures/weatherCambermet.eps} \\
+\includegraphics[scale=0.3]{figures/cgp-weatherCambermet.eps} &
+\includegraphics[scale=0.3]{figures/weatherCambermet.eps}
+\\
 \includegraphics[scale=0.3]{figures/slfm-weatherChimet.eps} &
+\includegraphics[scale=0.3]{figures/cgp-weatherChimet.eps} &
 \includegraphics[scale=0.3]{figures/weatherChimet.eps}

### 2014-03-11 src/gpsvi_predict.m
mod for easier initialization

@@ -33,6 +33,6 @@ for i=1:nbatch
   var_2 =  sum(Kms.*(Kmminv*Kms),1)';   
   vaar = [vaar; var_1 + Kss - var_2];
 end
-vaar = max(vaarr,1e-10); % remove numerical noise i.e. negative variance
+vaar = max(vaar,1e-10); % remove numerical noise i.e. negative variance
 return;
 

### 2014-03-11 src/init_params.m

@@ -1,4 +1,4 @@
-function params = init_params(x,y,M,nhyper,z0)
+function params = init_params(params,x,y,M,nhyper,z0)
 %INIT_PARAMS params = init_params(x,y,M)
 %   Init parameters of a single augmented sparse gp.
 %
@@ -16,8 +16,10 @@ params.m = m;
 params.S = S;
 params.z = z0;
 params.z0 = z0;
-loghyp = log(ones(nhyper,1));
-params.loghyp = loghyp;
+if ~isfield(params,'loghyp') || isempty(params.loghyp)
+  loghyp = log(ones(nhyper,1));
+  params.loghyp = loghyp;
+end
 params.delta_hyp = zeros(size(params.loghyp));
 params.delta_z = zeros(size(z0));
 end

### 2014-03-11 src/scripts/slfm_sarcos.m

@@ -1,89 +1,87 @@
-clear all; clc; %close all;
-rng(1110,'twister');
+clear all; %clc; %close all;
 
 % pre-process y
 load sarcos_inv
 load sarcos_inv_test
 fmae = @(ytrue,ypred) mean(abs(ypred-ytrue));
 frmse = @(ytrue,ypred) sqrt(mean((ypred-ytrue).^2));
-
 x = sarcos_inv(:,1:21);
 y = sarcos_inv(:,22:end);
+
 xtest = sarcos_inv_test(:,1:21);
 ytest = sarcos_inv_test(:,22:end);
-outputs = [2,3,4,7];
+outputs = [4,7];
 y = y(:,outputs);
 ytest = ytest(:,outputs);
 [x,xmean,xstd] = standardize(x,[],[]);
 xtest = standardize(xtest,xmean,xstd);
 [y,ymean,ystd] = standardize(y,[],[]);
+clear sarcos_inv
+clear sarcos_inv_test
 
 %% ssvi
 cf.covfunc_g  = 'covSEard';
+cf.covfunc_h = 'covSEard';
 cf.lrate      = 1e-2;
 cf.momentum   = 0.9;
 cf.lrate_hyp  = 1e-5;
 cf.lrate_beta = 1e-4;
 cf.momentum_w = 0.9;
-cf.lrate_w    = 1e-5;
-cf.learn_z    = false;
+cf.lrate_w    = 1e-6;
+cf.learn_z    = true;
 cf.momentum_z = 0.0;
 cf.lrate_z    = 1e-4;
-cf.maxiter = 100;
-cf.nbatch = 500;
-cf.beta = 1/1;
-cf.initz = 'kmeans';
+cf.maxiter = 1;
+cf.nbatch = 2000;
+cf.beta = 1/0.01;
+cf.initz = 'random';
+cf.monitor_elbo = 50;
+cf.w = ones(size(y,2),1);
+%cf.w = [1 0.01; 0.01 1];
+cf.fix_first = true;
 
 Q = 1;
-runs = 1;
-Ms = [100,200,500,700,1000];
-Ms = 500;
-for i=1:numel(Ms)
-  M = Ms(i);
-  disp(['M = ' num2str(M)])
-  for j=1:runs
-    par.g = cell(Q,1);
-    [elbo,par] = slfm2_learn(x,y,M,par,cf);
-    [mu,var,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest,size(y,2));
[…]

### 2014-03-11 src/scripts/ssvi_jura.m

@@ -1,64 +1,0 @@
-clear all; clc; %close all;
-rng(1110,'twister');
-
-[x,y,xtest,ytest] = read_juraCd();
-% pre-process y
-y0 = y;
-y = log(y);
-[y,ymean,ystd] = standardize(y,[],[]);
-
-%M = size(x,1);
-M.g = 20; M.h = 200;
-% ssvi
-cf.covfunc_g  = 'covSEard';
-cf.covfunc_h  = 'covSEard';
-cf.lrate      = 1e-2;
-cf.momentum   = 0.9;
-cf.lrate_hyp  = 1e-5;
-cf.lrate_beta = 1e-4;
-cf.momentum_w = 0.9;
-cf.lrate_w    = 1e-5;
-cf.learn_z    = true;
-cf.momentum_z = 0.0;
-cf.lrate_z    = 1e-4;
-cf.init_kmeans = false;
-cf.maxiter = 200;
-cf.nbatch = 10;
-
-[elbo,params] = ssvi_learn(x,y,M,cf,[]);
-[mu,var,mu_g,var_g] = ssvi_predict(cf.covfunc_g,cf.covfunc_h,params,xtest);
-mu = exp(mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1));
-disp('mae = ')
-disp(mean(abs(mu-ytest)))
-disp('smse = ')
-for i=1:size(ytest,2)
-  fprintf('%.4f\t',mysmse(ytest(:,i),mu(:,i),ymean(i)));
-end
-
-figure;
-semilogy(1:numel(elbo),-elbo);
-%plot(1:numel(elbo),elbo);
-ylabel('elbo')
-xlabel('iteration')
-title(['elbo vs. iteration, lrate = ' num2str(cf.lrate_z)])
-
-disp('elbo = ')
-disp(elbo(end))
-disp('learned w = ')
-disp(params.w)
-
-figure; hold on;
-scatter(x(:,1),x(:,2))
-scatter(params.g.z(:,1),params.g.z(:,2),'xr')
-scatter(params.g.z0(:,1),params.g.z0(:,2),'sm')
-title('inducing of g')
-
-% for i=1:size(ytest,2)
-% figure; hold on;
-% scatter(x(:,1),x(:,2))
-% scatter(params.task{i}.z(:,1),params.task{i}.z(:,2),'xr')
-% title(['inducing of task ' num2str(i)]);
-% end
-
-% standard full gp or standard gpsvi?
-

### 2014-03-11 src/slfm_learn.m

@@ -13,7 +13,7 @@ function [elbo,par] = slfm_learn(x,y,M,par,cf)
 [N D] = size(x);
 % get observation indice for each output
 P = numel(par.task); Q = numel(par.g);
-par.w = ones(P,Q);
+par.w = cf.w;
 par.delta_w = zeros(P,Q);
 par.beta = cf.beta*ones(P,1);
 par.delta_beta = zeros(P,1);
@@ -26,21 +26,16 @@ for i=1:P
 end
 ymean = mean(ytmp,2);
 for j=1:Q
-  par.g{j} = init_params(x,ymean,M.g,nhyper_g,initz(x,M.g,cf.initz));
+  par.g{j} = init_params(par.g{j},x,ymean,M.g,nhyper_g,initz(x,M.g,cf.initz));
   par.g{j}.m = zeros(size(par.g{j}.m));
 end
 nhyper_h = eval(feval(cf.covfunc_h));
 for i=1:P
-  par.task{i} = init_params(x,y(:,i),M.h,nhyper_h,initz(x,M.h,cf.initz));
-  %par.task{i}.m = zeros(size(par.task{i}.m));
-  if strcmp(cf.covfunc_h,'covNoise')
-    par.task{i}.loghyp = log(0.05); % learning of noise is hard so set it small here
-  end
-  if strcmp(cf.covfunc_h,'covSEard')
-    par.task{i}.loghyp(end) = log(0.1); % h_i should be small
-  end
+  par.task{i} = init_params(par.task{i},x,y(:,i),M.h,nhyper_h,initz(x,M.h,cf.initz));
+  par.task{i}.m = zeros(size(par.task{i}.m));
 end
-elbo = zeros(numel(cf.maxiter),1);
+
+elbo = zeros(cf.maxiter,1);
 for iter = 1:cf.maxiter
   idx = randperm(N, cf.nbatch);
   par.idx = ~isnan(y(idx,:));
@@ -49,65 +44,84 @@ for iter = 1:cf.maxiter
   % variational parameters of g_j
   par = slfm_update_g(xbatch,ybatch,par,cf);
   
-  % hyperparameters of g_j
-  [~,dbeta,dw,dloghyp] = slfm_elbo(xbatch,ybatch,par,cf);
-  for j=1:Q
-    [par.g{j}.loghyp,par.g{j}.delta_hyp] = stochastic_update(...
[…]

### 2014-03-11 src/svi_learn.m

@@ -14,9 +14,9 @@ function [mu,s2,elbo,par] = svi_learn(x,y,xtest,M,cf,z0)
 [N D] = size(x);
 nhyper = eval(feval(cf.covfunc));
 if isempty(z0)
-  par = init_params(x,y,M,nhyper,initz(x,M,cf.initz));
+  par = init_params([],x,y,M,nhyper,initz(x,M,cf.initz));
 else
-  par = init_params(x,y,M,nhyper,z0);
+  par = init_params([],x,y,M,nhyper,z0);
 end
 par.beta = cf.beta;
 par.delta_beta = 0;

### 2014-03-11 src/svi_update.m

@@ -38,7 +38,9 @@ theta1 = theta1_old + cf.lrate*(betaval*A'*y - theta1_old);
 [VV DD] = eig(theta2);
 invTheta2 = VV*diag(1./diag(DD))*VV';
 params.S = - 0.5*invTheta2;
-params.m = params.S*theta1;
-
+% a bit of hack to allow h_i to be input-dependent noise
+if ~strcmp(covfunc,'covNoise')
+  params.m = params.S*theta1;
+end
 return;
 

### 2014-03-12 paper/experiments.tex
description of the air temperature prediction

@@
 %todo: result for independent gps
-We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to that of CGP variants and independent GPs. 
-Our model outperforms both variants of the convolved GPs in terms of standardized mean-squared error (SMSE).
-It has lower test likelihood, as measured by the negative log predictive density (NLPD), which is mainly due to the less conservative predictive variance of the full CGP for the CAD currency.
+We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citet{alvarez-lawrence-nips-08} (CGP) and with approximation via the variational inducing kernels \citet{alvarez2010efficient}) (CGPVAR) in addition to using independent GPs (IGP, one for each output).
+Our model outperforms both of the CGP variants in terms of standardized mean-squared error (SMSE).
+It has lower test likelihood, as measured by the negative log predictive density (NLPD), which is mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
+The results are averaged across the 3 outputs over 5 repetitions.
+% no std because there are 3 outputs but variance is small
 Note that the SMSE of CGPVAR is taken from \citet{alvarez2010efficient} while the NLPD was not provided.
@@
 COGP  & \textbf{0.2125} & -0.8394 \\
-CGPVAR & 0.2795 & NA \\
 CGP & 0.2427 & \textbf{-2.9474} \\
+IGP & 0.5996 & 0.4082 \\
+CGPVAR & 0.2795 & NA 
 %ICM & 0.3927 &
@@
 \subsection{AIR TEMPERATURE PREDICTION}
-Two latent functions, inverse lengthscale 136 capturing global trend of increasing temperature from day 10 - 15 and inverse lengthscale of 0.5 to model the fluctuation within a day.
-Training for full cgp model with 2 latent functions took 3 hours compared to ... in our model.
+Next, we consider the task of predicting air temperature at 4 different locations in the south coast of England. 
+The data is gathered from a network of weather sensors (named Bramblemet, Sotonmet, Cambermet, and Chimet), each of which measures several environmental variables \citet{osborne2008towards}\footnote{Data is available at seperate web pages, see e.g.
+http://www.bramblemet.co.uk}.
+%The sensors are close geographically so we can expect correlation in air temperature.
+We selected the sensor signal for air temperature during the period from July 10 to July 15, 2013.
+The sensors record measurements every 5 minutes, resulting in a maximum of 4320 observations.
+However, there are missing data for Bramblemet (100 points), Chimet (15 points), and Sotonmet (1002 points), possibly due to network outages or hardware failures.
+We further simulated failure of the sensors by removing the observations from the time periods [10.2 - 10.8] for Cambermet and [13.5 - 14.2] for Chimet.
+The removed data comprises 375 data points, which is used for testing, and the remaining data consisting of 15,788 points is used for training.
+Similar to the previous experiment, the objective is to evaluate the ability of the model to use the signals from the functioning sensors to extrapolate the missing signals.
+
+For pre-processing we normalized the outputs to have zero mean and unit variance.
+The inducing inputs $(M = 200)$ are randomly selected from the training set and fixed throughout training.
+The number of latent processes is $Q = 2$ for both our model and the CGP model.
+
+The real data and the predictive distributions by 3 models (COGP, CGP with exact inference, and independent GPs) are shown in Figure \ref{fig:weather}.
+It is clear that the independent GPs model is clueless in the test regions and thus simply uses the average temperature as its prediction.
+For Cambermet, both COGP and CGP can capture the rising in temperature from the morning till the afternoon and the fall afterward.
+For Chimet, both models perform poorly but CGP more so as it falsely predicts  wild fluctuations that are non-existent in the data. 
+Note that the two latent processes learned by our model indeed correspond to different patterns in the data: one process has the inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has the inverse lengthscale of 0.5 to model the local variations within a single day.
+The comparative performance of the models are summarized in Table \ref{tab:air}, which further supports the qualitative analysis that our model outperforms CGP.
+All results are averaged of the 2 outputs over 5 repetitions.
+Training of our model took on average 5 minutes compared to 3 hours of CGP with exact inference.
 
 \begin{table}[h]
-\caption{Performance comparison by SMSE and NLPD on the air temperature dataset.}
-\label{tab:fx}
+\caption{Performance comparison on the air temperature dataset.}
+\label{tab:air}
 \begin{center}
 \begin{tabular}{c|c|c}
-method & smse & nlpd \\ \hline
+METHOD & SMSE & NLPD \\ \hline
 COGP & \textbf{0.1077} & \textbf{2.1712} \\

### 2014-03-12 paper/experiments.tex
move appendix to supplemental and re-format

@@
 The sensors record measurements every 5 minutes, resulting in a maximum of 4320 observations.
-However, there are missing data for Bramblemet (100 points), Chimet (15 points), and Sotonmet (1002 points), possibly due to network outages or hardware failures.
+There are missing data for Bramblemet (100 points), Chimet (15 points), and Sotonmet (1002 points), possibly due to network outages or hardware failures.
 We further simulated failure of the sensors by removing the observations from the time periods [10.2 - 10.8] for Cambermet and [13.5 - 14.2] for Chimet.
@@
 
-\subsection{ROBOT ARMS}
+\subsection{INVERSE DYNAMICS OF ROBOT ARM}
+Our last experiment is with a large scale dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm\footnote{Data is available at http://www.gaussianprocess.org/gpml/data/}.
+The data consists of 48,933 mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
+The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citet{vijayakumar2000locally}.
+Furthermore, exploratory analysis using standard GP with automatic relevance determination suggests that all of the 21 dimensions are relevant.
+We consider here the task of joint learning for the $4$th and $7$th outputs.
+Note that previous work (see e.g. \citet{rasmussen-williams-book},  \citet{vijayakumar2000locally}) on this dataset has only dealt with learning  a single output.
+We used the same training/testing split as in \citet{rasmussen-williams-book}, resulting in a training set of 88,968 observations and a test set of 
+8,898 observations.

### 2014-03-12 paper/inference.tex
move appendix to supplemental and re-format

@@
 The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
-Each individual term $l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ is computed using the identity in eq.
-\ref{eq:identity} and  given in eq \ref{eq:lin} (both are in the Appendix).
 \newcommand{\Ahi}{\A^h_i}
 \newcommand{\Zi}{\Z_i}
- Substituting $l_{in}$ into equation \ref{eq:elbo} and carrying out the integral using the identity in eq. \ref{eq:identity} again we get:
+The ELBO can be computed by first solving for the individual expectations 
+$l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ (see the supplementary material for detailed derivation), which are then substituted into equation \ref{eq:elbo} to give:
 \begin{align}

### 2014-03-12 paper/supplemental.tex
move appendix to supplemental and re-format

--- 2014-03-06 paper/appendix.tex
@@
-\subsection{Expectation of an individual likelihood term}
-The expectation of an individual likelihood term with respect to the posterior distribution is given by:
+% Sample LaTeX file for creating a paper in the Morgan Kaufmannn two
+% column, 8 1/2 by 11 inch proceedings format.
+
+\documentclass[]{article}
+\usepackage{proceed2e}
+\usepackage{hyperref}
+\usepackage{url}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{bm}
+\usepackage{hyperref}
+\usepackage{natbib}
+\usepackage{latexsym,amsbsy,amssymb,color,xspace, booktabs}
+
+\include{macros}
+\include{local_macros}
+
+\title{Collaborative Multioutput Gaussian Processes: \\ Supplementary Material}
+
+\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
+          % UAI  reviewing is double-blind.
+
+% The author names and affiliations should appear only in the accepted paper.
+%
+%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
+%alternate address.}} \\
+%Computer Science Dept. \\
+%Cranberry University\\
+%Pittsburgh, PA 15213 \\
+%\And
+%{\bf Coauthor}  \\
+%Affiliation          \\
+%Address \\
+%\And
+%{\bf Coauthor}   \\
+%Affiliation \\
+%Address    \\
+%(if needed)\\
+%}
+
+\begin{document}
+
+\maketitle
+
+\section{Gaussian Identities}
+Let $\y$, $\g = \{\g_j\}_{j=1}^q$, and $\h$ be random variables with multivariate Gaussian distributions: 
+$p(\y | \g, \h) = \Normal(\y; \sum_{j=1}^Q \W_j \g_j + \W \h, \beta^{-1} \I)$, $p(\g_j) = \Normal(\g_j; \m_j, \S_j)$, and $p(\h) = \Normal(\h; \m, \S)$.
+The following identity is important in deriving the evidence lower bound:
+\begin{align}
+\nonumber
+\int& \log p(\y | \g, \h) \prod_{j=1}^q p(\g_j) p(\h) \der \g \der \h \\
+\nonumber
+=& \log \Normal(\y; \sum_{j=1}^q \W_j \m_j + \W \m, \beta^{-1} \I) \\
+& - \frac{1}{2} \beta \trace \W^T \W \S - \frac{1}{2} \beta \trace \sum_{j=1}^q \W_j^T \W_j \S_j.
+ \label{eq:identity}
+\end{align}
+The identity can be proved by using this fact: 
+\begin{align}
+\nonumber
+&\int (\W\x - \Mu)^T \BigSigma^{-1} (\W \x - \Mu) \Normal(\x; \m, \S) \der \x \\
+&= (\Mu - \W\m)^T \BigSigma^{-1} (\Mu - \W\m) + \trace \W^T \BigSigma^{-1} \W \S.
+\end{align}
+
+\section{Expectation of an Individual Likelihood Term}
+Applying the identity in eq. \ref{eq:identity}, the expectation of an individual likelihood term with respect to the posterior distribution is given by:
 \begin{align}
@@
 \end{align}
-where $\tilde{k}_{nn} = (\tilde{\K})_{nn}, \tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}, \mu_{jn} = (\Mu_j)_n, \text{ and } \mu^h_{in} = (\Mu^h_i)_n$.
+where $\tilde{k}_{nn} = (\tilde{\K})_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$, and $\mu^h_{in} = (\Mu^h_i)_n$.
 
@@
 
-\subsection{Appendix 1: Derivatives}
-Here we derive the gradients of the lower bound wrt the  hyperparameters.
-As a template, consider the lower bound in Hensman et al (as a function of the hyperparameters):
+\section{Derivatives of the ELBO}
+For convenience, we derive the gradients of the lower bound for the case of a single GP (i.e. the bound in \citet{hensmangaussian}).
+The derivatives of the ELBO of the collaborative multioutput GPs model are  typically linear combination of the derivatives here.
+The lower bound as a function of all parameters is
 \begin{align}
@@
 \calL
-=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I)
- - \frac{1}{2} \beta \trace \tilde{\K}
+=& \log \Normal(\y; \K_{NM} \K_{MM}^{-1} \m, \beta^{-1}\I) \\
+\nonumber
+ &- \frac{1}{2} \beta \trace \tilde{\K}
  - \frac{1}{2} \beta \trace (\S\K_{MM}^{-1} \K_{MN} \K_{NM} \K_{MM}^{-1}) \\  \nonumber
@@
 =& \underbrace{\log \Normal(\y; \A \m, \beta^{-1}\I)}_{\calL_1}
- - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2}
- - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3} \\  
-&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4},
+ - \underbrace{\frac{1}{2} \beta \trace (\K_{NN} - \A\K_{MN})}_{\calL_2} \\
+ \nonumber
+&- \underbrace{\frac{1}{2} \left( \log |\K_{MM}| + \trace(\K_{MM}^{-1}(\m \m^T + \S)) \right)}_{\calL_4}\\
+ & - \underbrace{\frac{1}{2} \beta \trace (\S\A^T\A)}_{\calL_3},
 \end{align}
 where $\A = \K_{NM} \K_{MM}^{-1}$.
-Notice that we have re-written the sum of individual terms in matrix form which will make the derivation and also computation easier.
+Here $\K_{NM} = k(\X,\Z)$ is the cross-covariance matrix between the observed inputs and the inducing inputs and $\K_{MM} = k(\Z,\Z)$ is the auto-covariance matrix between the inducing inputs.
+Notice that we have re-written the sum of individual terms in matrix form which will make the derivation easier and also the computation faster via vectorization.
 
@@
 \begin{align}
-\deriv{\calL_1}{t} &= \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
-\deriv{\calL_2}{t} &= \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
-\deriv{\calL_3}{t} &= \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
-\deriv{\calL_4}{t} &= \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) - \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
+\deriv{\calL_1}{t} =& \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
+\deriv{\calL_2}{t} =& \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
+\deriv{\calL_3}{t} =& \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
+\nonumber
+\deriv{\calL_4}{t} =& \frac{1}{2}  \trace \left(\K_{MM}^{-1} \deriv{ \K_{MM}}{t}\right) \\
+&- \frac{1}{2} \trace \left(\K_{MM}^{-1} \deriv{\K_{MM}}{t} \K_{MM}^{-1} (\m \m^T + \S) \right) 
 \end{align}
@@
 \subsection{Derivatives of the Inducing Inputs}
-To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\z_m\}_{m=1}^M$ are also parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
+To compute the derivatives of $\calL$ wrt the inducing inputs, first notice that $\Z = \{\z_m\}_{m=1}^M$ can also be viewed as parameters of the covariance matrices $\K_{NM}$ and $\K_{MM}$.
 Hence the derivative wrt a single dimension of an inducing input, i.e. $z_{mj}$, is the same as that of $\deriv{ \calL}{t}$.
@@
 \noindent We re-write $\deriv{\calL_1}{t}, \deriv{\calL_2}{t}, \deriv{ \calL_3}{t}, \deriv{\calL_4}{t}$ by expanding $\deriv{\A}{t}$ (here $t = z_{mj}$):
+% dL1
 \begin{align}
 \nonumber
-% dL1
 \deriv{\calL_1}{t}
- &= \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
-&= \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} 
--\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t} \\
+=& \beta \trace (\y - \A\m)^T \left(\deriv{\K_{NM}}{t} -  \A \deriv{ \K_{MM}}{t}\right)\K_{MM}^{-1} \m \\
+\nonumber
+=& \beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \deriv{\K_{NM}}{t} \\
+&-\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A \deriv{\K_{MM}}{t}
+\end{align}
 %dL2
+\begin{align}
 \deriv{\calL_2}{t}
 &= - \beta \trace \A^T \deriv{\K_{NM}}{t}
- + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}  \\
+ + \frac{1}{2} \beta \trace \A^T \A \deriv{\K_{MM}}{t}
+ \end{align}
 % dL3
+ \begin{align}
 \deriv{\calL_3}{t}
 &= \beta \trace \K_{MM}^{-1} \S \A^T \deriv{\K_{NM}}{t}
- - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}\\
+ - \beta \trace \K_{MM}^{-1} \S \A^T \A \deriv{\K_{MM}}{t}
+ \end{align}
  % dL4
+ \begin{align}
 \deriv{\calL_4}{t}
- &= \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t}
-  - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
+ =& \frac{1}{2}  \trace \K_{MM}^{-1} \deriv{\K_{MM}}{t} \\
+ \nonumber
+& - \frac{1}{2} \trace \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1} \deriv{\K_{MM}}{t} 
 \end{align}
-From the above equations we get,
+From the above 4 equations we get,
 \begin{align}
@@
 \D_2 =& -\beta \trace \K_{MM}^{-1} \m (\y - \A\m)^T \A
- - \frac{1}{2} \beta \A^T \A
-  + \beta \K_{MM}^{-1} \S \A^T \A	 \\ 
-  &-\frac{1}{2} \K_{MM}^{-1} + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
+ - \frac{1}{2} \beta \A^T \A -\frac{1}{2} \K_{MM}^{-1} \\
+  &+ \beta \K_{MM}^{-1} \S \A^T \A	 
+   + \frac{1}{2} \K_{MM}^{-1} (\m \m^T + \S) \K_{MM}^{-1}
 \end{align}
+%TODO: give the correct cost
 Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
@@
 
-\subsection{Appendix 2: Gaussian Identity}
-Let $\y$, $\g = \{\g_j\}_{j=1}^q$, and $\h$ be random variables with multivariate Gaussian distributions: 
-$p(\y | \g, \h) = \Normal(\y; \sum_{j=1}^Q \W_j \g_j + \W \h, \beta^{-1} \I)$, $p(\g_j) = \Normal(\g_j; \m_j, \S_j)$, and $p(\h) = \Normal(\h; \m, \S)$.
-The following identity is important in deriving the evidence lower bound:
-\begin{align}
-\nonumber
-\int \log p(\y | \g, \h) \prod_{j=1}^q p(\g_j) p(\h) \der \g \der \h
-=& \log \Normal(\y; \sum_{j=1}^q \W_j \m_j + \W \m, \beta^{-1} \I) - \frac{1}{2} \beta \trace \W^T \W \S \\
-\label{eq:identity}
-&- \frac{1}{2} \beta \trace \sum_{j=1}^q \W_j^T \W_j \S_j.
-\end{align}
-The identity can be proved by using this fact: 
-\begin{align}
-\nonumber
-\int (\W\x - \Mu)^T \BigSigma^{-1} &(\W \x - \Mu) \Normal(\x; \m, \S) \der \x \\
-\nonumber
-&= (\Mu - \W\m)^T \BigSigma^{-1} (\Mu - \W\m) + \trace \W^T \BigSigma^{-1} \W \S.
-\end{align}
+\bibliographystyle{plainnat}
+\bibliography{references}
+
+\end{document}

### 2014-03-15 paper/experiments.tex
latest experiment results and paper style

@@
 It has lower test likelihood, as measured by the negative log predictive density (NLPD), which is mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
-The results are averaged across the 3 outputs over 5 repetitions.
+%The results are averaged across the 3 outputs over 5 repetitions.
 % no std because there are 3 outputs but variance is small
@@
 
-\begin{table}[h]
-\caption{Performance comparison on the foreign exchange rate dataset.}
+\setlength{\tabcolsep}{4pt}
+\begin{table}[t]
+\caption{Performance comparison on the foreign exchange rate dataset. Results are averaged of the 3 outputs over 5 repetitions.}
 \label{tab:fx}
 \begin{center}
-\begin{tabular}{c|c|c}
-METHOD & SMSE & NLPD \\ \hline
+\begin{tabular}{ccc}
+\toprule
+\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\ \hline
 COGP  & \textbf{0.2125} & -0.8394 \\
@@
 IGP & 0.5996 & 0.4082 \\
-CGPVAR & 0.2795 & NA 
+CGPVAR & 0.2795 & NA \\
 %ICM & 0.3927 &
+\bottomrule
 \end{tabular}
@@
 The comparative performance of the models are summarized in Table \ref{tab:air}, which further supports the qualitative analysis that our model outperforms CGP.
-All results are averaged of the 2 outputs over 5 repetitions.
+%All results are averaged of the 2 outputs over 5 repetitions.
 Training of our model took on average 5 minutes compared to 3 hours of CGP with exact inference.
 
-\begin{table}[h]
-\caption{Performance comparison on the air temperature dataset.}
+\setlength{\tabcolsep}{4pt}
+\begin{table}[t]
+\caption{Performance comparison on the air temperature dataset. Results are averaged of 2 outputs over 5 repetitions. }
 \label{tab:air}
 \begin{center}
-\begin{tabular}{c|c|c}
-METHOD & SMSE & NLPD \\ \hline
+\begin{tabular}{ccc}
+\toprule
+\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\
+\hline
 COGP & \textbf{0.1077} & \textbf{2.1712} \\
 CGP & 0.1125 & 2.2219 \\
-IGP & 0.8944 & 12.5319
+IGP & 0.8944 & 12.5319 \\
+\bottomrule
 \end{tabular}
@@
 \begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/slfm-weatherCambermet.eps} &
+\includegraphics[scale=0.3]{figures/cogp-weatherCambermet.eps} &
 \includegraphics[scale=0.3]{figures/cgp-weatherCambermet.eps} &
@@
 \\
-\includegraphics[scale=0.3]{figures/slfm-weatherChimet.eps} &
+\includegraphics[scale=0.3]{figures/cogp-weatherChimet.eps} &
 \includegraphics[scale=0.3]{figures/cgp-weatherChimet.eps} &
@@
 Note that previous work (see e.g. \citet{rasmussen-williams-book},  \citet{vijayakumar2000locally}) on this dataset has only dealt with learning  a single output.
-We used the same training/testing split as in \citet{rasmussen-williams-book}, resulting in a training set of 88,968 observations and a test set of 
-8,898 observations.
+We used the same training/testing split as in \citet{rasmussen-williams-book}, resulting in a training set of 88,968 observations and a test set of 8,898 observations.
+
+Since none of the existing multioutput models are applicable to this problem, we compare with two baselines which learn each output independently.
+The first is the subset of data approach where a subset of 1000 data points (for each output) is randomly sampled for training. 
+We call this method SOD1000.
+The second baseline is the sparse GP with stochastic variational inference \citet{hensmangaussian} using 500 inducing inputs, which we refer to as SVIGP.
+The outputs are normalized and the squared exponential with ARD covariance function is used for all methods.
+In case of COGP, this mean the shared process (we used $Q = 1$) and each of the independent processes have this same covariance function.
+
+The predictive performance of the methods are given in Table \ref{tab:robotarm}.
+The results are averaged of the 2 outputs over 5 repetitions.
+Both COGP and SVIGP using 500 inducing inputs significantly outperform SOD in terms of the mean absolute error (MAE) but have lower predictive likelihood.
+
+\begin{table}[t]
+\caption{Performance comparison on the robot arm dataset.
+In the two bottom lines, standard GP is applied to output 1 and the remaining method is applied to output 2.
+Results are averaged over 5 repetitions.}
+\label{tab:robotarm}
+\begin{center}
+\begin{tabular}{lcccc}
+\toprule
+& \multicolumn{2}{c}{\textbf{OUTPUT 1}} & \multicolumn{2}{c}{\textbf{OUTPUT 2}} \\ \cmidrule(r){2-5}
+\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} & \textbf{SMSE} & \textbf{NLPD}\\ 
+ \midrule
+COGP, learn z & \textbf{0.2631} & \textbf{3.0600} & 0.0127 & \textbf{0.8302} \\
+COGP, fix z & 0.2821& 3.2281 & 0.0131 & 0.8685 \\
+GP, SVIGP & 0.3119 & 3.2198 & \textbf{0.0101} & 1.1914 \\
+GP, SOD & 0.3119 & 3.2198 & 0.0104 & 1.9407 \\
+\bottomrule
+\end{tabular}
+\end{center}
+\end{table}
+
+\subsubsection{Impact of Optimizing the Inducing Inputs}
+% table for smses, nlpds, training tiem
+\begin{figure}
+\includegraphics[scale=0.4]{figures/sarcosTime.eps}
+\caption{The cost of optimizing versus fixing the inducing inputs.}
+\label{fig:time}
+\end{figure}

### 2014-03-15 paper/paper.tex
updated paper
clean figures
updated figures and description of experiment on the fx dataset
move appendix to supplemental and re-format
latest experiment results and paper style

@@
 \usepackage{natbib}
-\usepackage{latexsym,amsbsy,amssymb,color,xspace, booktabs}
+\usepackage{multirow}
+\usepackage{fancyhdr}
+\usepackage{latexsym,amsbsy,amssymb,color,xspace,booktabs}
 
@@
 
-\title{Stochastic Multiple Output Gaussian Processes}
+\title{Collaborative Multioutput Gaussian Processes}
 
@@
 
-\subsubsection{Citations in Text}
-
-Citations within the text should include the author's last name and
-year, e.g., (Cheesman, 1985). Reference style should follow the style
-that you are used to using, as long as the citation style is
-consistent.
-
 \section{MODEL DESCRIPTION}
@@
 \input{discussion}
-\newpage
-\section{APPENDIX}
-Perhaps this should go to a supplementary material.
-\input{appendix}
 
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{plainnat}
 \bibliography{references}

### 2014-03-15 paper/experiments.tex
experiments first complete draft

@@
-We evaluate the proposed approach using a set of four experiments.
-A toy problem is first used to study the transfer of learning between two related processes through the shared inducing points.
-We then compare the model with existing multioutput models on the tasks of predicting foreign exchange rates and air temperature.
-In the final experiment, we test the model on the prediction of robot arms dynamic on a large-scale dataset. 
-All experiments are executed on an Intel(R) Core(TM) i7-2600 3.40GHz CPU with 8GB of RAM using Matlab R2012a.
-%except for large scale?
+We evaluate the proposed approach with four experiments.
+A toy problem is first used to study the transfer of learning between two related processes via the shared inducing points.
+We then compare the model with existing multioutput models on the tasks of predicting foreign exchange rate and air temperature.
+In the final experiment, we show that joint learning under sparsity can lead to significant performance gain on a large scale dataset of inverse dynamics of a robot arm.
+
+Since we are using stochastic optimization, the learn rates need to be chosen carefully.
+We found that the rates used in \citet{hensmangaussian} also work well for our model.
+Specifically, we used the learn rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise hyperparameter, and inducing inputs.
+We also included a momentum term of $0.9$ for all of the parameters except the variational parameters and the inducing inputs.
+All of the experiments are executed on an Intel(R) Core(TM) i7-2600 3.40GHz CPU with 8GB of RAM using Matlab R2012a.
 
 \subsection{TOY PROBLEM}
-In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
+In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ and corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
 The first output has missing observations in the $(-7,-3)$ interval and the second output in the $(4,8)$ interval.
-The predictive distributions by our model ($Q = 1$) and independent stochastic GPs (one for each output) are shown in \ref{fig:toy}.
+Figure \ref{fig:toy} shows the predictive distributions by our model ($Q = 1$) and independent GPs with stochastic variational inference (GPSVI, one for each output).
 It can be seen that our multioutput model gives perfect prediction in the unobserved regions where the independent models fail to interpolate despite having the same set of inducing inputs.
-This confirms that transfer of information from one output to another via collaborative learning of the posterior over the shared inducing inputs.
-Furthermore, the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which precisely reflects the correlation between the two outputs.
+This confirms the effect of transfer of information from one output to another via collaborative learning of the inducing points.
+Furthermore, the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
 
@@
 \subsection{FOREIGN EXCHANGE RATE PREDICTION}
-The application considered here is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca/data.html}.
+The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca/data.html}.
 The setting of our experiment described here is identical to that in \citet{alvarez2010efficient}.
-The dataset contains all the data available for the 251 working days in the year of 2007.
+The dataset consists of all the data available for the 251 working days in the year of 2007.
 There are 9, 8, and 42 days of missing values  for gold, silver, and platinum, respectively.
 We remove from the data the exchange rate of CAD on days 50-100, JPY on day 100-150, and AUD on day 150-200.
-Note that these 3 currencies are from very different geographical locations. 
-The 153 points extracted is used for testing and the remaining 3051 data points is used for training.
-Since the missing data is long contiguous sections, the objective is to evaluate the capacity of the model to impute the missing currency values based on other currencies.
+Note that these 3 currencies are from very different geographical locations, making the problem more interesting. 
+The 153 points extracted is used for testing, and the remaining 3051 data points is used for training.
+Since the missing data correspond to long contiguous sections, the objective here is to evaluate the capacity of the model to impute the missing currency values based on other currencies.
 %todo: batch size, learn rate (maybe at the beginning of experiments)
@@
 Since the exchange rates are driven by a small number of latent market forces \citet{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
-We used the squared-exponential covariance function for the shared processes and a noise covariance function for the independent process of each output.
-The inducing inputs $(M = 100)$ are randomly selected from the training set and fixed throughout training.
+We used the squared-exponential covariance function for the shared processes and the noise covariance function for the independent process of each output.
+The inducing inputs $(M = 100)$ are randomly selected from the training data and fixed throughout training.
 
-The predictive distributions by our model (Figure \ref{fig:fx}) exhibit similar behavior to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
+The real data and predictive distributions by our model are shown in Figure \ref{fig:fx}.
+They exhibit similar behaviors to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
 In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
-Our investigation of the dataset found that 4 other currencies (GBP, NZD, KRW, MXN) also experienced the same trend during the days 150 - 200.
+Our investigation of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 150 - 200.
 This was effectively used by the model to extrapolate the values of the AUD.
@@
 We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citet{alvarez-lawrence-nips-08} (CGP) and with approximation via the variational inducing kernels \citet{alvarez2010efficient}) (CGPVAR) in addition to using independent GPs (IGP, one for each output).
-Our model outperforms both of the CGP variants in terms of standardized mean-squared error (SMSE).
+Our model outperforms both of the CGP variants in terms of standardized mean squared error (SMSE).
 It has lower test likelihood, as measured by the negative log predictive density (NLPD), which is mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
@@
 Note that the SMSE of CGPVAR is taken from \citet{alvarez2010efficient} while the NLPD was not provided.
-Training took only 10 minutes for our model compared to 1.4 hours of the full CGP model.
+Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.
 
@@
 
-The real data and the predictive distributions by 3 models (COGP, CGP with exact inference, and independent GPs) are shown in Figure \ref{fig:weather}.
+The real data and the predictive distributions by 3 models (COGP, CGP with exact inference \citet{alvarez-lawrence-nips-08}, and independent GPs) are shown in Figure \ref{fig:weather}.
 It is clear that the independent GPs model is clueless in the test regions and thus simply uses the average temperature as its prediction.
@@
 Note that the two latent processes learned by our model indeed correspond to different patterns in the data: one process has the inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has the inverse lengthscale of 0.5 to model the local variations within a single day.
-The comparative performance of the models are summarized in Table \ref{tab:air}, which further supports the qualitative analysis that our model outperforms CGP.
+The comparative performance of the models are summarized in Table \ref{tab:air}, from which it can be seen that our model outperforms CGP in terms of both test accuracy and confidence.
 %All results are averaged of the 2 outputs over 5 repetitions.
-Training of our model took on average 5 minutes compared to 3 hours of CGP with exact inference.
+It took 5 minutes on average to train our model compared to 3 hours of CGP with exact inference.
 
@@
 \subsection{INVERSE DYNAMICS OF ROBOT ARM}
-Our last experiment is with a large scale dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm\footnote{Data is available at http://www.gaussianprocess.org/gpml/data/}.
+Our last experiment is with a dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm\footnote{Data is available at http://www.gaussianprocess.org/gpml/data/}.
 The data consists of 48,933 mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
-The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citet{vijayakumar2000locally}.
-Furthermore, exploratory analysis using standard GP with automatic relevance determination suggests that all of the 21 dimensions are relevant.
-We consider here the task of joint learning for the $4$th and $7$th outputs.
-Note that previous work (see e.g. \citet{rasmussen-williams-book},  \citet{vijayakumar2000locally}) on this dataset has only dealt with learning  a single output.
-We used the same training/testing split as in \citet{rasmussen-williams-book}, resulting in a training set of 88,968 observations and a test set of 8,898 observations.
+%The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citet{vijayakumar2000locally}.
+%Furthermore, exploratory analysis using standard GP with automatic relevance determination suggests that all of the 21 dimensions are relevant.
+It has been used in previous work (see e.g. \citet{rasmussen-williams-book},  \citet{vijayakumar2000locally}) but only for single task learning.
+Here we consider joint learning for the 4th and 7th torques where the first has only 2,000 points  while the second has the full data of 44,484 points for training.
+The test set consists of 8,898 observations equally divided between the two outputs.
+Note that the data was collected at 100Hz for 7.5 minutes in total, hence ideal for this experiment which evaluates the effectiveness of collaborative learning under the assumption of sparsity in the output spaces.
 
-Since none of the existing multioutput models are applicable to this problem, we compare with two baselines which learn each output independently.
-The first is the subset of data approach where a subset of 1000 data points (for each output) is randomly sampled for training. 
-We call this method SOD1000.
-The second baseline is the sparse GP with stochastic variational inference \citet{hensmangaussian} using 500 inducing inputs, which we refer to as SVIGP.
-The outputs are normalized and the squared exponential with ARD covariance function is used for all methods.
-In case of COGP, this mean the shared process (we used $Q = 1$) and each of the independent processes have this same covariance function.
+Since none of the existing multioutput models are applicable to problems of this scale, we compare with independent models that learn each output separately.
+Standard GP is applied to the first output as it has only 2,000 observations for training.
+For the second output, we used two baselines.
+The first is the subset of data (SOD) approach where 2,000 data points are randomly selected for training with a standard GP model.
+The second is the sparse GP with stochastic variational inference (SVIGP) using 500 inducing inputs and a batch size of 1,000.
+In case of COGP, we also used a batch size of 1,000 and 500 inducing points for the shared process ($Q = 1$) and each of the independent processes.
+The squared exponential with automatic relevance determination (SEard) covariance function is used for all processes of all methods.
+Finally, the outputs are normalized to have zero mean and unit variance.
 
-The predictive performance of the methods are given in Table \ref{tab:robotarm}.
-The results are averaged of the 2 outputs over 5 repetitions.
-Both COGP and SVIGP using 500 inducing inputs significantly outperform SOD in terms of the mean absolute error (MAE) but have lower predictive likelihood.
+The performance of all methods, in terms of SMSE and NLPD averaged over 5 repetitions, is given in Table \ref{tab:robotarm}.
+The benefits of learning the two outputs jointly are evident, as can be seen by the significantly higher test accuracy and confidence of COGP compared to the full GP for the first output (4th torque).
+While the SMSE of the second output is essentially the same by all methods, the predictive likelihood of our multioutput model is better than that of the independent SVIGP model which has the same amount of training data for this torque.
+The results thus confirm the impact of collaborative learning even when sparsity assumption is made in the model.
 
@@
 
-\subsubsection{Impact of Optimizing the Inducing Inputs}
+Also from Table \ref{tab:robotarm}, it seems that optimizing the inducing inputs can lead to better performance than fixing them.
+More importantly, this comes with slight overhead in computation, as demonstrated by the training time for the two options in Figure \ref{fig:time}.
+For instance, the total training time with 500 inducing inputs is 1.9 hours for learning versus 1.6 hours for keeping them fixed.
+Recall that this dataset is 21-dimensional so this confirms our analysis in the inference section that the cost of optimization of the inducing inputs does not scale linearly with the dimensionality of the problem.
+Thus, for datasets of high dimension like this one, learning of the inducing inputs is a practical option.
+
 % table for smses, nlpds, training tiem

### 2014-03-16 paper/inference.tex
latest experiment results and paper style
model description and inference

@@
-\subsection{Variational Inference in Sparse GPs Revisited}
-In this section we review inference in sparse GPs (as presented in Titsias).
-The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
-The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g.
+%\subsection{Variational Inference in Sparse GPs Revisited}
+%
+%In this section we review inference in sparse GPs (as presented in Titsias).
+%The posterior in an augmented model is $p(\f, \u | \y) = p(\f | \u, \y) p(\u | \y)$.
+%The key property of such augmented GPs  is the notion of \textit{sufficient statistics}: given the inducing points $\u$, the latent values $\f$ are independent with any other set of latent values (e.g.
 the test set).
-In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
-This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
-Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
+%In the optimal setting when $\u$ is the sufficient statistics of $\f$, it should hold that $p(\f | \u, \y) = p(\f | \u)$ as $\y$ is only the noisy version of $\f$.
+%This leads to choosing a variational approximation of the posterior which factorizes as $q(\f, \u | \y) = p(\f | \u) q(\u | \y)$.
+%Since the conditional $p(\f | \u)$ is known, variational inference becomes learning an optimal posterior $q(\u | \y)$ only. \\
+%
+%\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
+%\begin{align}
+%\nonumber
+%p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
+%\nonumber
+%&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
+%&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
+%\label{eq:sorprediction}
+%&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
+%\end{align}
+%where,
+%\begin{align}
+%\nonumber
+%\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
+%\nonumber
+%\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
+%\end{align}
+% Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.
 
-\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
-\begin{align}
-\nonumber
-p(\vfstar | \y, \X_*) &= \int p(\vfstar | \f, \u, \X_*) q(\f, \u | \y) \der \f \der \u \\
-\nonumber
-&= \int p(\vfstar | \u) q(\u| \y) p(\f | \u) \der \f \der \u \\ \nonumber
-&= \int p(\vfstar | \u) q(\u| \y) \der \u \\
-\label{eq:sorprediction}
-&= \Normal(\vfstar; \bs{\mu_*},\vec{s_*})
-\end{align}
-where,
-\begin{align}
-\nonumber
-\bs{\mu_*} &= \K_{*z} \K_{zz}^{-1}\m \\ 
-\nonumber
-\S_* &= \K_{**} - \K_{*z} \left(\K_{zz}^{-1} - \K_{zz}^{-1} \S \K_{zz}^{-1} \right) \K_{*z}^T.
-\end{align}
- Here $\K_{*z}$ is the covariance matrix between test and inducing inputs, $\K_{zz}$ is the covariance matrix of the inducing inputs.
+Inference in this model is done with variational inference.
+In section \ref{sec:variationalLowerBound}, we derive a lower bound of the marginal likelihood which has the novel property of factorizing over the data points.
+The next section takes advantage of this factorization to derive stochastic variational inference which allows the model to scale to massive datasets.
+Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.
 
-\subsection{Variational Inference}
+\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
 \newcommand{\ug}{\u_g}
@@
 \newcommand{\Shi}{\S^h_i}
-Our goal of inference is to find the posterior $p(\g, \h, \u, \v | \y)$. 
-Following the previous discussion, we assume a variational distribution which factorizes as:
+Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
+Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as,
 \begin{align}
-\nonumber
+p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
+\end{align}
+Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
+This motivates the use of a distribution where the true posteriors over $\g$ and $\h$ are replaced by their conditional distributions given the inducing variables:
+\begin{align}
 q(\g, \h, \u, \v | \y)
-\nonumber
- &= p(\g|\u) p(\h|\v) q(\u,\v)  \\
- &= p(\g|\u) p(\h|\v) \prod_{j=1}^Q q(\u_j) \prod_{i=1}^P  q(\v_i)
+= p(\g|\u) p(\h|\v) q(\u,\v),
 \end{align}
-Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given, we need only  find the optimum $q(\u_j) = \Normal(\u_j; \mgj, \Sgj)$ and $q(\v_i) = q(\v_i) = \Normal(\v_i; \mhi, \Shi)$.
+with
+\begin{align}
+q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
+\end{align}
+This technique has been used in \citep{titsiasvariational,hensmangaussian} to derive variational inference for the single output case.
+Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given (Equation \ref{eq:gu} and \ref{eq:hv}), inference requires only learning of $q(\u, \v)$ that minimize the divergence between the approximate and true posteriors, which becomes small when the sufficiency condition holds with greater certainty.
 
@@
 \end{align}
-where, with the help of auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  
-$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$, 
+where, with the help of auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  $\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$, 
 \begin{align}
@@
 
-\subsubsection{Variational Parameters Derivatives}
+\subsection{STOCHASTIC VARIATIONAL INFERENCE}
 \newcommand{\oi}{\vec{o}_i}
 % some notation
-Before diving into the details, we first define the indexing operator of a matrix: $\B(\vec{r},\vec{c})$ extracts the submatrix in rows $\vec{r}$ and columns $\vec{c}$ of $\B$.
-To index all columns we use $\B(\vec{r},:)$ and similarly for all rows $\B(:,\vec{c})$.
-Readers familiar with this operator will recognize that this is the MATLAB indexing operator.
 
@@
 
+\subsection{INDUCING INPUTS AND HYPER-PARAMETERS \label{sec:hyperparameters}}
 %\subsubsection{Update Equations}
 % and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
-\subsection{Prediction}
+\subsection{PREDICTION}
 The predictive distribution of the $i$-th output for a test input $\x_*$ is 

### 2014-03-16 paper/intro.tex
model description and inference

% paper on multioutput so this should appear first in the introduction
%Gaussian processes (GPs) \citep{rasmussen-williams-book} are perhaps the prior of choice in nonparametric Bayesian regression not only in the standard setting but also in multioutput learning \citep{} .
% it is beauty but also a beast
%Unfortunately, given a dataset of size $N$, exact inference for GP regression requires $\calO(N^3)$ for computation and $\calO(N^2)$ for memory, which can only be applied to problems in non-realistic settings.
% cite titsias
%Consequently, various approximate methods \citep{lawrence2002fast,seeger2003bayesian,seeger2003fast, smola2001sparse,snelson2006sparse,williams2001using} have been proposed to overcome this limitation.
%Taking insights from the seminal work of \citep{quinonero2005unifying}, most of these approximations can be formulated around the \emph{inducing points}.
%These come from the same latent function that generates the data and can be viewed as representing its sufficient statistics \citet{titsias2009variational}.
%Using these approaches, the complexity reduces to $\bigO(NB^2)$ in computation and $\bigO(NB)$ in memory where $B \ll N$ is the number of inducing points.
%Recent development in stochastic variational inference for GPs \citep[GPSVI,][]{hensmangaussian} has led to further reduction of these costs to $\bigO(B^3)$ and $\bigO(B^2)$ respectively.

% While the important advance by GPSVI has made possible the application of GPs to datasets of millions of data points, the assumption that the statistics of the data is sufficiently captured by a, typically small, set of inducing variables might be too strong and limiting.
%This is addressed in \citep{nguyen2014fast} by using local sets of inducing points, as opposed to a global set, such that each local inducing sets sufficiently support a different local region of the input space.
%Yet another approach to this challenge is multi-task / multi-output learning, which is the focus of this paper.
%Specifically we pose the question: \emph{under the sparsity constraint of individual outputs}, can multiple output learning be effective?
%We believe this to be an interesting and  important research problem toward scalable multioutput GP models.
%Here we emphasize the word \emph{sparsity}, as it has been evident from the vast literature on GP approximations that  sparsity is the impeccable assumption to deal with large scale datasets.
%When learning things together, the hope is that information from different tasks can be used collaboratively to compensate for the sparsity enforcement to the individual outputs.

% can we do joint learning of multiple output under the sparsity assumption (meantioned above) for

% why is is this important?
% active research in multioutput learning GP-based, but on very small scale (show table)
% because existing multiple output does not approach the problem from this perspective (and only applicable to small scale)
% when we consider large scale, it is impeccable that the assumption of sparsity for each output must still hold.

% our solution: the modeling assumption and philosophy solution
%In this paper we propose a ...

% contribution from inference for large scale

% Experiments and summary of results

### 2014-03-16 paper/model.tex
model description and inference

--- 2014-03-06 paper/model.tex
@@
-We describe a multiple-output regression model.
-Suppose that we have inputs $\X = \{\x_n\}_{n=1}^N$ and outputs $\y = \{\y_n\}_{n=1}^N$, with $\x_n \in \calR^D$ and $\y_n \in \calR^P$, i.e. the output is $P$-dimensional.
-For ease of exposition, we assume that there are no missing values in any output dimension.
-The most dominant approach in GP-based multiple output regression is to compose the different outputs as some \textit{linear} mixing (combination) of some basic processes (random functions).
-Our model is also underpinned by this composition approach.
+Before diving into details of the specification, we discuss the philosophy and motivation behind our collaborative multioutput model.
+To learn the outputs jointly, we need a mechanism through which information can be transferred among the outputs.
+This is achieved in the model by allowing the outputs to share multiple sets of inducing variables.
+For exposition, consider a set of inducing points for a moment.
+Because they serve as the sufficient statistics of the individual outputs, a pattern common to the outputs can be encapsulated by these points when learning is done jointly. 
+Different patterns in the data can be captured by having  multiple set of inducing variables.
+It should be evident that these variables play a double pivotal role in the model: they represent the sufficient statistics that is crucial for sparse approximation and they collaborate sharing of information across the outputs.
 
-\noindent Consider the semiparametric latent factor models (SLFM) \cite{teh-et-al-aistats-05}.
-It assumes $Q$ basic independent processes, $g_j(\x) \sim \GP(0, k(\x,\x';\vectheta_j))$, where $\vectheta_j$ is the hyperparameters of the covariance function of $g_j(\x)$.
-Each (latent) output is a weighted combination of these basic processes, i.e.
-\begin{align}
-f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x), \quad i = 1 \hdots P,
-\end{align}
-In \cite{seeger2005semiparametric}, an extra degree of freedom is given by allowing each output $i$ to have its own process $h_i(\x) \sim \GP(0, k(\x,\x'; \vectheta^h_i)$ leading to
-\begin{align}
-f_i(\x) = \sum_{j=1}^Q w_{ij} g_j(\x) + h_i(\x), 
-\end{align}
-We emphasize that, although $k$ denotes covariance functions in general, each independent process can have its own covariance function.
-
-\noindent 
-\textbf{Augmented sparse GPs}
-Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
-This adds much flexibility to the model as each of the shared process $g_j(\x)$ can model a different pattern in the data with its own covariance function and inducing inputs.
-The roles of $g_j(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
-Furthermore, $g_j(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be a subspace of the input.
-
-%However, to really address the issue of scalability, this model is built upon sparse GPs from the ground up.
-
-\subsection{Prior and Likelihood}
-In this section we specify the prior and likelihood of this model. 
 \newcommand{\Zj}{\Z_j}
 \newcommand{\Zhi}{\Z^h_i}
-Let the set of inducing inputs for $g_j(\x)$ and $h_i(\x)$ be $\Z_j$ and $\Z^h_i$, and the corresponding inducing points $\u_j$ and $\v_i$, respectively.
+Consider multioutput regression of $P$ tasks with inputs $\X = \{\x_n \in \calR^D\}_{n=1}^N$ and outputs $\y = \{\y_i\}_{i=1}^P$ where $\y_i = \{y_{in}\}_{n=1}^N$. 
+Assume $Q$ latent functions with independent Gaussian process priors, $g_j(\x) \sim \GP(0, k_j(\cdot,\cdot))$, $j= 1 \hdots Q$.
+We introduce the sets of \emph{inducing variables} $\u_j$, each of which comes from $g_j(\x)$, i.e., $\u_j$ contains the function values at the inducing inputs $\Z_j$, which live in the same space as $\X$.
+These processes and inducing sets are shared by all of the outputs.
+To accommodate greater flexibility, we further allow each output to have its own independent process $h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$ whose the sets of inducing inputs and variables are $\Z^h_i$ and $\v_i$ respectively, where $i = 1 \hdots P$.
+% information is transfered via the inducing variables
 We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
 The subscripts are $i = 1 \hdots P$, $j = 1 \hdots Q$, and $n = 1 \hdots N$.
+For convenience, we shall assume the same number of inducing points, $M$, for all of the processes, however this is not imposed in practice.
 
-\noindent
-\textbf{Prior}
-The prior of the augmented model is given by:
+Following the definition of the GPs and the independence of the processes, the \emph{prior} of the multioutput model can be written as:
 \begin{align}
+\label{eq:gu}
 p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
 p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
+\label{eq:hv}
 p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
-p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)),
+p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)), \\
+ \BigMu_j &= k(\X,\Zj)k(\Zj,\Zj)^{-1}\u_j \\
+\BigMu^h_i &= k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}\v_i \\
+\tilde{\K}_j &= k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X) \\
+\tilde{\K}^h_i &= k^h_i(\X,\X) - k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}k(\Zhi,\X).
 \end{align}
-where $\BigMu_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}\u$ and $\tilde{\K}_j = k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X)$ and $\BigMu^h_i, \tilde{\K}^h_i$ are similarly defined.
-To avoid notational clutter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.
+In the above equations and hereafter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.
 
-\noindent
-\textbf{Likelihood}
-The likelihood as usual follows the standard iid Gaussian likelihood,
+The \emph{likelihood} with standard iid Gaussian likelihood is given by:
 \begin{align}
-p(\y | \g, \h ) = \prod_{i=1}^P p( \y_i ; \g, \h_i) = \prod_{i=1}^P \Normal( \y_i ; \sum_{j=1}^Q w_{ij} \g_j + \h_i, \beta_i^{-1} \I).
+p(\y | \g, \h ) = \prod_{i=1}^P \prod_{n=1}^N \Normal( y_{in} ; \sum_{j=1}^Q w_{ij} g_j(\x_n) + h_i(\x_n), \beta_i^{-1}),
 \end{align}
+which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
+As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
+We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor models \citep{teh-et-al-aistats-05,seeger2005semiparametric}.
+However, doing so is against the purpose of this model which encourage sharing of outputs via the inducing outputs.
+Furthermore, as we shall see in the next section, explicit representation of these variables is instrumental to scalable inference of this model.
+
+%\textbf{Augmented sparse GPs}
+%Toward scalable modeling, we replace standard GPs with sparse GPs augmented with \textit{different} set of inducing inputs.
+%This adds much flexibility to the model as each of the shared process $g_j(\x)$ can model a different pattern in the data with its own covariance function and inducing inputs.
+The roles of $g_j(\x)$ and $h_i(\x)$ can be quite different, so it is necessary that each has its own inducing inputs.
+%Furthermore, $g_j(\x)$ can be seen as a \textit{global} function operating on the entire input space (of all output dimensions), while each $h_i(\x)$ operates only on the inputs of the $i$-th output, which can be a subspace of the input.

### 2014-03-17 paper/experiments.tex
introduction

@@
 The first output has missing observations in the $(-7,-3)$ interval and the second output in the $(4,8)$ interval.
-Figure \ref{fig:toy} shows the predictive distributions by our model ($Q = 1$) and independent GPs with stochastic variational inference (GPSVI, one for each output).
+Figure \ref{fig:toy} shows the predictive distributions by our model ($Q = 1$) and independent GPs with stochastic variational inference (SVIGP, one for each output).
 It can be seen that our multioutput model gives perfect prediction in the unobserved regions where the independent models fail to interpolate despite having the same set of inducing inputs.
@@
 \includegraphics[scale=0.2]{figures/toy-svigp-y2.eps}
-\label{fig:toy}
 \end{tabular}
-\caption{Predictive distributions of the multioutput GPs (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the  toy problem.
+\caption{Real data and predictive distributions of our method (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the toy problem.
 Solid black line: predictive mean; grey bar: two standard deviations; magenta dots: real observations; blue dots: missing data.
@@
 \end{tabular}
-\caption{Real observations and predictive distributions for CAD (left), JPY (middle), and AUD (right). The color coding scheme is the same as in figure \ref{fig:toy}.}
+\caption{Real observations and predictive distributions for CAD (left), JPY (middle), and AUD (right). The color coding scheme is the same as in Figure \ref{fig:toy}.}
 \label{fig:fx}
@@
 \end{tabular}
-\caption{Predictive distributions of the multioutput GPs (left column) and full independent GPs (right column) for the air temperature problem.}
+\caption{Real data and predictive distributions by our method (COGP, left figures), the convolved GP method with exact inference (CGP, middle figures), and full independent GPs (right figures) for the air temperature problem.
+The color coding is the same as in Figure \ref{fig:toy}}
 \label{fig:weather}

### 2014-03-17 paper/inference.tex
introduction

@@
 %
+
+Inference in this model is done with variational inference.
+In section \ref{sec:variationalLowerBound}, we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points.
+The next section takes advantage of this factorization to derive stochastic variational inference which allows the model to scale to massive datasets.
+Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.
+
+\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
+\newcommand{\ug}{\u_g}
+\newcommand{\uh}{\u^h}
+\newcommand{\mgj}{\m_j}
+\newcommand{\mhi}{\m^h_i}
+\newcommand{\Sgj}{\S_j}
+\newcommand{\Shi}{\S^h_i}
+Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
+Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as,
+\begin{align}
+p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
+\end{align}
+Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
+This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution that of the form
+\begin{align}
+q(\g, \h, \u, \v | \y)
+= p(\g|\u) p(\h|\v) q(\u,\v),
+\end{align}
+with
+\begin{align}
+q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
+\end{align}
+This technique has been used in \citep{titsiasvariational,hensmangaussian} to derive variational inference for the single output case.
+Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given (eq.
+\ref{eq:gu} and \ref{eq:hv}), inference requires only learning of $q(\u, \v)$ that minimize the divergence between the approximate and true posteriors.
+This underlines the pivotal role of the inducing variables in this model as previously discussed.
+
+\noindent
+To find the best distribution $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
+\begin{align}
+%\nonumber
+%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
+%\nonumber
+%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
+%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
+\nonumber
+&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
+&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
+\label{eq:elbo}
+\end{align}
+which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
+Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ complexity.
+To compute the expected likelihood term in the ELBO we first see that
+\begin{align}
+\nonumber
+\log \text{ } &p(\y | \u, \v)
+% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
+% \nonumber
+\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
+&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g | \u) p(\h_i | \v_i)} 
+\end{align}
+where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
+The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
+\newcommand{\Ahi}{\A^h_i}
+\newcommand{\Zi}{\Z_i}
+The ELBO can be computed by first solving for the individual expectations 
+$l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ (see the supplementary material), which are then substituted into eq. \ref{eq:elbo} to give:
+\begin{align}
+\nonumber
+&\log p(\y)
+\ge \sum_{i,n}
+\bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
+          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{jnn} \\ \nonumber
+         &- \frac{1}{2} \beta_i \tilde{k}^h_{inn}
+         - \frac{1}{2} \beta_i \sum_{j=1}^Q \trace w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
+\bigg) \\
+\nonumber
+& - \sum_{j=1}^Q \bigg( \frac{1}{2} \log |\K_{jzz} \S_j^{-1}| + \frac{1}{2} \trace \K_{jzz}^{-1}  (\m_j \m_j^T + \S_j)\bigg) \\
+\nonumber
+& - \sum_{i=1}^P  \bigg( \frac{1}{2} \log |\K_{izz}(\Shi)^{-1}| + \frac{1}{2} \trace \K_{izz}^{-1} (\mhi (\mhi)^T + \Shi) \bigg) \\
+  &\define \calL,
+  \label{eq:L}
+\end{align}
+where $\K_{jzz} = k(\Zj, \Zj)$, $\K_{izz} = k(\Zhi, \Zhi)$, and with the help of auxiliary matrices $\A_j = k(\X,\Zj) \K_{jzz}^{-1}$ and $\Ahi = k(\X_i,\Zhi)\K_{izz}^{-1}$, 
+\begin{align}
+\tilde{\mu}_{in}
+%&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
+&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \\
+\mat{\Lambda}_{jn}
+%&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
+&= \A_j(n,:)^T \A_j(n,:)  \\
+\mat{\Lambda}_{in}
+%&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
+&= \Ahi(n,:)^T \Ahi(n,:) 
+\end{align}
+where $\tilde{k}_{jnn} = (\tilde{\K}_j)_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$,  $\mu^h_{in} = (\Mu^h_i)_n$, and $\A_j(n,:)$ selects the $n$-th row vector of $\A_j$.
+Notice that this ELBO clearly generalizes the bound for standard GP regression derived in \citep{hensmangaussian}, which can be recovered by setting $P = Q = 1$, $w_{ij} = 1$ and $h_i(\x) = 0$.
+
+The novelty of the derived variational lower bound is that it decomposes across both of the outputs and inputs.
+This enables the use of stochastic optimization methods which makes this model applicable to much larger datasets compared to the existing GP-based multiple regression models.
+
+\subsection{STOCHASTIC VARIATIONAL INFERENCE}
+\newcommand{\oi}{\vec{o}_i}
+% some notation
+
+So far in the description of the model and inference, it is implicitly assumed that every output has full observations at all of the inputs $\X$.
+To discern where learning occurs for each output, we make the missing data scenario more explicit.
+Specifically, each output $i$ can have observations at a different set of inputs $\X_i$. 
+We shall use $\oi$ to denote the indice of $\X_i$ (in the set $\X$) and use the indexing operator $\B(\oi)$ to select the rows corresponding to $\oi$ from any arbitrary matrix $\B$.
+We also overload $\y_i$ as the observed targets of output $i$.
+
+%------------------------------------------
+% derivatives of q(u_j)
+\newcommand{\Lgj}{\calL^g_j}
+\newcommand{\ynoj}{\y_i^{\backslash j}}
+\newcommand{\Kjzz}{\mat{K}_{jzz}}
+From the expression of the ELBO (eq. \ref{eq:L}), we can obtain the derivatives of the variational parameters to carry out stochastic optimization.
+The derivatives of $\calL$ w.r.t the parameters of $q(\u_j)$, $\m_j$ and $\S_j$, are given by:
+\begin{align}
+\nonumber
+\deriv{\calL}{\m_j} 
+=& \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj \\
+&- \bigg[\Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi) \bigg] \m_j \text{,} \\
+\deriv{\calL}{\S_j} 
+=& \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi) \bigg],
+\end{align}
+where $\y_i^{\backslash j} = \y_i - \Ahi(\oi) \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi) \m_{j'}$.
+%\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
+%\begin{align}
+%\nonumber
+%\calL^g_j \define&
+% \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
+% \nonumber
+% &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
+% + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
+% \\
+% &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
+%\end{align}
+%where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
+%$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\
+
+%-------------------------------------------
+%  derivatives of q(v_i)
+\newcommand{\Lhi}{\calL^h_i}
+\newcommand{\Kizz}{\mat{K}_{izz}}
+\noindent The derivatives of $\calL$ w.r.t the parameters of $q(\v_i)$, $\mhi$ and $\Shi$, are given by:
+\newcommand{\ynoh}{\y_i^{\backslash h}}
+\begin{align}
+\nonumber
+\deriv{\calL}{\mhi}
+= & \beta_i \Ahi(\oi)^T \ynoh  \\
+&- \bigg[\Kizz^{-1} +  \beta_i \Ahi(\oi)^T \Ahi(\oi) \bigg] \m_i \text{,} \\
+\deriv{\calL}{\Shi} 
+=& \frac{1}{2} \S_i^{-1} - \frac{1}{2} \bigg[ \Kizz^{-1} + \beta_i \Ahi(\oi)^T \Ahi(\oi) \bigg] ,
+\end{align}
+where $\ynoh = \y_i - \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j$.
+%\noindent As a function of the parameters of $q(\v_i)$, the lower bound $\calL$ is:
+%\begin{align}
+%\nonumber
+%\Lhi \define&
+% \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)
+% - \frac{1}{2} \beta_i \trace \tilde{\K}^h_i(\oi,\oi)
+% - \frac{1}{2} \beta_i \trace \Shi (\Ahi)^T \Ahi
+% \\
+%  &- \frac{1}{2} \log |k(\Zhi,\Zhi) (\Shi)^{-1}| -\frac{1}{2} \trace k(\Zhi,\Zhi)^{-1} (\mhi (\mhi)^T + \Shi) ,
+%\end{align}
+
+% comment on computation
+\noindent It can be seen that the derivatives of the parameters of $q(\v_i)$ only involve the observations of the output $i$.
+The derivatives of the parameters of $q(\u_j)$ involve the observations across all outputs but is a sum of contributions from individual outputs.
+Therefore, computation of the derivatives can be distributed or parallelized easily.
+This key property allows the model to potentially scale to problems with a very large number of inputs and outputs.
+
+Since the optimal distributions $q(\u_j)$ and $q(\v_i)$ are in the exponential family, it is more convenient to use stochastic variational inference \citep{hensman2012fast,hensmangaussian} to perform update of their canonical parameters.
+This works by taking a step of length $l$ in the direction of the natural gradient approximated by mini-batches of the data.
+For instance, consider $q(\u_j)$ whose canonical parameters are $\Phi_1 = \S_j^{-1} \m_j$ and $\Phi_2 = -\frac{1}{2}\S_j^{-1}$.
+Their stochastic update equations at time $t + 1$ are given by:
+\begin{align}
+\Phi_{1(t+1)} &= \S_{j(t)}^{-1} \m_{j(t)} + l \bigg(\sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj - \S_{j(t)}^{-1} \m_{j(t)} \bigg) \\
+\Phi_{2(t+1)} &= -\frac{1}{2} \S_{j(t)}^{-1} + l \bigg(\frac{1}{2} \S_{j(t)}^{-1} - \frac{1}{2} \mat{\Lambda} \bigg),
+\end{align}
+where $\mat{\Lambda} = \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi)$.
+
+The cost of this update is $\calO(PM^3)$ or $\calO(PN_bM^2)$, depending on which is larger between the number of inducing inputs (per process) and the mini-batch size.
+For large scale problems, we tend to use a large batch size (e.g.
+$N_b = 1000$) to reduce the stochasticity of the optimization, hence we consider the model to have a computational complexity of $\calO(PN_bM^2)$.
+Following this assumption, the memory demand is $\calO(PN_bM)$.
+%TODO table for complexity
+\subsection{INDUCING INPUTS AND HYPER-PARAMETERS \label{sec:hyperparameters}}
+To learn the hyperparameters, which in this model include the mixing weights, the covariance hyperparameters of the latent processes, and the noise precision of each output, we follow standard practice in GP inference.
+For this model, this involves taking derivatives of the ELBO and applying standard stochastic gradient descent in alternative steps with the variational parameters, much like a variational EM algorithm.
+The derivatives are given in the supplementary material.
+
+Learning of the inducing inputs, which was not considered in the single output case in \citep{hensmangaussian}, is also possible in our stochastic optimization approach.
+In the supplementary material, we show that with the help of vectorized operations, the derivatives of the lower bound w.r.t the inducing inputs can be computed with a cost \emph{independent} of the input dimension.
+This makes optimizing the inducing locations a practical option, which can be important for e.g. in high-dimensional problems where fixing the inducing inputs may hinder the performance.
+Indeed, our experiments on a large scale multioutput problem shows that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation.
+
+\subsection{PREDICTION}
+The predictive distribution of the $i$-th output for a test input $\x_*$ is given by:
+\begin{align}
+p(\fstar | \y, \x_*) = \Normal(\fstar; \sum_{j=1}^Q w_{ij} \mu_{j*} + \mu^h_{i*}, w_{ij}^2 s_{j*} + s^h_{i*}), 
+\end{align}
+where $\mu_{j*}$ and $s_{j*}$ are the mean and variance of the prediction for $g_{j*} = g_j(\x_*)$, i.e. $p(g_{j*} | \y, \x_*) = \Normal(g_{j*}; \mu_{j*}, s_{j*})$.
+Likewise, $\mu^h_{i*}$ and $s^h_{i*}$ are the mean and variance of the prediction for $h_{i*} = h_i(\x_*)$, $p(h_{i*} | \y, \x_*) = \Normal(h_{i*}; \mu^h_{i*}, s^h_{i*})$.
+Since the predictions for $g_{j*}$ and $h_{i*}$ have similar form, we only provide here the mean and variance for $\mu_{j*}$ and $s_{j*}$: 
+\begin{align}
+\mu_{j*} &= \vec{k}_{j*z} \K_{jzz}^{-1} \m_j \\ s_{j*} &= k_{j**} - \vec{k}_{j*z} \left(\K_{jzz}^{-1} - \K_{jzz}^{-1} \S_j \K_{jzz}^{-1} \right) \vec{k}_{j*z}^T
+\end{align}
+where $k_{j**} = k_j(\x_*, \x_*)$ and $\vec{k}_{j*z}$ is the covariance between the test point and the inducing inputs $\Zj$.
+
 %\noindent Also as a consequence of sufficient statistics, the approximate prediction for test targets $\vfstar$ at test inputs $\X_*$ is
@@
 
-Inference in this model is done with variational inference.
-In section \ref{sec:variationalLowerBound}, we derive a lower bound of the marginal likelihood which has the novel property of factorizing over the data points.
-The next section takes advantage of this factorization to derive stochastic variational inference which allows the model to scale to massive datasets.
-Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.
-
-\subsection{VARIATIONAL LOWER BOUND \label{sec:variationalLowerBound}}
-\newcommand{\ug}{\u_g}
-\newcommand{\uh}{\u^h}
-\newcommand{\mgj}{\m_j}
-\newcommand{\mhi}{\m^h_i}
-\newcommand{\Sgj}{\S_j}
-\newcommand{\Shi}{\S^h_i}
-Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
-Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as,
-\begin{align}
-p(\g, \h, \u, \v | \y) = p(\g | \u, \y) p(\h | \v, \y) p(\u, \v | \y).
-\end{align}
-Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
-This motivates the use of a distribution where the true posteriors over $\g$ and $\h$ are replaced by their conditional distributions given the inducing variables:
-\begin{align}
-q(\g, \h, \u, \v | \y)
-= p(\g|\u) p(\h|\v) q(\u,\v),
-\end{align}
-with
-\begin{align}
-q(\u,\v) &= \prod_{j=1}^Q \underbrace{\Normal(\u_j; \mgj, \Sgj)}_{q(\u_j)} \prod_{i=1}^P  \underbrace{\Normal(\v_i; \mhi, \Shi)}_{q(\v_i)}.
-\end{align}
-This technique has been used in \citep{titsiasvariational,hensmangaussian} to derive variational inference for the single output case.
-Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given (Equation \ref{eq:gu} and \ref{eq:hv}), inference requires only learning of $q(\u, \v)$ that minimize the divergence between the approximate and true posteriors, which becomes small when the sufficiency condition holds with greater certainty.
-
-\noindent
-In variational inference, this is done by optimizing the evidence lower bound (ELBO) of the log marginal:
-\begin{align}
-%\nonumber
-%\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
-%\nonumber
-%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v 
-%+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
-\nonumber
-&\log p(\y) \ge \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
-&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
-\label{eq:elbo}
-\end{align}
-which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
-Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ computation.
-To compute the expected likelihood term in the ELBO we first see that
-\begin{align}
-\nonumber
-\log \text{ } &p(\y | \u, \v)
-% &= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
-% \nonumber
-\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
-&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_{in} | \v_i)} 
-\end{align}
-where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
-The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
-\newcommand{\Ahi}{\A^h_i}
-\newcommand{\Zi}{\Z_i}
-The ELBO can be computed by first solving for the individual expectations 
-$l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ (see the supplementary material for detailed derivation), which are then substituted into equation \ref{eq:elbo} to give:
-\begin{align}
-\nonumber
-&\log p(\y)
-\ge \sum_{i,n}
-\bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
-          - \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} \\ \nonumber
-         &- \frac{1}{2} \beta_i \tilde{k}^h_{inn}
-         - \frac{1}{2} \beta_i \trace  \sum_{j=1}^Q w_{ij}^2 \S_j \mat{\Lambda}_{jn} - \beta_i \frac{1}{2} \trace \S^h_i \mat{\Lambda}_{in} 
-\bigg) \\
-&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)]  \define \calL,
-\end{align}
-where, with the help of auxiliary matrices $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$ and  $\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$, 
-\begin{align}
-\tilde{\mu}_{in}
-%&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
-&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \\
-\mat{\Lambda}_{jn}
-%&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
-&= \A_j(n,:)^T \A_j(n,:)  \\
-\mat{\Lambda}_{in}
-%&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
-&= \Ahi(n,:)^T \Ahi(n,:) 
-\end{align}
-where $\tilde{k}_{nn} = (\tilde{\K})_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$,  $\mu^h_{in} = (\Mu^h_i)_n$, and $\A_j(n,:)$ selects the $n$-th row vector of $\A_j$.
-
-% More on this elbo term
-\noindent Notice that this ELBO clearly generalizes the standard GP regression.
-In particular, setting $P = Q = 1$, $w_i = 1$ and $h_i(\x) = 0$ we recover the bound in Hensman et al \citet{hensmangaussian}.
-Due to the decomposition of this bound, we can use stochastic gradient descent to learn the variational parameters.
-
-\subsection{STOCHASTIC VARIATIONAL INFERENCE}
-\newcommand{\oi}{\vec{o}_i}
-% some notation
-
-The optimal posteriors are found by setting the gradients of the lowerbound $\calL$ wrt to the parameters of $q(\u_j)$ and $q(\v_i)$.
-Recall that in the model, different outputs can be observed at different inputs (i.e. the case of missing values).
-Let $\oi$ be the indice of the observed inputs of the output dimension $i$.
-We denote its set of observed inputs and targets as: $\X_i = \X(\oi,:)$ and $\y_i = y_{i}(\oi)$.
-
-%\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
 %\begin{align}
-%\nonumber
-%\calL^g_j \define&
-% \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
-% \nonumber
-% &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
-% + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
-% \\
-% &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
+%p(\fstar | \y, \x_*) = \int \Normal(\fstar; \sum_{j=1}^Q w_j g_{j*} + h_{i*}, 0) p(\g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der \g_* \der h_{i*},
 %\end{align}
-%where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
-%$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\
-
-%------------------------------------------
-% derivatives of q(u_j)
-\newcommand{\Lgj}{\calL^g_j}
-\newcommand{\ynoj}{\y_i^{\backslash j}}
-\newcommand{\Kjzz}{\mat{K}_{jzz}}
-\noindent The derivatives of $\calL$ wrt $\m_j$ and $\S_j$ are given by:
-\begin{align}
-\nonumber
-\deriv{\calL}{\m_j} 
-=& \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi,:)^T \ynoj \\
-&- \bigg[\Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg] \m_j \\
-\deriv{\calL}{\S_j} 
-=& \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi,:)^T \A_j(\oi,:) \bigg],
-\end{align}
-where $\Kjzz = k(\Zj,\Zj)$ and $\y_i^{\backslash j} = \y_i - \Ahi(\oi,:) \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi,:) \m_{j'}$.
-
-%-------------------------------------------
-%  derivatives of q(v_i)
-%\noindent As a function of the parameters of $q(\v_i)$, the lower bound $\calL$ is:
-%\begin{align}
-%\nonumber
-%\Lhi \define&
-% \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)
-% - \frac{1}{2} \beta_i \trace \tilde{\K}^h_i(\oi,\oi)
-% - \frac{1}{2} \beta_i \trace \Shi (\Ahi)^T \Ahi
-% \\
-%  &- \frac{1}{2} \log |k(\Zhi,\Zhi) (\Shi)^{-1}| -\frac{1}{2} \trace k(\Zhi,\Zhi)^{-1} (\mhi (\mhi)^T + \Shi) ,
-%\end{align}
-\newcommand{\Lhi}{\calL^h_i}
-\newcommand{\Kizz}{\mat{K}_{izz}}
-\noindent The derivatives of $\calL$ wrt $\mhi$ and $\Shi$ are given by:
-\newcommand{\ynoh}{\y_i^{\backslash h}}
-\begin{align}
-\nonumber
-\deriv{\calL}{\mhi}
-= & \beta_i (\Ahi(\oi,:))^T \ynoh \\ 
-& - \bigg[\Kizz^{-1} +  \beta_i (\Ahi(\oi,:))^T \Ahi(\oi,:) \bigg] \m_i \\
-\deriv{\calL}{\Shi} 
-&= \frac{1}{2} \S_i^{-1} - \frac{1}{2} \bigg[ \Kizz^{-1} + \beta_i \Ahi(\oi,:))^T \Ahi(\oi,:) \bigg] ,
-\end{align}
-where $\ynoh = \y_i - \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j$ and $\Kizz = k(\Zhi, \Zhi)$.
-
-% comment on computation
-\noindent It can be seen that the derivatives of the parameters of $q(\v_i)$ only involve the observations of the output dimension $i$.
-The derivatives of the parameters of $q(\u_j)$ involve the observations across all output dimensions but decompose as a sum of contributions from individual outputs.
-Therefore, computation of the derivatives (and hence the update equations) can be distributed or parallelized easily.
-This attractive property allows the model to scale to a very large number of inputs and outputs.
-
-% Comment on other hyperparameters
-
-\subsection{INDUCING INPUTS AND HYPER-PARAMETERS \label{sec:hyperparameters}}
-%\subsubsection{Update Equations}
-% and comment on the intuition of the update equations: e.g. y(x) - g(x) for h(x) 
-\subsection{PREDICTION}
-The predictive distribution of the $i$-th output for a test input $\x_*$ is 
-\begin{align}
-p(\fstar | \y, \x_*) = \int \Normal(\fstar; \sum_{j=1}^Q w_j g_{j*} + h_{i*}, 0) p(\g_* | \y, \x_*) p(h_{i*} | \y, \x_*) \der \g_* \der h_{i*},
-\end{align}
-where $p(\g_* | \y, \x_*) = \prod_{j=1}^Q \Normal(g_{j*}; \mu_{j*}, s_{j*})$ and $p(h_{i*} | \y, \x_*) = p(h_{i*}; \mu^h_{i*}, s^h_{i*})$ are the predictive distributions of the sparse GPs as given in eq.
-\ref{eq:sorprediction}.
-Therefore we have:
-\begin{align}
-p(\fstar | \y, \x_*) = \Normal(\fstar; \sum_{j=1}^Q w_{ij} \mu_{j*} + \mu_{i*}, w_{ij}^2 s_{j*} + s_{i*}). 
-\end{align}
-
-%\begin{linenomath}
-%\begin{align}
-%\sum_{i,h} \lambda_i \lambda_h cov [f(\x, \x')]
-%&= \sum_{i,h}   \sum_{j,j'} \lambda_i g(\x_i - \vs_j) k(\vs_j, \vs_{j'}) \lambda_h g(\x_h - \vs_j') \\
-%\end{align}
-%\end{linenomath}

### 2014-03-17 paper/intro.tex
introduction

@@
 % paper on multioutput so this should appear first in the introduction
+Gaussian processes \citep[GPs,][]{rasmussen-williams-book} are perhaps the dominant approach for learning functions in non-parametric Bayesian regression.
+Unfortunately, the prohibitive complexity when performing exact inference in the GP framework severely hinders its application to real world problems. 
+In particular, when learning $P$ multiple correlated functions over $N$ data points, the complexity is $\calO(N^3P^3)$ in computation and $\calO(N^2P^2)$ in memory.
+An example of a real-world application where multioutput regression may be useful is the tracking of  movements of a robot arm using 2 or more joint torques. 
+If one of the robot motors malfunctions and fails to record a torque, data collected from the other motors may be used to infer the missing torque values.
+However, taking 100 measurements per second already results in a total of over 40,000 data points in just 7 minutes.
+Clearly this problem is well beyond the capability of conventional multioutput GPs.
+Building scalable multioutput GPs model that can learn correlated tasks at the scale of the application just considered and potentially much larger is thus the main focus of this paper.
+
+% traditional approach 
+The traditional approach in multiple GP regression has been to mix independent latent processes to generate correlated functions.
+The mixing can be a weighted linear combinations with fixed \citep{teh-et-al-aistats-05,bonilla-et-al-nips-08} or input-dependent \citep{wilson-et-al-icml-12,nguyen2013efficient} coefficients or convolution of processes \citep{boyle-frean-nips-05,alvarez-lawrence-nips-08}.
+The correlations of the outputs are induced through sharing of these base processes.
+% sparse approximation technique and limitatio
+Sparse approximate techniques such as the informative vector machine (IVM) \cite{lawrence2002fast} or low-rank approximations which can be formulated around the \emph{inducing points} framework \citep{quinonero2005unifying},
+are then employed to reduce the computational and storage demands of the full models.
+Nevertheless, as shown in Table \ref{tab:complexity}, most of these models have complexity of at least $\calO(PNM^2)$ in computation and $\calO(PNM)$ thus prevents their applications to the  large scale problems we are targeting.
+
+\setlength{\tabcolsep}{4pt}
+\begin{table}[t]
+\caption{Comparison of the time and storage complexity of (approximate) inference of GP-based multioutput regression models.
+CGP is the convolved GPs model \citep{alvarez-lawrence-nips-08}.
+For all methods, $P$, $N$, $Q$ and $M$ are the number of outputs, inputs, latent processes, and inducing points, respectively.
+Only our method has complexity independent of the size of the inputs and thus can scale to massive datasets.}
+\label{tab:complexity}
+\begin{center}
+\begin{tabular}{lcc}
+\toprule
+\textbf{METHOD} & \textbf{TIME} & \textbf{STORAGE} \\ \hline
+COGP, this paper  & $\calO(PM^3)$ & $\calO(PM^2)$ \\
+SLFM, \citep{teh-et-al-aistats-05} & $\calO(QNM)$ & $\calO(QNM)$ \\
+MTGP, \citep{bonilla-et-al-nips-08} & $\calO(PNM^2)$ & $\calO(PNM)$\\ 
+CGP & $\calO(PNM^2)$ & $\calO(PNM)$ \\
+GPRN, \citep{wilson-et-al-icml-12} & $\calO(PQN^3)$ & $\calO(PQN^2)$ \\
+\bottomrule
+\end{tabular}
+\end{center}
+\end{table}
+
+% we approach from different perspective 
+Toward scalable multioutput Gaussian processes, we approach the problem from a different perspective based on several observations.
+First, \emph{sparsity} is an inevitable assumption in very large scale GPs, as evident by the vast amount of research on sparse approximations in the single output setting. % can cite ref here 
+Second, the \emph{inducing variables} are the key agent for effective learning under sparsity.
+In particular, they capture the \emph{sufficient statistics} of a dataset and lead to optimization of sparse processes that converge to the exact GP model \citep{titsias2009variational}.
+This is further exploited in \cite{hensmangaussian} where by explicitly representing a distribution over the inducing variables, stochastic variational inference can be used to work with millions of data points.
+
+% can we do joint learning of multiple output under the sparsity assumption (meantioned above) for 
+Motivated by these observations, we propose the collaborative multioutput GPs (COGP) model in which output correlations are fostered via sharing of the inducing points.
+The basic building blocks of the model are sparse processes.
+Each process has its own inducing points that sufficiently contain the main signals of the process. 
+Allowing the outputs to share these different inducing sets then creates a medium through which multiple sparse processes can be learned jointly.
+
+% contribution from inference for large scale
+This formulation around inducing points and sparse processes enables scalable inference of the model.
+In particular, we derive a variational lower bound of the model that factorizes across both of the inputs and outputs.
+This factorization makes possible the application of stochastic variational inference, thus allowing the model to handle even 'big data' problems.
+Furthermore, learning of all of the  hyperparameters, including the inducing inputs, can be done in a consistent stochastic optimization framework.
+
+The effectiveness of the model is first verified on a toy problem where the inducing variables are shown to be conducive to the sharing of information between two related tasks. 
+Further experiments are conducted on two moderate-sized multioutput problems in which the model shows competitive and better performance compared to other baselines.
+A large scale experiment confirms the modeling assumption that joint learning via sparse processes and their inducing inputs can lead to substantial performance gain.
+
+% Experiments and summary of results
+The rest of the paper is organized as follows.
+We give description of the model in Section \ref{sec:model}, followed by the inference procedure in Section \ref{sec:inference}. 
+Section \ref{sec:experiments} presents empirical evaluation.
+We conclude with some discussion in Section \ref{sec:discussion}.
+
 %Gaussian processes (GPs) \citep{rasmussen-williams-book} are perhaps the prior of choice in nonparametric Bayesian regression not only in the standard setting but also in multioutput learning \citep{} .
@@
 %When learning things together, the hope is that information from different tasks can be used collaboratively to compensate for the sparsity enforcement to the individual outputs.
-
-% can we do joint learning of multiple output under the sparsity assumption (meantioned above) for
-
-% why is is this important?
-% active research in multioutput learning GP-based, but on very small scale (show table)
-% because existing multiple output does not approach the problem from this perspective (and only applicable to small scale)
-% when we consider large scale, it is impeccable that the assumption of sparsity for each output must still hold.
-
-% our solution: the modeling assumption and philosophy solution
-%In this paper we propose a ...
-
-% contribution from inference for large scale
-
-% Experiments and summary of results

### 2014-03-17 paper/model.tex
introduction

@@
 Different patterns in the data can be captured by having  multiple set of inducing variables.
-It should be evident that these variables play a double pivotal role in the model: they represent the sufficient statistics that is crucial for sparse approximation and they collaborate sharing of information across the outputs.
+It should be evident that these variables play a double pivotal role in the model: they represent the sufficient statistics of the sparse processes and they collaborate sharing of information across the outputs.
 
@@
 We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor models \citep{teh-et-al-aistats-05,seeger2005semiparametric}.
-However, doing so is against the purpose of this model which encourage sharing of outputs via the inducing outputs.
-Furthermore, as we shall see in the next section, explicit representation of these variables is instrumental to scalable inference of this model.
+However, doing so is against the purpose of this model which encourage sharing of outputs via the inducing variables.
+Furthermore, as we shall see in the next section, explicit representation of these variables is instrumental to scalable inference of the model.
 

### 2014-03-18 paper/abstract.tex
add abstract

% limitation of learning with multiple GPs
Learning of correlated functions in the popular Gaussian processes (GPs) framework is often performed by mixing latent processes.
However, when these constituent processes are standard GPs, the applicability of this approach is severely inhibited by their intensive computational and memory demands.
% our approach
To address this problem, we propose the collaborative multioutput GPs for multitask regression at very large scale.
The building blocks of the model are \emph{sparse} processes underpinned by the \emph{inducing variables}.
These variables captures the sufficient statistics of a pattern (i.e. sparse process) in the data, and dependencies of the outputs are fostered via sharing different sets of inducing points.
This formulation enables extremely scalable inference of the model via stochastic variational inference.
Furthermore, all of the model hyperparameters including the inducing inputs can be optimized in the same stochastic approach.
Experiments on a toy problem and three real-world datasets verify the effectiveness of collaborative learning of sparse processes via the inducing variables promoted by the model.
They also demonstrate the computational superiority and scalability of the model compared to conventional multiple GPs regression.

### 2014-03-18 paper/intro.tex
add abstract

@@
 % sparse approximation technique and limitatio
-Sparse approximate techniques such as the informative vector machine (IVM) \cite{lawrence2002fast} or low-rank approximations which can be formulated around the \emph{inducing points} framework \citep{quinonero2005unifying},
+Sparse approximate techniques such as the informative vector machine (IVM) \cite{lawrence2002fast} or low-rank approximations which can be formulated around the inducing pointsframework \citep{quinonero2005unifying},
 are then employed to reduce the computational and storage demands of the full models.

### 2014-03-18 paper/discussion.tex
add discussion and fixes references

We have presented scalable multioutput GPs for learning of correlated functions.
The formulation around the inducing variables and their encapsulation of sufficient statistics was shown to be conducive to effective joint learning in the model. 
The sparsity induced by these variables crucially allows inference to be performed with stochastic optimization.
We note that although our large scale experiment was done with over 40,000 observations, which is the largest publicly available multioutput dataset we could find, the model can easily handle much bigger datasets.

We discuss several interesting extensions to this work.
First, consider the model for the single GP case, i.e., $P = 1$. By using $Q > 1$ sparse processes, the GP can now have multiple set of inducing variables.
These can be make local to capture varying characteristics of the latent process in different local input regions, for e.g. similar to \citep{nguyen2014fast}, leading to scalable GPs with non-stationary properties.

A further extension is to replace the constant mixing weights in the model with input-dependent coefficients like in the Gaussian process regression networks \citep[GPRN][]{wilson-et-al-icml-12} to accommodate adaptive dependencies among the outputs.
We leave the details of these extensions to future work.

### 2014-03-18 paper/experiments.tex
1. detailed description of toy problem 2. fix citation style 3. misc edits

@@
 Since we are using stochastic optimization, the learn rates need to be chosen carefully.
-We found that the rates used in \citet{hensmangaussian} also work well for our model.
+We found that the rates used in \citep{hensmangaussian} also work well for our model.
 Specifically, we used the learn rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise hyperparameter, and inducing inputs.
@@
 In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ and corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
-The first output has missing observations in the $(-7,-3)$ interval and the second output in the $(4,8)$ interval.
-Figure \ref{fig:toy} shows the predictive distributions by our model ($Q = 1$) and independent GPs with stochastic variational inference (SVIGP, one for each output).
-It can be seen that our multioutput model gives perfect prediction in the unobserved regions where the independent models fail to interpolate despite having the same set of inducing inputs.
-This confirms the effect of transfer of information from one output to another via collaborative learning of the inducing points.
-Furthermore, the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
+Each output is given 200 observations with missing values in the $(-7,-3)$ interval for the first output and the $(4,8)$ interval for the second output.
+We used $Q = 1$ latent sparse process, $h_1(x) = h_2(x) = 0$, $M = 15$ inducing inputs for our model.
+
+Figure \ref{fig:toy} shows the predictive distributions by COGP and independent GPs with stochastic variational inference (SVIGP, one for each output).
+The locations of the inducing inputs are fixed and identical for both methods.
+It is apparent from the figure that the independent GPs fail to predict the functions in the unobserved regions, especially for output 1.
+In contrast, prediction by COGP is perfect in these regions by using information from the observed intervals of one output to interpolate the missing signal of the other.
+Thus, this confirms the effectiveness of collaborative learning of sparse processes via the shared inducing variables. 
+Also note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
 
@@
 \end{tabular}
-\caption{Real data and predictive distributions of our method (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the toy problem.
+\caption{Real data and predictive distributions of by COGP (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the toy problem.
 Solid black line: predictive mean; grey bar: two standard deviations; magenta dots: real observations; blue dots: missing data.
-The black crosses show the locations of the inducing inputs.}
+The black crosses show the locations of the inducing inputs.
+By sharing inducing points across the outputs, COGP accurately interpolates the missing function values.}
 \label{fig:toy}
@@
 \subsection{FOREIGN EXCHANGE RATE PREDICTION}
-The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca/data.html}.
-The setting of our experiment described here is identical to that in \citet{alvarez2010efficient}.
+The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca}.
+The setting of our experiment described here is identical to that in \citep{alvarez2010efficient}.
 The dataset consists of all the data available for the 251 working days in the year of 2007.
@@
 Note that these 3 currencies are from very different geographical locations, making the problem more interesting. 
-The 153 points extracted is used for testing, and the remaining 3051 data points is used for training.
-Since the missing data correspond to long contiguous sections, the objective here is to evaluate the capacity of the model to impute the missing currency values based on other currencies.
+The 153 points are used for testing, and the remaining 3051 data points are used for training.
+Since the missing data corresponds to long contiguous sections, the objective here is to evaluate the capacity of the model to impute the missing currency values based on other currencies.
 %todo: batch size, learn rate (maybe at the beginning of experiments)
@@
 For preprocessing we normalized the outputs to have zero mean and unit variance.
-Since the exchange rates are driven by a small number of latent market forces \citet{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
+Since the exchange rates are driven by a small number of latent market forces \citep{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
 We used the squared-exponential covariance function for the shared processes and the noise covariance function for the independent process of each output.
-The inducing inputs $(M = 100)$ are randomly selected from the training data and fixed throughout training.
+The inducing inputs $(M = 100, \text{per sparse process})$  are randomly selected from the training data and fixed throughout training.
 
 The real data and predictive distributions by our model are shown in Figure \ref{fig:fx}.
-They exhibit similar behaviors to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
+They exhibit similar behaviors to those by the convolved model with inducing kernels in \citep{alvarez2010efficient}.
 In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
@@
 %todo: result for independent gps
-We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citet{alvarez-lawrence-nips-08} (CGP) and with approximation via the variational inducing kernels \citet{alvarez2010efficient}) (CGPVAR) in addition to using independent GPs (IGP, one for each output).
+We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citep{alvarez-lawrence-nips-08} (CGP) and with approximation via the variational inducing kernels \citep{alvarez2010efficient}) (CGPVAR) in addition to using independent GPs (IGP, one for each output).
 Our model outperforms both of the CGP variants in terms of standardized mean squared error (SMSE).
@@
 % no std because there are 3 outputs but variance is small
-Note that the SMSE of CGPVAR is taken from \citet{alvarez2010efficient} while the NLPD was not provided.
+Note that the SMSE of CGPVAR is taken from \citep{alvarez2010efficient} while the NLPD was not provided.
 Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.
@@
 Next, we consider the task of predicting air temperature at 4 different locations in the south coast of England. 
-The data is gathered from a network of weather sensors (named Bramblemet, Sotonmet, Cambermet, and Chimet), each of which measures several environmental variables \citet{osborne2008towards}\footnote{Data is available at seperate web pages, see e.g.
+The data is gathered from a network of weather sensors (named Bramblemet, Sotonmet, Cambermet, and Chimet), each of which measures several environmental variables \citep{osborne2008towards}\footnote{Data is available at seperate web pages, see e.g.
 http://www.bramblemet.co.uk}.
@@
 
-The real data and the predictive distributions by 3 models (COGP, CGP with exact inference \citet{alvarez-lawrence-nips-08}, and independent GPs) are shown in Figure \ref{fig:weather}.
+The real data and the predictive distributions by 3 models (COGP, CGP with exact inference \citep{alvarez-lawrence-nips-08}, and independent GPs) are shown in Figure \ref{fig:weather}.
 It is clear that the independent GPs model is clueless in the test regions and thus simply uses the average temperature as its prediction.
@@
 The data consists of 48,933 mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
-%The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citet{vijayakumar2000locally}.
+%The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citep{vijayakumar2000locally}.
 %Furthermore, exploratory analysis using standard GP with automatic relevance determination suggests that all of the 21 dimensions are relevant.
-It has been used in previous work (see e.g. \citet{rasmussen-williams-book},  \citet{vijayakumar2000locally}) but only for single task learning.
+It has been used in previous work (see e.g. \citep{rasmussen-williams-book},  \citep{vijayakumar2000locally}) but only for single task learning.
 Here we consider joint learning for the 4th and 7th torques where the first has only 2,000 points  while the second has the full data of 44,484 points for training.

### 2014-03-18 paper/inference.tex
add discussion and fixes references
1. detailed description of toy problem 2. fix citation style 3. misc edits

@@
 \end{align}
-This technique has been used in \citep{titsiasvariational,hensmangaussian} to derive variational inference for the single output case.
+This technique has been used in \citep{titsias2009variational,hensmangaussian} to derive variational inference for the single output case.
 Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given (eq.
@@
 Their stochastic update equations at time $t + 1$ are given by:
+\small
 \begin{align}
@@
 \end{align}
+\normalsize
 where $\mat{\Lambda} = \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi)$.
@@
 Likewise, $\mu^h_{i*}$ and $s^h_{i*}$ are the mean and variance of the prediction for $h_{i*} = h_i(\x_*)$, $p(h_{i*} | \y, \x_*) = \Normal(h_{i*}; \mu^h_{i*}, s^h_{i*})$.
-Since the predictions for $g_{j*}$ and $h_{i*}$ have similar form, we only provide here the mean and variance for $\mu_{j*}$ and $s_{j*}$: 
+These predictive means and variances are given by: 
 \begin{align}
-\mu_{j*} &= \vec{k}_{j*z} \K_{jzz}^{-1} \m_j \\ s_{j*} &= k_{j**} - \vec{k}_{j*z} \left(\K_{jzz}^{-1} - \K_{jzz}^{-1} \S_j \K_{jzz}^{-1} \right) \vec{k}_{j*z}^T
+\mu_{j*} &= \vec{k}_{j*z} \K_{jzz}^{-1} \m_j \\ s_{j*} &= k_{j**} - \vec{k}_{j*z} \left(\K_{jzz}^{-1} - \K_{jzz}^{-1} \S_j \K_{jzz}^{-1} \right) \vec{k}_{j*z}^T \\
+\mu^h_{i*} &= \vec{k}_{i*z} \K_{izz}^{-1} \mhi \\
+s^h_{i*} &= k_{i**} - \vec{k}_{i*z} \left(\K_{izz}^{-1} - \K_{izz}^{-1} \S_j \K_{izz}^{-1} \right) \vec{k}_{i*z}^T
 \end{align}
-where $k_{j**} = k_j(\x_*, \x_*)$ and $\vec{k}_{j*z}$ is the covariance between the test point and the inducing inputs $\Zj$.
+where $k_{j**} = k_j(\x_*, \x_*)$, $k_{i**} = k^h_i(\x_*, \x_*)$,  $\vec{k}_{j*z}$ is the covariance between $\x_*$ and $\Zj$, and $\vec{k}_{i*z}$ is the covariance between $\x_*$ and $\Zhi$.
 

### 2014-03-18 paper/discussion.tex
add complexity analysis and edit various parts

@@
 We have presented scalable multioutput GPs for learning of correlated functions.
-The formulation around the inducing variables and their encapsulation of sufficient statistics was shown to be conducive to effective joint learning in the model. 
-The sparsity induced by these variables crucially allows inference to be performed with stochastic optimization.
-We note that although our large scale experiment was done with over 40,000 observations, which is the largest publicly available multioutput dataset we could find, the model can easily handle much bigger datasets.
+The formulation around the inducing variables and their encapsulation of sufficient statistics was shown to be conducive to not only effective but also scalable joint learning under sparsity. 
+We note that although our large scale experiment was done with over 40,000 observations -- the largest publicly available multioutput dataset we could find, the model can easily handle much bigger datasets.
 

### 2014-03-18 paper/experiments.tex
add complexity analysis and edit various parts

@@
 Each output is given 200 observations with missing values in the $(-7,-3)$ interval for the first output and the $(4,8)$ interval for the second output.
-We used $Q = 1$ latent sparse process, $h_1(x) = h_2(x) = 0$, $M = 15$ inducing inputs for our model.
+We used $Q = 1$ latent sparse process with squared exponential kernel, $h_1(x) = h_2(x) = 0$, and $M = 15$ inducing inputs for our model.
 
-Figure \ref{fig:toy} shows the predictive distributions by COGP and independent GPs with stochastic variational inference (SVIGP, one for each output).
+Figure \ref{fig:toy} shows the predictive distributions by our model (COGP) and independent GPs with stochastic variational inference (SVIGP, one for each output).
 The locations of the inducing inputs are fixed and identical for both methods.
 It is apparent from the figure that the independent GPs fail to predict the functions in the unobserved regions, especially for output 1.
-In contrast, prediction by COGP is perfect in these regions by using information from the observed intervals of one output to interpolate the missing signal of the other.
-Thus, this confirms the effectiveness of collaborative learning of sparse processes via the shared inducing variables. 
+In contrast, by using information from the observed intervals of one output to interpolate the missing signal of the other, COGP makes perfect prediction for both outputs.
+This confirms the effectiveness of collaborative learning of sparse processes via the shared inducing variables. 
 Also note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
@@
 For preprocessing we normalized the outputs to have zero mean and unit variance.
-Since the exchange rates are driven by a small number of latent market forces \citep{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
-We used the squared-exponential covariance function for the shared processes and the noise covariance function for the independent process of each output.
+Since the exchange rates are driven by a small number of latent market forces \citep[see e.g.][]{alvarez2010efficient}, we tried different values of $Q = {1,2,3}$ and selected $Q = 2$ which gave the best model evidence (ELBO).
+We used the squared-exponential covariance function for the shared processes and the noise covariance function for the individual process of each output.
 The inducing inputs $(M = 100, \text{per sparse process})$  are randomly selected from the training data and fixed throughout training.
+
+\begin{figure*}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.3]{figures/fxCAD.eps} &
+\includegraphics[scale=0.3]{figures/fxJPY.eps} &
+\includegraphics[scale=0.3]{figures/fxAUD.eps}
+\end{tabular}
+\caption{Real observations and predictive distributions for CAD (left), JPY (middle), and AUD (right). The color coding scheme is the same as in Figure \ref{fig:toy}.}
+\label{fig:fx}
+\end{figure*}
 
@@
 Our investigation of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 150 - 200.
-This was effectively used by the model to extrapolate the values of the AUD.
+This information from these currencies was effectively used by the model to extrapolate the values of the AUD.
 
 %todo: result for independent gps
-We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citep{alvarez-lawrence-nips-08} (CGP) and with approximation via the variational inducing kernels \citep{alvarez2010efficient}) (CGPVAR) in addition to using independent GPs (IGP, one for each output).
-Our model outperforms both of the CGP variants in terms of standardized mean squared error (SMSE).
-It has lower test likelihood, as measured by the negative log predictive density (NLPD), which is mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
-%The results are averaged across the 3 outputs over 5 repetitions.
+We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citep[CGP,][]{alvarez-lawrence-nips-08} and independent GPs (IGP, one for each output).
+Our model outperforms both of CGP and IGP in terms of the standardized mean squared error (SMSE).
+CGP has lower negative log predictive density (NLPD), mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
 % no std because there are 3 outputs but variance is small
-Note that the SMSE of CGPVAR is taken from \citep{alvarez2010efficient} while the NLPD was not provided.
+For reference, the convolved GPs with approximation via the variational inducing kernels  \citep[CGPVAR,][]{alvarez2010efficient} has the SMSE of 0.2795 while the NLPD was not provided.
 Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.
@@
 \begin{table}[t]
-\caption{Performance comparison on the foreign exchange rate dataset. Results are averaged of the 3 outputs over 5 repetitions.}
+\caption{Performance comparison on the foreign exchange rate dataset. Results are averaged of the 3 outputs over 5 repetitions. Smaller figures are better.}
 \label{tab:fx}
@@
 
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/fxCAD.eps} &
-\includegraphics[scale=0.3]{figures/fxJPY.eps} &
-\includegraphics[scale=0.3]{figures/fxAUD.eps}
-\end{tabular}
-\caption{Real observations and predictive distributions for CAD (left), JPY (middle), and AUD (right). The color coding scheme is the same as in Figure \ref{fig:toy}.}
-\label{fig:fx}
-\end{figure*}
-
 \subsection{AIR TEMPERATURE PREDICTION}
@@
 For pre-processing we normalized the outputs to have zero mean and unit variance.
+We used $Q = 2$ sparse processes with the squared exponential covariance function and individual processes with the noise covariance function.
 The inducing inputs $(M = 200)$ are randomly selected from the training set and fixed throughout training.
-The number of latent processes is $Q = 2$ for both our model and the CGP model.
 
-The real data and the predictive distributions by 3 models (COGP, CGP with exact inference \citep{alvarez-lawrence-nips-08}, and independent GPs) are shown in Figure \ref{fig:weather}.
+\begin{figure*}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.3]{figures/cogp-weatherCambermet.eps} &
+\includegraphics[scale=0.3]{figures/cgp-weatherCambermet.eps} &
+\includegraphics[scale=0.3]{figures/weatherCambermet.eps}
+\\
+\includegraphics[scale=0.3]{figures/cogp-weatherChimet.eps} &
+\includegraphics[scale=0.3]{figures/cgp-weatherChimet.eps} &
+\includegraphics[scale=0.3]{figures/weatherChimet.eps}
+\end{tabular}
+\caption{Real data and predictive distributions by our method (COGP, left figures), the convolved GP method with exact inference (CGP, middle figures), and full independent GPs (right figures) for the air temperature problem.
+The coding color scheme is the same as in Figure \ref{fig:toy}.}
+\label{fig:weather}
+\end{figure*}
+
+The real data and the predictive distributions by our model, CGP with exact inference \citep{alvarez-lawrence-nips-08}, and independent GPs are shown in Figure \ref{fig:weather}.
 It is clear that the independent GPs model is clueless in the test regions and thus simply uses the average temperature as its prediction.
 For Cambermet, both COGP and CGP can capture the rising in temperature from the morning till the afternoon and the fall afterward.
-For Chimet, both models perform poorly but CGP more so as it falsely predicts  wild fluctuations that are non-existent in the data. 
-Note that the two latent processes learned by our model indeed correspond to different patterns in the data: one process has the inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has the inverse lengthscale of 0.5 to model the local variations within a single day.
-The comparative performance of the models are summarized in Table \ref{tab:air}, from which it can be seen that our model outperforms CGP in terms of both test accuracy and confidence.
+%For Chimet, both models perform poorly but CGP more so as it falsely predicts  wild fluctuations that are non-existent in the data. 
+The comparative performance of the models are summarized in Table \ref{tab:air}, from which it can be seen that our model outperforms CGP in terms of both SMSE and NLPD.
 %All results are averaged of the 2 outputs over 5 repetitions.
@@
 
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/cogp-weatherCambermet.eps} &
-\includegraphics[scale=0.3]{figures/cgp-weatherCambermet.eps} &
-\includegraphics[scale=0.3]{figures/weatherCambermet.eps}
-\\
-\includegraphics[scale=0.3]{figures/cogp-weatherChimet.eps} &
-\includegraphics[scale=0.3]{figures/cgp-weatherChimet.eps} &
-\includegraphics[scale=0.3]{figures/weatherChimet.eps}
-\end{tabular}
-\caption{Real data and predictive distributions by our method (COGP, left figures), the convolved GP method with exact inference (CGP, middle figures), and full independent GPs (right figures) for the air temperature problem.
-The color coding is the same as in Figure \ref{fig:toy}}
-\label{fig:weather}
-\end{figure*}
+It is also worth noting the characteristics of the sparse processes learned by our model as they indeed correspond to different patterns in the data.
+In particular, one process has the inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has the inverse lengthscale of 0.5 to model the local variations within a single day.
 
-\subsection{INVERSE DYNAMICS OF ROBOT ARM}
-Our last experiment is with a dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm\footnote{Data is available at http://www.gaussianprocess.org/gpml/data/}.
+\subsection{ROBOT INVERSE DYNAMICS}
+Our last experiment is with a dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm \citep{vijayakumar2000locally}.
 The data consists of 48,933 mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
@@
 %Furthermore, exploratory analysis using standard GP with automatic relevance determination suggests that all of the 21 dimensions are relevant.
-It has been used in previous work (see e.g. \citep{rasmussen-williams-book},  \citep{vijayakumar2000locally}) but only for single task learning.
-Here we consider joint learning for the 4th and 7th torques where the first has only 2,000 points  while the second has the full data of 44,484 points for training.
+It has been used in previous work \citep[see e.g.][]{rasmussen-williams-book,vijayakumar2000locally} but only for single task learning. 
+In \citep{chai2008multi}, multitask learning of robot inverse dynamics was considered but on a different and much smaller dataset.
+
+Here we consider joint learning for the 4th and 7th torques where the first has 2,000 points  while the second has the full data of 44,484 points for training.
 The test set consists of 8,898 observations equally divided between the two outputs.
-Note that the data was collected at 100Hz for 7.5 minutes in total, hence ideal for this experiment which evaluates the effectiveness of collaborative learning under the assumption of sparsity in the output spaces.
+Note that the data was collected at 100Hz for 7.5 minutes in total, hence ideal for this experiment which evaluates the effectiveness of joint learning under the assumption of sparsity in the output spaces.
 
@@
 The performance of all methods, in terms of SMSE and NLPD averaged over 5 repetitions, is given in Table \ref{tab:robotarm}.
-The benefits of learning the two outputs jointly are evident, as can be seen by the significantly higher test accuracy and confidence of COGP compared to the full GP for the first output (4th torque).
-While the SMSE of the second output is essentially the same by all methods, the predictive likelihood of our multioutput model is better than that of the independent SVIGP model which has the same amount of training data for this torque.
-The results thus confirm the impact of collaborative learning even when sparsity assumption is made in the model.
+The benefits of learning the two outputs jointly are evident, as can be seen by the significantly lower SMSE and NLPD of COGP compared to the full GP for the first output (4th torque).
+While the SMSE of the second output is essentially the same for all methods, the NLPD of COGP is substantially better than that of the independent SVIGP model which has the same amount of training data for this torque.
+The results validate the impact of collaborative learning even when under sparsity assumption, thus open up new opportunities for improvement over single task learning with independent sparse processes.
 
 \begin{table}[t]
-\caption{Performance comparison on the robot arm dataset.
-In the two bottom lines, standard GP is applied to output 1 and the remaining method is applied to output 2.
-Results are averaged over 5 repetitions.}
+\caption{Performance comparison on the robot inverse dynamics dataset.
+In the two bottom lines, standard GP is applied to output 1 and the other method is applied to output 2.
+Results are averaged over 5 repetitions.
+Smaller figures are better.}
 \label{tab:robotarm}
@@
 Also from Table \ref{tab:robotarm}, it seems that optimizing the inducing inputs can lead to better performance than fixing them.
-More importantly, this comes with slight overhead in computation, as demonstrated by the training time for the two options in Figure \ref{fig:time}.
-For instance, the total training time with 500 inducing inputs is 1.9 hours for learning versus 1.6 hours for keeping them fixed.
-Recall that this dataset is 21-dimensional so this confirms our analysis in the inference section that the cost of optimization of the inducing inputs does not scale linearly with the dimensionality of the problem.
-Thus, for datasets of high dimension like this one, learning of the inducing inputs is a practical option.
+More importantly, the overhead in computation is slight, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
+For instance, the total training time is only 1.9 hours when learning with 500 inducing inputs compared to 1.6 hours when fixing them.
+As this dataset is 21-dimensional, the small difference in training time confirms our analysis that the cost of optimizing the inducing inputs only scales sublinearly with the dimensionality of the problem.
+Thus, for datasets of high dimension, learning of the inducing inputs is a practical option.
 

### 2014-03-18 paper/inference.tex
add complexity analysis and edit various parts

@@
 which is derived using Jensen's inequality and the fact that both of $q(\u, \v)$ and $p(\u, \v)$ fully factorize.
-Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergences are analytically tractable and require $\calO(M^3)$ complexity.
+Since $q(\u_j), q(\v_i), p(\u_j), p(\v_i)$ are all multivariate Gaussian distributions, the KL divergence terms are analytically tractable.
 To compute the expected likelihood term in the ELBO we first see that
@@
 
-The novelty of the derived variational lower bound is that it decomposes across both of the outputs and inputs.
-This enables the use of stochastic optimization methods which makes this model applicable to much larger datasets compared to the existing GP-based multiple regression models.
+The novelty of the derived variational lower bound is that it decomposes across not only the outputs but also the inputs.
+This enables the use of stochastic optimization methods which makes this model applicable to much larger datasets compared to the existing GP-based multioutput models.
 
@@
 \newcommand{\Kjzz}{\mat{K}_{jzz}}
-From the expression of the ELBO (eq. \ref{eq:L}), we can obtain the derivatives of the variational parameters to carry out stochastic optimization.
-The derivatives of $\calL$ w.r.t the parameters of $q(\u_j)$, $\m_j$ and $\S_j$, are given by:
+From (eq. \ref{eq:L}), we can obtain the derivatives of the ELBO wrt the variational parameters for optimization.
+The derivatives of $\calL$ wrt the parameters of $q(\u_j)$ are given by:
 \begin{align}
-\nonumber
 \deriv{\calL}{\m_j} 
 =& \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj \\
+\nonumber
 &- \bigg[\Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi) \bigg] \m_j \text{,} \\
@@
 \newcommand{\Kizz}{\mat{K}_{izz}}
-\noindent The derivatives of $\calL$ w.r.t the parameters of $q(\v_i)$, $\mhi$ and $\Shi$, are given by:
+\noindent The derivatives of $\calL$ wrt the parameters of $q(\v_i)$ are given by:
 \newcommand{\ynoh}{\y_i^{\backslash h}}
@@
 Therefore, computation of the derivatives can be distributed or parallelized easily.
-This key property allows the model to potentially scale to problems with a very large number of inputs and outputs.
+This property further enhances the scalability of the model, allowing it to handle big data problems with a very large number of inputs and outputs.
 
@@
 Their stochastic update equations at time $t + 1$ are given by:
-\small
 \begin{align}
@@
 
-The cost of this update is $\calO(PM^3)$ or $\calO(PN_bM^2)$, depending on which is larger between the number of inducing inputs (per process) and the mini-batch size.
-For large scale problems, we tend to use a large batch size (e.g.
-$N_b = 1000$) to reduce the stochasticity of the optimization, hence we consider the model to have a computational complexity of $\calO(PN_bM^2)$.
-Following this assumption, the memory demand is $\calO(PN_bM)$.
 %TODO table for complexity
@@
 Indeed, our experiments on a large scale multioutput problem shows that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation.
+
+\subsection{COMPLEXITY ANALYSIS}
+In this section we analyze the complexity of the model and compare it to existing multioutput approaches.
+For consistency, we first unify common notations used for all models.
+We use $P$ as the number of outputs, $N$ as the number of inputs, $Q$ as the number of shared latent processes, $M_t$ as the \emph{total} number of inducing inputs.
+It is worth noting that $M_t = (P + Q) \times M$ in our COGP model, assuming that each sparse process has equal number of inducing points. 
+Also, COGP have $P$ additional individual processes, one for each output.
+
+The complexity of COGP can be read off by inspecting the ELBO (eq. \ref{eq:L}), with a key observation that it contains a sum over the outputs as well the inputs.
+This means a mini-batch containing a small subset of the inputs and outputs can be used for stochastic optimization.
+Technically, the cost is $\calO(M^3)$ or $\calO(N_bM^2)$, where $N_b$ is the size of the mini-batches, depending on which is larger between $M$ and $N_b$. 
+Though in practice we may use $N_b > M$ as a large batch size (e.g. $N_b = 1000$) helps reduce stochasticity of the optimization.
+However, here we use $\calO(M^3)$ here for easier comparison with existing models whose computational and storage demands are given in \ref{tab:complexity}. 
+As can be seen, COGP has superior complexity which is independent of the size of the inputs and outputs, thus is the only method capable of handling large scale problems.
+
+\setlength{\tabcolsep}{4pt}
+\begin{table}[t]
+\caption{Comparison of the time and storage complexity of (approximate) inference of GP-based multioutput regression models.
+CGP-FITC is the convolved GPs model with sparse approximation\citep{alvarez-lawrence-nips-08}.
+COGP is the only method with complexity \emph{independent} of the number of inputs and outputs, thus can scale to massive datasets.}
+\label{tab:complexity}
+\begin{center}
+\small
+\begin{tabular}{lcc}
+\toprule
+\textbf{METHOD} & \textbf{TIME} & \textbf{STORAGE} \\ \hline
+COGP, this paper  & $\calO(M^3)$ & $\calO(M^2)$ \\
+SLFM, \citep{teh-et-al-aistats-05} & $\calO(QNM_t)$ & $\calO(QNM_t)$ \\
+MTGP, \citep{bonilla-et-al-nips-08} & $\calO(PNM_t^2)$ & $\calO(PNM_t)$\\ 
+CGP-FITC & $\calO(PNM_t^2)$ & $\calO(PNM_t)$ \\
+GPRN, \citep{wilson-et-al-icml-12} & $\calO(PQN^3)$ & $\calO(PQN^2)$ \\
+\bottomrule
+\end{tabular}
+\end{center}
+\end{table}
 

### 2014-03-18 paper/intro.tex
1. detailed description of toy problem 2. fix citation style 3. misc edits
add complexity analysis and edit various parts

--- 2014-03-16 paper/intro.tex
@@
 % paper on multioutput so this should appear first in the introduction
+Gaussian processes \citep[GPs,][]{rasmussen-williams-book} are perhaps the dominant approach for learning functions in non-parametric Bayesian regression.
+Unfortunately, the prohibitive complexity when performing exact inference in the GP framework severely hinders its application to real world problems. 
+In particular, when learning $P$ multiple correlated functions over $N$ data points, the complexity is $\calO(N^3P^3)$ in computation and $\calO(N^2P^2)$ in memory.
+An example of a real-world application where multioutput regression may be useful is the tracking of  movements of a robot arm using 2 or more joint torques. 
+If one of the robot motors malfunctions and fails to record a torque, data collected from the other motors may be used to infer the missing torque values.
+However, taking 100 measurements per second already results in a total of over 40,000 data points per torque in just 7 minutes.
+Clearly this problem is well beyond the capability of conventional multiple output GPs.
+Building scalable multioutput GPs model that can learn correlated tasks at the scale of the application just considered and much larger is thus the main focus of this paper.
+
+% traditional approach 
+The traditional approach in multiple GP regression has been to mix independent latent processes to generate correlated functions.
+The mixing can be a weighted linear combinations with fixed \citep{teh-et-al-aistats-05,bonilla-et-al-nips-08} or input-dependent \citep{wilson-et-al-icml-12,nguyen2013efficient} coefficients or convolution of processes \citep{boyle-frean-nips-05,alvarez-lawrence-nips-08}.
+The correlations of the outputs are induced through sharing of these base processes.
+% sparse approximation technique and limitatio
+Sparse approximate techniques such as the informative vector machine (IVM) \cite{lawrence2002fast} or low-rank approximations based around the inducing points framework \citep{quinonero2005unifying},
+are then employed to reduce the computational and storage demands of the full models.
+Nevertheless, most of these models have complexity of at least $\calO(PNM^2)$ in computation and $\calO(PNM)$ in memory, where $M$ is the number of inducing points.
+thus prevents their applications to the  large scale problems we are targeting.
+
+% we approach from different perspective 
+In this work, we approach the challenge of building scalable multioutput Gaussian processes based on several observations.
+First, \emph{sparsity} is an inevitable assumption in very large scale GPs, as evident by the vast amount of research on sparse approximations in the single output setting. % can cite ref here 
+Second, the \emph{inducing variables} are the key catalyst for effective and scalable learning under sparsity.
+In particular, they capture the \emph{sufficient statistics} of a dataset and lead to optimization of sparse processes that converge to the exact GP model \citep{titsias2009variational}.
+This is further exploited in \cite{hensmangaussian} where by explicitly representing a distribution over the inducing variables, stochastic variational inference can be used to work with millions of data points.
+
+% can we do joint learning of multiple output under the sparsity assumption (meantioned above) for 
+Motivated by these observations, we propose the collaborative multioutput GPs (COGP) model where each output is a weighted combination of sparse processes. 
+Accompanying each process is its own set of inducing points that sufficiently encapsulates the process signals and characteristics.
+Correlations among the outputs are fostered via sharing of multiple inducing sets which in turn create a medium through which joint inference happens.
+
+% contribution from inference for large scale
+This formulation around inducing points and sparse processes enables scalable inference of the model.
+In particular, we derive a variational lower bound of the model evidence that factorizes across not only the outputs but also the inputs.
+This factorization makes possible the application of stochastic variational inference, thus allowing the model to handle large scale problems, in terms of both of the number of inputs and outputs.
+Also, learning of all of the  hyperparameters, including the inducing inputs, can be done in a consistent stochastic optimization framework.
+
+The effectiveness of the model is first verified on a toy problem where the inducing variables are shown to be conducive to the sharing of information between two related tasks. 
+Further experiments are conducted on two moderate-sized datasets in which the model shows better performance compared to other baselines.
+Our large scale experiment confirms the substantial benefits of collaborative learning brought about by the model, 
+thus opening up new opportunities for improving the performance over learning with single sparse process.
+
+% Experiments and summary of results
+The rest of the paper is organized as follows.
+We give description of the model in Section \ref{sec:model}, followed by the inference procedure in Section \ref{sec:inference}. 
+Section \ref{sec:experiments} presents empirical evaluation.
+We conclude with some discussion in Section \ref{sec:discussion}.
+
 %Gaussian processes (GPs) \citep{rasmussen-williams-book} are perhaps the prior of choice in nonparametric Bayesian regression not only in the standard setting but also in multioutput learning \citep{} .
@@
 %Taking insights from the seminal work of \citep{quinonero2005unifying}, most of these approximations can be formulated around the \emph{inducing points}.
-%These come from the same latent function that generates the data and can be viewed as representing its sufficient statistics \citet{titsias2009variational}.
+%These come from the same latent function that generates the data and can be viewed as representing its sufficient statistics \citep{titsias2009variational}.
 %Using these approaches, the complexity reduces to $\bigO(NB^2)$ in computation and $\bigO(NB)$ in memory where $B \ll N$ is the number of inducing points.

### 2014-03-18 paper/model.tex
add discussion and fixes references
add complexity analysis and edit various parts

@@
-Before diving into details of the specification, we discuss the philosophy and motivation behind our collaborative multioutput model.
+Before diving into details of the specification, we discuss the modeling philosophy behind our collaborative multioutput Gaussian processes.
 To learn the outputs jointly, we need a mechanism through which information can be transferred among the outputs.
-This is achieved in the model by allowing the outputs to share multiple sets of inducing variables.
-For exposition, consider a set of inducing points for a moment.
-Because they serve as the sufficient statistics of the individual outputs, a pattern common to the outputs can be encapsulated by these points when learning is done jointly. 
-Different patterns in the data can be captured by having  multiple set of inducing variables.
-It should be evident that these variables play a double pivotal role in the model: they represent the sufficient statistics of the sparse processes and they collaborate sharing of information across the outputs.
+This is achieved in the model by allowing the outputs to share multiple sets of inducing variables, each of which captures a different pattern common to the outputs.
+%For exposition, consider a set of inducing points for a moment.
+%Because they serve as the sufficient statistics of the individual outputs, a pattern common to the outputs can be encapsulated in these points when learning is done jointly. 
+%Different patterns in the data can be captured by having  multiple set of inducing variables.
+These variables indeed play a double pivotal role in the model: they collaborate sharing of information across the outputs and they provide the notion of  sufficient statistics that induces sparse processes.
 
@@
 \newcommand{\Zhi}{\Z^h_i}
-Consider multioutput regression of $P$ tasks with inputs $\X = \{\x_n \in \calR^D\}_{n=1}^N$ and outputs $\y = \{\y_i\}_{i=1}^P$ where $\y_i = \{y_{in}\}_{n=1}^N$. 
-Assume $Q$ latent functions with independent Gaussian process priors, $g_j(\x) \sim \GP(0, k_j(\cdot,\cdot))$, $j= 1 \hdots Q$.
-We introduce the sets of \emph{inducing variables} $\u_j$, each of which comes from $g_j(\x)$, i.e., $\u_j$ contains the function values at the inducing inputs $\Z_j$, which live in the same space as $\X$.
-These processes and inducing sets are shared by all of the outputs.
-To accommodate greater flexibility, we further allow each output to have its own independent process $h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$ whose the sets of inducing inputs and variables are $\Z^h_i$ and $\v_i$ respectively, where $i = 1 \hdots P$.
+Consider multiple output regression of $P$ tasks with inputs $\X = \{\x_n \in \calR^D\}_{n=1}^N$ and outputs $\y = \{\y_i\}_{i=1}^P$ where $\y_i = \{y_{in}\}_{n=1}^N$.
+We shall model each output as a weighted combination of $Q$ shared latent functions, plus an individual latent function unique to that output for greater flexibility.
+The $Q$ shared functions have independent Gaussian process priors, $g_j(\x) \sim \GP(0, k_j(\cdot,\cdot))$, $j= 1 \hdots Q$.
+As we wish to use sparse processes, we introduce the set of \emph{inducing variables} $\u_j$ for  each $g_j(\x)$, i.e., $\u_j$ contains the values of $g_j(\x)$ at the inducing inputs $\Z_j$.
+Similarly, each individual function of an output also has GP prior, i.e., $h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$ with inducing inputs $\Zhi$ and inducing values $\v_i$.
+The inducing inputs live in the same space as the inputs $\X$, which should be apparent since their corresponding values come from the processes defined on the input space.
+For convenience, we shall assume all of the processes have the same number of inducing points, $M$, however we emphasize that this is \emph{not} imposed in practice.
+
 % information is transfered via the inducing variables
-We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
-The subscripts are $i = 1 \hdots P$, $j = 1 \hdots Q$, and $n = 1 \hdots N$.
-For convenience, we shall assume the same number of inducing points, $M$, for all of the processes, however this is not imposed in practice.
 
-Following the definition of the GPs and the independence of the processes, the \emph{prior} of the multioutput model can be written as:
+As notations we denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$.
+Note that we reserve subscript $i$ for indexing the outputs and their corresponding individual processes ($i = 1 \hdots P$), $j$ for the latent processes ($j = 1 \hdots Q$), and $n$ for the inputs ($n = 1 \hdots N$).
+
+From the definition of the GPs and the independence of the processes, the \emph{prior} of the multioutput model can be written as:
 \begin{align}
@@
 p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
+\label{eq:u}
 p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
@@
 p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
-p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)), \\
+\label{eq:v}
+p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)),
+\end{align}
+where
+\begin{align}
  \BigMu_j &= k(\X,\Zj)k(\Zj,\Zj)^{-1}\u_j \\
@@
 \end{align}
-In the above equations and hereafter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.
+In the equations and hereafter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.
+Here, equations \ref{eq:u} and \ref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (eq.
+\ref{eq:gu} and \ref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
+Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the important role of the inducing variables in the model.
 
@@
 As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
-We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor models \citep{teh-et-al-aistats-05,seeger2005semiparametric}.
+We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor models \citep{teh-et-al-aistats-05}.
 However, doing so is against the purpose of this model which encourage sharing of outputs via the inducing variables.

### 2014-03-19 paper/intro.tex
first pass to intro

@@
-% paper on multioutput so this should appear first in the introduction
-Gaussian processes \citep[GPs,][]{rasmussen-williams-book} are perhaps the dominant approach for learning functions in non-parametric Bayesian regression.
-Unfortunately, the prohibitive complexity when performing exact inference in the GP framework severely hinders its application to real world problems. 
-In particular, when learning $P$ multiple correlated functions over $N$ data points, the complexity is $\calO(N^3P^3)$ in computation and $\calO(N^2P^2)$ in memory.
-An example of a real-world application where multioutput regression may be useful is the tracking of  movements of a robot arm using 2 or more joint torques. 
-If one of the robot motors malfunctions and fails to record a torque, data collected from the other motors may be used to infer the missing torque values.
-However, taking 100 measurements per second already results in a total of over 40,000 data points per torque in just 7 minutes.
+% Gaussian processes are good for bayesian regression, Very flexible in multi-output settings, for e.g. bonilla, alvarez, wilson
+Gaussian process models  \citep[GPs,][]{rasmussen-williams-book}  are a popular choice in 
+Bayesian regression due to their ability to capture complex dependencies and non-linearities in data.
+In particular, when having multiple outputs or tasks they have proved effective in modeling 
+the dependencies between the tasks  and  they have been shown to outperform competitive baselines and 
+single output learners  \citep{bonilla-et-al-nips-08,teh-et-al-aistats-05,alvarez-lawrence-nips-08,wilson-et-al-icml-12}.
+% However they are costly
+However, the prohibitive cost of performing exact inference in  GP models severely hinders 
+their application to  large-scale multi-output problems. For example, na\"{i}ve inference 
+in a fully coupled Gaussian process model over $P$  outputs and $N$ data-points can have 
+a complexity of $\calO(N^3P^3)$  and $\calO(N^2P^2)$ in time and memory, respectively.
+
+% Motivating example is good
+A motivating  example of a large scale multi-output application 
+is the tracking of  movements of a robot arm using 2 or more joint torques. 
+If one of the robot motors malfunctions and fails to record a torque, data collected from the 
+other motors may be used to infer the missing torque values.
+However, taking 100 measurements per second already results in a total of over 40,000 data points 
+per torque in just 7 minutes.
 Clearly this problem is well beyond the capability of conventional multiple output GPs.
-Building scalable multioutput GPs model that can learn correlated tasks at the scale of the application just considered and much larger is thus the main focus of this paper.
+Building scalable multi-output GP models that can learn correlated tasks at the scale of 
+these types of problems  is thus the main focus of this paper.
 
-% traditional approach 
-The traditional approach in multiple GP regression has been to mix independent latent processes to generate correlated functions.
-The mixing can be a weighted linear combinations with fixed \citep{teh-et-al-aistats-05,bonilla-et-al-nips-08} or input-dependent \citep{wilson-et-al-icml-12,nguyen2013efficient} coefficients or convolution of processes \citep{boyle-frean-nips-05,alvarez-lawrence-nips-08}.
-The correlations of the outputs are induced through sharing of these base processes.
-% sparse approximation technique and limitatio
-Sparse approximate techniques such as the informative vector machine (IVM) \cite{lawrence2002fast} or low-rank approximations based around the inducing points framework \citep{quinonero2005unifying},
-are then employed to reduce the computational and storage demands of the full models.
-Nevertheless, most of these models have complexity of at least $\calO(PNM^2)$ in computation and $\calO(PNM)$ in memory, where $M$ is the number of inducing points.
-thus prevents their applications to the  large scale problems we are targeting.
+% Tranditional approaches in single output setting, based e..g. on inducing points are not good enough 
+In the single output setting most approximation techniques for GP inference can be understood 
+within a single probabilistic framework that uses a set of \emph{inducing points}  in order to 
+obtain an approximate process (or a low-rank approximate covariance) over which inference can be 
+performed more efficiently \citep{quinonero2005unifying}. These models have been referred 
+to in the literature as sparse models. 
+Nevertheless, straightforward application of such approximate techniques will yield a computational 
+cost of at least $\calO(PNM^2)$ in time and  $\calO(PNM)$ in memory, where 
+$M$ is the number of inducing points. This high complexity still prevent us from applying GP models 
+to the large scale multi-output problems we are targeting.
 
-% we approach from different perspective 
-In this work, we approach the challenge of building scalable multioutput Gaussian processes based on several observations.
-First, \emph{sparsity} is an inevitable assumption in very large scale GPs, as evident by the vast amount of research on sparse approximations in the single output setting. % can cite ref here 
-Second, the \emph{inducing variables} are the key catalyst for effective and scalable learning under sparsity.
-In particular, they capture the \emph{sufficient statistics} of a dataset and lead to optimization of sparse processes that converge to the exact GP model \citep{titsias2009variational}.
-This is further exploited in \cite{hensmangaussian} where by explicitly representing a distribution over the inducing variables, stochastic variational inference can be used to work with millions of data points.
+% Useful observations
+In this work we approach the challenge of building large scale multi-output Gaussian process models  
+based on the following observations.
+Firstly, \emph{inducing variables} are the key catalyst
+for achieving sparsity and dealing with large-scale problems in Gaussian process models.
+In particular, they capture the \emph{sufficient statistics} of a dataset allowing the construction of sparse 
+processes that can approximate arbitrarily well the exact GP model \citep{titsias2009variational}.
+Secondly, the use of global latent variables (such as the inducing points) allows us to 
+induce dependencies in a highly correlated model efficiently. 
+This observation is exploited in \cite{hensmangaussian} for single output GP regression models 
+where by explicitly representing a distribution over the inducing variables, stochastic variational inference 
+can be used to work with millions of data points.
+Finally, the key to multi-output and multi-task learning is to model dependencies between the outputs  
+based on realistic assumptions of what can be shared across the tasks. It turns out that is 
+reasonable to assume that multiple tasks can share some sparsity structure.
 
-% can we do joint learning of multiple output under the sparsity assumption (meantioned above) for 
-Motivated by these observations, we propose the collaborative multioutput GPs (COGP) model where each output is a weighted combination of sparse processes. 
-Accompanying each process is its own set of inducing points that sufficiently encapsulates the process signals and characteristics.
-Correlations among the outputs are fostered via sharing of multiple inducing sets which in turn create a medium through which joint inference happens.
+% Our approach starts from mixing latent process
+% having inducing points for sparsity
+% sharing inducing points across the latent process
+Based on these observations, we propose the collaborative multi-output Gaussian Process (COGP) model 
+where each output is a weighted combination of sparse latent processes.
+ Each process has its own set of inducing points that sufficiently encapsulates the process signal and characteristics.
+Correlations among the outputs are fostered via sharing of multiple inducing point sets 
+which in turn create a medium through which joint inference happens efficiently.
+% exploit stochastic variational inference
+In particular, we derive a variational lower bound of the model evidence that decomposes across 
+inputs and outputs. 
+This decomposition makes possible the application of stochastic variational inference, 
+thus allowing the model to handle a large number of observations per output and a large number of outputs.
+Furthermore, learning of all of the  parameters in the model, including
+kernel hyperparameters and inducing inputs, can be done in a consistent stochastic optimization framework.
 
-% contribution from inference for large scale
-This formulation around inducing points and sparse processes enables scalable inference of the model.
-In particular, we derive a variational lower bound of the model evidence that factorizes across not only the outputs but also the inputs.
-This factorization makes possible the application of stochastic variational inference, thus allowing the model to handle large scale problems, in terms of both of the number of inputs and outputs.
-Also, learning of all of the  hyperparameters, including the inducing inputs, can be done in a consistent stochastic optimization framework.
-
-The effectiveness of the model is first verified on a toy problem where the inducing variables are shown to be conducive to the sharing of information between two related tasks. 
-Further experiments are conducted on two moderate-sized datasets in which the model shows better performance compared to other baselines.
-Our large scale experiment confirms the substantial benefits of collaborative learning brought about by the model, 
-thus opening up new opportunities for improving the performance over learning with single sparse process.
+We analyze our multi-out model on a toy problem  where the inducing variables are shown to be 
+conducive to the sharing of information between two related tasks. 
+Additionally, we  evaluate  our model on two moderate-sized datasets in which we show that it
+can outperform previous non-scalable multi-output approaches as well as single output baselines.
+Finally, on a  large scale experiment regarding the learning of robot inverse dynamics we show 
+the substantial benefits of collaborative learning provided by our model.
 
@@
 
-%Gaussian processes (GPs) \citep{rasmussen-williams-book} are perhaps the prior of choice in nonparametric Bayesian regression not only in the standard setting but also in multioutput learning \citep{} .
-% it is beauty but also a beast
-%Unfortunately, given a dataset of size $N$, exact inference for GP regression requires $\calO(N^3)$ for computation and $\calO(N^2)$ for memory, which can only be applied to problems in non-realistic settings.
-% cite titsias
-%Consequently, various approximate methods \citep{lawrence2002fast,seeger2003bayesian,seeger2003fast, smola2001sparse,snelson2006sparse,williams2001using} have been proposed to overcome this limitation.
-%Taking insights from the seminal work of \citep{quinonero2005unifying}, most of these approximations can be formulated around the \emph{inducing points}.
-%These come from the same latent function that generates the data and can be viewed as representing its sufficient statistics \citep{titsias2009variational}.
-%Using these approaches, the complexity reduces to $\bigO(NB^2)$ in computation and $\bigO(NB)$ in memory where $B \ll N$ is the number of inducing points.
-%Recent development in stochastic variational inference for GPs \citep[GPSVI,][]{hensmangaussian} has led to further reduction of these costs to $\bigO(B^3)$ and $\bigO(B^2)$ respectively.
-
-% While the important advance by GPSVI has made possible the application of GPs to datasets of millions of data points, the assumption that the statistics of the data is sufficiently captured by a, typically small, set of inducing variables might be too strong and limiting.
-%This is addressed in \citep{nguyen2014fast} by using local sets of inducing points, as opposed to a global set, such that each local inducing sets sufficiently support a different local region of the input space.
-%Yet another approach to this challenge is multi-task / multi-output learning, which is the focus of this paper.
-%Specifically we pose the question: \emph{under the sparsity constraint of individual outputs}, can multiple output learning be effective?
-%We believe this to be an interesting and  important research problem toward scalable multioutput GP models.
-%Here we emphasize the word \emph{sparsity}, as it has been evident from the vast literature on GP approximations that  sparsity is the impeccable assumption to deal with large scale datasets.
-%When learning things together, the hope is that information from different tasks can be used collaboratively to compensate for the sparsity enforcement to the individual outputs.
+% Some citations that can be brought back somewhere
+%The traditional approach in multiple GP regression has been to mix independent latent processes to generate correlated functions.
+%The mixing can be a weighted linear combinations with fixed \citep{teh-et-al-aistats-05,bonilla-et-al-nips-08} or input-dependent \citep{wilson-et-al-icml-12,nguyen2013efficient} coefficients or convolution of processes \citep{boyle-frean-nips-05,alvarez-lawrence-nips-08}.
+%The correlations of the outputs are induced through sharing of these base processes.

### 2014-03-19 paper/intro.tex
mods to intro

@@
 Clearly this problem is well beyond the capability of conventional multiple output GPs.
-Building scalable multi-output GP models that can learn correlated tasks at the scale of 
+Building  multi-output GP models that can learn correlated tasks at the scale of 
 these types of problems  is thus the main focus of this paper.
@@
 % Tranditional approaches in single output setting, based e..g. on inducing points are not good enough 
-In the single output setting most approximation techniques for GP inference can be understood 
+In the single output setting previous attempts to scale up  GP inference 
+resort to approximate inference. Most  approximation techniques 
+ can be understood 
 within a single probabilistic framework that uses a set of \emph{inducing points}  in order to 
@@
 This observation is exploited in \cite{hensmangaussian} for single output GP regression models 
-where by explicitly representing a distribution over the inducing variables, stochastic variational inference 
+where, by explicitly representing a distribution over the inducing variables, stochastic variational inference 
 can be used to work with millions of data points.
 Finally, the key to multi-output and multi-task learning is to model dependencies between the outputs  
-based on realistic assumptions of what can be shared across the tasks. It turns out that is 
-reasonable to assume that multiple tasks can share some sparsity structure.
+based on realistic assumptions of what can be shared across the tasks. It turns out that 
+sharing  ``sparsity structure'' can not only be a reasonable assumption but also a crucial component
+when modeling dependencies between different related tasks.
 
@@
 where each output is a weighted combination of sparse latent processes.
- Each process has its own set of inducing points that sufficiently encapsulates the process signal and characteristics.
+ Each latent process has its own set of inducing points that sufficiently encapsulates the process signal and characteristics.
 Correlations among the outputs are fostered via sharing of multiple inducing point sets 
@@
 thus allowing the model to handle a large number of observations per output and a large number of outputs.
-Furthermore, learning of all of the  parameters in the model, including
+Furthermore, learning of all  the  parameters in the model, including
 kernel hyperparameters and inducing inputs, can be done in a consistent stochastic optimization framework.

### 2014-03-19 paper/supplemental.tex
update supplemental

@@
 \section{Gaussian Identities}
-Let $\y$, $\g = \{\g_j\}_{j=1}^q$, and $\h$ be random variables with multivariate Gaussian distributions: 
+Let $\g = \{\g_j\}_{j=1}^q$, $\h$, and $\y$ be random variables with multivariate Gaussian distributions: 
 $p(\y | \g, \h) = \Normal(\y; \sum_{j=1}^Q \W_j \g_j + \W \h, \beta^{-1} \I)$, $p(\g_j) = \Normal(\g_j; \m_j, \S_j)$, and $p(\h) = \Normal(\h; \m, \S)$.
@@
 
-\section{Expectation of an Individual Likelihood Term}
-Applying the identity in eq. \ref{eq:identity}, the expectation of an individual likelihood term with respect to the posterior distribution is given by:
+\section{Derivation of the Variational Lower Bound}
+The variational lower bound of the log marginal (eq. 13 in the main text) is given by:
+\begin{align}
+\nonumber
+\log p(\y) \ge& \int q(\u, \v) \log \frac{p(\y | \u, \v) p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
+\nonumber
+%=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v  \\ 
+%\nonumber
+&+ \int q(\u, \v) \log \frac{p(\u, \v)}{q(\u, \v)} \der \u \der \v \\
+\nonumber
+=& \int q(\u, \v) \log p(\y | \u, \v)  \der \u \der \v \\
+\nonumber
+&- \sum_{j=1}^Q \KL[q(\u_j) || p(\u_j)] - \sum_{i=1}^P \KL[q(\v_i) || p(\v_i)],
+\end{align}
+
+Since the KL terms are analytically tractable, we compute the first term in the above sum. 
+This is done first by deriving a lower bound to the likelihood $p(\y | \u, \v)$ (eq. 14 in the main text).
+\begin{align}
+\nonumber
+\log \text{ } p(\y | \u, \v)
+&= \log \Eb{p(\y | \g, \h)}_{p(\g,\h | \u, \v)} \\
+\nonumber
+&\ge \Eb{\log p(\y | \g, \h)}_{p(\g,\h | \u, \v)}  \\
+\nonumber
+&= \sum_{i=1}^P \sum_{n=1}^N \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g | \u) p(\h_i | \v_i)} 
+\end{align}
+Applying the identity in eq. \ref{eq:identity}, 
+the expectation of an individual likelihood term with respect to the posterior distribution is given by: 
 \begin{align}
@@
 =& \log \Normal(y_{in}; \sum_{j=1}^Q w_{ij} \mu_{jn}+ \mu^h_{in}, \beta_i^{-1}) \\
-&- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{nn} 
+&- \frac{1}{2} \beta_i \sum_{j=1}^Q w_{ij}^2 \tilde{k}_{jnn} 
 - \frac{1}{2} \beta_i \tilde{k}^h_{inn},
-\label{eq:lin}
 \end{align}
-where $\tilde{k}_{nn} = (\tilde{\K})_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$, and $\mu^h_{in} = (\Mu^h_i)_n$.
+where $\tilde{k}_{jnn} = (\tilde{\K}_j)_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$, and $\mu^h_{in} = (\Mu^h_i)_n$.
+
+Substituting $l_{in}$ into the expression for the lower bound of $\log p(\y | \u, \v)$ (again, this is eq.
+14 in the main text), and applying the identity in eq.
+\ref{eq:identity} to carry out the integral we obtain the lower bound as given in the main text.
 
@@
 \section{Derivatives of the ELBO}
-For convenience, we derive the gradients of the lower bound for the case of a single GP (i.e. the bound in \citet{hensmangaussian}).
+For exposition, we derive the gradients of the lower bound for the case of a single GP (i.e. the bound in \citet{hensmangaussian}).
 The derivatives of the ELBO of the collaborative multioutput GPs model are typically linear combination of the derivatives here.
@@
 \deriv{\calL_1}{t} =& \beta (\y - \A\m)^T \deriv{\A}{t} \m \\
-\deriv{\calL_2}{t} =& \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right) \\
+\deriv{\calL_2}{t} =& \frac{1}{2}\beta \trace \left(\deriv{\K_{NN}}{t} - \A \deriv{\K_{MN}}{t} - \deriv{\A}{t} \K_{MN}\right)
+\end{align}
+\begin{align}
 \deriv{\calL_3}{t} =& \beta \trace \left(\A \S \deriv{\A^T}{t} \right) \\
@@
 Notice that $\D_1$ and $\D_2$ can be pre-computed with a cost of $\calO(M^3)$ (or $\calO(N_bM^2)$ if the minibatch size $N_b > M$).
-The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two $\trace$ operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
-This fact can be further exploited to perform vectorized operation, for e.g. in Matlab, such that the iteration over all inducing inputs can be avoided.
+The computational cost of taking derivatives of $MD$ inducing parameters is thus $\calO(M^3 + MDM) = \calO(M^3)$ as the cost of the two trace operators is $\calO(M)$ due to the fact that only $\calO(M)$ elements of $\deriv{\K_{MM}}{t}$ or $\deriv{\K_{NM}}{t}$ are non-zero.
+
+For implementation with MATLAB, a loop over $M \times D$ parameters of the inducing inputs can be very slow for even moderate values of $M$ and $D$.
+The aforementioned fact about $\deriv{\K_{MM}}{t}$ and $\deriv{\K_{NM}}{t}$ can be used to perform vectorized operations that compute the derivatives of all $M$ parameters given a specific dimension at a cost of $\calO(M^2)$.
+The loop is the executed over the input dimension $D$, leading to a complexity of still $\calO(DM^2)$ only.
 

### 2014-03-19 paper/abstract.tex
first pass to abstract

@@
-% limitation of learning with multiple GPs
-Learning of correlated functions in the popular Gaussian processes (GPs) framework is often performed by mixing latent processes.
-However, when these constituent processes are standard GPs, the applicability of this approach is severely inhibited by their intensive computational and memory demands.
-% our approach
-To address this problem, we propose the collaborative multioutput GPs for multitask regression at very large scale.
-The building blocks of the model are \emph{sparse} processes underpinned by the \emph{inducing variables}.
-These variables captures the sufficient statistics of a pattern (i.e. sparse process) in the data, and dependencies of the outputs are fostered via sharing different sets of inducing points.
-This formulation enables extremely scalable inference of the model via stochastic variational inference.
-Furthermore, all of the model hyperparameters including the inducing inputs can be optimized in the same stochastic approach.
-Experiments on a toy problem and three real-world datasets verify the effectiveness of collaborative learning of sparse processes via the inducing variables promoted by the model.
-They also demonstrate the computational superiority and scalability of the model compared to conventional multiple GPs regression.
+We propose the collaborative multi-output Gaussian process (GP)  model 
+where each output is a weighted combination of sparse latent processes.
+ Each latent process has its own set of inducing points that sufficiently encapsulates the process signal and characteristics.
+Correlations among the outputs are fostered via sharing of multiple inducing point sets, 
+which facilitate  the  application of variational inference and the derivation of 
+an evidence lower bound  that decomposes across inputs and outputs. 
+We learn all the parameters of the model from data in a  single stochastic optimization framework
+that scales to a large number of observations per output and a large number of outputs.
+%
+We analyze our multi-output model on a toy problem  and evaluate its performance on 
+two medium-sized datasets and a large dataset showing the benefits of collaborative learning 
+and the superior performance of our approach compared to single output learning and previous 
+multi-output GP models.
+
+%% limitation of learning with multiple GPs
+%Learning of correlated functions in the popular Gaussian processes (GPs) framework is often performed by mixing latent processes.
+%However, when these constituent processes are standard GPs, the applicability of this approach is severely inhibited by their intensive computational and memory demands.
+%% our approach
+%To address this problem, we propose the collaborative multioutput GPs for multitask regression at very large scale.
+%The building blocks of the model are \emph{sparse} processes underpinned by the \emph{inducing variables}.
+%These variables captures the sufficient statistics of a pattern (i.e. sparse process) in the data, and dependencies of the outputs are fostered via sharing different sets of inducing points.
+%This formulation enables extremely scalable inference of the model via stochastic variational inference.
+%Furthermore, all of the model hyperparameters including the inducing inputs can be optimized in the same stochastic approach.
+%Experiments on a toy problem and three real-world datasets verify the effectiveness of collaborative learning of sparse processes via the inducing variables promoted by the model.
+%They also demonstrate the computational superiority and scalability of the model compared to conventional multiple GPs regression.

### 2014-03-19 paper/model.tex
model done

@@
-Before diving into details of the specification, we discuss the modeling philosophy behind our collaborative multioutput Gaussian processes.
+Before diving into technical details of the model specification, 
+we discuss the modeling philosophy behind our collaborative multi-output Gaussian processes.
 To learn the outputs jointly, we need a mechanism through which information can be transferred among the outputs.
-This is achieved in the model by allowing the outputs to share multiple sets of inducing variables, each of which captures a different pattern common to the outputs.
-%For exposition, consider a set of inducing points for a moment.
-%Because they serve as the sufficient statistics of the individual outputs, a pattern common to the outputs can be encapsulated in these points when learning is done jointly. 
-%Different patterns in the data can be captured by having  multiple set of inducing variables.
-These variables indeed play a double pivotal role in the model: they collaborate sharing of information across the outputs and they provide the notion of  sufficient statistics that induces sparse processes.
+This is achieved in the model by allowing the outputs to share multiple sets of inducing variables, 
+each of which captures a different pattern common to the outputs.
+These variables  play a double pivotal role in the model: they collaboratively share  information 
+across the outputs and  provide \emph{sufficient statistics}  so as to induce sparse processes.
 
@@
 \newcommand{\Zhi}{\Z^h_i}
-Consider multiple output regression of $P$ tasks with inputs $\X = \{\x_n \in \calR^D\}_{n=1}^N$ and outputs $\y = \{\y_i\}_{i=1}^P$ where $\y_i = \{y_{in}\}_{n=1}^N$.
-We shall model each output as a weighted combination of $Q$ shared latent functions, plus an individual latent function unique to that output for greater flexibility.
-The $Q$ shared functions have independent Gaussian process priors, $g_j(\x) \sim \GP(0, k_j(\cdot,\cdot))$, $j= 1 \hdots Q$.
-As we wish to use sparse processes, we introduce the set of \emph{inducing variables} $\u_j$ for  each $g_j(\x)$, i.e., $\u_j$ contains the values of $g_j(\x)$ at the inducing inputs $\Z_j$.
-Similarly, each individual function of an output also has GP prior, i.e., $h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$ with inducing inputs $\Zhi$ and inducing values $\v_i$.
+Consider the joint regression of $P$ tasks with inputs $\X = \{\x_n \in \calR^D\}_{n=1}^N$ and outputs $\y = \{\y_i\}_{i=1}^P$ where $\y_i = \{y_{in}\}_{n=1}^N$.
+We  model each output as a weighted combination of $Q$ shared latent functions $\{ g_{j}\}_{j=1}^{Q}$, plus an individual latent 
+function $\{ h_{i}\}_{i=1}^{P}$ unique to that output for greater flexibility.
+The $Q$ shared functions have independent Gaussian process priors  $g_j(\x) \sim \GP(0, k_j(\cdot,\cdot))$. %, with $j= 1, \ldots Q$.
+Similarly, each individual function of an output also has GP prior, i.e.~$h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$. %with $i=1 ,\ldots P$.
+
+As we wish to sparsify these processes for scalability, 
+we introduce the set of \emph{shared inducing variables} $\u_j$ for  each $g_j(\x)$, i.e. $\u_j$ contains the values of 
+$g_j(\x)$ at the inducing inputs $\Z_j$. Likewise, we have 
+individual inducing variables corresponding to each  $h_i(\x)$,  which we denote with with $\v_i$ and their 
+corresponding inducing inputs $\Zhi$.
+%
 The inducing inputs live in the same space as the inputs $\X$, which should be apparent since their corresponding values come from the processes defined on the input space.
-For convenience, we shall assume all of the processes have the same number of inducing points, $M$, however we emphasize that this is \emph{not} imposed in practice.
+For convenience, we shall assume all processes have the same number of inducing points, $M$, however we emphasize that this is not
+ imposed in practice.
 
@@
 
-As notations we denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$.
+We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
 Note that we reserve subscript $i$ for indexing the outputs and their corresponding individual processes ($i = 1 \hdots P$), $j$ for the latent processes ($j = 1 \hdots Q$), and $n$ for the inputs ($n = 1 \hdots N$).
-
-From the definition of the GPs and the independence of the processes, the \emph{prior} of the multioutput model can be written as:
+%
+\subsection{Prior Model}
+From the definition of the GPs and the conditional independence of the processes given the inducing variables, 
+the \emph{prior} of the multi-output model can be written as:
 \begin{align}
@@
 Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the important role of the inducing variables in the model.
-
-The \emph{likelihood} with standard iid Gaussian likelihood is given by:
+%
+\subsection{Likelihood Model}
+As mentioned above, we assume that  observations for each output 
+are  (noisy) linear combinations of the $Q$ latent functions $g_{j}(\vec{x})$ plus
+an independent function $h_{i}(\vec{x})$. Hence we have that the likelihood 
+with standard iid Gaussian noise is given by:
 \begin{align}
@@
 \end{align}
-which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
+where $w_{ij}$ are the corresponding weights and $\beta_i$ is the precision of each Gaussian.
+%which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
 As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
-We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor models \citep{teh-et-al-aistats-05}.
-However, doing so is against the purpose of this model which encourage sharing of outputs via the inducing variables.
-Furthermore, as we shall see in the next section, explicit representation of these variables is instrumental to scalable inference of the model.
+We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor model \citep{teh-et-al-aistats-05}.
+However, doing so is against the purpose of our model which encourages sharing of outputs via the inducing variables.
+Furthermore, as we shall see in the next section, explicit representation of these variables is fundamental to scalable inference of the model.
 

### 2014-03-19 paper/inference.tex
3.1 done

@@
 %
-
-Inference in this model is done with variational inference.
-In section \ref{sec:variationalLowerBound}, we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points.
-The next section takes advantage of this factorization to derive stochastic variational inference which allows the model to scale to massive datasets.
+We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ across
+multiple outputs using variational inference.
+In section \ref{sec:variationalLowerBound} we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points and outputs.
+Section \ref{sec:svi}  takes advantage of this factorization to
+carry out stochastic variational inference which allows the model to scale to very large datasets.
 Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.
@@
 Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
-Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as,
+Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as:
 \begin{align}
@@
 Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
-This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution that of the form
+This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution that of the form:
 \begin{align}
@@
 \end{align}
-This technique has been used in \citep{titsias2009variational,hensmangaussian} to derive variational inference for the single output case.
-Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are given (eq.
-\ref{eq:gu} and \ref{eq:hv}), inference requires only learning of $q(\u, \v)$ that minimize the divergence between the approximate and true posteriors.
+This technique has been used by \citet{titsias2009variational} and \citet{hensmangaussian} 
+to derive variational inference algorithms for the single output case.
+Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \ref{eq:gu} and \ref{eq:hv}), 
+we only need to 
+learn $q(\u, \v)$ so as to minimize the divergence between the approximate posterior and
+the true posteriors.
 This underlines the pivotal role of the inducing variables in this model as previously discussed.
@@
 \newcommand{\Zi}{\Z_i}
+
 The ELBO can be computed by first solving for the individual expectations 
-$l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ (see the supplementary material), which are then substituted into eq. \ref{eq:elbo} to give:
+%$l_{in} \define \Eb{\log p(y_{in} | \g_n, h_{in}) }_{p(\g_n | \u) p(\h_in | \v_i)}$ 
+$ \Eb{\log p(y_{in} | \g_n, h_{in}) }$ over ${p(\g | \u) p(\h_i | \v_i)}$ and 
+then substituting these into Equation \ref{eq:elbo}. (Please see 
+the supplementary material for the details of the derivation.)  
+Hence the resulting lower bound is given by:
 \begin{align}
 \nonumber
-&\log p(\y)
-\ge \sum_{i,n}
+%&\log p(\y)
+ & \calL = 
+\sum_{i,n}
 \bigg( \log  \Normal(y_{in}; \tilde{\mu}_{in}, \beta_i^{-1})
@@
 \nonumber
-& - \sum_{j=1}^Q \bigg( \frac{1}{2} \log |\K_{jzz} \S_j^{-1}| + \frac{1}{2} \trace \K_{jzz}^{-1}  (\m_j \m_j^T + \S_j)\bigg) \\
+& - \sum_{j=1}^Q \bigg( \frac{1}{2} \log |\K_{jzz} \S_j^{-1}| + \frac{1}{2} \trace \K_{jzz}^{-1}  \left( \m_j \m_j^T + \S_j \right)\bigg) \\
 \nonumber
-& - \sum_{i=1}^P  \bigg( \frac{1}{2} \log |\K_{izz}(\Shi)^{-1}| + \frac{1}{2} \trace \K_{izz}^{-1} (\mhi (\mhi)^T + \Shi) \bigg) \\
-  &\define \calL,
+& - \sum_{i=1}^P  \bigg( \frac{1}{2} \log |\K_{izz}(\Shi)^{-1}| \\
+& \quad \qquad + \frac{1}{2} \trace \K_{izz}^{-1} \left( \mhi (\mhi)^T   +   \Shi \right) \bigg)  \text{,}
   \label{eq:L}
 \end{align}
-where $\K_{jzz} = k(\Zj, \Zj)$, $\K_{izz} = k(\Zhi, \Zhi)$, and with the help of auxiliary matrices $\A_j = k(\X,\Zj) \K_{jzz}^{-1}$ and $\Ahi = k(\X_i,\Zhi)\K_{izz}^{-1}$, 
+%
+where $\K_{jzz} = k(\Zj, \Zj)$, $\K_{izz} = k(\Zhi, \Zhi)$, and:
 \begin{align}
@@
 %&= \sum_{j=1}^Q w_{ij} k(\x_n, \Zj)k(\Zj,\Zj)^{-1}\m_j + k(\x_n, \Zhi)k(\Zhi,\Zhi)^{-1}\mhi \\
-&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \\
+&= \sum_{j=1}^Q w_{ij} \A_j(n,:) \m_j + \Ahi(n,:) \mhi \text{,}\\
 \mat{\Lambda}_{jn}
 %&= k(\Zj,\Zj)^{-1} k(\Zj, \x_n) k(\x_n, \Zj) k(\Zj,\Zj)^{-1} \\
-&= \A_j(n,:)^T \A_j(n,:)  \\
+&= \A_j(n,:)^T \A_j(n,:)  \text{,}\\
 \mat{\Lambda}_{in}
 %&= k(\Zhi,\Zhi)^{-1} k(\Zhi, \x_n) k(\x_n, \Zhi) k(\Zhi,\Zhi)^{-1}
-&= \Ahi(n,:)^T \Ahi(n,:) 
+&= \Ahi(n,:)^T \Ahi(n,:)  \text{,}
 \end{align}
-where $\tilde{k}_{jnn} = (\tilde{\K}_j)_{nn}$, $\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$, $\mu_{jn} = (\Mu_j)_n$,  $\mu^h_{in} = (\Mu^h_i)_n$, and $\A_j(n,:)$ selects the $n$-th row vector of $\A_j$.
-Notice that this ELBO clearly generalizes the bound for standard GP regression derived in \citep{hensmangaussian}, which can be recovered by setting $P = Q = 1$, $w_{ij} = 1$ and $h_i(\x) = 0$.
+with $\tilde{k}_{jnn} = (\tilde{\K}_j)_{nn}$; 
+$\tilde{k}^h_{inn} = (\tilde{\K}^h_i)_{nn}$; 
+$\mu_{jn} = (\Mu_j)_n$;
+ $\mu^h_{in} = (\Mu^h_i)_n$; 
+and we have defined the auxiliary matrices $\A_j = k(\X,\Zj) \K_{jzz}^{-1}$ and $\Ahi = k(\X_i,\Zhi)\K_{izz}^{-1}$ and 
+used $\A_j(n,:)$ to denote the $n$-th row vector of $\A_j$. 
+Notice that this ELBO clearly generalizes the bound for standard GP regression derived in \cite{hensmangaussian}, which can be recovered by setting $P = Q = 1$, $w_{ij} = 1$ and $h_i(\x) = 0$.
 
-The novelty of the derived variational lower bound is that it decomposes across not only the outputs but also the inputs.
-This enables the use of stochastic optimization methods which makes this model applicable to much larger datasets compared to the existing GP-based multioutput models.
-
-\subsection{STOCHASTIC VARIATIONAL INFERENCE}
+The novelty of the  variational lower bound in Equation  \eqref{eq:L} is that it decomposes across both inputs and outputs.
+This enables the use of stochastic optimization methods, which 
+allow the model to handle  ver large datasets for which existing GP-based multi-output models are simply 
+impractical.  
+%
+\subsection{STOCHASTIC VARIATIONAL INFERENCE \label{sec:svi}}
 \newcommand{\oi}{\vec{o}_i}

### 2014-03-19 paper/inference.tex
3.2 finished

@@
 
-So far in the description of the model and inference, it is implicitly assumed that every output has full observations at all of the inputs $\X$.
+So far in the description of the model and inference method we have
+ implicitly assumed that every output has full observations at all inputs $\X$.
 To discern where learning occurs for each output, we make the missing data scenario more explicit.
 Specifically, each output $i$ can have observations at a different set of inputs $\X_i$. 
-We shall use $\oi$ to denote the indice of $\X_i$ (in the set $\X$) and use the indexing operator $\B(\oi)$ to select the rows corresponding to $\oi$ from any arbitrary matrix $\B$.
+We shall use $\oi$ to denote the indices of $\X_i$ (in the set $\X$) and use the indexing operator $\B(\oi)$ to select the rows corresponding to $\oi$ from any arbitrary matrix $\B$.
 We also overload $\y_i$ as the observed targets of output $i$. 
-
+%
+\subsubsection{Learning the Parameters of the Variational Distribution}
 %------------------------------------------
@@
 \newcommand{\Kjzz}{\mat{K}_{jzz}}
-From (eq. \ref{eq:L}), we can obtain the derivatives of the ELBO wrt the variational parameters for optimization.
+ We can obtain the derivatives of the ELBO in Equation  \eqref{eq:L}
+  wrt the variational parameters for optimization.
 The derivatives of $\calL$ wrt the parameters of $q(\u_j)$ are given by:
@@
 \begin{align}
-\Phi_{1(t+1)} &= \S_{j(t)}^{-1} \m_{j(t)} + l \bigg(\sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj - \S_{j(t)}^{-1} \m_{j(t)} \bigg) \\
-\Phi_{2(t+1)} &= -\frac{1}{2} \S_{j(t)}^{-1} + l \bigg(\frac{1}{2} \S_{j(t)}^{-1} - \frac{1}{2} \mat{\Lambda} \bigg),
+\nonumber
+& \Phi_{1(t+1)} = \S_{j(t)}^{-1} \m_{j(t)}  \\
+ & \quad \qquad +  l \bigg(\sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj - \S_{j(t)}^{-1} \m_{j(t)} \bigg) \\
+& \Phi_{2(t+1)} = -\frac{1}{2} \S_{j(t)}^{-1} + l \bigg(\frac{1}{2} \S_{j(t)}^{-1} - \frac{1}{2} \mat{\Lambda} \bigg),
 \end{align}
@@
 where $\mat{\Lambda} = \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi)$.
-
-%TODO table for complexity
-\subsection{INDUCING INPUTS AND HYPER-PARAMETERS \label{sec:hyperparameters}}
+%
+\subsubsection{Inducing Inputs and Hyper-parameters \label{sec:hyperparameters}}
 To learn the hyperparameters, which in this model include the mixing weights, the covariance hyperparameters of the latent processes, and the noise precision of each output, we follow standard practice in GP inference.
-For this model, this involves taking derivatives of the ELBO and applying standard stochastic gradient descent in alternative steps with the variational parameters, much like a variational EM algorithm.
+For this model this involves taking derivatives of the ELBO and applying standard stochastic gradient descent in alternative steps with the variational parameters, much like a variational EM algorithm.
 The derivatives are given in the supplementary material.
 
-Learning of the inducing inputs, which was not considered in the single output case in \citep{hensmangaussian}, is also possible in our stochastic optimization approach.
-In the supplementary material, we show that with the help of vectorized operations, the derivatives of the lower bound w.r.t the inducing inputs can be computed with a cost \emph{independent} of the input dimension.
-This makes optimizing the inducing locations a practical option, which can be important for e.g. in high-dimensional problems where fixing the inducing inputs may hinder the performance.
-Indeed, our experiments on a large scale multioutput problem shows that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation.
-
+Learning of the inducing inputs, which was not considered in the single output case in \citet{hensmangaussian}, is also possible in our stochastic optimization approach.
+In the supplementary material, we show that with the help of vectorized operations, the derivatives of the lower bound wrt the inducing inputs can be computed with a cost \emph{independent of the input dimensionality}.
+This makes optimizing the inducing locations a practical option that can be important, for example, in high-dimensional problems where fixing the inducing inputs may hinder the performance of the model.
+Indeed, our experiments on a large scale multi-output problem show that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation. 
+%
 \subsection{COMPLEXITY ANALYSIS}

### 2014-03-19 paper/model.tex
minor mods to model
3.2 finished

@@
 %
-\subsection{Prior Model}
+\subsection{PRIOR MODEL}
 From the definition of the GPs and the conditional independence of the processes given the inducing variables, 
@@
 \end{align}
-where
+where the corresponding means and covariances of the Gaussians are given by:
 \begin{align}
@@
 %
-\subsection{Likelihood Model}
+\subsection{LIKELIHOOD MODEL}
 As mentioned above, we assume that  observations for each output 

### 2014-03-19 paper/inference.tex
section 3 finished

@@
 \subsection{COMPLEXITY ANALYSIS}
-In this section we analyze the complexity of the model and compare it to existing multioutput approaches.
+In this section we analyze the complexity of the model and compare it to existing multi-output approaches.
 For consistency, we first unify common notations used for all models.
@@
 
-The complexity of COGP can be read off by inspecting the ELBO (eq. \ref{eq:L}), with a key observation that it contains a sum over the outputs as well the inputs.
+The complexity of COGP can be read off by inspecting the ELBO  in Equation \eqref{eq:L}, with a key observation that it contains 
+a sum over the outputs as well as over the inputs.
 This means a mini-batch containing a small subset of the inputs and outputs can be used for stochastic optimization.
 Technically, the cost is $\calO(M^3)$ or $\calO(N_bM^2)$, where $N_b$ is the size of the mini-batches, depending on which is larger between $M$ and $N_b$. 
-Though in practice we may use $N_b > M$ as a large batch size (e.g. $N_b = 1000$) helps reduce stochasticity of the optimization.
-However, here we use $\calO(M^3)$ here for easier comparison with existing models whose computational and storage demands are given in \ref{tab:complexity}. 
-As can be seen, COGP has superior complexity which is independent of the size of the inputs and outputs, thus is the only method capable of handling large scale problems.
-
+Though in practice we may use $N_b > M$ as a large batch size (e.g.~$N_b = 1000$) helps reduce stochasticity of the optimization.
+However, here we use $\calO(M^3)$  for easier comparison with existing models whose computational and storage demands are 
+given in Table \ref{tab:complexity}. 
+We see that COGP has 
+a computational  complexity that is independent of the size of the inputs and outputs, 
+thus is the only method capable of handling large scale problems.
+%
 \setlength{\tabcolsep}{4pt}
 \begin{table}[t]
-\caption{Comparison of the time and storage complexity of (approximate) inference of GP-based multioutput regression models.
-CGP-FITC is the convolved GPs model with sparse approximation\citep{alvarez-lawrence-nips-08}.
-COGP is the only method with complexity \emph{independent} of the number of inputs and outputs, thus can scale to massive datasets.}
+\caption{Comparison of the time and storage complexity of approximate inference of  multi-output GP regression models. 
+A\&W, 2009 refers to \citet{alvarez-lawrence-nips-08}. COGP is the only method with complexity \emph{independent} of the
+ number of inputs $N$ and outputs $P$, thus it can scale to very large datasets.}
 \label{tab:complexity}
@@
 COGP, this paper  & $\calO(M^3)$ & $\calO(M^2)$ \\
-SLFM, \citep{teh-et-al-aistats-05} & $\calO(QNM_t)$ & $\calO(QNM_t)$ \\
+SLFM, \citep{teh-et-al-aistats-05} & $\calO(QNM_t^{2})$ & $\calO(QNM_t)$ \\
 MTGP, \citep{bonilla-et-al-nips-08} & $\calO(PNM_t^2)$ & $\calO(PNM_t)$\\ 
-CGP-FITC & $\calO(PNM_t^2)$ & $\calO(PNM_t)$ \\
+CGP-FITC (A\&W, 2009)& $\calO(PNM_t^2)$ & $\calO(PNM_t)$ \\
 GPRN, \citep{wilson-et-al-icml-12} & $\calO(PQN^3)$ & $\calO(PQN^2)$ \\
@@
 \begin{align}
-\mu_{j*} &= \vec{k}_{j*z} \K_{jzz}^{-1} \m_j \\ s_{j*} &= k_{j**} - \vec{k}_{j*z} \left(\K_{jzz}^{-1} - \K_{jzz}^{-1} \S_j \K_{jzz}^{-1} \right) \vec{k}_{j*z}^T \\
-\mu^h_{i*} &= \vec{k}_{i*z} \K_{izz}^{-1} \mhi \\
-s^h_{i*} &= k_{i**} - \vec{k}_{i*z} \left(\K_{izz}^{-1} - \K_{izz}^{-1} \S_j \K_{izz}^{-1} \right) \vec{k}_{i*z}^T
+\mu_{j*} &= \vec{k}_{j*z} \K_{jzz}^{-1} \m_j  \text{,}\\ 
+s_{j*} &= k_{j**} - \vec{k}_{j*z} \left(\K_{jzz}^{-1} - \K_{jzz}^{-1} \S_j \K_{jzz}^{-1} \right) \vec{k}_{j*z}^T  \text{,}\\
+\mu^h_{i*} &= \vec{k}_{i*z} \K_{izz}^{-1} \mhi \text{,} \\
+s^h_{i*} &= k_{i**} - \vec{k}_{i*z} \left(\K_{izz}^{-1} - \K_{izz}^{-1} \S_j \K_{izz}^{-1} \right) \vec{k}_{i*z}^T \text{,}
 \end{align}

### 2014-03-19 paper/experiments.tex
sec 4 done

@@
 
-Since we are using stochastic optimization, the learn rates need to be chosen carefully.
-We found that the rates used in \citep{hensmangaussian} also work well for our model.
-Specifically, we used the learn rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise hyperparameter, and inducing inputs.
+Since we are using stochastic optimization, the learning rates need to be chosen carefully.
+We found that the rates used in \cite{hensmangaussian} also work well for our model.
+Specifically, we used the learning rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise hyperparameter, and inducing inputs.
 We also included a momentum term of $0.9$ for all of the parameters except the variational parameters and the inducing inputs.
 All of the experiments are executed on an Intel(R) Core(TM) i7-2600 3.40GHz CPU with 8GB of RAM using Matlab R2012a.
-
+%
 \subsection{TOY PROBLEM}
-In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ and corrupted by independent noises: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
+In this toy problem, two related outputs are simulated from the same latent function $sin(x)$ and corrupted by independent noise: $y_1(x) = sin(x) + \epsilon$ and $y_2(x) = -sin(x) + \epsilon$, $\epsilon \sim \Normal(0,0.01)$.
 Each output is given 200 observations with missing values in the $(-7,-3)$ interval for the first output and the $(4,8)$ interval for the second output.
@@
 This confirms the effectiveness of collaborative learning of sparse processes via the shared inducing variables. 
-Also note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
+Also we note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
 
@@
 There are 9, 8, and 42 days of missing values  for gold, silver, and platinum, respectively.
-We remove from the data the exchange rate of CAD on days 50-100, JPY on day 100-150, and AUD on day 150-200.
+We remove from the data the exchange rate of CAD on days 50--100, JPY on day 100--150, and AUD on day 150--200.
 Note that these 3 currencies are from very different geographical locations, making the problem more interesting. 
@@
 We used the squared-exponential covariance function for the shared processes and the noise covariance function for the individual process of each output.
-The inducing inputs $(M = 100, \text{per sparse process})$  are randomly selected from the training data and fixed throughout training.
+$M = 100$ inducing inputs (\text{per sparse process}) were randomly selected from the training data and fixed throughout training.
 
@@
 The real data and predictive distributions by our model are shown in Figure \ref{fig:fx}.
-They exhibit similar behaviors to those by the convolved model with inducing kernels in \citep{alvarez2010efficient}.
+They exhibit similar behaviors to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
 In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
-Our investigation of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 150 - 200.
+Further analysis of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 
+150 -- 200.
 This information from these currencies was effectively used by the model to extrapolate the values of the AUD.
@@
 %todo: result for independent gps
-We also report in Table \ref{tab:fx} the predictive accuracy of our model compared to the convolved GPs model with exact inference \citep[CGP,][]{alvarez-lawrence-nips-08} and independent GPs (IGP, one for each output).
+We also report in Table \ref{tab:fx} the predictive performance of our model compared to the convolved GPs model with exact inference \citep[CGP,][]{alvarez-lawrence-nips-08} and independent GPs (IGP, one for each output).
 Our model outperforms both of CGP and IGP in terms of the standardized mean squared error (SMSE).
@@
 \begin{table}[t]
-\caption{Performance comparison on the foreign exchange rate dataset. Results are averaged of the 3 outputs over 5 repetitions. Smaller figures are better.}
+\caption{Performance comparison on the foreign exchange rate dataset. Results are averages of the 3 outputs over 5 repetitions. Smaller figures are better.}
 \label{tab:fx}
@@
 IGP & 0.5996 & 0.4082 \\
-CGPVAR & 0.2795 & NA \\
+%CGPVAR & 0.2795 & NA \\
 %ICM & 0.3927 &
@@
 We used $Q = 2$ sparse processes with the squared exponential covariance function and individual processes with the noise covariance function.
-The inducing inputs $(M = 200)$ are randomly selected from the training set and fixed throughout training.
-
+$M=200$  inducing inputs  were randomly selected from the training set and fixed throughout training.
+%
 \begin{figure*}
@@
 It took 5 minutes on average to train our model compared to 3 hours of CGP with exact inference.
-
+%
 \setlength{\tabcolsep}{4pt}
@@
 \end{table}
-
+%
 It is also worth noting the characteristics of the sparse processes learned by our model as they indeed correspond to different patterns in the data.
@@
 It has been used in previous work \citep[see e.g.][]{rasmussen-williams-book,vijayakumar2000locally} but only for single task learning. 
-In \citep{chai2008multi}, multitask learning of robot inverse dynamics was considered but on a different and much smaller dataset.
+\citet{chai2008multi} considered multitask learning of robot inverse dynamics  but on a different and much smaller dataset.
 
@@
 Also from Table \ref{tab:robotarm}, it seems that optimizing the inducing inputs can lead to better performance than fixing them.
-More importantly, the overhead in computation is slight, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
+More importantly, the overhead in computation is small, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
 For instance, the total training time is only 1.9 hours when learning with 500 inducing inputs compared to 1.6 hours when fixing them.

### 2014-03-19 paper/discussion.tex
sec 5 done

@@
 First, consider the model for the single GP case, i.e., $P = 1$. By using $Q > 1$ sparse processes, the GP can now have multiple set of inducing variables.
-These can be make local to capture varying characteristics of the latent process in different local input regions, for e.g.
-similar to \citep{nguyen2014fast}, leading to scalable GPs with non-stationary properties.
+These can be made local to capture varying characteristics of the latent process in different  input regions, for e.g. similar to the 
+work by \citet{nguyen2014fast}, leading to scalable GPs with non-stationary properties.
 
-A further extension is to replace the constant mixing weights in the model with input-dependent coefficients like in the Gaussian process regression networks \citep[GPRN][]{wilson-et-al-icml-12} to accommodate adaptive dependencies among the outputs.
-We leave the details of these extensions to future work.
+A further extension is to replace the constant mixing weights in the model with input-dependent coefficients as in the Gaussian process regression networks framework \citep[GPRN,][]{wilson-et-al-icml-12} to accommodate adaptive dependencies among the outputs.
+We leave  these extensions to future work.

### 2014-03-19 paper/intro.tex
got things into shape

@@
 $M$ is the number of inducing points. This high complexity still prevent us from applying GP models 
-to the large scale multi-output problems we are targeting.
+to  large scale multi-output problems.
 
@@
 Furthermore, learning of all  the  parameters in the model, including
-kernel hyperparameters and inducing inputs, can be done in a consistent stochastic optimization framework.
+kernel hyperparameters and inducing inputs, can be done in a single stochastic optimization framework.
 
@@
 % Experiments and summary of results
-The rest of the paper is organized as follows.
-We give description of the model in Section \ref{sec:model}, followed by the inference procedure in Section \ref{sec:inference}. 
-Section \ref{sec:experiments} presents empirical evaluation.
-We conclude with some discussion in Section \ref{sec:discussion}.
+%The rest of the paper is organized as follows.
+%We give description of the model in Section \ref{sec:model}, followed by the inference procedure in Section \ref{sec:inference}. 
+%Section \ref{sec:experiments} presents empirical evaluation.
+%We conclude with some discussion in Section \ref{sec:discussion}.
 

### 2014-03-19 paper/model.tex
got things into shape

@@
 
-As we wish to sparsify these processes for scalability, 
+As we want to sparsify these processes, 
 we introduce the set of \emph{shared inducing variables} $\u_j$ for  each $g_j(\x)$, i.e. $\u_j$ contains the values of 
@@
 %
-The inducing inputs live in the same space as the inputs $\X$, which should be apparent since their corresponding values come from the processes defined on the input space.
-For convenience, we shall assume all processes have the same number of inducing points, $M$, however we emphasize that this is not
+The inducing inputs lie in the same space as the inputs $\X$.
+% which should be apparent since their corresponding values come from the processes defined on the input space.
+For convenience, we  assume all processes have the same number of inducing points, $M$, however we emphasize that this is not
  imposed in practice.

### 2014-03-19 paper/discussion.tex
fix eq references and multioutput

@@
-We have presented scalable multioutput GPs for learning of correlated functions.
+We have presented scalable multi-output GPs for learning of correlated functions.
 The formulation around the inducing variables and their encapsulation of sufficient statistics was shown to be conducive to not only effective but also scalable joint learning under sparsity. 
-We note that although our large scale experiment was done with over 40,000 observations -- the largest publicly available multioutput dataset we could find, the model can easily handle much bigger datasets.
+We note that although our large scale experiment was done with over 40,000 observations -- the largest publicly available multi-output dataset we could find, the model can easily handle much bigger datasets.
 

### 2014-03-19 paper/experiments.tex
fix eq references and multioutput

@@
 A toy problem is first used to study the transfer of learning between two related processes via the shared inducing points.
-We then compare the model with existing multioutput models on the tasks of predicting foreign exchange rate and air temperature.
+We then compare the model with existing multi-output models on the tasks of predicting foreign exchange rate and air temperature.
 In the final experiment, we show that joint learning under sparsity can lead to significant performance gain on a large scale dataset of inverse dynamics of a robot arm.
@@
 
-Since none of the existing multioutput models are applicable to problems of this scale, we compare with independent models that learn each output separately.
+Since none of the existing multi-output models are applicable to problems of this scale, we compare with independent models that learn each output separately.
 Standard GP is applied to the first output as it has only 2,000 observations for training.

### 2014-03-19 paper/inference.tex
got things into shape
fix eq references and multioutput

@@
 carry out stochastic variational inference which allows the model to scale to very large datasets.
-Section \ref{sec:hyperparameters} discusses learning of the inducing inputs and other hyperparameters.
+Section \ref{sec:complexity} compares the complexity of the model with previous multi-output methods.
 
@@
 Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
-Before choosing the form of the approximate posterior, observe that the posterior distribution can be written as:
+We first observe that true the posterior distribution can be written as:
 \begin{align}
@@
 to derive variational inference algorithms for the single output case.
-Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \ref{eq:gu} and \ref{eq:hv}), 
+Since the conditionals $p(\g | \u)$ and $p(\h | \v)$ are known (Equations \eqref{eq:gu} and \eqref{eq:hv}), 
 we only need to 
@@
 the true posteriors.
-This underlines the pivotal role of the inducing variables in this model as previously discussed.
+The quality of aproximation depends entirely on the posterior over the inducing variables, thus underlining their pivotal role in the model as previously discussed.
 
@@
 $ \Eb{\log p(y_{in} | \g_n, h_{in}) }$ over ${p(\g | \u) p(\h_i | \v_i)}$ and 
-then substituting these into Equation \ref{eq:elbo}. (Please see 
+then substituting these into Equation \eqref{eq:elbo}. (Please see 
 the supplementary material for the details of the derivation.)  
@@
 
-So far in the description of the model and inference method we have
+So far in the description of the model and inference  we have
  implicitly assumed that every output has full observations at all inputs $\X$.
@@
 \noindent It can be seen that the derivatives of the parameters of $q(\v_i)$ only involve the observations of the output $i$.
-The derivatives of the parameters of $q(\u_j)$ involve the observations across all outputs but is a sum of contributions from individual outputs.
+The derivatives of the parameters of $q(\u_j)$ involve the observations of all outputs but is a sum of contributions from individual outputs.
 Therefore, computation of the derivatives can be distributed or parallelized easily.
@@
 %
-\subsection{COMPLEXITY ANALYSIS}
+\subsection{COMPLEXITY ANALYSIS \label{sec:complexity}}
 In this section we analyze the complexity of the model and compare it to existing multi-output approaches.

### 2014-03-19 paper/model.tex
fix eq references and multioutput

@@
 % which should be apparent since their corresponding values come from the processes defined on the input space.
-For convenience, we  assume all processes have the same number of inducing points, $M$, however we emphasize that this is not
+For convenience, we  assume all processes have the same number of inducing points, $M$.
+However we emphasize that this is not
  imposed in practice.
@@
 We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
-Note that we reserve subscript $i$ for indexing the outputs and their corresponding individual processes ($i = 1 \hdots P$), $j$ for the latent processes ($j = 1 \hdots Q$), and $n$ for the inputs ($n = 1 \hdots N$).
+Note that we reserve subscript $i$ for indexing the outputs and their corresponding individual processes ($i = 1 \hdots P$), $j$ for the shared latent processes ($j = 1 \hdots Q$), and $n$ for the inputs ($n = 1 \hdots N$).
 %
 \subsection{PRIOR MODEL}
-From the definition of the GPs and the conditional independence of the processes given the inducing variables, 
+From the definition of the GPs and the independence of the processes, 
 the \emph{prior} of the multi-output model can be written as:
@@
 In the equations and hereafter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.
-Here, equations \ref{eq:u} and \ref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (eq.
-\ref{eq:gu} and \ref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
-Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the important role of the inducing variables in the model.
+
+Equations \eqref{eq:u} and \eqref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (Equations  \eqref{eq:gu} and \eqref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
+Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the \emph{sufficient statistics} role of $\u$ and $\v$ in the model.
+Here we use sufficient statistics in the same sense as in \citet{titsias2009variational}, i.e. 
+for any sparse process (say $g_j$), any other set of function values is independent with $\g_j$ given the inducing variables $\u_j$.
 %

### 2014-03-19 paper/experiments.tex
perhaps final edits

@@
 We found that the rates used in \cite{hensmangaussian} also work well for our model.
-Specifically, we used the learning rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise hyperparameter, and inducing inputs.
+Specifically, we used the learning rates of $0.01$ for the variational parameters, $1 \times 10^{-5}$ for the  covariance hyperparameters, and $1 \times 10^{-4}$ for the weights, noise precisions, and inducing inputs.
 We also included a momentum term of $0.9$ for all of the parameters except the variational parameters and the inducing inputs.
@@
 The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca}.
-The setting of our experiment described here is identical to that in \citep{alvarez2010efficient}.
+The setting of our experiment described here is identical to that in \citet{alvarez2010efficient}.
 The dataset consists of all the data available for the 251 working days in the year of 2007.
@@
 We further simulated failure of the sensors by removing the observations from the time periods [10.2 - 10.8] for Cambermet and [13.5 - 14.2] for Chimet.
-The removed data comprises 375 data points, which is used for testing, and the remaining data consisting of 15,788 points is used for training.
+The removed data comprises 375 data points and  is used for testing. 
+The remaining data consisting of 15,788 points is used for training.
 Similar to the previous experiment, the objective is to evaluate the ability of the model to use the signals from the functioning sensors to extrapolate the missing signals.
@@
 \begin{table}[t]
-\caption{Performance comparison on the air temperature dataset. Results are averaged of 2 outputs over 5 repetitions. }
+\caption{Performance comparison on the air temperature dataset. Results are averages of 2 outputs over 5 repetitions. }
 \label{tab:air}
@@
 The second is the sparse GP with stochastic variational inference (SVIGP) using 500 inducing inputs and a batch size of 1,000.
-In case of COGP, we also used a batch size of 1,000 and 500 inducing points for the shared process ($Q = 1$) and each of the independent processes.
+In case of COGP, we also used a batch size of 1,000 and 500 inducing points for the shared process ($Q = 1$) and each of the individual processes.
 The squared exponential with automatic relevance determination (SEard) covariance function is used for all processes of all methods.
@@
 
-The performance of all methods, in terms of SMSE and NLPD averaged over 5 repetitions, is given in Table \ref{tab:robotarm}.
-The benefits of learning the two outputs jointly are evident, as can be seen by the significantly lower SMSE and NLPD of COGP compared to the full GP for the first output (4th torque).
-While the SMSE of the second output is essentially the same for all methods, the NLPD of COGP is substantially better than that of the independent SVIGP model which has the same amount of training data for this torque.
-The results validate the impact of collaborative learning even when under sparsity assumption, thus open up new opportunities for improvement over single task learning with independent sparse processes.
-
 \begin{table}[t]
 \caption{Performance comparison on the robot inverse dynamics dataset.
-In the two bottom lines, standard GP is applied to output 1 and the other method is applied to output 2.
+In the last two lines, standard GP is applied to output 1 and the other method is applied to output 2.
 Results are averaged over 5 repetitions.
-Smaller figures are better.}
+}
 \label{tab:robotarm}
@@
 
-Also from Table \ref{tab:robotarm}, it seems that optimizing the inducing inputs can lead to better performance than fixing them.
-More importantly, the overhead in computation is small, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
-For instance, the total training time is only 1.9 hours when learning with 500 inducing inputs compared to 1.6 hours when fixing them.
-As this dataset is 21-dimensional, the small difference in training time confirms our analysis that the cost of optimizing the inducing inputs only scales sublinearly with the dimensionality of the problem.
-Thus, for datasets of high dimension, learning of the inducing inputs is a practical option.
+The performance of all methods in terms of SMSE and NLPD is given in Table \ref{tab:robotarm}.
+The benefits of learning the two outputs jointly are evident, as can be seen by the significantly lower SMSE and NLPD of COGP compared to the full GP for the first output (4th torque).
+While the SMSE of the second output is essentially the same for all methods, the NLPD of COGP is substantially better than that of the independent SVIGP model which has the same amount of training data for this torque.
+The results validate the impact of collaborative learning even when under sparsity assumption, thus open up new opportunities for improvement over single task learning with independent sparse processes.
 
@@
 \end{figure}
+
+Also from Table \ref{tab:robotarm}, we see that optimizing the inducing inputs can lead to better performance than fixing them.
+More importantly, the overhead in computation is small, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
+For instance, the total training time is only 1.9 hours when learning with 500 inducing inputs compared to 1.6 hours when fixing them.
+As this dataset is 21-dimensional, the small difference in training time confirms that learning of the inducing inputs is a practical option even when dealing with problems of high dimensions.
+%As this dataset is 21-dimensional, the small difference in training time our analysis that the cost of optimizing the inducing inputs only scales sublinearly with the dimensionality of the problem.
+%Thus, for datasets of high dimension, learning of the inducing inputs is a practical option.

### 2014-03-19 paper/inference.tex
perhaps final edits

@@
 %
-We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ across
-multiple outputs using variational inference.
+We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ using variational inference.
 In section \ref{sec:variationalLowerBound} we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points and outputs.
 Section \ref{sec:svi}  takes advantage of this factorization to
-carry out stochastic variational inference which allows the model to scale to very large datasets.
+derive stochastic variational inference, allowing the model to scale to very large datasets.
 Section \ref{sec:complexity} compares the complexity of the model with previous multi-output methods.
@@
 $ \Eb{\log p(y_{in} | \g_n, h_{in}) }$ over ${p(\g | \u) p(\h_i | \v_i)}$ and 
-then substituting these into Equation \eqref{eq:elbo}. (Please see 
-the supplementary material for the details of the derivation.)  
+then substituting these into Equation \eqref{eq:elbo} (see the supplementary material for details). 
 Hence the resulting lower bound is given by:
@@
 The derivatives of the parameters of $q(\u_j)$ involve the observations of all outputs but is a sum of contributions from individual outputs.
-Therefore, computation of the derivatives can be distributed or parallelized easily.
-This property further enhances the scalability of the model, allowing it to handle big data problems with a very large number of inputs and outputs.
+Computation of the derivatives can therefore be easily distributed or parallelized.
+%Therefore, computation of the derivatives can be distributed or parallelized easily.
+%This property further enhances the scalability of the model, allowing it to handle big data problems with a very large number of inputs and outputs.
 
@@
 Learning of the inducing inputs, which was not considered in the single output case in \citet{hensmangaussian}, is also possible in our stochastic optimization approach.
-In the supplementary material, we show that with the help of vectorized operations, the derivatives of the lower bound wrt the inducing inputs can be computed with a cost \emph{independent of the input dimensionality}.
+In the supplementary material, we show that 
+%with the help of vectorized operations, 
+the cost of computing the derivatives of the lower bound wrt the inducing inputs is subdominant to that  of updating the variational parameters.
+%can be computed with a cost \emph{independent of the input dimensionality}.
 This makes optimizing the inducing locations a practical option that can be important, for example, in high-dimensional problems where fixing the inducing inputs may hinder the performance of the model.
@@
 Technically, the cost is $\calO(M^3)$ or $\calO(N_bM^2)$, where $N_b$ is the size of the mini-batches, depending on which is larger between $M$ and $N_b$. 
-Though in practice we may use $N_b > M$ as a large batch size (e.g.~$N_b = 1000$) helps reduce stochasticity of the optimization.
-However, here we use $\calO(M^3)$  for easier comparison with existing models whose computational and storage demands are 
+In practice, we may use $N_b > M$ as a large batch size (e.g.~$N_b = 1000$) helps reduce stochasticity of the optimization.
+However, here we use $\calO(M^3)$  for easier comparison with other models whose time and storage demands are 
 given in Table \ref{tab:complexity}. 
@@
 \begin{table}[t]
-\caption{Comparison of the time and storage complexity of approximate inference of  multi-output GP regression models. 
+\caption{Comparison of the time and storage complexity of approximate inference of  multi-output GP models. 
 A\&W, 2009 refers to \citet{alvarez-lawrence-nips-08}. COGP is the only method with complexity \emph{independent} of the
@@
 \mu^h_{i*} &= \vec{k}_{i*z} \K_{izz}^{-1} \mhi \text{,} \\
-s^h_{i*} &= k_{i**} - \vec{k}_{i*z} \left(\K_{izz}^{-1} - \K_{izz}^{-1} \S_j \K_{izz}^{-1} \right) \vec{k}_{i*z}^T \text{,}
+s^h_{i*} &= k_{i**} - \vec{k}_{i*z} \left(\K_{izz}^{-1} - \K_{izz}^{-1} \S_i \K_{izz}^{-1} \right) \vec{k}_{i*z}^T \text{,}
 \end{align}

### 2014-03-19 paper/model.tex
perhaps final edits

@@
 $g_j(\x)$ at the inducing inputs $\Z_j$. Likewise, we have 
-individual inducing variables corresponding to each  $h_i(\x)$,  which we denote with with $\v_i$ and their 
+individual inducing variables corresponding to each  $h_i(\x)$,  which we denote with $\v_i$ and their 
 corresponding inducing inputs $\Zhi$.
@@
 Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the \emph{sufficient statistics} role of $\u$ and $\v$ in the model.
-Here we use sufficient statistics in the same sense as in \citet{titsias2009variational}, i.e. 
-for any sparse process (say $g_j$), any other set of function values is independent with $\g_j$ given the inducing variables $\u_j$.
+Here by sufficient statistics we mean, for any sparse process (say $g_j$), any other set of function values is independent with $\g_j$ given the inducing variables $\u_j$.
 %

### 2014-03-19 paper/discussion.tex
perhaps final edits

@@
 We have presented scalable multi-output GPs for learning of correlated functions.
-The formulation around the inducing variables and their encapsulation of sufficient statistics was shown to be conducive to not only effective but also scalable joint learning under sparsity. 
-We note that although our large scale experiment was done with over 40,000 observations -- the largest publicly available multi-output dataset we could find, the model can easily handle much bigger datasets.
+The formulation around the inducing variables was shown to be conducive to effective and scalable joint learning under sparsity. 
+We note that although our large scale experiments were done with over 40,000 observations -- the largest publicly available multi-output dataset found, the model can easily handle much bigger datasets.
 
@@
 
-A further extension is to replace the constant mixing weights in the model with input-dependent coefficients as in the Gaussian process regression networks framework \citep[GPRN,][]{wilson-et-al-icml-12} to accommodate adaptive dependencies among the outputs.
+A further extension is to replace the constant mixing weights in the model with input-dependent coefficients as in the Gaussian process regression networks framework \citep{wilson-et-al-icml-12} to accommodate adaptive dependencies among the outputs.
 We leave  these extensions to future work.

### 2014-03-19 paper/intro.tex
perhaps final edits

@@
 However, the prohibitive cost of performing exact inference in  GP models severely hinders 
-their application to  large-scale multi-output problems. For example, na\"{i}ve inference 
+their application to  large scale multi-output problems. For example, na\"{i}ve inference 
 in a fully coupled Gaussian process model over $P$  outputs and $N$ data-points can have 
@@
 cost of at least $\calO(PNM^2)$ in time and  $\calO(PNM)$ in memory, where 
-$M$ is the number of inducing points. This high complexity still prevent us from applying GP models 
+$M$ is the number of inducing points. This high complexity still prevents us from applying GP models 
 to  large scale multi-output problems.
@@
 % Useful observations
-In this work we approach the challenge of building large scale multi-output Gaussian process models  
+In this work we approach the challenge of building scalable multi-output Gaussian process models  
 based on the following observations.
 Firstly, \emph{inducing variables} are the key catalyst
-for achieving sparsity and dealing with large-scale problems in Gaussian process models.
+for achieving sparsity and dealing with large scale problems in Gaussian process models.
 In particular, they capture the \emph{sufficient statistics} of a dataset allowing the construction of sparse 

### 2014-06-11 paper/experiments.tex
perhaps final edits
move table for better format
1. add author names 2. fix references 3. some typos

@@
 \end{tabular}
-\caption{Real data and predictive distributions of by COGP (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the toy problem.
+\caption{Simulated data and predictive distributions of by COGP (first and third figure) and independent GPs using stochastic variational inference (second and last figure) for the toy problem.
 Solid black line: predictive mean; grey bar: two standard deviations; magenta dots: real observations; blue dots: missing data.
@@
 The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca}.
-The setting of our experiment described here is identical to that in \citet{alvarez2010efficient}.
+The setting of our experiment described here is identical to that in \citep{alvarez2010efficient}.
 The dataset consists of all the data available for the 251 working days in the year of 2007.

### 2014-06-11 paper/inference.tex
1. add author names 2. fix references 3. some typos

@@
 Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
-We first observe that true the posterior distribution can be written as:
+We first observe that the true posterior distribution can be written as:
 \begin{align}
@@
 Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
-This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution that of the form:
+This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution of the form:
 \begin{align}
@@
 the true posteriors.
-The quality of aproximation depends entirely on the posterior over the inducing variables, thus underlining their pivotal role in the model as previously discussed.
+The quality of approximation depends entirely on the posterior over the inducing variables, thus underlining their pivotal role in the model as previously discussed.
 
@@
 This enables the use of stochastic optimization methods, which 
-allow the model to handle  ver large datasets for which existing GP-based multi-output models are simply 
+allow the model to handle  very large datasets for which existing GP-based multi-output models are simply 
 impractical.  

### 2014-06-11 paper/abstract.tex
mods to abstrat
fix eq references and multioutput
abstract for better novelty claim

@@
-We propose the collaborative multi-output Gaussian process (GP)  model 
-where each output is a weighted combination of sparse latent processes.
- Each latent process has its own set of inducing points that sufficiently encapsulates the process signal and characteristics.
-Correlations among the outputs are fostered via sharing of multiple inducing point sets, 
-which facilitate  the  application of variational inference and the derivation of 
+We introduce the collaborative multi-output Gaussian process (GP) model for learning dependent tasks with very large datasets. 
+The model fosters task correlations by mixing sparse processes and sharing multiple sets of inducing points. 
+This facilitates  the  application of variational inference and the derivation of 
 an evidence lower bound  that decomposes across inputs and outputs. 
-We learn all the parameters of the model from data in a  single stochastic optimization framework
+We learn all the parameters of the model in a  single stochastic optimization framework
 that scales to a large number of observations per output and a large number of outputs.
 %
-We analyze our multi-output model on a toy problem  and evaluate its performance on 
-two medium-sized datasets and a large dataset showing the benefits of collaborative learning 
-and the superior performance of our approach compared to single output learning and previous 
-multi-output GP models.
+We demonstrate our approach on a toy problem, two medium-sized datasets and a large dataset.
+The model achieves superior performance compared to single output learning and previous multi-output GP models, confirming the 
+benefits of correlating sparsity structure of the outputs via the inducing points.
 
@@
 %% our approach
-%To address this problem, we propose the collaborative multioutput GPs for multitask regression at very large scale.
+%To address this problem, we propose the collaborative multi-output GPs for multitask regression at very large scale.
 %The building blocks of the model are \emph{sparse} processes underpinned by the \emph{inducing variables}.

### 2014-06-11 paper/intro.tex
related work

@@
 % Some citations that can be brought back somewhere
-%The traditional approach in multiple GP regression has been to mix independent latent processes to generate correlated functions.
-%The mixing can be a weighted linear combinations with fixed \citep{teh-et-al-aistats-05,bonilla-et-al-nips-08} or input-dependent \citep{wilson-et-al-icml-12,nguyen2013efficient} coefficients or convolution of processes \citep{boyle-frean-nips-05,alvarez-lawrence-nips-08}.
-%The correlations of the outputs are induced through sharing of these base processes.
+\paragraph{Related work}
+Most GP-based multi-output models create correlated outputs by mixing a set of one or more independent latent processes.
+The mixing can be a linear combinations with fixed coefficients \citep[see e.g.][]{teh-et-al-aistats-05,bonilla-et-al-nips-08}; this is known in the geostatistics community as ''linear model of coregionalization'' \citep{goovaerts1997geostatistics}.
+Such models may also be reformulated in a common Bayesian framework, for example by placing a spike and slab prior over the coefficients \citep{titsias2011spike}.
+More complex dependencies can be induced by using 
+ input-dependent coefficients \citep{wilson-et-al-icml-12,nguyen2013efficient}  or convolving processes \citep{boyle-frean-nips-05,alvarez-lawrence-nips-08,alvarez2010efficient}.
+
+While we also use the mixing construction, the key difference in our model is the role of inducing variables.
+In particular, when used in previous models to reduce the computational costs \citep[see e.g.][]{alvarez-lawrence-nips-08,alvarez2010efficient}, the inducing points are integrated out or collapsed. 
+In contrast, our model maintains an explicit representation of the posterior over the inducing variables that is learned using data from all outputs.
+The explicit representation furthermore facilitates scalable learning in a similar fashion to \citep{hensmangaussian}, making it applicable to much larger datasets.

### 2014-06-11 paper/inference.tex
other minor fixes

@@
 %
-We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ using variational inference.
+We approximate the posterior over the latent variables $\vec{g}, \vec{h}, \vec{u}, \vec{v}$ given observations $\vec{y}$ using variational inference \citep{jordan-variational-99}.
 In section \ref{sec:variationalLowerBound} we derive a lower bound of the marginal likelihood which has the key property of factorizing over the data points and outputs.
@@
 \newcommand{\Shi}{\S^h_i}
-Our inference task is to find the posterior $p(\g, \h, \u, \v | \y)$, for which we use variational inference \citep{jordan-variational-99}. 
+In variational inference, we find the "closest" approximate distribution to the true posterior in terms of the $\KL$ divergence.
 We first observe that the true posterior distribution can be written as:
@@
 \noindent
-To find the best distribution $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
+To find the "best" $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
 \begin{align}

### 2014-06-11 paper/model.tex
other minor fixes

@@
 These variables  play a double pivotal role in the model: they collaboratively share  information 
-across the outputs and  provide \emph{sufficient statistics}  so as to induce sparse processes.
+across the outputs and  provide sufficient statistics so as to induce sparse processes.
 

### 2014-06-11 paper/discussion.tex
squeeze things for 9 pages

@@
 
-We discuss several interesting extensions to this work.
-First, consider the model for the single GP case, i.e., $P = 1$. By using $Q > 1$ sparse processes, the GP can now have multiple set of inducing variables.
-These can be made local to capture varying characteristics of the latent process in different  input regions, for e.g. similar to the 
-work by \citet{nguyen2014fast}, leading to scalable GPs with non-stationary properties.
+%Trung: commented this out for space
+%We discuss several interesting extensions to this work.
+%First, consider the model for the single GP case, i.e., $P = 1$. By using $Q > 1$ sparse processes, the GP can now have multiple set of inducing variables.
+%These can be made local to capture varying characteristics of the latent process in different  input regions, for e.g. similar to the 
+%work by \citet{nguyen2014fast}, leading to scalable GPs with non-stationary properties.
 
-A further extension is to replace the constant mixing weights in the model with input-dependent coefficients as in the Gaussian process regression networks framework \citep{wilson-et-al-icml-12} to accommodate adaptive dependencies among the outputs.
-We leave  these extensions to future work.
+%A further extension is to replace the constant mixing weights in the model with input-dependent coefficients as in the Gaussian process regression networks framework \citep{wilson-et-al-icml-12} to accommodate adaptive dependencies among the outputs.
+%We leave  these extensions to future work.

### 2014-06-11 paper/experiments.tex
squeeze things for 9 pages

@@
 \subsection{FOREIGN EXCHANGE RATE PREDICTION}
+\begin{figure*}
+\centering
+\begin{tabular}{ccc}
+\includegraphics[scale=0.28]{figures/fxCAD.eps} &
+\includegraphics[scale=0.28]{figures/fxJPY.eps} &
+\includegraphics[scale=0.28]{figures/fxAUD.eps}
+\end{tabular}
+\caption{Real observations and predictive distributions for CAD (left), JPY (middle), and AUD (right). The color coding scheme is the same as in Figure \ref{fig:toy}.}
+\label{fig:fx}
+\end{figure*}
+
 The first real world application we consider is to predict the foreign exchange rate w.r.t the US dollar of the top 10 international currencies (CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN) and 3 precious metals (gold, silver, and platinum)\footnote{Data is available at http://fx.sauder.ubc.ca}.
@@
 $M = 100$ inducing inputs (\text{per sparse process}) were randomly selected from the training data and fixed throughout training.
-
-\begin{figure*}
-\centering
-\begin{tabular}{ccc}
-\includegraphics[scale=0.3]{figures/fxCAD.eps} &
-\includegraphics[scale=0.3]{figures/fxJPY.eps} &
-\includegraphics[scale=0.3]{figures/fxAUD.eps}
-\end{tabular}
-\caption{Real observations and predictive distributions for CAD (left), JPY (middle), and AUD (right). The color coding scheme is the same as in Figure \ref{fig:toy}.}
-\label{fig:fx}
-\end{figure*}
-
-The real data and predictive distributions by our model are shown in Figure \ref{fig:fx}.
-They exhibit similar behaviors to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
-In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
-Further analysis of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 
-150 -- 200.
-This information from these currencies was effectively used by the model to extrapolate the values of the AUD.
-
-%todo: result for independent gps
-We also report in Table \ref{tab:fx} the predictive performance of our model compared to the convolved GPs model with exact inference \citep[CGP,][]{alvarez-lawrence-nips-08} and independent GPs (IGP, one for each output).
-Our model outperforms both of CGP and IGP in terms of the standardized mean squared error (SMSE).
-CGP has lower negative log predictive density (NLPD), mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
-% no std because there are 3 outputs but variance is small
-For reference, the convolved GPs with approximation via the variational inducing kernels  \citep[CGPVAR,][]{alvarez2010efficient} has the SMSE of 0.2795 while the NLPD was not provided.
-Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.
 
@@
 
+The real data and predictive distributions by our model are shown in Figure \ref{fig:fx}.
+They exhibit similar behaviors to those by the convolved model with inducing kernels in \citet{alvarez2010efficient}.
+In particular, both models perform better at capturing the strong depreciation of the AUD than the fluctuations of the CAD and JPY currency.
+Further analysis of the dataset found that 4 other currencies (GBP, NZD, KRW, and MXN) also experienced the same trend during the days 
+150 -- 200.
+This information from these currencies was effectively used by the model to extrapolate the values of the AUD.
+
+%todo: result for independent gps
+We also report in Table \ref{tab:fx} the predictive performance of our model compared to the convolved GPs model with exact inference \citep[CGP,][]{alvarez-lawrence-nips-08} and independent GPs (IGP, one for each output).
+Our model outperforms both of CGP and IGP in terms of the standardized mean squared error (SMSE).
+CGP has lower negative log predictive density (NLPD), mainly due to the less conservative predictive variance of the exact CGP for the CAD currency.
+% no std because there are 3 outputs but variance is small
+For reference, the convolved GPs with approximation via the variational inducing kernels  \citep[CGPVAR,][]{alvarez2010efficient} has the SMSE of 0.2795 while the NLPD was not provided.
+Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.
+
 \subsection{AIR TEMPERATURE PREDICTION}
+%
+\setlength{\tabcolsep}{4pt}
+\begin{table}[t]
+\caption{Performance comparison on the air temperature dataset. Results are averages of 2 outputs over 5 repetitions. }
+\label{tab:air}
+\begin{center}
+\begin{tabular}{ccc}
+\toprule
+\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\
+\hline
+COGP & \textbf{0.1077} & \textbf{2.1712} \\
+CGP & 0.1125 & 2.2219 \\
+IGP & 0.8944 & 12.5319 \\
+\bottomrule
+\end{tabular}
+\end{center}
+\end{table}
+%
 Next, we consider the task of predicting air temperature at 4 different locations in the south coast of England. 
-The data is gathered from a network of weather sensors (named Bramblemet, Sotonmet, Cambermet, and Chimet), each of which measures several environmental variables \citep{osborne2008towards}\footnote{Data is available at seperate web pages, see e.g.
+The air temperatures are recorded by a network of weather sensors (named Bramblemet, Sotonmet, Cambermet, and Chimet) during the period from July 10 to July 15, 2013.
+Measurements were taken every 5 minutes, resulting in a maximum of 4320 observations.
+%The data is gathered from a network of weather sensors (named Bramblemet, Sotonmet, Cambermet, and Chimet), each of which measures several environmental variables \citep{osborne2008towards}\footnote{Data is available at seperate web pages, see e.g.
 http://www.bramblemet.co.uk}.
 %The sensors are close geographically so we can expect correlation in air temperature.
-We selected the sensor signal for air temperature during the period from July 10 to July 15, 2013.
-The sensors record measurements every 5 minutes, resulting in a maximum of 4320 observations.
+%We selected the sensor signal for air temperature during the period from July 10 to July 15, 2013.
+%The sensors record measurements every 5 minutes, resulting in a maximum of 4320 observations.
 There are missing data for Bramblemet (100 points), Chimet (15 points), and Sotonmet (1002 points), possibly due to network outages or hardware failures.
@@
 It took 5 minutes on average to train our model compared to 3 hours of CGP with exact inference.
-%
-\setlength{\tabcolsep}{4pt}
-\begin{table}[t]
-\caption{Performance comparison on the air temperature dataset. Results are averages of 2 outputs over 5 repetitions. }
-\label{tab:air}
-\begin{center}
-\begin{tabular}{ccc}
-\toprule
-\textbf{METHOD} & \textbf{SMSE} & \textbf{NLPD} \\
-\hline
-COGP & \textbf{0.1077} & \textbf{2.1712} \\
-CGP & 0.1125 & 2.2219 \\
-IGP & 0.8944 & 12.5319 \\
-\bottomrule
-\end{tabular}
-\end{center}
-\end{table}
-%
+
 It is also worth noting the characteristics of the sparse processes learned by our model as they indeed correspond to different patterns in the data.
@@
 The test set consists of 8,898 observations equally divided between the two outputs.
-Note that the data was collected at 100Hz for 7.5 minutes in total, hence ideal for this experiment which evaluates the effectiveness of joint learning under the assumption of sparsity in the output spaces.
+%Note that the data was collected at 100Hz for 7.5 minutes in total, hence ideal for this experiment which evaluates the effectiveness of joint learning under the assumption of sparsity in the output spaces.
 
@@
 In case of COGP, we also used a batch size of 1,000 and 500 inducing points for the shared process ($Q = 1$) and each of the individual processes.
-The squared exponential with automatic relevance determination (SEard) covariance function is used for all processes of all methods.
-Finally, the outputs are normalized to have zero mean and unit variance.
+%The squared exponential with automatic relevance determination (SEard) covariance function is used for all processes of all methods.
+%Finally, the outputs are normalized to have zero mean and unit variance.
 
@@
 
-% table for smses, nlpds, training tiem
-\begin{figure}
-\includegraphics[scale=0.4]{figures/sarcosTime.eps}
-\caption{The cost of optimizing versus fixing the inducing inputs.}
-\label{fig:time}
-\end{figure}
-
 Also from Table \ref{tab:robotarm}, we see that optimizing the inducing inputs can lead to better performance than fixing them.
@@
 %Thus, for datasets of high dimension, learning of the inducing inputs is a practical option.
+% table for smses, nlpds, training tiem
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{figures/sarcosTime.eps}
+\caption{The cost of optimizing versus fixing the inducing inputs.}
+\label{fig:time}
+\end{figure}

### 2014-06-16 paper/experiments.tex
squeezed in acks

@@
 %For Chimet, both models perform poorly but CGP more so as it falsely predicts  wild fluctuations that are non-existent in the data. 
-The comparative performance of the models are summarized in Table \ref{tab:air}, from which it can be seen that our model outperforms CGP in terms of both SMSE and NLPD.
+The  performance of the models are summarized in Table \ref{tab:air},  which shows that our model outperforms CGP in terms of both SMSE and NLPD.
 %All results are averaged of the 2 outputs over 5 repetitions.
@@
 
-It is also worth noting the characteristics of the sparse processes learned by our model as they indeed correspond to different patterns in the data.
-In particular, one process has the inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has the inverse lengthscale of 0.5 to model the local variations within a single day.
+It is also worth noting the characteristics of the sparse processes learned by our model as they  correspond to different patterns in the data.
+In particular, one process has an inverse lengthscale of 136 which captures the global increase in temperature during the training period while the other has an inverse lengthscale of 0.5 to model the local variations within a single day.
 

### 2014-06-16 paper/experiments.tex
quite a few things here and there

@@
 We then compare the model with existing multi-output models on the tasks of predicting foreign exchange rate and air temperature.
-In the final experiment, we show that joint learning under sparsity can lead to significant performance gain on a large scale dataset of inverse dynamics of a robot arm.
+In the final experiment, we show that joint learning under sparsity can yield significant performance gain on a large scale dataset of inverse dynamics of a robot arm.
 
@@
 This confirms the effectiveness of collaborative learning of sparse processes via the shared inducing variables. 
-Also we note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
+Additionally, we note that the inference procedure learned that the weights are $w_{11} = 1.07$ and $w_{21} = -1.06$ which accurately reflects the correlation between the two outputs.
 
@@
 % no std because there are 3 outputs but variance is small
-For reference, the convolved GPs with approximation via the variational inducing kernels  \citep[CGPVAR,][]{alvarez2010efficient} has the SMSE of 0.2795 while the NLPD was not provided.
+For reference, the convolved GPs with approximation via the variational inducing kernels  \citep[CGPVAR,][]{alvarez2010efficient} 
+has an SMSE of 0.2795 while the NLPD was not provided.
 Training took only 10 minutes for our model compared to 1.4 hours for the full CGP model.
@@
 %
-Next, we consider the task of predicting air temperature at 4 different locations in the south coast of England. 
+Next we consider the task of predicting air temperature at 4 different locations in the south coast of England. 
 The air temperatures are recorded by a network of weather sensors (named Bramblemet, Sotonmet, Cambermet, and Chimet) during the period from July 10 to July 15, 2013.
@@
 
-For pre-processing we normalized the outputs to have zero mean and unit variance.
+We normalized the outputs to have zero mean and unit variance.
 We used $Q = 2$ sparse processes with the squared exponential covariance function and individual processes with the noise covariance function.
@@
 The real data and the predictive distributions by our model, CGP with exact inference \citep{alvarez-lawrence-nips-08}, and independent GPs are shown in Figure \ref{fig:weather}.
-It is clear that the independent GPs model is clueless in the test regions and thus simply uses the average temperature as its prediction.
-For Cambermet, both COGP and CGP can capture the rising in temperature from the morning till the afternoon and the fall afterward.
+It is clear that the independent GP model is clueless in the test regions and thus simply uses the average temperature as its prediction.
+For Cambermet, both COGP and CGP can capture the rising in temperature from the morning until the afternoon and the fall afterwards.
 %For Chimet, both models perform poorly but CGP more so as it falsely predicts  wild fluctuations that are non-existent in the data. 
@@
 Our last experiment is with a dataset relating to an inverse dynamics model of a 7-degree-of-freedom anthropomorphic robot arm \citep{vijayakumar2000locally}.
-The data consists of 48,933 mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
+The data consists of 48,933 datapoints mapping from a 21-dimensional input space (7 joints positions, 7 joint velocities, 7 joint accelerations) to the corresponding 7 joint torques.
 %The problem is strongly nonlinear due to complex superpositions of sine and cosine functions in robot dynamics \citep{vijayakumar2000locally}.
@@
 
-Here we consider joint learning for the 4th and 7th torques where the first has 2,000 points  while the second has the full data of 44,484 points for training.
+Here we consider joint learning for the 4th and 7th torques, where the former has 2,000 points  
+while the latter has  44,484 points for training.
 The test set consists of 8,898 observations equally divided between the two outputs.
@@
 While the SMSE of the second output is essentially the same for all methods, the NLPD of COGP is substantially better than that of the independent SVIGP model which has the same amount of training data for this torque.
-The results validate the impact of collaborative learning even when under sparsity assumption, thus open up new opportunities for improvement over single task learning with independent sparse processes.
+These results validate the impact of collaborative learning under sparsity assumptions,  opening up new opportunities for improvement over single task learning with independent sparse processes.
 
-Also from Table \ref{tab:robotarm}, we see that optimizing the inducing inputs can lead to better performance than fixing them.
-More importantly, the overhead in computation is small, as demonstrated by the training time using the two options in Figure \ref{fig:time}.
+Finally, we see on Table \ref{tab:robotarm}  that optimizing the inducing inputs can yield better performance than fixing them.
+More importantly, the overhead in computation is small, as demonstrated by the training times shown in Figure \ref{fig:time}.
 For instance, the total training time is only 1.9 hours when learning with 500 inducing inputs compared to 1.6 hours when fixing them.
-As this dataset is 21-dimensional, the small difference in training time confirms that learning of the inducing inputs is a practical option even when dealing with problems of high dimensions.
+As this dataset is 21-dimensional, this small difference in training time confirms that learning of the inducing inputs is a practical option even when dealing with problems of high dimensions.
 %As this dataset is 21-dimensional, the small difference in training time our analysis that the cost of optimizing the inducing inputs only scales sublinearly with the dimensionality of the problem.

### 2014-06-16 paper/inference.tex
quite a few things here and there

@@
 \newcommand{\Shi}{\S^h_i}
-In variational inference, we find the "closest" approximate distribution to the true posterior in terms of the $\KL$ divergence.
+In variational inference, we find the ``closest'' approximate distribution to the true posterior in terms of the $\KL$ divergence.
 We first observe that the true posterior distribution can be written as:
@@
 \end{align}
-Here we recall the modeling assumption that each set of the inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
+Here we recall the modeling assumption that each set of  inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
 This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution of the form:
@@
 \noindent
-To find the "best" $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
+To find the best $q(\u, \v)$, we  optimize the evidence lower bound (ELBO) of the log marginal:
 \begin{align}
@@
 where $\g_n = \{g_{jn} = (\g_j)_n\}_{j=1}^Q$.
-The inequality is due to Jensen's inequality and the equality is due to the fact that the likelihood fully factorizes.
+The inequality is due to Jensen's inequality and the equality is due to %the fact that the likelihood fully factorizes.
+the factorization of the likelihood. 
 \newcommand{\Ahi}{\A^h_i}
@@
 used $\A_j(n,:)$ to denote the $n$-th row vector of $\A_j$. 
-Notice that this ELBO clearly generalizes the bound for standard GP regression derived in \cite{hensmangaussian}, which can be recovered by setting $P = Q = 1$, $w_{ij} = 1$ and $h_i(\x) = 0$.
+Notice that this ELBO  generalizes the bound for standard GP regression derived in \cite{hensmangaussian}, which can be recovered by setting $P = Q = 1$, $w_{ij} = 1$ and $h_i(\x) = 0$.
 
@@
 %with the help of vectorized operations, 
-the cost of computing the derivatives of the lower bound wrt the inducing inputs is subdominant to that  of updating the variational parameters.
+the additional cost of computing the derivatives of the lower bound wrt the inducing inputs is not significantly 
+higher than the cost of updating the variational parameters.
 %can be computed with a cost \emph{independent of the input dimensionality}.
-This makes optimizing the inducing locations a practical option that can be important, for example, in high-dimensional problems where fixing the inducing inputs may hinder the performance of the model.
+This makes optimizing the inducing locations a practical option, which can be critical 
+in high-dimensional problems. %where fixing the inducing inputs may hinder the performance of the model.
 Indeed, our experiments on a large scale multi-output problem show that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation. 
@@
 
-The complexity of COGP can be read off by inspecting the ELBO  in Equation \eqref{eq:L}, with a key observation that it contains 
+The complexity of COGP can be read off by inspecting the ELBO  in Equation \eqref{eq:L}, with the key observation that it contains 
 a sum over the outputs as well as over the inputs.
@@
 a computational  complexity that is independent of the size of the inputs and outputs, 
-thus is the only method capable of handling large scale problems.
+which makes it the only method capable of handling large scale problems.
 %

### 2014-06-16 paper/intro.tex
quite a few things here and there

@@
 per torque in just 7 minutes.
-Clearly this problem is well beyond the capability of conventional multiple output GPs.
+Clearly this problem is well beyond the capabilities  of conventional multiple output GPs.
 Building  multi-output GP models that can learn correlated tasks at the scale of 
@@
 % Some citations that can be brought back somewhere
-\paragraph{Related work}
-Most GP-based multi-output models create correlated outputs by mixing a set of one or more independent latent processes.
-The mixing can be a linear combinations with fixed coefficients \citep[see e.g.][]{teh-et-al-aistats-05,bonilla-et-al-nips-08}; this is known in the geostatistics community as ''linear model of coregionalization'' \citep{goovaerts1997geostatistics}.
+\paragraph{Related work.}
+Most GP-based multi-output models create correlated outputs by mixing a set of independent latent processes.
+The mixing can be a linear combination with fixed coefficients \citep[see e.g.][]{teh-et-al-aistats-05,bonilla-et-al-nips-08}.
+This is known in the geostatistics community as  the ``linear model of coregionalization'' \citep{goovaerts1997geostatistics}.
 Such models may also be reformulated in a common Bayesian framework, for example by placing a spike and slab prior over the coefficients \citep{titsias2011spike}.
@@
 In contrast, our model maintains an explicit representation of the posterior over the inducing variables that is learned using data from all outputs.
-The explicit representation furthermore facilitates scalable learning in a similar fashion to \citep{hensmangaussian}, making it applicable to much larger datasets.
+This explicit representation  facilitates scalable learning in a similar fashion to 
+the approach in \citet{hensmangaussian}, making it applicable to very large datasets.

### 2014-06-16 paper/model.tex
quite a few things here and there

@@
 The $Q$ shared functions have independent Gaussian process priors  $g_j(\x) \sim \GP(0, k_j(\cdot,\cdot))$. %, with $j= 1, \ldots Q$.
-Similarly, each individual function of an output also has GP prior, i.e.~$h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$. %with $i=1 ,\ldots P$.
+Similarly, each individual function of an output also has a GP prior, i.e.~$h_i(\x) \sim \GP(0, k^h_i(\cdot,\cdot))$. %with $i=1 ,\ldots P$.
 
 As we want to sparsify these processes, 
-we introduce the set of \emph{shared inducing variables} $\u_j$ for  each $g_j(\x)$, i.e. $\u_j$ contains the values of 
+we introduce a set of \emph{shared inducing variables} $\u_j$ for  each $g_j(\x)$, i.e. $\u_j$ contains the values of 
 $g_j(\x)$ at the inducing inputs $\Z_j$. Likewise, we have 
@@
 Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the \emph{sufficient statistics} role of $\u$ and $\v$ in the model.
-Here by sufficient statistics we mean, for any sparse process (say $g_j$), any other set of function values is independent with $\g_j$ given the inducing variables $\u_j$.
+Here by sufficient statistics we mean, for any sparse process (say $g_j$), any other set of function values is 
+independent of $\g_j$ given the inducing variables $\u_j$.
 %

### 2014-06-16 paper/inference.tex
final edits

@@
 \end{align}
-Here we recall the modeling assumption that each set of  inducing variables is the \emph{sufficient statistics} of the corresponding latent process. 
+Here we recall the modeling assumption that each set of  inducing variables is the sufficient statistics of the corresponding latent process. 
 This motivates replacing the true posteriors over $\g$ and $\h$ with their conditional distributions given the inducing variables, leading to a distribution of the form:
@@
 For consistency, we first unify common notations used for all models.
-We use $P$ as the number of outputs, $N$ as the number of inputs, $Q$ as the number of shared latent processes, $M_t$ as the \emph{total} number of inducing inputs.
+We use $P$ as the number of outputs; $N$ as the number of inputs; $Q$ as the number of shared latent processes; and $M_t$ as the \emph{total} number of inducing inputs.
 It is worth noting that $M_t = (P + Q) \times M$ in our COGP model, assuming that each sparse process has equal number of inducing points. 
-Also, COGP have $P$ additional individual processes, one for each output.
+Also, COGP has $P$ additional individual processes, one for each output.
 

### 2014-06-16 paper/intro.tex
final edits

@@
 In particular, when having multiple outputs or tasks they have proved effective in modeling 
-the dependencies between the tasks  and  they have been shown to outperform competitive baselines and 
+the dependencies between the tasks, outperforming  competitive baselines and 
 single output learners  \citep{bonilla-et-al-nips-08,teh-et-al-aistats-05,alvarez-lawrence-nips-08,wilson-et-al-icml-12}.
@@
 their application to  large scale multi-output problems. For example, na\"{i}ve inference 
-in a fully coupled Gaussian process model over $P$  outputs and $N$ data-points can have 
+in a fully coupled Gaussian process model over $P$  outputs and $N$ data points can have 
 a complexity of $\calO(N^3P^3)$  and $\calO(N^2P^2)$ in time and memory, respectively.
@@
 In the single output setting previous attempts to scale up  GP inference 
-resort to approximate inference. Most  approximation techniques 
+resort to approximate inference. Most  approximation methods  
  can be understood 
@@
 based on the following observations.
-Firstly, \emph{inducing variables} are the key catalyst
+Firstly, inducing variables are the key catalyst
 for achieving sparsity and dealing with large scale problems in Gaussian process models.
@@
 Based on these observations, we propose the collaborative multi-output Gaussian Process (COGP) model 
-where each output is a weighted combination of sparse latent processes.
- Each latent process has its own set of inducing points that sufficiently encapsulates the process signal and characteristics.
-Correlations among the outputs are fostered via sharing of multiple inducing point sets 
-which in turn create a medium through which joint inference happens efficiently.
+where latent processes are mixed to generate dependent outputs. 
+Each process is sparse and characterized by its own set of inducing points.
+The sparsity structure enabling output correlations is thus created via the shared inducing sets. 
+To learn this structure, we maintain an explicit representation of the posterior over the inducing points which in turn allows inference to be carried out efficiently.
 % exploit stochastic variational inference
-In particular, we derive a variational lower bound of the model evidence that decomposes across 
+In particular, we obtain a variational lower bound of the model evidence that decomposes across 
 inputs and outputs. 

### 2014-06-16 paper/model.tex
final edits

@@
 Equations \eqref{eq:u} and \eqref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (Equations  \eqref{eq:gu} and \eqref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
-Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the \emph{sufficient statistics} role of $\u$ and $\v$ in the model.
+Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the sufficient statistics role of $\u$ and $\v$ in the model.
 Here by sufficient statistics we mean, for any sparse process (say $g_j$), any other set of function values is 
